---
title: "Workshop: Multiple Correspondence Analysis — Part 1: Foundations and Simulation"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---

```{r import_libraries}
#| echo: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(furrr)
library(glue)
library(fs)
library(FactoMineR)
library(factoextra)
library(ca)
library(dendextend)
library(cluster)
library(broom)

source("lib_utils.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "FactoMineR", "factoextra", "ca")
)

options(
  width    = 80L,
  warn     = 1,
  mc.cores = parallelly::availableCores()
)

set.seed(42)

theme_set(theme_cowplot())

plan(multisession)
```

# The Big Idea: Making Sense of Categorical Data

> *"The greatest value of a picture is when it forces us to notice what we never expected to see."*
> — John W. Tukey

Imagine you're sitting in a café and you observe customers ordering coffee. You notice:

- Some people always order espresso, always at 8 AM, always on weekdays.
- Others order cappuccino on weekends, usually in the afternoon.
- A few order decaf late at night, regardless of the day.

These are **patterns in categorical data** — combinations of choices across multiple dimensions. Each customer exhibits a *lifestyle* or *preference profile*, and you want to uncover the natural groupings: which combinations of choices go together? What defines a "morning espresso type" versus an "afternoon cappuccino person"?

This is the problem that **Multiple Correspondence Analysis (MCA)** solves.

## The Challenge with Categorical Data

For decades, statistics has been built around numerical data. We have well-established techniques for continuous variables: correlation matrices, principal component analysis, linear regression. But **categorical data** — the choices we make, the categories we belong to, the responses on a survey — plays by different rules.

Consider a simple dataset: 100 people surveyed on three questions:

- **Favorite coffee type**: Espresso, Cappuccino, Flat White, Decaf
- **Time of day**: Morning, Afternoon, Evening, Night
- **Day of week**: Weekday, Weekend

This yields $4 \times 4 \times 2 = 32$ possible combinations. Some combinations might be common ("Espresso + Morning + Weekday"); others might be rare ("Decaf + Night + Weekday"); and some might barely exist ("Cappuccino + Night + Weekend").

If you try to use traditional correlation analysis, you'll run into problems:

1. **There is no natural ordering** — "Espresso" is not "less than" Cappuccino; they are distinct categories.
2. **Dummy coding creates artificial structure** — encoding Espresso as 1 and Cappuccino as 0 suggests a relationship that doesn't exist.
3. **Standard PCA treats all categories equally** — it has no concept that "Espresso at 8 AM on a weekday" is a meaningful combination that often appears together.

**Multiple Correspondence Analysis** solves these problems by asking: *What are the natural associations between categories?* It finds dimensions that best separate and organize the categories in a way that respects the contingency structure — the actual co-occurrence patterns in your data.

## Workshop Materials

All code and materials for this workshop are available on GitHub:
**https://github.com/kaybenleroll/data_workshops**

This workshop requires the following R packages:

- `FactoMineR`: Core MCA implementation and utilities
- `factoextra`: Beautiful visualization of MCA results
- `ca`: Classical correspondence analysis (biplot methods)
- `cluster` / `dendextend`: Clustering and dendrogram visualization
- `tidyverse`: Data manipulation and visualization
- `cowplot`: Plot theming and arrangement

## How to Read This Workshop

This is not a dry mathematical treatise. Instead, we'll build intuition through three complementary lenses:

1. **Geometric intuition**: Pictures of how categories are positioned relative to each other.
2. **Contingency tables**: Understanding co-occurrence patterns directly.
3. **Code simulations**: Generating synthetic data so you can see how different patterns emerge in MCA space.

By the end, you'll understand:

- Why MCA is the right tool for categorical data
- How to run MCA and interpret its output
- How to use it for exploratory analysis, clustering, and dimensionality reduction
- When to use advanced techniques like supplementary variables, mixed-data MCA, and post-MCA clustering

---

# Part 1: Building Blocks — From Tables to Geometry

## The Humble Contingency Table

Let's start with the simplest interesting case: a **2 × 2 contingency table**. Suppose we survey 200 people on two questions:

- **Coffee type**: Espresso vs. Cappuccino
- **Time of day**: Morning vs. Afternoon

We get this observed table:

|                | Morning | Afternoon |
|---|---|---|
| **Espresso**   | 60      | 20        |
| **Cappuccino** | 15      | 105       |

What does this tell us? Visually, it's already clear: Espresso is strongly associated with Morning, and Cappuccino with Afternoon. But *how do we quantify this association?* And how do we visualize it so the relationship jumps out geometrically?

## Chi-Square Distance: Measuring Association

The trick is to ask: **if coffee type and time of day were independent, what would we expect?** If they were unrelated, we'd expect:

$$
E_{ij} = \frac{(\text{row total}_i) \times (\text{col total}_j)}{\text{grand total}}
$$

For Espresso + Morning:

$$
E_{1,1} = \frac{(60 + 20) \times (60 + 15)}{200} = \frac{80 \times 75}{200} = 30
$$

But we *observed* 60, not 30. That's a big discrepancy. Across all cells:

| | Morning | Afternoon |
|---|---|---|
| **Espresso** | Obs: 60, Exp: 30 | Obs: 20, Exp: 50 |
| **Cappuccino** | Obs: 15, Exp: 45 | Obs: 105, Exp: 75 |

This mismatch is captured by the **chi-square statistic**:

$$
\chi^2 = \sum_{i,j} \frac{(O_{ij} - E_{ij})^2}{E_{ij}}
$$

In our example, $\chi^2 \approx 133.33$, which is huge — far beyond what random chance would produce. This table has *strong* association.

But here's the key insight: we can use chi-square distance to define a **metric** between categories. The distance between "Espresso" and "Cappuccino" in "space" can be measured using the chi-square distance between their row profiles (their distributions across all columns).

```{r chisq_demo}
#| echo: true
#| fig-cap: "A 2×2 contingency table with a clear association: Espresso→Morning, Cappuccino→Afternoon."

# Observed frequencies
obs_table <- matrix(c(60, 15, 20, 105), nrow = 2, ncol = 2,
                    dimnames = list(
                      c("Espresso", "Cappuccino"),
                      c("Morning", "Afternoon")
                    ))

# Expected frequencies under independence
row_totals <- rowSums(obs_table)
col_totals <- colSums(obs_table)
grand_total <- sum(obs_table)

exp_table <- outer(row_totals, col_totals) / grand_total

# Chi-square statistic
chisq_stat <- sum((obs_table - exp_table)^2 / exp_table)

# Create a nice visualization
contingency_tbl <- as_tibble(obs_table, rownames = "coffee") |>
  pivot_longer(cols = -coffee, names_to = "time", values_to = "count")

ggplot(contingency_tbl, aes(x = time, y = coffee, fill = count)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(aes(label = count), colour = "white", fontface = "bold", size = 6) +
  scale_fill_gradient(low = "#FFF44F", high = "#D81B60") +
  labs(
    title    = "Coffee Preference by Time of Day",
    subtitle = glue("χ² = {round(chisq_stat, 2)} (very strong association)"),
    x        = "Time of Day",
    y        = "Coffee Type",
    fill     = "Count"
  ) +
  theme(
    axis.text = element_text(size = 12),
    legend.position = "right"
  )
```

### Row and Column Profiles

To perform correspondence analysis, we need to convert this table to *proportions*. For each row, we ask: **what is its profile across the columns?**

For Espresso: $(60/80, 20/80) = (0.75, 0.25)$ — "75% morning, 25% afternoon"  
For Cappuccino: $(15/120, 105/120) = (0.125, 0.875)$ — "12.5% morning, 87.5% afternoon"

These profiles are dramatically different. If we could visualize them in 2D space, the two points would be far apart.

```{r row_profiles}
#| echo: true
#| fig-cap: "Row profiles (coffee types as distributions over time of day). The distance between profiles reflects the strength of association."

# Row profiles
row_profiles <- obs_table / rowSums(obs_table)

profile_tbl <- as_tibble(row_profiles, rownames = "coffee") |>
  pivot_longer(cols = -coffee, names_to = "time", values_to = "proportion")

ggplot(profile_tbl, aes(x = time, y = proportion, fill = coffee, group = coffee)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_y_continuous(labels = percent) +
  scale_fill_manual(values = c("steelblue", "darkorange")) +
  labs(
    title    = "Row Profiles: Coffee Type Distribution by Time of Day",
    x        = "Time of Day",
    y        = "Proportion",
    fill     = "Coffee Type"
  )
```

## The Magic: Eigenvector Decomposition

Here's where the magic happens. We want to find a low-dimensional representation (like 2D) that preserves these distances as much as possible. This is solved via the **singular value decomposition (SVD)** of the chi-square-standardized table.

The process:

1. Compute row profiles and column profiles.
2. Standardize by dividing by expected frequencies (chi-square rescaling).
3. Decompose the matrix into eigenvectors and eigenvalues.
4. The eigenvectors become coordinates in MCA space.

The mathematics is elegant: the first eigenvector points in the direction of *maximum chi-square distance* between row/column profiles; the second eigenvector is orthogonal and captures the *next-most variation*; and so on.

For our 2×2 table, there is only one meaningful dimension (we have 4 categories, so at most 3 dimensions, but one is trivial). Let me show this with a simple visualization:

```{r simple_ca_visualization}
#| echo: true
#| fig-cap: "Correspondence analysis of the 2×2 table. A 2×2 table yields only one meaningful dimension — points are placed on a number line where distance reflects chi-square distance."

# Perform CA
ca_fit <- ca::ca(obs_table)

# A 2x2 table has only one meaningful CA dimension (rank-1 after centering).
# We plot on a number line rather than a biplot.
all_coords <- tibble(
  category = c(rownames(ca_fit$rowcoord), rownames(ca_fit$colcoord)),
  score    = c(ca_fit$rowcoord[, 1], ca_fit$colcoord[, 1]),
  type     = c(rep("Coffee Type", nrow(ca_fit$rowcoord)),
               rep("Time of Day",  nrow(ca_fit$colcoord)))
)

pct_inertia <- round(ca_fit$sv[1]^2 / sum(ca_fit$sv^2) * 100, 1)

ggplot(all_coords, aes(x = score, y = type, colour = type, label = category)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_point(size = 5, alpha = 0.85) +
  geom_text(nudge_y = 0.25, size = 3.5, fontface = "bold") +
  scale_colour_manual(values = c("steelblue", "darkorange"), guide = "none") +
  labs(
    title    = "Correspondence Analysis — Dimension 1",
    subtitle = glue("Single axis captures {pct_inertia}% of inertia"),
    x        = "CA Score (Dimension 1)",
    y        = NULL
  ) +
  theme(
    panel.grid.minor = element_blank()
  )
```

**Key insight**: Espresso and Morning are on the same side (they attract); Cappuccino and Afternoon are on the opposite side. The origin of the plot is at the "grand average" — the center of mass of all categories weighted by frequency.

---

# Part 2: Moving to Multiple Variables — MCA

In the real world, you don't have two variables; you have many. A customer survey might have 15 questions. A marketing dataset might have dozens of categorical features. How do we extend this simple 2×2 correspondence analysis to handle multiple variables?

## The Conceptual Shift: Indicator Matrix

The key is to convert your multi-variable dataset into an **indicator matrix** (also called a Burt matrix variant).

Suppose we have 10 people and 3 variables:

- **Coffee**: Espresso, Cappuccino
- **Time**: Morning, Afternoon
- **Day**: Weekday, Weekend

Instead of a single contingency table, we create a matrix where:
- Each row is a person (observation)
- Each column is a *category* (not a variable)
- Each cell is 1 if the person has that category, 0 otherwise

```{r indicator_matrix_demo}
#| echo: true
#| fig-cap: "The indicator matrix: each person (row) indicates which categories they belong to."

# Create sample data
sample_data <- tibble(
  ID = 1:10,
  Coffee = c("Espresso", "Espresso", "Cappuccino", "Cappuccino", "Espresso",
             "Cappuccino", "Espresso", "Cappuccino", "Cappuccino", "Espresso"),
  Time = c("Morning", "Morning", "Afternoon", "Afternoon", "Afternoon",
           "Morning", "Evening", "Afternoon", "Morning", "Evening"),
  Day = c("Weekday", "Weekday", "Weekday", "Weekend", "Weekend",
          "Weekday", "Weekend", "Weekday", "Weekend", "Weekday")
)

# Create indicator matrix
indicator_mat <- sample_data |>
  select(-ID) |>
  model.matrix(~ . - 1, data = _) |>
  as_tibble(rownames = "ID")

# Display a nicely formatted version
indicator_mat |>
  head(5) |>
  knitr::kable(caption = "Indicator Matrix (first 5 people)")
```

This 10 × 6 indicator matrix is the input to MCA. Each row is a person's profile across all categories. MCA will then find dimensions that best separate *people* (and simultaneously position *categories*) so that categories that co-occur frequently are close together, and categories that rarely co-occur are far apart.

## Why the Indicator Matrix is Singular (and Why That is OK)

This is an important technical point: the complete indicator matrix is **rank-deficient by construction**.

For any variable with $K$ categories, each row has exactly one 1 and $K-1$ zeros, so the category columns for that variable satisfy:

$$
\sum_{k=1}^{K} x_{ik} = 1 \quad \text{for every row } i
$$

That linear constraint means those $K$ columns only contain $K-1$ independent degrees of freedom. In other words, each variable contributes one exact dependency.

If we have $Q$ categorical variables with category counts $K_1, K_2, \dots, K_Q$, and total indicator columns

$$
J = \sum_{q=1}^{Q} K_q,
$$

then the effective column rank is at most

$$
\sum_{q=1}^{Q} (K_q - 1) = J - Q
$$

(and of course cannot exceed $n-1$ from row-centering).

So yes: the full disjunctive matrix is singular. That is expected, and it is exactly analogous to the dummy-variable trap in regression.

## Why We Still Use All $K$ Columns

In regression, we usually drop one dummy column to avoid perfect collinearity. In MCA, we intentionally keep the full disjunctive coding because the goal is different: we want a **symmetric geometric representation of categories**.

Keeping all categories explicit is useful because:

- It treats every category level equally (no arbitrary "reference" level).
- Distances and contributions are interpreted directly at the category level.
- Barycentric relationships (individual points as weighted averages of their categories) are preserved cleanly.
- Supplementary projections are simpler: new observations are encoded the same way as active ones.

If we dropped one category per variable, the geometry would depend on which level was omitted, which makes interpretation less stable and less transparent.

**Why this is fine in MCA**:

- MCA is built around this coding; you do **not** need to drop a reference category manually.
- The SVD/eigendecomposition naturally yields zero (or near-zero) eigenvalues in redundant directions.
- Interpretation should focus on the non-zero principal dimensions, which correspond to the true degrees of freedom in categorical structure.

## Inertia: The Total Spread of the Data

In MCA, we measure total variability using **inertia**, which is proportional to the chi-square statistic divided by the total sample size:

$$
\text{Total Inertia} = \frac{\chi^2}{n}
$$

This is analogous to the "total variance" in PCA. It tells us: *how much association is there in the data overall?* A high inertia means categories are strongly associated; a low inertia means they're scattered randomly.

When we extract the first MCA dimension, we *reduce* the inertia — we're explaining some of the total variability. The **eigenvalue** of each dimension tells us what fraction of the total inertia it explains. Typically:

- Dimension 1 might explain 15–30% of inertia
- Dimension 2 might explain 10–20%
- And so on, with each successive dimension explaining less

```{r inertia_scree_example}
#| echo: true
#| fig-cap: "A scree plot from a real MCA: each dimension explains a fraction of total inertia. Typically we retain dimensions where the scree 'bends'."

# Create a synthetic scree plot (realistic structure)
n_dims <- 12
eigenvalues <- c(0.45, 0.28, 0.18, 0.12, 0.08, 0.05, 0.03, 0.02, 0.01, 0.005, 0.003, 0.002)
eigenvalues <- eigenvalues / sum(eigenvalues)  # Normalize to proportions

scree_tbl <- tibble(
  Dimension = paste0("Dim ", 1:n_dims),
  Inertia = eigenvalues,
  Cumulative = cumsum(eigenvalues)
)

ggplot(scree_tbl |> head(8), aes(x = reorder(Dimension, 1:8), y = Inertia)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_line(aes(group = 1), colour = "firebrick", linewidth = 1) +
  geom_point(colour = "firebrick", size = 3) +
  scale_y_continuous(labels = percent) +
  labs(
    title    = "Scree Plot: Inertia Explained by Each MCA Dimension",
    subtitle = "Note the 'elbow' around dimension 3–4",
    x        = "Dimension",
    y        = "Inertia Explained (%)"
  )
```

**Rule of thumb**: Retain dimensions until the scree "bends" noticeably or until you've explained 60–80% of total inertia.

## Contributions: What Each Category Contributes to Each Dimension

Here's another crucial concept: **contributions**. For each dimension, we ask: which categories are *pulling* the most in that direction?

For example, if Dimension 1 is a "Espresso vs. Cappuccino" axis, then "Espresso" might contribute 40% to Dimension 1 (it's a major player), and "Morning" might contribute 25%, while "Weekend" contributes only 3%.

Contributions help you *interpret* what each dimension means. If a dimension is driven by categories like "Young," "Urban," "High-Income," you might label it "Urban Affluence."

$$
\text{Contribution of category } i = \frac{(\text{distance from origin}) \times (\text{frequency})}{(\text{inertia of dimension})}
$$

We'll see this in action when we work with real data.

---

# Part 3: Building Intuition with Simulation

Let me create a synthetic dataset where the patterns are *obvious*, so you can see exactly how MCA reveals structure.

## Scenario: The Coffee Shop Personality Types

Imagine a café with three distinct customer archetypes:

1. **Morning Espresso Regulars**: Come on weekdays, always morning, always espresso, often buy pastries.
2. **Afternoon Cappuccino Social Club**: Come on weekends, afternoon, cappuccino, eat lunch.
3. **Late-Night Decaf Readers**: Come evenings, any day, decaf, buy desserts.

We'll simulate 300 customers with these archetypes (with some noise), then show how MCA recovers the structure.

```{r simulate_coffee_data}
#| echo: true
#| fig-cap: "Simulated customer profiles: three archetypes with specific preference patterns."

set.seed(42)

# Archetype 1: Morning Espresso Regulars (100 people)
type1 <- tibble(
  Archetype = "Morning Espresso",
  Coffee = sample(c("Espresso", "Americano"), 100, replace = TRUE, prob = c(0.8, 0.2)),
  Time = sample(c("Morning", "Afternoon"), 100, replace = TRUE, prob = c(0.9, 0.1)),
  Day = sample(c("Weekday", "Weekend"), 100, replace = TRUE, prob = c(0.85, 0.15)),
  Pastry = sample(c("Yes", "No"), 100, replace = TRUE, prob = c(0.7, 0.3))
)

# Archetype 2: Afternoon Cappuccino Social (120 people)
type2 <- tibble(
  Archetype = "Afternoon Cappuccino",
  Coffee = sample(c("Cappuccino", "Latte"), 120, replace = TRUE, prob = c(0.75, 0.25)),
  Time = sample(c("Afternoon", "Morning"), 120, replace = TRUE, prob = c(0.85, 0.15)),
  Day = sample(c("Weekend", "Weekday"), 120, replace = TRUE, prob = c(0.7, 0.3)),
  Pastry = sample(c("Yes", "No"), 120, replace = TRUE, prob = c(0.5, 0.5))
)

# Archetype 3: Late-Night Decaf Readers (80 people)
type3 <- tibble(
  Archetype = "Late-Night Decaf",
  Coffee = sample(c("Decaf", "Tea"), 80, replace = TRUE, prob = c(0.8, 0.2)),
  Time = sample(c("Evening", "Afternoon"), 80, replace = TRUE, prob = c(0.85, 0.15)),
  Day = sample(c("Weekday", "Weekend"), 80, replace = TRUE, prob = c(0.6, 0.4)),
  Pastry = sample(c("Yes", "No"), 80, replace = TRUE, prob = c(0.6, 0.4))
)

# Combine and add ID
coffee_data <- bind_rows(type1, type2, type3) |>
  mutate(ID = row_number()) |>
  select(ID, everything())

# Display summary
cat("Customer Archetypes Summary:\n\n")
coffee_data |>
  group_by(Archetype) |>
  summarise(
    n = n(),
    Coffee = paste(names(sort(table(Coffee), decreasing = TRUE))[1:2], collapse = ", "),
    Time = paste(names(sort(table(Time), decreasing = TRUE))[1:2], collapse = ", "),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Three customer archetypes with distinct preference patterns")

# Show actual distribution
dist_tbl <- coffee_data |>
  count(Archetype, Coffee) |>
  pivot_wider(names_from = Coffee, values_from = n, values_fill = 0)

cat("\nCoffee distribution by archetype:\n")
print(dist_tbl)
```

Now let's apply MCA to this dataset and see if it recovers the three archetypes:

```{r mca_simulation}
#| echo: true
#| fig-cap: "MCA applied to the coffee dataset: the three archetypes naturally separate on the first two dimensions."

# Perform MCA
mca_fit <- MCA(coffee_data |> select(-ID, -Archetype), 
               graph = FALSE, 
               ncp = 5)

# Extract individual factor scores
ind_coords <- mca_fit$ind$coord |>
  as_tibble(rownames = "ID") |>
  rename_with(~gsub(" ", ".", .x, fixed = TRUE)) |>
  mutate(ID = as.numeric(ID)) |>
  left_join(
    coffee_data |> select(ID, Archetype),
    by = "ID"
  )

# Plot individuals colored by archetype
ggplot(ind_coords, aes(x = Dim.1, y = Dim.2, colour = Archetype)) +
  geom_point(size = 2, alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", alpha = 0.5) +
  scale_colour_manual(
    values = c(
      "Morning Espresso" = "#FF6B35",
      "Afternoon Cappuccino" = "#004E89",
      "Late-Night Decaf" = "#1F77B4"
    )
  ) +
  labs(
    title    = "MCA: Coffee Shop Customer Archetypes",
    subtitle = "Three distinct groups emerge naturally on first two dimensions",
    x        = glue("Dimension 1 ({round(mca_fit$eig[1,2], 1)}% of inertia)"),
    y        = glue("Dimension 2 ({round(mca_fit$eig[2,2], 1)}% of inertia)"),
    colour   = "Archetype"
  )
```

Excellent! MCA has automatically separated the three archetypes. Now let's see *why* they've separated — which categories define each dimension:

```{r mca_category_biplot}
#| echo: true
#| fig-cap: "Category biplot: categories that define the archetypes are positioned accordingly."

# FactoMineR stores category coords with raw level names as row names (e.g., "Espresso", "Morning")
# Build a level→variable lookup from the input data
mca_input <- coffee_data |> select(-ID, -Archetype)
level_to_var <- imap_dfr(mca_input, \(col, var) {
  tibble(Variable = var, Level = as.character(unique(col)))
})

# Extract category coordinates
cat_coords <- mca_fit$var$coord |>
  as_tibble(rownames = "Level") |>
  rename_with(~gsub(" ", ".", .x, fixed = TRUE)) |>
  left_join(level_to_var, by = "Level")

# Plot categories
ggplot(cat_coords, aes(x = Dim.1, y = Dim.2, colour = Variable, label = Level)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_text(nudge_x = 0.05, nudge_y = 0.05, size = 3, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", alpha = 0.5) +
  scale_colour_brewer(palette = "Set1") +
  labs(
    title    = "Category Positions: What Defines Each Archetype",
    subtitle = "Categories on the same side of the plot co-occur frequently",
    x        = "Dimension 1",
    y        = "Dimension 2",
    colour   = "Variable"
  ) +
  theme(
    legend.position = "right"
  )
```

**Read this carefully**: The left side is dominated by "Espresso," "Morning," and "Weekday." The right side has "Cappuccino," "Afternoon," and "Weekend." The bottom has "Decaf" and "Evening." This perfectly matches our three archetypes!

This is the power of MCA: it finds *what goes together* without being explicitly told the archetypes exist.

## The Scree Plot: How Much Structure Did We Find?

```{r mca_scree}
#| echo: true
#| fig-cap: "Scree plot of the simulated coffee data: inertia drops off after dimension 2–3, suggesting three main patterns."

scree_tbl <- mca_fit$eig |>
  as_tibble(rownames = "Dimension") |>
  rename(Eigenvalue = 2, InertiaPercent = 3, CumulativePercent = 4) |>
  mutate(
    InertiaPercent = InertiaPercent / 100,
    CumulativePercent = CumulativePercent / 100
  ) |>
  mutate(Dimension = row_number()) |>
  filter(Dimension <= 8)

ggplot(scree_tbl, aes(x = Dimension, y = InertiaPercent)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_line(aes(group = 1), colour = "firebrick", linewidth = 1) +
  geom_point(colour = "firebrick", size = 3) +
  scale_x_continuous(breaks = 1:8) +
  scale_y_continuous(labels = percent) +
  labs(
    title    = "Scree Plot: How Much Variability Does Each Dimension Explain?",
    subtitle = "Strong drop after dimension 2–3",
    x        = "Dimension",
    y        = "Inertia Explained (%)"
  )
```

The scree plot shows that the first two dimensions capture a substantial portion of the structure. This confirms that our three archetypes are *real* patterns, not noise.

---

