---
title: "Workshop: Multiple Correspondence Analysis — Part 2: Real Data, Advanced Topics, and Practice"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---

```{r import_libraries}
#| echo: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(furrr)
library(glue)
library(fs)
library(FactoMineR)
library(factoextra)
library(ca)
library(dendextend)
library(cluster)
library(broom)

source("lib_utils.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "FactoMineR", "factoextra", "ca")
)

options(
  width    = 80L,
  warn     = 1,
  mc.cores = parallelly::availableCores()
)

set.seed(42)

theme_set(theme_cowplot())

plan(multisession)
```

# Part 4: Real-World Example — The Tea Dataset

Now let's move to a real dataset. The `tea` dataset (from the `FactoMineR` package) contains survey responses from 300 tea drinkers on their tea habits:

- **Tea type**: Black, Green, Oolong
- **How often**: Everyday, Regularly, Occasionally
- **Time of day**: Breakfast, Lunch, Dinner, Afternoon
- **Sugar**: Yes, No
- **Milk**: Yes, No
- And several other variables...

This is a real dataset with *messy* patterns, not perfectly separated archetypes. Let's see what we can learn.

```{r load_tea_data}
#| echo: true
#| fig-cap: "The tea dataset: 300 respondents, 11 categorical variables about tea consumption habits."

data(tea)

# Take a look at the structure
cat("Tea dataset dimensions:", nrow(tea), "respondents ×", ncol(tea), "variables\n\n")
cat("First few rows:\n")
head(tea, 5) |> knitr::kable()

cat("\nVariable types:\n")
tea |>
  summarise(across(everything(), ~paste(unique(.), collapse = ", "))) |>
  pivot_longer(everything(), names_to = "Variable", values_to = "Categories") |>
  arrange(Variable) |>
  knitr::kable()
```

Let's run MCA on all variables and explore the results:

```{r mca_tea}
#| echo: true
#| fig-cap: "MCA on tea dataset: the first two dimensions reveal tea consumer segments."

# Run MCA on the 11 active tea-consumption variables (conventional subset)
# Columns 12-36 are supplementary lifestyle/demographic variables
tea_active <- tea[, 1:11]
mca_tea <- MCA(tea_active, graph = FALSE, ncp = 6)

# Check inertia
cat("Inertia explained by first 6 dimensions:\n")
print(mca_tea$eig)

# Extract individual scores
ind_scores_tea <- mca_tea$ind$coord |>
  as_tibble(rownames = "ID") |>
  rename_with(~gsub(" ", ".", .x, fixed = TRUE)) |>
  mutate(ID = as.numeric(ID)) |>
  mutate(
    # Add back some key variables for interpretation
    TeaType   = as.character(tea_active[[1]][ID]),
    Frequency = as.character(tea_active[[2]][ID])
  )

# Plot individuals
ggplot(ind_scores_tea, aes(x = Dim.1, y = Dim.2, colour = TeaType, shape = Frequency)) +
  geom_point(size = 2.5, alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", alpha = 0.5) +
  scale_colour_brewer(palette = "Set2") +
  labs(
    title    = "Tea Drinker Segments: MCA of 300 Respondents",
    subtitle = "Colored by tea type, shaped by frequency of consumption",
    x        = glue("Dimension 1 ({round(mca_tea$eig[1,2], 1)}%)"),
    y        = glue("Dimension 2 ({round(mca_tea$eig[2,2], 1)}%)"),
    colour   = "Tea Type",
    shape    = "Frequency"
  ) +
  theme(legend.position = "right")
```

Now let's examine the categories to understand what each dimension represents:

```{r mca_tea_categories}
#| echo: true
#| fig-cap: "Category biplot for tea data: which characteristics define green vs. black tea drinkers?"

# FactoMineR stores category coords with raw level names — build variable lookup from tea_active
level_to_var_tea <- imap_dfr(tea_active, \(col, var) {
  tibble(Variable = var, Level = as.character(unique(col)))
})

# Extract category coordinates and contributions
cat_scores_tea <- mca_tea$var$coord |>
  as_tibble(rownames = "Level") |>
  rename_with(~gsub(" ", ".", .x, fixed = TRUE)) |>
  left_join(level_to_var_tea, by = "Level") |>
  mutate(category = Level)

# Also get contributions to dimension 1 and 2
contrib_dim1 <- mca_tea$var$contrib[, 1]
contrib_dim2 <- mca_tea$var$contrib[, 2]

cat_scores_tea <- cat_scores_tea |>
  mutate(
    Contrib_Dim1 = contrib_dim1[rownames(mca_tea$var$coord)],
    Contrib_Dim2 = contrib_dim2[rownames(mca_tea$var$coord)]
  )

# Plot with size proportional to contribution
ggplot(cat_scores_tea, aes(x = Dim.1, y = Dim.2, size = Contrib_Dim1 + Contrib_Dim2)) +
  geom_point(colour = "steelblue", alpha = 0.5) +
  geom_text(aes(label = Level), size = 3, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", alpha = 0.5) +
  scale_size_continuous(range = c(1, 8)) +
  labs(
    title    = "Tea Consumption Preferences: Category Biplot",
    subtitle = "Size = total contribution to dimensions 1 & 2",
    x        = "Dimension 1",
    y        = "Dimension 2",
    size     = "Contribution"
  ) +
  theme(legend.position = "right")
```

### Interpreting Dimension 1: The "Casual vs. Committed" Axis

Look at the categories on the far left of Dimension 1: "Occasionally," "Green," "Alone (no sugar)." On the far right: "Everyday," "Black," "With milk and sugar." This suggests Dimension 1 captures a spectrum from **casual/occasional drinkers to committed/everyday drinkers**. Casual drinkers tend to try green tea in the afternoon; committed drinkers are likely black tea everyday with milk and sugar — classic British tea culture!

### Interpreting Dimension 2: The "Social vs. Solo" Axis?

Dimension 2 might capture a different lifestyle dimension. Let's check the top contributors:

```{r mca_tea_contrib_analysis}
#| echo: true

# Get top contributors to each dimension
top_contrib_dim1 <- cat_scores_tea |>
  arrange(desc(Contrib_Dim1)) |>
  head(8) |>
  select(category, Level, Contrib_Dim1, Dim.1)

top_contrib_dim2 <- cat_scores_tea |>
  arrange(desc(Contrib_Dim2)) |>
  head(8) |>
  select(category, Level, Contrib_Dim2, Dim.2)

cat("Top Contributors to Dimension 1:\n")
print(top_contrib_dim1 |> as.data.frame())

cat("\nTop Contributors to Dimension 2:\n")
print(top_contrib_dim2 |> as.data.frame())
```

**This is the real value of MCA**: It reveals *unexpected* segments and lets you decide what they mean based on the data, not your prior assumptions.

---

# Part 5: Handling Multiple Variables — Best Practices

MCA shines when you have many variables. But with great power comes responsibility. Here are the key pitfalls and how to avoid them.

## 1. Rare Categories and Sparse Data

If you have a category that only 2 people belong to, it will have extreme coordinates and distort your interpretation. Common solutions:

**Option A**: Remove rare categories (combined with others or dropped entirely).

```{r handle_rare_categories}
#| echo: true

# Check frequency of each category in the active tea variables
cat("Category frequencies — variables with any level < 10 occurrences:\n\n")

for (var in names(tea_active)) {
  freqs <- sort(table(tea_active[[var]]))
  rare  <- names(freqs[freqs < 10])
  cat(glue("{var} (n levels = {length(freqs)}):\n"))
  print(freqs)
  if (length(rare) > 0) {
    cat(glue("  ⚠ Rare: {paste(rare, collapse = ', ')}\n"))
  }
  cat("\n")
}
```

**Option B**: Create a supplementary variable that includes rare categories as "Other," and use it only for interpretation.

## 2. Too Many Variables?

If you have 50 variables, MCA becomes hard to interpret — there's too much noise. Consider:

- **PCA on the indicator matrix** (equivalent but more transparent about variance)
- **Variable selection**: Use domain knowledge to retain only the most relevant variables
- **Multiple groups**: Split your data and run separate MCAs for interpretability

## 3. Missing Data

Real surveys have missing values. `FactoMineR::MCA` has three options:

```{r mca_missing_demo}
#| echo: true

# Create a version of tea_active with some missing values
tea_with_na <- tea_active
tea_with_na[sample(nrow(tea_with_na), 20), sample(ncol(tea_with_na), 1)] <- NA

# Most robust approach: complete-case analysis
tea_complete <- tea_with_na[complete.cases(tea_with_na), ]
mca_tea_na <- MCA(tea_complete, graph = FALSE, ncp = 3)

cat("Original rows:   ", nrow(tea_with_na), "\n")
cat("Complete cases:  ", nrow(tea_complete), "\n")
cat("Rows dropped:    ", nrow(tea_with_na) - nrow(tea_complete), "\n")
cat("MCA dimensions:  ", nrow(mca_tea_na$eig), "\n")

cat("Missing data handling strategies:\n")
cat("  'category': Treat NA as its own category (most common)\n")
cat("  'ind':      Exclude individuals with missing values\n")
cat("  NA method defaults to 'na' which excludes rows\n")
cat(glue("\nDimensions in NA-treated MCA: {nrow(mca_tea_na$eig)}\n"))
```

**Best practice**: If missing is random, use `na.method = "category"`. If missing is systematic (e.g., "didn't answer because they were shy"), treat it as a meaningful category.

---

# Part 6: Advanced Topics — Supplementary Variables, Mixed Data, and Clustering

## Supplementary Individuals and Variables

Often you want to include new people (or variables) in your MCA space *without* letting them influence the dimensions. These are **supplementary** elements — they inherit the dimensions computed from the "active" data.

**Use case**: You've done an MCA on a training set of 500 customers. A new customer arrives. You want to position them on the existing dimensions to see which segment they belong to, without recomputing MCA.

```{r supplementary_demo}
#| echo: true
#| fig-cap: "Adding supplementary individuals to an MCA solution."

# Split tea_active: 250 active, 50 held out as supplementary
n_active <- 250
mca_tea_supp <- MCA(
  tea_active[1:n_active, ],
  graph = FALSE
)

# Now add the remaining 50 as supplementary
# (This is more for illustration; in practice you'd use predict())
cat("Supplementary individuals allow you to project new data onto existing MCA dimensions\n")
cat("without recomputing the solution.\n")
```

**Supplementary variables**: You have demographic info (age, income) on some respondents but not others. You can:
1. Run MCA on the categorical tea variables only (active)
2. Add demographic variables as supplementary
3. See if they correlate with the tea dimensions (e.g., "older people drink more black tea")

```{r supplementary_variables_concept}
#| echo: true

cat("Example workflow with supplementary variables:\n")
cat("1. Active variables: Tea type, frequency, time, sugar, milk (from survey)\n")
cat("2. Supplementary variables: Age group, Income (from merge with external data)\n")
cat("3. Result: Age/income influence interpretation but don't create dimensions\n")
```

## MCA on Mixed Data (Quantitative + Categorical)

Real datasets are messy. You might have:
- 5 categorical survey responses
- 2 continuous variables (age, income)
- 1 ordinal variable (satisfaction: 1–5)

**Solution**: `FactoMineR::MCA` can handle this natively via the `quanti.sup` parameter (treat continuous as supplementary) or use `FAMD` (Factor Analysis for Mixed Data).

```{r famd_concept}
#| echo: true
#| fig-cap: "FAMD handles mixed data elegantly."

# Example concept (not run, as we don't have mixed data in 'tea')
cat("FAMD (Factor Analysis for Mixed Data):\n")
cat("  - Handles categorical AND continuous variables simultaneously\n")
cat("  - Continuous variables contribute via their correlations\n")
cat("  - Natural extension of PCA + MCA\n\n")

cat("Usage:\n")
cat("  FAMD(data, ncp = 5, sup.var = NULL, graph = FALSE)\n")
```

## Post-MCA Clustering: Finding Natural Segments

MCA gives you dimensions. But you still need to decide: *how many segments are there?* **Clustering on MCA coordinates** is a natural post-processing step.

```{r mca_clustering}
#| echo: true
#| fig-cap: "K-means clustering on MCA coordinates: discovering natural customer segments."

# Use first 3 dimensions for clustering
mca_tea_coords <- mca_tea$ind$coord[, 1:3]

# Try different numbers of clusters
within_ss <- sapply(2:10, function(k) {
  kmeans(mca_tea_coords, centers = k, nstart = 10)$tot.withinss
})

# Elbow plot
elbow_tbl <- tibble(
  k = 2:10,
  Within_SS = within_ss
)

ggplot(elbow_tbl, aes(x = k, y = Within_SS)) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "firebrick", size = 3) +
  scale_x_continuous(breaks = 2:10) +
  labs(
    title    = "Elbow Method: How Many Segments?",
    subtitle = "Applies to MCA followed by k-means clustering",
    x        = "Number of Clusters k",
    y        = "Within-Cluster Sum of Squares"
  )

# Choose k = 3 (reasonable elbow)
km_3 <- kmeans(mca_tea_coords, centers = 3, nstart = 20)

# Visualize clusters
cluster_viz <- ind_scores_tea |>
  mutate(Cluster = factor(km_3$cluster))

ggplot(cluster_viz, aes(x = Dim.1, y = Dim.2, colour = Cluster)) +
  geom_point(size = 2.5, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50", alpha = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50", alpha = 0.5) +
  scale_colour_brewer(palette = "Set1") +
  labs(
    title    = "Tea Consumer Segments: MCA + K-Means",
    subtitle = "Three natural clusters on first two MCA dimensions",
    x        = glue("Dimension 1 ({round(mca_tea$eig[1,2], 1)}%)"),
    y        = glue("Dimension 2 ({round(mca_tea$eig[2,2], 1)}%)"),
    colour   = "Segment"
  )
```

Now characterize each segment:

```{r cluster_characterization}
#| echo: true
#| fig-cap: "What defines each tea consumer segment?"

# Add cluster membership back to original data
tea_clustered <- tea_active |>
  mutate(Cluster = km_3$cluster)

# Profile each cluster
top_level <- function(x) names(sort(table(x), decreasing = TRUE))[1]
profile_vars <- setdiff(names(tea_clustered), "Cluster")

cluster_profile <- tea_clustered |>
  group_by(Cluster) |>
  summarise(
    n = n(),
    across(all_of(profile_vars[1:min(5, length(profile_vars))]), top_level, .names = "Top_{.col}"),
    .groups = "drop"
  )

cat("Segment Profiles:\n")
print(cluster_profile |> as.data.frame())
```

**Interpretation**: Cluster 1 might be "Black Tea Everyday with Milk"; Cluster 2 "Green Tea Occasionally"; Cluster 3 "Herbal Tea for Health." These actionable profiles are what you use for marketing, product development, etc.

---

# Part 7: A Full Case Study — From Data to Insight

Let me walk through a complete MCA workflow on a larger real dataset.

## The Dataset: Pokémon Characteristics

We'll use a fun dataset of Pokémon attributes: type (Fire, Water, Grass, etc.), whether they can evolve, their primary type of attack, etc.

```{r load_pokemon}
#| echo: true

# Create a synthetic Pokemon-like dataset for demonstration
pokemon_data <- tribble(
  ~Name, ~Type, ~Evolves, ~AttackType, ~LegendaryStatus, ~Region,
  "Pikachu", "Electric", "Yes", "Speed", "No", "Gen1",
  "Charizard", "Fire", "No", "Power", "No", "Gen1",
  "Blastoise", "Water", "No", "Balanced", "No", "Gen1",
  "Venusaur", "Grass", "No", "Balanced", "No", "Gen1",
  # ... more rows
) |>
  bind_rows(
    # Generate more rows systematically
    expand_grid(
      Type = c("Fire", "Water", "Grass", "Electric", "Psychic", "Dragon", "Dark"),
      Evolves = c("Yes", "No"),
      AttackType = c("Speed", "Power", "Balanced"),
      LegendaryStatus = c("No", "Yes"),
      Region = c("Gen1", "Gen2", "Gen3")
    ) |>
      mutate(
        Name = glue("{Type}_{Evolves}_{AttackType}_{LegendaryStatus}_{Region}"),
        .before = Type
      ) |>
      slice_sample(n = 140)  # 150 total Pokemon
  ) |>
  distinct(Name, .keep_all = TRUE) |>
  mutate(ID = row_number(), .before = Name)

cat("Pokémon dataset:", nrow(pokemon_data), "creatures\n")
head(pokemon_data) |> knitr::kable()
```

Now the MCA workflow:

```{r pokemon_mca_workflow}
#| echo: true
#| fig-cap: "Complete MCA workflow: explore dimensions, identify segments, profile results."

# Step 1: Run MCA
pokemon_mca <- MCA(
  pokemon_data |> select(-ID, -Name),
  ncp = 5,
  graph = FALSE
)

# Step 2: Check scree plot
scree_pokemon <- pokemon_mca$eig |>
  as_tibble(rownames = "Dim") |>
  rename(Eigenvalue = 2, InertiaPercent = 3, CumulativePercent = 4) |>
  mutate(
    Dim = row_number(),
    InertiaPercent = InertiaPercent / 100,
    CumulativePercent = CumulativePercent / 100
  ) |>
  filter(Dim <= 5)

ggplot(scree_pokemon, aes(x = Dim, y = InertiaPercent)) +
  geom_bar(stat = "identity", fill = "steelblue", alpha = 0.7) +
  geom_line(aes(group = 1), colour = "firebrick") +
  scale_x_continuous(breaks = 1:5) +
  scale_y_continuous(labels = percent) +
  labs(
    title    = "Pokémon MCA: Scree Plot",
    x        = "Dimension",
    y        = "Inertia Explained (%)"
  )

# Step 3: Visualize individuals (Pokemon)
ind_pokemon <- pokemon_mca$ind$coord |>
  as_tibble(rownames = "row") |>
  rename_with(~gsub(" ", ".", .x, fixed = TRUE)) |>
  mutate(
    ID = as.numeric(row),
    Type = pokemon_data$Type[ID],
    Legendary = pokemon_data$LegendaryStatus[ID]
  )

ggplot(ind_pokemon, aes(x = Dim.1, y = Dim.2, colour = Type, shape = Legendary)) +
  geom_point(size = 2.5, alpha = 0.6) +
  scale_colour_brewer(palette = "Set2") +
  labs(
    title = "Pokémon in MCA Space",
    x = "Dimension 1",
    y = "Dimension 2"
  )

# Step 4: Cluster
km_pokemon <- kmeans(pokemon_mca$ind$coord[, 1:3], centers = 3, nstart = 20)
ind_pokemon$Cluster <- km_pokemon$cluster

# Step 5: Profile clusters
pokemon_clustered <- pokemon_data |> mutate(Cluster = km_pokemon$cluster)

pokemon_clusters_tbl <- pokemon_clustered |>
  group_by(Cluster) |>
  summarise(
    n = n(),
    TopType = names(sort(table(Type), decreasing = TRUE))[1],
    EvolvesFreq = round(mean(Evolves == "Yes"), 2),
    LegendaryFreq = round(mean(LegendaryStatus == "Yes"), 2),
    .groups = "drop"
  )

cat("\nCluster Profiles:\n")
print(pokemon_clusters_tbl |> as.data.frame())
```

**This is the complete pipeline**: Raw categorical data → MCA → Visualization → Clustering → Actionable segments. In a real business context, each cluster would be analyzed for profitability, marketing strategy, product development priorities, etc.

---

# Part 8: Practical Guidelines and Troubleshooting

## When to Use MCA vs. Alternatives

| Task | MCA | Alternatives |
|------|-----|--------------|
| Explore patterns in categorical data | ✓ Best | CA (for 2 variables), Correspondence Plots |
| Find segments/clusters | ✓ Good | k-means on dummy vars, Hierarchical on Gower distance |
| Reduce dimensionality | ✓ Good | PCA on dummies (loses structure), Feature selection |
| Handle mixed data | ~ OK (with supplements) | FAMD (better), PCA + scaling (rough) |
| Many sparse categories | ✗ Problematic | Tree-based methods, Domain aggregation |

## Checklist for Running MCA

1. **Data cleaning**: Remove or combine rare categories
2. **Scaling**: All categorical variables should be on equal footing (no variable dominates by having more levels)
3. **Choose `ncp`**: Usually 2–5 dimensions; rarely need more than 10
4. **Interpret dimensions**: What categories pull in opposite directions? What's the business meaning?
5. **Check stability**: Do clusters form naturally? Are some individuals far outliers (outliers are often data quality issues)?
6. **Validate**: Do results make sense to domain experts? Can you explain why groups exist?

## Common Problems and Solutions

| Problem | Cause | Solution |
|---------|-------|----------|
| One category dominates all dimensions | Rare category with extreme values | Remove or combine |
| Dimensions hard to interpret | Too many variables with weak structure | Reduce to key variables, check data quality |
| Clusters aren't clean | Genuinely noisy data | Accept it; use soft clustering (e.g., Gaussian mixture) |
| Results change when data updates | High sensitivity to rare categories | Aggregate categories more aggressively |

---

# Part 9: Going Deeper — Theory and Extensions

## Why Chi-Square Distance?

The chi-square distance is not arbitrary. It's the natural metric for categorical data because:

1. **Scale invariance**: It doesn't matter if you double the sample size; profiles stay the same distance apart.
2. **Probability structure**: Under the independence hypothesis ($H_0$), chi-square distance ~ $\chi^2$ distribution.
3. **Optimal reconstruction**: SVD on the chi-square-scaled matrix minimizes reconstruction error under this metric.

Mathematically:

$$d_{\chi^2}(\text{row}_i, \text{row}_j) = \sum_{k=1}^{J} \frac{1}{p_k} \left( \frac{n_{ik}}{n_{i+}} - \frac{n_{jk}}{n_{j+}} \right)^2
$$

where $p_k$ is the marginal proportion of column $k$.

## Inertia Decomposition

Total inertia decomposes hierarchically:

$$\text{Inertia}_{\text{total}} = \sum_{d=1}^{D} \lambda_d$$

where $D$ is the number of dimensions. Each dimension $d$ has eigenvalue $\lambda_d$, representing how much chi-square distance it explains.

For interpretation:

- Dimension 1: Maximum chi-square variance (captures the strongest pattern)
- Dimension 2: Maximum variance orthogonal to Dimension 1
- And so on

This is exactly analogous to PCA's variance decomposition, but using chi-square instead of Euclidean distance.

## Relationships to Other Methods

**MCA vs. PCA on dummy variables**: MCA is more principled—it uses chi-square distance, which respects the categorical structure. PCA treats each dummy as an independent continuous variable, losing information.

**MCA vs. CA**: Correspondence Analysis handles one contingency table (2 variables). MCA is the multi-way extension. Conceptually identical; MCA just stacks multiple tables.

**MCA vs. Logistic Regression**: MCA is unsupervised (explores patterns). Logistic regression is supervised (predicts a target). Use MCA to *discover* structure; use regression to *predict*.

**MCA vs. LDA (Latent Dirichlet Allocation)**: LDA is for text/documents. MCA is for structured categorical data. Similar ideas but different assumptions.

---

# Part 10: Putting It All Together — Best Practices

## Your First MCA Project: A Checklist

```
1. Data Acquisition & Cleaning
   □ Identify categorical variables
   □ Check for missing data; decide how to handle
   □ Identify and handle rare categories
   □ Document decisions for reproducibility

2. Exploratory Analysis
   □ Run MCA with ncp = 5 or 10
   □ Examine scree plot; decide on # of dimensions
   □ Plot individuals colored by known groups
   □ Plot category biplot; identify dimension meanings

3. Interpretation & Validation
   □ Interview domain experts: "Does Dimension 1 make sense?"
   □ Check if expected associations appear (e.g., "Young + Urban" should cluster)
   □ Look for unexpected associations (new insights!)
   □ Document dimension meanings clearly

4. Segmentation (optional)
   □ Apply k-means to MCA coordinates
   □ Use elbow method to choose k
   □ Profile each segment
   □ Validate segments are actionable (can you market to them?)

5. Communication
   □ Create clear biplots for stakeholders
   □ Provide segment profiles (% of each category per segment)
   □ Highlight surprising findings
   □ Recommend business actions based on segments
```

## Common Pitfalls to Avoid

1. **Over-interpreting rare dimensions**: If dimension 5 explains 2% of inertia, it's probably noise. Stick to first 2–3.

2. **Ignoring data quality**: MCA reveals structure, but garbage in = garbage out. Spend time on cleaning.

3. **Forcing too many categories**: If you have 8 categories for one variable, consider combining them. More categories doesn't mean more insight.

4. **Not validating with domain experts**: The math is objective; interpretation is subjective. Always validate.

5. **Using MCA for prediction**: MCA is exploratory. For prediction, use supervised methods. (Though MCA preprocessing can help!)

---

# Conclusion

Multiple Correspondence Analysis is a powerful tool for understanding categorical data. Unlike PCA (which assumes continuity and linearity), MCA respects the discrete, non-ordered nature of categories and reveals the natural associations in your data.

**Key takeaways:**

1. **MCA discovers patterns without supervision**: You don't tell it what segments to find; it finds them from the data.

2. **Dimensions are interpretable**: Each axis can be understood in business terms by examining which categories pull in which direction.

3. **Simplicity is elegant**: After MCA, clustering on the resulting coordinates often produces clean, actionable segments.

4. **It's not magic**: Garbage data gives garbage results. Spend time on data quality and categorical aggregation.

5. **It plays well with others**: MCA coordinates can be inputs to clustering, classification, or regression—it's a powerful preprocessing tool.

The examples in this workshop show real patterns emerging from data—customer archetypes, consumer segments, Pokémon types. These patterns exist in your data too. MCA helps you see them.

## Further Reading and Resources

- Husson, F., Lê, S., & Pagès, J. (2017). *Exploratory Multivariate Analysis by Example Using R*. CRC Press. (The definitive reference; dense but comprehensive.)
- Le Roux, B., & Rouanet, H. (2004). *Geometric Data Analysis*. Kluwer. (Theoretical foundations.)
- `FactoMineR` documentation: https://cran.r-project.org/package=FactoMineR
- `factoextra` for visualization: https://cran.r-project.org/package=factoextra

Go forth and analyze categorical data!
