---
title: "Dublin Data Science Workshop on Time Series Analysis"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Monday, 21 January 2019"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float:
      smooth_scroll: FALSE
  pdf_document: default
---

```{r import_packages, echo=FALSE}


```


```{r knit_opts, include = FALSE}
knitr::opts_chunk$set(tidy  = FALSE
                     ,cache = FALSE
                     ,message = FALSE
                     ,warning = FALSE
                     ,fig.height =  8
                     ,fig.width  = 11)

options(width = 80L
       ,warn  = 1
        )

set.seed(42)


```



# Basic Concepts

Time series occur in almost any field of study that produces quantitative data.
Whenever quantities are measured over time, those measurements form a
time-series, or more formally, a discrete-time stochastic process.

One famous example of a time-series is count of airline passengers in the US,
as shown in the figure below. This is a simple time-series, with
measurements taken on a monthly basis over a number of years, each datum
consisting of a single number - i.e. this time-series is univariate.

```{r plot_air_passenger_timeseries, echo=TRUE}
plot(AirPassengers)
```

Before we begin trying to analyse data such as this, we need to first create a
mathematical framework to work in. Fortunately, we do not need anything too
complicated, and for a finite time-series of length $N$, we model the time
series as a sequence of $N$ random variables, $X_i$, with $i = 1, 2, ..., N$.

Realise that each individual $X_i$ is a wholly separate random variable --
analysing time series statistically is unusual as we only ever have a single
measurement for each random variable from which we can do inference. In many
cases we simplify this much further, but it is important to understand and
appreciate that such simplifications are just that, and is often the reason why
time series are difficult to analyse.

Before we get to any of that though, and before we try to build any kind of
models for the data, we always start with visualising the data. Often, a simple
plot of the data helps use pick out aspects to analyse and incorporate into the
models. For time series, one of the first things to do is the *time plot*, a
simple plot of the data over time.

For the passenger data, a few aspects stand out that are very common in time
series. It is apparent that the numbers increase over time, and this systematic
change in the data is called the *trend*. Often, approximating the trend as a
linear function of time is adequate for many data sets.

A repeating pattern in the data that occurs over the period of the data (in
this case, each year), is called the *seasonal variation*, though a more
general concept of 'season' is implied --- it often will not coincide with the
seasons of the calendar.

A slightly more generalised concept from the seasonality is that of *cycles*,
repeating patterns in the data that do not correspond to the natural fixed
periods of the model. None of these are apparent in the air passenger data, and
accounting for them are beyond the scope of this introductory tutorial.

Finally, another important benefit of visualising the data is that it helps
identify possible *outliers* and *erroneous* data.





# R Environment

```{r show_session_info, echo=TRUE, message=TRUE}
devtools::session_info()
```
