---
title: "Exploring the P/NBD Model on the Synthetic Dataset"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: TRUE
    code_folding: hide
    fig_caption: TRUE

  html_document:
    fig_caption: yes
    theme: spacelab #sandstone #spacelab #flatly
    highlight: pygments
    number_sections: TRUE
    toc: TRUE
    toc_depth: 2
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---


```{r import_libraries, echo=FALSE, message=FALSE}
library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(directlabels)
library(magrittr)
library(rlang)
library(fs)
library(purrr)
library(furrr)
library(glue)
library(cmdstanr)
library(brms)
library(posterior)
library(bayesplot)
library(tidybayes)


source("lib_utils.R")
source("lib_btyd.R")


conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "MASS",
    "fitdistrplus")
  )


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallel::detectCores()
  )

theme_set(theme_cowplot())

set.seed(42)

plan(multisession)
```

In this workbook we introduce the various different BTYD models, starting with
a discussion of the underlying theory.



# Fit Synthetic Data Model

We take our transaction data and transform the data into a format amenable for
the BTYD models.

We first ensure that all transactions on the same day or combined into a single
transaction, and then convert this data into the $(x, t_x, T)$ triplet-format.

To mimic the training / test split to help validate real-world models, we do
the same thing here.

```{r set_training_data_date, echo=TRUE}
synth_training_date <- as.Date("2016-12-31")
```

We also wish to load the generated data.

```{r load_parameter_data, echo=TRUE}
customer_synth_cohort_tbl <- read_rds("data/customer_synth_050000_cohort_tbl.rds")
customer_synth_cohort_tbl %>% glimpse()

customer_synth_simparams_tbl <- read_rds("data/customer_synth_050000_simparams_tbl.rds")
customer_synth_simparams_tbl %>% glimpse()

customer_synth_transactions_tbl <- read_rds("data/customer_synth_050000_transactions_tbl.rds")
customer_synth_transactions_tbl %>% glimpse()
```


We now convert the transaction data into the CBS format as an input the BTYD
models, aggregating the CBS data to construct our CLV input data.


```{r construct_clvinput_data, echo=TRUE}
customer_synth_cbs_tbl <- customer_synth_transactions_tbl %>%
  calculate_transaction_cbs_data(
    last_date = synth_training_date
    ) %>%
  filter(
    first_tnx_date <= synth_training_date,
    x > 0
    )

customer_synth_cbs_tbl %>% glimpse()
```

We now write that CBS data to disk for later use.

```{r write_synth_cbs_disk, echo=TRUE}
customer_synth_cbs_tbl %>% write_rds("data/customer_synth_cbs_tbl.rds")
```

We may want to fit our models on reduced datasets, so we construct a sample of
5,000 customers to use for our model, and save those `customer_id` to ensure
we use the same reduced dataset each time we fit our models.

```{r sample_reduced_customerid_values, echo=TRUE}
reduced_ids <- customer_synth_cbs_tbl %>%
  slice_sample(n = 5000) %>%
  arrange(first_tnx_date) %>%
  pull(customer_id)

reduced_ids %>% glimpse()
```


## Construct Fixed Hyper-parameter Model

```{r compile_pnbd_synth_fixed_model, echo=TRUE}
pnbd_synth_fixed_stanmodel <- cmdstan_model(
  "stan_code/clv_pnbd_fixed.stan",
  include_paths = "stan_code/",
  pedantic      =  TRUE,
  dir           = "stan_models/"
  )
```


```{r fit_pnbd_synth_fixed_stan_model, echo=TRUE}
stan_data_lst <- customer_synth_cbs_tbl %>%
  filter(customer_id %in% reduced_ids) %>%
  select(-first_tnx_date, -last_tnx_date) %>%
  compose_data(
    r     =  1,
    alpha =  3,
    
    s     =  1,
    beta  = 10,
    )

pnbd_synth_fixed_stanfit <- pnbd_synth_fixed_stanmodel$sample(
  data            =                stan_data_lst,
  chains          =                            4,
  iter_warmup     =                         1000,
  iter_sampling   =                         1000,
  seed            =                          421,
  output_dir      =                "stan_models",
  output_basename =           "pnbd_synth_fixed"
  )
```

We first want to check the HMC diagnostics.

```{r calculate_pnbd_synth_fixed_hmc_diagnostics, echo=TRUE}
pnbd_synth_fixed_stanfit$cmdstan_diagnose()
```



We use the functionality given by `tidybayes` to extract the posterior
distribution of the parameters.

```{r extract_synth_pnbd_posterior_params, echo=TRUE}
customer_synth_pnbd_fixed_draws <- pnbd_synth_fixed_stanfit$draws()

customer_synth_pnbd_fixed_posterior_tbl <- customer_synth_pnbd_fixed_draws %>%
  recover_types(customer_synth_cbs_tbl) %>%
  spread_draws(mu[customer_id], lambda[customer_id]) %>%
  ungroup()

customer_synth_pnbd_fixed_posterior_tbl %>% glimpse()
```


```{r construct_synth_pnbd_posterior_validation, echo=TRUE}
customer_posterior_tbl <- customer_synth_pnbd_fixed_posterior_tbl %>%
  group_by(customer_id) %>%
  summarise(
    posterior_mu     = list(mu),
    posterior_lambda = list(lambda)
    )

customer_postcheck_tbl <- customer_synth_simparams_tbl %>%
  select(customer_id, customer_mu, customer_lambda, customer_tau) %>%
  inner_join(customer_posterior_tbl, by = "customer_id") %>%
  mutate(
    post_mu_qval     = map2_dbl(customer_mu,     posterior_mu,     ~ ecdf(.y)(.x)),
    post_lambda_qval = map2_dbl(customer_lambda, posterior_lambda, ~ ecdf(.y)(.x))
    )

customer_postcheck_tbl %>% glimpse()
```

We now plot the uniformity of the data.

```{r plot_bsic_pnbd_uniformity, echo=TRUE}
ggplot(customer_postcheck_tbl) +
  geom_histogram(aes(x = post_mu_qval), bins = 50) +
  labs(
    x = expression(mu),
    y = "Count",
    title = "Plot of Customer q-values for mu"
    )


ggplot(customer_postcheck_tbl) +
  geom_histogram(aes(x = post_lambda_qval), bins = 50) +
  labs(
    x = expression(lambda),
    y = "Count",
    title = "Plot of Customer q-values for lambda"
    )
```

As we see, our posterior does a reasonable job of discovering the dropout rate
$\mu$ but has a low bias on the transaction rate $\lambda$.


```{r plot_posterior_params_facets, echo=TRUE}
postcheck_facet_plotdata_tbl <- customer_synth_pnbd_fixed_posterior_tbl %>%
  ungroup() %>%
  group_nest(customer_id) %>%
  slice_sample(n = 20) %>%
  unnest(data) %>%
  inner_join(customer_synth_simparams_tbl, by = "customer_id")

ggplot(postcheck_facet_plotdata_tbl) +
  geom_histogram(aes(x = mu), bins = 50) +
  geom_vline(aes(xintercept = customer_mu), colour = "red") +
  facet_wrap(facets = "customer_id", scales = "free")
```


### Write Model to Disk

```{r write_pnbd_synth_fixed_disk, echo=TRUE}
pnbd_synth_fixed_stanmodel %>% write_rds("data/pnbd_synth_fixed_stanmodel.rds")
pnbd_synth_fixed_stanfit   %>% write_rds("data/pnbd_synth_fixed_stanfit.rds")
```

```{r, echo=FALSE}
knitr::knit_exit()
```


## Construct Hyper-parameter Model

We now want to improve the model by adding priors around the hyper-parameters
for how the $\mu$ and $\lambda$ parameters are distributed. To do this, we
try using Gamma distributed hyper-priors.

```{r compile_synth_prior_model, echo=TRUE}
pnbd_synth_prior_stanmodel <- cmdstan_model(
  "stan_code/clv_pnbd_prior.stan",
  include_paths = "stan_code/",
  pedantic      =  TRUE,
  dir           = "stan_models/"
  )
```


```{r fit_synth_prior_stan_model, echo=TRUE, warning=TRUE}
stan_data_lst <- customer_synth_cbs_tbl %>%
  filter(customer_id %in% reduced_ids) %>%
  select(-first_tnx_date, -last_tnx_date) %>%
  compose_data(
    lb_mean_a = -1.6000,   lb_mean_b =  1.0000,
    lb_cov_a  = -2.1000,   lb_cov_b  =  1.0000,

    mu_mean_a = -0.0200,   mu_mean_b =  0.2000,
    mu_cov_a  = -0.0002,   mu_cov_b  =  0.0200
    )

pnbd_synth_prior_stanfit <- pnbd_synth_prior_stanmodel$sample(
  data            =                stan_data_lst,
  chains          =                            4,
  iter_warmup     =                         1000,
  iter_sampling   =                         1000,
  seed            =                          422,
  output_dir      =                "stan_models",
  output_basename =           "pnbd_synth_prior"
  )

pnbd_synth_prior_stanfit$summary(
  c("lb_mean", "mu_mean", "mu_cov", "r", "alpha", "s", "beta")
  )
```


We first want to check the HMC diagnostics.

```{r calculate_synth_prior_hmc_diagnostics, echo=TRUE, message=TRUE}
pnbd_synth_prior_stanfit$cmdstan_diagnose()
```



```{r extract_synth_pnbd_prior_posterior_params, echo=TRUE}
customer_synth_pnbd_prior_draws <- pnbd_synth_prior_stanfit$draws()

customer_synth_pnbd_prior_posterior_tbl <- customer_synth_pnbd_prior_draws %>%
  recover_types(customer_synth_cbs_tbl) %>%
  spread_draws(mu[customer_id], lambda[customer_id]) %>%
  ungroup()

customer_synth_pnbd_prior_posterior_tbl %>% glimpse()
```


## Fit Sample with Altenative Prior Parameters

```{r fit_synth_prior_altprior_stan_model, echo=TRUE, warning=TRUE}
stan_data_lst <- customer_synth_cbs_tbl %>%
  filter(customer_id %in% reduced_ids) %>%
  select(-first_tnx_date, -last_tnx_date) %>%
  compose_data(
    lb_mean_a = -1.6000,   lb_mean_b =  1.0000,
    lb_cov_a  = -2.1000,   lb_cov_b  =  1.0000,

    mu_mean_a = -0.0500,   mu_mean_b =  1.0000,
    mu_cov_a  = -0.1250,   mu_cov_b  =  0.5000
    )

pnbd_synth_prior_altprior_stanfit <- pnbd_synth_prior_stanmodel$sample(
  data            =                stan_data_lst,
  chains          =                            4,
  iter_warmup     =                         1000,
  iter_sampling   =                         1000,
  seed            =                          422,
  output_dir      =                "stan_models",
  output_basename =  "pnbd_synth_prior_altprior"
  )

pnbd_synth_prior_altprior_stanfit$cmdstan_diagnose()

pnbd_synth_prior_altprior_stanfit$summary(
  c("lb_mean", "mu_mean", "mu_cov", "r", "alpha", "s", "beta")
  )
```


## Construct Fixed-Coefficient of Variance Model

In this model we fix the coefficient of variation for the hyperparameters and
model the means - we then sample fixing different values for the coefficient of
variance.

```{r compile_synth_prior_fixedcov_model, echo=TRUE}
pnbd_synth_fixedcov_stanmodel <- cmdstan_model(
  "stan_code/clv_pnbd_fixedcov.stan",
  include_paths = "stan_code/",
  pedantic      =  TRUE,
  dir           = "stan_models/"
  )
```


### CoV = 1.0

```{r fit_synth_fixedcov10_stan_model, echo=TRUE, warning=TRUE}
stan_data_lst <- customer_synth_cbs_tbl %>%
  filter(customer_id %in% reduced_ids) %>%
  select(-first_tnx_date, -last_tnx_date) %>%
  compose_data(
    mu_cov = 1.0,
    lb_cov = 1.0
    )

pnbd_synth_fixedcov10_stanfit <- pnbd_synth_fixedcov_stanmodel$sample(
  data            =                stan_data_lst,
  chains          =                            4,
  iter_warmup     =                         1000,
  iter_sampling   =                         1000,
  seed            =                          423,
  output_dir      =                "stan_models",
  output_basename =      "pnbd_synth_fixedcov10",
  )

pnbd_synth_fixedcov10_stanfit$summary(
  c("lb_mean", "mu_mean", "r", "alpha", "s", "beta")
  )
```

We first want to check the HMC diagnostics.

```{r calculate_synth_fixedcov10_hmc_diagnostics, echo=TRUE, warning=TRUE}
pnbd_synth_fixedcov10_stanfit$cmdstan_diagnose()
```




### CoV = 2.0

We now refit the model with a higher level of coefficient-of-variance.

```{r fit_synth_fixedcov20_stan_model, echo=TRUE, warning=TRUE}
stan_data_lst <- customer_synth_cbs_tbl %>%
  select(-first_tnx_date, -last_tnx_date) %>%
  compose_data(
    mu_cov = 2.0,
    lb_cov = 2.0
    )

pnbd_synth_fixedcov20_stanfit <- pnbd_synth_fixedcov_stanmodel$sample(
  data            =                stan_data_lst,
  chains          =                            4,
  iter_warmup     =                         1000,
  iter_sampling   =                         1000,
  seed            =                          424,
  output_dir      =                "stan_models",
  output_basename =      "pnbd_synth_fixedcov20"
  )

pnbd_synth_fixedcov20_stanfit$summary(
  c("lb_mean", "mu_mean", "r", "alpha", "s", "beta")
  )
```



We first want to check the HMC diagnostics.

```{r calculate_synth_fixedcov_hmc_diagnostics, echo=TRUE, warning=TRUE}
pnbd_synth_fixedcov20_stanfit$cmdstan_diagnose()
```


### CoV = 0.5

```{r fit_synth_fixedcov05_stan_model, echo=TRUE, warning=TRUE}
stan_data_lst <- customer_synth_cbs_tbl %>%
  select(-first_tnx_date, -last_tnx_date) %>%
  compose_data(
    mu_cov = 0.5,
    lb_cov = 0.5
    )

pnbd_synth_fixedcov05_stanfit <- pnbd_synth_fixedcov_stanmodel$sample(
  data            =                stan_data_lst,
  chains          =                            4,
  iter_warmup     =                         1000,
  iter_sampling   =                         1000,
  seed            =                          425,
  output_dir      =                "stan_models",
  output_basename =      "pnbd_synth_fixedcov05"
  )

pnbd_synth_fixedcov05_stanfit$summary(
  c("lb_mean", "mu_mean", "r", "alpha", "s", "beta")
  )
```



We first want to check the HMC diagnostics.

```{r calculate_synth_fixedcov05_hmc_diagnostics, echo=TRUE, warning=TRUE}
pnbd_synth_fixedcov05_stanfit$cmdstan_diagnose()
```


### Compare CoV Samples

We now combine all the posterior data and compare the parameter estimates for
each of the different parameter values.

```{r compare_different_posterior_parameters, echo=TRUE}
post05_customer_tbl <- pnbd_synth_fixedcov05_stanfit$draws() %>%
  recover_types(customer_synth_cbs_tbl) %>%
  spread_draws(mu[customer_id], lambda[customer_id]) %>%
  ungroup()


post10_customer_tbl <- pnbd_synth_fixedcov10_stanfit$draws() %>%
  recover_types(customer_synth_cbs_tbl) %>%
  spread_draws(mu[customer_id], lambda[customer_id]) %>%
  ungroup()


post20_customer_tbl <- pnbd_synth_fixedcov20_stanfit$draws() %>%
  recover_types(customer_synth_cbs_tbl) %>%
  spread_draws(mu[customer_id], lambda[customer_id]) %>%
  ungroup()


post_custdata_tbl <- list(
    `CoV 0.5` = post05_customer_tbl,
    `CoV 1.0` = post10_customer_tbl,
    `CoV 2.0` = post20_customer_tbl
    ) %>%
  bind_rows(.id = "label")

post_custdata_tbl %>% glimpse()
```

```{r plot_sample_customer_boxplots, echo=TRUE}
qn_prob  <- c(0.05, 0.10, 0.25, 0.50, 0.75, 0.90, 0.95)
qn_label <- sprintf("p%02d", qn_prob * 100)

summdata_tbl <- post_custdata_tbl %>%
  group_by(label, customer_id) %>%
  summarise(
    .groups = "drop",
    
    qn     = glue("{qn_label}"),
    mu     = quantile(mu,     probs = qn_prob),
    lambda = quantile(lambda, probs = qn_prob)
    )

plot_tbl <- summdata_tbl %>%
  group_nest(customer_id, .key = "cust_data") %>%
  slice_sample(n = 20) %>%
  unnest(cust_data) %>%
  pivot_wider(
    id_cols = c(label, customer_id),
    names_from = "qn",
    values_from = c(mu, lambda)
    )

cust_data_tbl <- customer_synth_simparams_tbl %>%
  semi_join(plot_tbl, by = "customer_id")


ggplot() +
  geom_errorbar(
    aes(x = customer_id, ymin = mu_p25, ymax = mu_p75, colour = label),
    data = plot_tbl, width = 0, size = 2.0, position = position_dodge(width = 0.5)) +
  geom_errorbar(
    aes(x = customer_id, ymin = mu_p10, ymax = mu_p90, colour = label),
    data = plot_tbl, width = 0, size = 0.5, position = position_dodge(width = 0.5)) +
  geom_point(
    aes(x = customer_id, y = customer_mu),
    data = cust_data_tbl, size = 2) +
  labs(
    x      = "Customer ID",
    y      = "Posterior mu",
    colour = "Data"
    ) +
  ggtitle("Comparison Plot of Posterior Parameters") +
  theme(axis.text.x = element_text(angle = 20, vjust = 0.5))
```


```{r}
# pnbd_synth_fixedcov_stanmodel %>% write_rds("data/pnbd_synth_fixedcov_stanmodel.rds")
# pnbd_synth_fixedcov05_stanfit %>% write_rds("data/pnbd_synth_fixedcov05_stanfit.rds")
# pnbd_synth_fixedcov10_stanfit %>% write_rds("data/pnbd_synth_fixedcov10_stanfit.rds")
# pnbd_synth_fixedcov20_stanfit %>% write_rds("data/pnbd_synth_fixedcov20_stanfit.rds")
```



## Construct Fixed Single CoV Parameter Model


```{r compile_synth_fixedlbcov_model, echo=TRUE}
pnbd_synth_fixedlbcov_stanmodel <- cmdstan_model(
  "stan_code/clv_pnbd_fixedlbcov.stan",
  include_paths = "stan_code/",
  pedantic      =  TRUE,
  dir           = "stan_models/"
  )
```


```{r fit_synth_fixedlbcov_stan_model, echo=TRUE}
stan_data_lst <- customer_synth_cbs_tbl %>%
  select(-first_tnx_date, -last_tnx_date) %>%
  compose_data(
    lb_cov = 1
    )

pnbd_synth_fixedlbcov_stanfit <- pnbd_synth_fixedlbcov_stanmodel$sample(
  data          =  stan_data_lst,
  chains        =              4,
  iter_warmup   =           1000,
  iter_sampling =           1000,
  seed          =            426
  )

pnbd_synth_fixedlbcov_stanfit$summary(
  c("lb_mean", "mu_mean", "mu_cov", "r", "alpha", "s", "beta")
  )
```


We first want to check the HMC diagnostics.

```{r calculate_synth_fixedlbcov_hmc_diagnostics, echo=TRUE}
pnbd_synth_fixedlbcov_stanfit$cmdstan_diagnose()
```



# Compare Posterior Distributions


```{r construct_posterior_ecdf_function, echo=TRUE}
construct_posterior_ecdf_function <- function(customer_params_tbl) {
  calculate_data_probs <- function(posterior_tbl) {
    customer_posterior_tbl <- posterior_tbl %>%
      group_by(customer_id) %>%
      summarise(
        posterior_mu     = list(mu),
        posterior_lambda = list(lambda)
        )

    customer_checkdata_tbl <- customer_params_tbl %>%
      select(customer_id, customer_mu, customer_lambda, customer_tau) %>%
      inner_join(customer_posterior_tbl, by = "customer_id") %>%
      mutate(
        post_mu_qval     = map2_dbl(customer_mu,     posterior_mu,     ~ ecdf(.y)(.x)),
        post_lambda_qval = map2_dbl(customer_lambda, posterior_lambda, ~ ecdf(.y)(.x))
        ) %>%
      select(customer_id, customer_mu, customer_lambda, post_mu_qval, post_lambda_qval)

    return(customer_checkdata_tbl)      
  }
  
  return(calculate_data_probs)
}
```

```{r calculate_posterior_uniform_probabilities, echo=TRUE}
calculate_posterior_data_probs <- customer_synth_simparams_tbl %>%
  construct_posterior_ecdf_function()

posterior_uniformcheck_tbl <- list(
    fixed = customer_synth_pnbd_fixed_posterior_tbl,
    prior = customer_synth_pnbd_prior_posterior_tbl,
    synth = customer_synth_pnbd_prior_synth_posterior_tbl,
    cov05 = post05_customer_tbl,
    cov10 = post10_customer_tbl,
    cov20 = post20_customer_tbl
    ) %>%
  enframe(name = "label", value = "post_data") %>%
  mutate(
    check_data = map(post_data, calculate_posterior_data_probs)
    )

plotdata_tbl <- posterior_uniformcheck_tbl %>%
  select(label, check_data) %>%
  unnest(check_data)

ggplot(plotdata_tbl) +
  geom_histogram(aes(x = post_lambda_qval), bins = 50) +
  facet_wrap(vars(label), scale = "free") +
  xlab("Cumulative Probability") +
  ggtitle("Cumulative Probability of lambda Values")

ggplot(plotdata_tbl) +
  geom_histogram(aes(x = post_mu_qval), bins = 50) +
  facet_wrap(vars(label), scale = "free") +
  xlab("Cumulative Probability") +
  ggtitle("Cumulative Probability of mu Values")
```




# R Environment

```{r show_session_info, echo=TRUE, message=TRUE}
options(width = 120L)
sessioninfo::session_info()
options(width = 80L)
```
