---
title: "UCC Data & Analytics Society: Retrieving Data"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Wednesday, November 3 2021"
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: TRUE
    code_folding: hide
    fig_caption: TRUE

  html_document:
    fig_caption: yes
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---


```{r knit_opts, include = FALSE}
library(conflicted)
library(tidyverse)
library(magrittr)
library(rlang)
library(furrr)
library(cowplot)
library(scales)
library(glue)
library(DT)
library(tidygraph)


resolve_conflicts <- function(pkg_priority) {
  get_index <- function(pkg_name) {
    idx <- str_which(pkg_priority, pkg_name)

    if(length(idx) == 0) {
      idx <- 0L
    }

    return(idx)
  }

  conflict_lst <- conflict_scout()

  for(func_name in names(conflict_lst)) {
    pkg_index <- map_int(conflict_lst[[func_name]], get_index)

    pkg_index <- pkg_index[pkg_index > 0]

    if(length(pkg_index) == 0) {
      pkg_use <- conflict_lst[[func_name]][1]
    } else {
      pkg_use <- pkg_index %>%
        min() %>%
        pkg_priority[.]

    }

    conflict_prefer(func_name, pkg_use)
  }

  return(conflict_lst)
}

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2")
  )


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallel::detectCores()
  )

set.seed(42)

theme_set(theme_cowplot())

plan(multisession)
```



---


All code and data for this workshop is available at the following URL:

https://github.com/kaybenleroll/data_workshops

Code is available in the `ws_ucc_intro_202111/` directory.


# Load Data

```{r load_dota2_data, echo=TRUE}
matchdata_tbl <- read_rds("data/matchdata_wide_tbl.rds")
matchdata_tbl %>% glimpse()

match_teamdata_tbl <- read_rds("data/match_teamdata_tbl.rds")
match_teamdata_tbl %>% glimpse()

hero_data_tbl <- read_rds("data/hero_data_tbl.rds")
hero_data_tbl %>% glimpse()

cluster_region_tbl <- read_rds("data/cluster_region_tbl.rds")
cluster_region_tbl %>% glimpse()
```



## Construct Fields

We want to construct additional fields.

```{r construct_additional_fields, echo=TRUE}
hero_data_tbl <- hero_data_tbl %>%
  mutate(
    hero_id = fct_reorder(hero_id, as.numeric(hero_id))
    )

matchdata_tbl <- matchdata_tbl %>%
  left_join(cluster_region_tbl, by = "cluster")

matchdata_tbl %>% glimpse()


match_teamdata_tbl <- match_teamdata_tbl %>%
  mutate(
    team_status = if_else(result * team_comp < 0, "LOSING TEAM", "WINNING TEAM")
    ) %>%
  left_join(hero_data_tbl, by = "hero_id") %>%
  left_join(cluster_region_tbl, by = "cluster")

match_teamdata_tbl %>% glimpse()
```

Just to confirm this data has merged properly, we want to check if any of the
joined tables are blank.

```{r perform_basic_data_checks, echo=TRUE}
matchdata_tbl %>%
  filter(are_na(region)) %>%
  glimpse()


match_teamdata_tbl %>%
  filter(are_na(region)) %>%
  glimpse()
```


# Produce Basic Categorical Barplots

First we want to look at how some of the various categorical variables are
distributed, and illustrate this by producing barplots of the counts.

We start with the region of the match.

```{r plot_region_barplot, echo=TRUE}
plotdata_tbl <- matchdata_tbl %>%
  count(region, name = "match_count")

ggplot(plotdata_tbl) +
  geom_col(aes(x = region, y = match_count)) +
  labs(
    x = "Region",
    y = "Match Count"
    ) +
  scale_y_log10(labels = label_comma()) +
  ggtitle("Barplot of Matches by Region") +
  theme(axis.text.x = element_text(angle = 20, vjust = 0.5))
```


Now we look at the `game_type`:

```{r plot_gametype_barplot, echo=TRUE}
plotdata_tbl <- matchdata_tbl %>%
  count(game_type, name = "match_count")

ggplot(plotdata_tbl) +
  geom_col(aes(x = game_type, y = match_count)) +
  labs(
    x = "Game Type",
    y = "Match Count"
    ) +
  scale_y_log10(labels = label_comma()) +
  ggtitle("Barplot of Matches by Game Type")
```


Finally, we look at the distribution of `game_mode`.

```{r plot_gamemode_barplot, echo=TRUE}
plotdata_tbl <- matchdata_tbl %>%
  count(game_mode, name = "match_count")

ggplot(plotdata_tbl) +
  geom_col(aes(x = game_mode, y = match_count)) +
  labs(
    x = "Game Mode",
    y = "Match Count"
    ) +
  scale_y_log10(labels = label_comma()) +
  ggtitle("Barplot of Matches by Game Mode")
```

## Plot Hero Appearances and Win-Rate

We want to look at how often the various heroes appear in the data set and
what their individual win rate is.

```{r construct_hero_appearance_winrate_data, echo=TRUE}
hero_winrate_tbl <- match_teamdata_tbl %>%
  count(hero_id, name, team_status, name = "n_appear") %>%
  group_by(name) %>%
  mutate(
    n_matches = sum(n_appear),
    prop      = n_appear / sum(n_appear)
    ) %>%
  ungroup()

hero_winrate_tbl %>% glimpse()
```

First we want to get a sense of how win rates are distributed, as well as how
often heroes appear in the data.

```{r plot_hero_appearance_winrate, echo=TRUE}
plot_tbl <- hero_winrate_tbl %>%
  filter(team_status == "WINNING TEAM")

ggplot(plot_tbl) +
  geom_histogram(aes(x = n_matches), bins = 25) +
  scale_x_continuous(labels = label_comma()) +
  labs(
    x = "Appearances",
    y = "Count"
    ) +
  ggtitle("Histogram of Hero Appearances")
```



# Construct Mixed-Node Graph

We have a number of different was of thinking about how we might construct
graphs from this data. The first and most obvious one is to convert both
individual matches and the heroes as nodes on the graph.

```{r construct_mixed_type_graph, echo=TRUE}
nodes_match_tbl <- match_teamdata_tbl %>%
  transmute(node_id = match_id, node_type = "Match", game_mode, game_type, region) %>%
  distinct()

nodes_hero_tbl <- match_teamdata_tbl %>%
  transmute(node_id = hero_id, node_type = "Hero", name, localized_name) %>%
  distinct()

nodes_tbl <- list(
    nodes_match_tbl,
    nodes_hero_tbl
    ) %>%
  bind_rows()

edges_tbl <- match_teamdata_tbl %>%
  select(match_id, hero_id, team_status)


match_heroes_tblgraph <- tbl_graph(
  nodes    = nodes_tbl,
  edges    = edges_tbl,
  directed = FALSE,
  node_key = "node_id"
  )
```



# Construct Undirected Hero-Node Graph

```{r construct_heronode_data, echo=TRUE}
calculate_pairwise_weights <- function(matchdata_tbl) {
  pairs_tbl <- expand_grid(
      one = matchdata_tbl %>% select(h1 = hero_id, st1 = team_status),
      two = matchdata_tbl %>% select(h2 = hero_id, st2 = team_status)
      ) %>%
    unnest(c(one, two)) %>%
    mutate(
      weight = case_when(
        (st1 == "WINNING TEAM" & st2 == "WINNING TEAM") ~  1,
        (st1 == "LOSING TEAM"  & st2 == "LOSING TEAM")  ~ -1,
        TRUE ~ 0
      )
    )

  return(pairs_tbl)
}


hero_pairs_tbl <- match_teamdata_tbl %>%
  group_nest(match_id, .key = "match_data") %>%
  mutate(
    pairs_data = future_map(
      match_data, calculate_pairwise_weights,
      
      .progress = TRUE
      )
    )

edges_tbl <- hero_pairs_tbl %>%
  select(pairs_data) %>% 
  unnest(pairs_data) %>%
  group_by(h1, h2) %>%
  summarise(
    .groups = "drop",
    
    n_appear    = n(),
    team_weight = sum(weight)
    ) %>%
  mutate(
    h1 = h1 %>% fct_reorder(h1 %>% as.numeric()),
    h2 = h2 %>% fct_reorder(h2 %>% as.numeric())
    ) %>%
  arrange(h1, h2)

edges_tbl %>% glimpse()
```


# R Environment

```{r show_session_info, echo=TRUE, message=TRUE}
sessioninfo::session_info()
```
