---
title: "Workshop: Introduction to Copulas"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---

```{r import_libraries}
#| echo: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(furrr)
library(glue)
library(fs)
library(copula)
library(copulaData)
library(VineCopula)
library(rvinecopulib)
library(GGally)
library(lattice)
library(scatterplot3d)
library(ggraph)
library(tidygraph)
library(broom)
library(arrow)
library(qs2)

source("lib_utils.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "copula", "rvinecopulib")
)

options(
  width    = 80L,
  warn     = 1,
  mc.cores = parallelly::availableCores()
)

set.seed(42)

theme_set(theme_cowplot())

plan(multisession)
```

# Foundations of Dependency: Beyond Simple Correlation

In many real-world applications, we are interested in how multiple random variables 
behave together. While we often have a good handle on individual variables—like 
the height of a wave or the number of insurance claims in a year—it is the 
**joint behavior** that usually creates the most risk.

## The "Aggregation Problem" and the Lego Analogy

Traditional statistics often forces us into "one-size-fits-all" multivariate 
distributions. For example, the Multivariate Normal distribution requires that 
every variable follows a Normal distribution. In reality, data is messy: you 
might have a Gamma-distributed frequency of events linked to a Lognormal-distributed 
severity of costs.

**Copulas** solve this through the **Separation Principle**. Think of your 
multivariate model like a set of Lego bricks:

1.  **The Bricks (Marginals)**: You choose the best "brick" (distribution) for 
    each individual variable.

2.  **The Connectors (Copula)**: The copula is the specialized "peg" that 
    describes exactly how those bricks snap together, independent of their 
    shape or color.

## The Engine: Probability Integral Transform (PIT)

The magic that allows us to separate the "bricks" from the "connectors" is the 
**Probability Integral Transform (PIT)**. It acts as a universal translator. 

If you take any continuous data and pass it through its own Cumulative 
Distribution Function (CDF), you end up with a value between 0 and 1—essentially 
its "percentile rank." By converting everything to this uniform scale, we can 
link variables that have completely different units (like "Degrees Celsius" and 
"Megawatts") without the units getting in the way.

The following plots demonstrate this "uniformization" and show what a high-correlation 
relationship looks like once we've moved into this "Copula Space."

```{r theory_viz}
#| echo: true
#| fig-cap: "PIT and Copula Visualisation"

n_sim <- 5000

# PIT Demonstration: Transforming Gamma to Uniform
x <- rgamma(n_sim, shape = 2, rate = 0.5)
u <- pgamma(x, shape = 2, rate = 0.5)

plot_pit <- ggplot(tibble(u = u), aes(x = u)) +
  geom_histogram(bins = 30, fill = "darkorange", alpha = 0.7) +
  geom_hline(yintercept = n_sim / 30, color = "red", linetype = "dashed") +
  labs(
    x     = "U = F(X) (Percentile Rank)",
    y     = "Count",
    title = "PIT: Gamma to Uniform"
  )

# Dependence Visualization: Gaussian Copula
sigma <- matrix(c(1, 0.8, 0.8, 1), 2, 2)
z     <- MASS::mvrnorm(n_sim, mu = c(0, 0), Sigma = sigma)
u_dep <- pnorm(z) |> as_tibble(.name_repair = "unique")

plot_dep <- ggplot(u_dep, aes(x = ...1, y = ...2)) +
  geom_point(alpha = 0.2, size = 0.5, color = "steelblue") +
  labs(
    x     = "U1 (Variable A Rank)",
    y     = "U2 (Variable B Rank)",
    title = "Dependency in Copula Space (rho = 0.8)"
  )

plot_grid(plot_pit, plot_dep, ncol = 2)
```

# The Copula Zoo: Families and their "Tail Signatures"

Not all dependencies are created equal. Correlation tells us how variables 
move *on average*, but in risk management, we care about the **extremes**. 
This is where different "families" of copulas come into play.

## Rank Correlation: Measuring the "Order" of Things

Traditional correlation (Pearson’s $r$) is obsessed with straight lines. If 
your relationship is curved, Pearson's $r$ will underestimate its strength. 
In copula modelling, we use **Rank Correlation** (like Kendall’s Tau). It 
doesn't care about the scale or the units; it only cares about the *order* of 
the values. It is invariant to transformations—whether you model loss in 
Dollars or Log-Dollars, the rank correlation stays the same.

```{r rank_correlation_viz}
#| echo: true
#| fig-cap: "Pearson vs. Kendall Correlation under Non-Linear Transformation"

# Exponential relationship
x_lin <- seq(1, 10, length.out = 100)
y_exp <- exp(x_lin) + rnorm(100, sd = 1000)

cor_p <- cor(x_lin, y_exp, method = "pearson")
cor_k <- cor(x_lin, y_exp, method = "kendall")

tibble(x = x_lin, y = y_exp) |>
  ggplot(aes(x = x, y = y)) +
  geom_point(color = "darkblue") +
  labs(
    title    = "The Limit of Linear Correlation",
    subtitle = glue("Pearson r = {round(cor_p, 3)} (Low) vs. Kendall tau = {round(cor_k, 3)} (High)"),
    caption  = "Pearson misses the strength of the curved relationship; Kendall sees the perfect order."
  )
```

## Tail Dependence: "Ride-or-Die" vs. "Fair-Weather Friends"

This is perhaps the most critical concept in copula theory. It describes how 
likely two variables are to "crash" or "boom" together.

*   **Fair-Weather Friends (Gaussian Copula)**: These variables look correlated 
    in normal times, but as you move to the extreme edges, they actually become 
    independent. This "asymptotic independence" is why Gaussian models 
    famously failed during the 2008 financial crisis—they assumed that 
    just because two markets hadn't crashed together *recently*, they 
    *couldn't* crash together in the extreme.

*   **Ride-or-Die (t-Copula, Gumbel, Clayton)**: These exhibit **Tail Dependence**. 
    If Variable A is in its worst 1%, Variable B is highly likely to be there 
    too. 

### Visualizing the "Corners" of Risk

The plots below show how different families concentrate their risk in different 
"corners" of the uniform space.

```{r archimedean_viz}
#| echo: true
#| fig-cap: "Tail Dependence Signatures in Archimedean Copulas"

n_plot <- 2000

# Helper to simulate and plot
plot_arch_comp <- function(family, param_val, title_text, color_val) {
  if (family == "clayton") {
    cop <- claytonCopula(param = param_val, dim = 2)
  } else if (family == "gumbel") {
    cop <- gumbelCopula(param = param_val, dim = 2)
  } else {
    cop <- frankCopula(param = param_val, dim = 2)
  }

  u_sim_tbl <- rCopula(n_plot, cop) |>
    as_tibble()

  ggplot(u_sim_tbl, aes(x = V1, y = V2)) +
    geom_point(alpha = 0.3, color = color_val, size = 0.5) +
    labs(
      x     = "Rank U1",
      y     = "Rank U2",
      title = title_text
    ) +
    theme_minimal()
}

# Clayton (Lower Tail - Droughts/Crashes)
p1 <- plot_arch_comp("clayton", 2, "Clayton (Lower Tail Concentration)", "darkgreen")
# Gumbel (Upper Tail - Heatwaves/Booms)
p2 <- plot_arch_comp("gumbel", 2, "Gumbel (Upper Tail Concentration)", "firebrick")
# Frank (Symmetric - No Tail dependence)
p3 <- plot_arch_comp("frank", 5, "Frank (Symmetric / No Tails)", "purple")
# Student's t (Both Tails - Heavy Tails)
cop_t <- tCopula(param = 0.5, df = 2, dim = 2)
u_t   <- rCopula(n_plot, cop_t) |> as_tibble()
p4    <- ggplot(u_t, aes(x = V1, y = V2)) +
  geom_point(alpha = 0.3, color = "blue", size = 0.5) +
  labs(title = "Student's t (Both Tails)", x = "U1", y = "U2") +
  theme_minimal()

plot_grid(p1, p2, p3, p4, ncol = 2)
```

## The Critical Role of Degrees of Freedom ($df$)

In the **Student's t-Copula**, the $df$ parameter is our "dial" for tail risk. 
Low $df$ means extreme events happen together very often. As $df$ increases, 
the t-copula slowly turns into a Gaussian copula, and that joint tail risk 
evaporates.

```{r t_copula_viz}
#| echo: true
#| fig-cap: "The 'Normalization' of the t-Copula as df Increases"

n_plot  <- 2000
rho_fix <- 0.5

# Helper to simulate and plot
plot_t_comp <- function(df_val, title_text, color_val) {
  if (is.infinite(df_val)) {
    cop <- normalCopula(param = rho_fix, dim = 2)
  } else {
    cop <- tCopula(param = rho_fix, df = df_val, dim = 2)
  }

  u_sim_tbl <- rCopula(n_plot, cop) |>
    as_tibble(.name_repair = "unique")

  ggplot(u_sim_tbl, aes(x = ...1, y = ...2)) +
    geom_point(alpha = 0.4, color = color_val, size = 0.6) +
    labs(x = "U1", y = "U2", title = title_text) +
    theme_minimal()
}

p1_t <- plot_t_comp(1, "t (df=1): Heavy Tails", "firebrick")
p2_t <- plot_t_comp(30, "t (df=30): Almost Gaussian", "darkblue")

plot_grid(p1_t, p2_t, ncol = 2)
```

# Real-World Scenarios: Modelling Compounding Risk

In risk management, **compounding risk** occurs when multiple "bad" things 
happen at once. Copulas allow us to model these dependencies explicitly, 
even when the types of data are very different.

## Case Study: Insurance - Frequency and Severity

Imagine a year with a high number of storm events (Frequency). If those 
storms are also individually more severe (Severity), the total annual loss 
balloons. This is compounding risk.

```{r compounding_risk_sim}
#| echo: true

simulate_total_costs <- function(n_years, rho, lambda, shp, rt) {
  cop      <- normalCopula(param = rho, dim = 2)
  u_sim    <- rCopula(n_years, cop)
  n_events <- qpois(u_sim[, 1], lambda = lambda)

  total_costs <- map_dbl(
    1:n_years,
    function(i) {
      n <- n_events[i]
      if (n == 0) return(0)
      sev_p    <- u_sim[i, 2]
      year_sev <- qgamma(sev_p, shape = shp, rate = rt)
      return(n * year_sev)
    }
  )
  return(total_costs)
}

# Run simulations across correlation levels
n_years     <- 10000
costs_indep <- simulate_total_costs(n_years, 0.0, 10, 2, 0.02)
costs_high  <- simulate_total_costs(n_years, 0.7, 10, 2, 0.02)

cost_data_tbl <- tibble(
  correlation = rep(c("0.0 (Indep)", "0.7 (High)"), each = n_years),
  total_cost  = c(costs_indep, costs_high)
)
```

### The Impact on the "Tail"

Observe how high correlation doesn't just change the average; it dramatically 
increases the probability of extreme, "bankrupting" years.

```{r cost_density}
#| echo: true
#| fig-cap: "Compounding Risk: Impact of Correlation on Annual Loss"

ggplot(cost_data_tbl, aes(x = total_cost, fill = correlation)) +
  geom_density(alpha = 0.4) +
  scale_x_log10(labels = dollar) +
  labs(x = "Total Annual Cost (Log Scale)", title = "Log Scale: Seeing the Tail Risk")
```

## Diverse Coupling Examples

Copulas can link any two metrics that share a common environmental or 
economic driver.

### Public Health: Correlating Discrete Events

High pharmacy sales of flu medication often coincide with high ER admissions. 
Modeling these as independent would underestimate the peak load on the city’s 
healthcare system.

```{r public_health_discrete}
#| echo: true
#| fig-cap: "Correlated Public Health Indicators (Discrete-Discrete)"

# Moderate-High Correlation
rho_health <- 0.65
cop_health <- normalCopula(param = rho_health, dim = 2)
u_health   <- rCopula(2000, cop_health)

# Map to Marginals (Negative Binomial and Poisson)
sales  <- qnbinom(u_health[, 1], mu = 200, size = 20)
er_adm <- qpois(u_health[, 2], lambda = 15)

health_data_tbl <- tibble(Sales = sales, ER = er_adm)

ggplot(health_data_tbl, aes(x = Sales, y = ER)) +
  geom_jitter(alpha = 0.3, color = "darkred") +
  labs(x = "Daily Pharmacy Sales", y = "ER Admissions", title = "Public Health Coupling")
```

### Agriculture: Discrete-Continuous Coupling (Drought Risk)

In agriculture, we might model the number of rainy days (Discrete) against 
the final crop yield percentage (Continuous). Low rainfall is strongly 
linked to low yields, a classic "lower tail" dependency.

```{r agriculture_coupling}
#| echo: true
#| fig-cap: "Rainfall Frequency vs. Crop Yield (Discrete-Continuous)"

# Clayton Copula captures 'Lower Tail Dependence'
cop_agri <- claytonCopula(param = 1.5, dim = 2)
u_agri   <- rCopula(2000, cop_agri)

rain_days  <- qpois(u_agri[, 1], lambda = 12)
crop_yield <- qbeta(u_agri[, 2], shape1 = 5, shape2 = 2)

agri_data_tbl <- tibble(Rain = rain_days, Yield = crop_yield * 100)

ggplot(agri_data_tbl, aes(x = Rain, y = Yield)) +
  geom_jitter(alpha = 0.3, color = "darkgreen") +
  labs(x = "Rainy Days per Month", y = "Crop Yield (% of Max)", title = "Agriculture Coupling: Drought Risk")
```

### Energy: Continuous-Continuous Coupling (Heatwaves)

Extreme heatwaves drive non-linear spikes in electricity demand as air 
conditioning units work harder. This relationship is concentrated in the 
"upper tail."

```{r energy_coupling}
#| echo: true
#| fig-cap: "Energy Coupling: Heatwaves and Power Demand"

# Gumbel Copula for 'Upper Tail Dependence' (Heat -> Demand)
cop_energy <- gumbelCopula(param = 2.5, dim = 2)
u_energy   <- rCopula(2000, cop_energy)

temp   <- qgamma(u_energy[, 1], shape = 20, rate = 0.8)
demand <- qlnorm(u_energy[, 2], meanlog = 5, sdlog = 0.3)

energy_data_tbl <- tibble(Temp = temp, Demand = demand)

ggplot(energy_data_tbl, aes(x = Temp, y = Demand)) +
  geom_point(alpha = 0.2, color = "orange") +
  labs(x = "Peak Temp (°C)", y = "Demand (MW)", title = "Physical Coupling: Heat vs. Demand")
```

# The Empirical Detective: Fitting Copulas to Data

In practice, we don't start with the copula; we start with raw data. 
Finding the right dependency structure is a three-step detective process 
called **Inference Functions for Margins (IFM)**.

## The IFM Workflow

1.  **Marginal Estimation**: Fit the best distributions to each variable 
    independently (the "bricks").

2.  **Transformation (PIT)**: Use the fitted CDFs to move the data into 
    uniform "Copula Space."

3.  **Selection and Fitting**: Choose the copula family that best describes 
    the "corners" of the data.

## A "Recovery Test": Proving the Method

Let's see if we can "recover" the true dependency from our earlier 
Energy example. We know the truth was a **Gumbel Copula**.

```{r recovery_pit}
#| echo: true
#| fig-cap: "The Copula Space: Revealing the Hidden Structure"

# Transform raw data back to uniform scale (pseudo-observations)
u_energy_tbl <- pobs(energy_data_tbl) |>
  as_tibble() |>
  set_colnames(c("U1", "U2"))

ggplot(u_energy_tbl, aes(x = U1, y = U2)) +
  geom_point(alpha = 0.2, color = "darkblue") +
  labs(x = "Temp Rank", y = "Demand Rank", title = "Copula Space: Pure Dependency")
```

### Automated Selection (AIC/BIC)

Using `rvinecopulib`, we can automatically test all families and select 
the best one.

```{r recovery_fitting}
#| echo: true

# Select best bivariate copula
fit_energy <- bicop(data = as.matrix(u_energy_tbl), family_set = "all")
print(fit_energy)
```

The algorithm correctly identifies the Gumbel family, proving we can 
extract dependency structures from raw observations.

## The Danger of Choosing the Wrong Model

What if we chose a Gaussian model when the truth was Gumbel? We would 
severely underestimate how often extreme heat and extreme demand happen 
together.

```{r model_comparison_energy}
#| echo: true

# Fit various families to the same data
fit_gauss   <- bicop(as.matrix(u_energy_tbl), family_set = "gaussian")
fit_frank   <- bicop(as.matrix(u_energy_tbl), family_set = "gaussian")

# Compare AIC (Lower is better)
model_comparison_tbl <- tibble(
  Model  = c("Gumbel (True)", "Gaussian", "Frank"),
  AIC    = c(AIC(fit_energy), AIC(fit_gauss), AIC(fit_frank))
) |>
  arrange(AIC)

knitr::kable(model_comparison_tbl, digits = 1, caption = "Model Selection via AIC")
```

# Scaling Complexity: Multivariate and Vine Copulas

The real world is rarely bivariate. In finance, engineering, and 
environmental science, we often need to model the joint behavior of 
dozens of variables. This introduces the **Curse of Dimensionality**: 
as the number of variables grows, the complexity of capturing every 
pairwise and joint interaction increases exponentially.

## Multivariate Continuous Data: The Uranium Case Study

A classic high-dimensional example is the **Uranium exploration dataset** 
(Cook & Johnson, 1986). It contains concentrations of 7 chemical elements 
in 655 water samples. Capturing the "fingerprint" of these relationships 
allows geologists to identify anomalous patterns indicative of mineral 
deposits.

```{r uranium_analysis}
#| echo: true
#| fig-cap: "Pairwise Dependencies in 7D Uranium Data"

data(uranium, package = "copula")

# Focus on a few elements for clarity: Uranium, Lithium, Cobalt, Titanium
uranium_sub <- uranium |>
  as_tibble() |>
  select(U, Li, Co, Ti)

# Visualize the dependencies using ranks
ggpairs(
  pobs(uranium_sub),
  lower = list(continuous = wrap("points", alpha = 0.3, size = 0.5, color = "darkblue")),
  title = "Copula Space (Ranks) for Uranium Data"
)
```

Notice how different pairs exhibit different structures—some are tight and 
symmetric, others are sparse. A standard multivariate Gaussian copula 
would force a single, symmetric "average" structure on all these pairs, 
potentially missing the physical constraints of the geochemistry.

## Financial Contagion: The DAX Index

In finance, we care about **contagion**—the phenomenon where assets that 
are uncorrelated during normal times suddenly crash together during a 
crisis. The `daxreturns` dataset (daily log-returns of major DAX 
components) from the `VineCopula` package illustrates this.

```{r dax_analysis}
#| echo: true

data(daxreturns, package = "VineCopula")

# Focus on first few assets and give them cleaner names
dax_sub <- daxreturns[, 1:5]
colnames(dax_sub) <- c("Allianz", "BASF", "Bayer", "BMW", "Daimler")

# Fit a multivariate Student-t copula vs. Gaussian using rvinecopulib
# The Student-t captures the 'Ride-or-Die' crashing together
fit_gauss_multi <- bicop(pobs(dax_sub[, 1:2]), family_set = "gaussian")
fit_t_multi     <- bicop(pobs(dax_sub[, 1:2]), family_set = "t")

model_comp_dax <- tibble(
  Model = c("Gaussian", "Student-t"),
  AIC   = c(AIC(fit_gauss_multi), AIC(fit_t_multi))
)

knitr::kable(model_comp_dax, caption = "Contagion Check: Gaussian vs. t-Copula (DAX)")
```

The superior fit of the Student-t model confirms that market components 
exhibit significant tail dependence—they crash together far more often 
than a Gaussian model would predict.

## Vine Copulas: The Modular Solution

How do we scale this to 50 variables without losing the unique 
"tail signatures" of each pair? We use **Vine Copulas**.

Vine Copulas break down a high-dimensional distribution into a 
hierarchy of **Bivariate Copulas**. This is called **Pair-Copula 
Construction (PCC)**.

### The Pair-Copula Construction (PCC)

Think of it like building a complex house (your high-dimensional joint 
distribution) out of simple bricks (bivariate copulas). Each brick 
connects just two variables, but those connections can be conditional 
on other variables.

A 5-dimensional problem might involve:

1.  **Level 1**: Linking (1,2), (2,3), (3,4), (4,5) — unconditional pairs.
2.  **Level 2**: Linking (1,3 | 2), (2,4 | 3), (3,5 | 4) — conditional pairs.
3.  **Level 3**: Linking (1,4 | 2,3), (2,5 | 3,4) — higher-order conditional pairs.

### Vine Structures: C-Vines and D-Vines

This hierarchical decomposition allows for incredible flexibility. The 
way we organize these bivariate copulas defines the "vine" structure.

To understand why we need these structures, let's look at an **Energy 
Supply Chain Risk** scenario. Imagine we are a global retailer 
trying to model how shocks in energy markets propagate through our 
costs:

1.  **Crude Oil Prices**: The primary global energy benchmark.
2.  **Global Freight Rates**: The cost of shipping goods (fueled by oil).
3.  **Retail Food Prices**: Final costs to consumers (driven by both 
    shipping costs and direct fuel inputs for farms).

#### Step 1: The First Tree (Direct Relationships)

We start by modeling the most direct physical connections. In a 
**D-Vine** (Drawable Vine), we assume a logical sequence: 
**Oil $\rightarrow$ Freight $\rightarrow$ Food**.

```{r vine_step_1}
#| echo: true

# Hypothetical data for clarity (we map these from DAX for the demo)
# BMW (Auto/Fuel) -> DBK (Finance/Freight) -> SAP (Software/Retail)
u_chain <- pobs(daxreturns[, c("BMW.DE", "DBK.DE", "SAP.DE")])
colnames(u_chain) <- c("Crude_Oil", "Global_Freight", "Retail_Food")

# Fit bivariate copulas for the first links in the chain
fit_oil_freight   <- bicop(u_chain[, 1:2], family_set = "all")
fit_freight_food  <- bicop(u_chain[, 2:3], family_set = "all")

cat("Oil-Freight Link:", fit_oil_freight$family, "\n")
cat("Freight-Food Link:", fit_freight_food$family, "\n")
```

#### Step 2: Moving to the Second Tree (Conditional Dependency)

We now want to know: **Is there a relationship between Oil and Food that is 
not already explained by Freight?**

To find out, we must "filter out" the effect of Freight. We use the 
**h-function** to transform our ranks into **Conditional Ranks**. This 
tells us how Oil and Food behave *given* a specific level of Freight cost.

```{r vine_step_2}
#| echo: true

# Transform Oil and Food to be conditional on Freight
# (Using the h-function to 'standardize' the residuals)
u_oil_given_freight  <- hbicop(u_chain[, 1:2], cond_var = 2, fit_oil_freight)
u_food_given_freight <- hbicop(u_chain[, 2:3], cond_var = 1, fit_freight_food)

u_cond_pair <- cbind(u_oil_given_freight, u_food_given_freight)
```

#### Step 3: Fitting the "Residual" Dependence

Finally, we fit a copula to these conditional ranks. This is the 
"Level 2" tree of our vine.

```{r vine_step_3}
#| echo: true

fit_oil_food_given_freight <- bicop(u_cond_pair, family_set = "all")

# Calculate Kendall's Tau for comparison
tau_direct <- par_to_ktau(fit_oil_freight$family, fit_oil_freight$rotation, fit_oil_freight$parameters)
tau_resid  <- par_to_ktau(fit_oil_food_given_freight$family, fit_oil_food_given_freight$rotation, fit_oil_food_given_freight$parameters)

cat("Direct Dependence (Oil-Freight):", round(tau_direct, 3), "\n")
cat("Conditional Dependence (Oil-Food | Freight):", round(tau_resid, 3), "\n")
```


#### Why do we justify this complexity?

By breaking the system into these pieces, we can see exactly where 
the risk lies:

1.  **Asymmetric Shocks**: We might find that Oil and Freight "crash 
    together" (Clayton Copula), but the relationship between Food and 
    Oil (once Freight is accounted for) is totally independent.
2.  **Structural Insights**:
    *   **C-Vines (Canonical)**: Use if you believe **Crude Oil** is the 
        "master driver" that affects Freight, Food, and everything else 
        simultaneously.
    *   **D-Vines (Drawable)**: Use if you believe shocks move through 
        a sequence (like a physical pipeline or supply chain).

### Automated Vine Selection: Building the Risk Map

In reality, for 10 or 20 variables, the number of possible "vines" is 
astronomical. We use algorithms (like those in `rvinecopulib`) to 
automatically find the "Maximum Spanning Tree" (MST).

#### Decoding the Vine Plot

When we visualize a fitted vine, we are looking at the "Tree 1"—the 
most important foundation of our model. Think of this plot as a 
**Dependency Highway Map** that shows the most critical paths through 
which risk travels between our variables.

The nodes in the plot represent the individual variables (in this case, 
our five DAX stocks). The lines connecting them, called **Edges**, 
represent the strongest unconditional relationships found by the 
algorithm. When you look at an edge, you'll often see two pieces of 
information: the **Copula Family** (the "connector" type) and the 
**Kendall's Tau** (the strength of that connection). A thick line 
indicates a strong, direct dependency, while a thinner line shows a 
weaker link.

The layout of these connections is determined by a **Maximum Spanning 
Tree (MST)** algorithm. In a 5-dimensional problem, there are 
thousands of ways we could link these variables together. The MST 
algorithm "hunts" for the pairs with the highest dependency first. 
It builds the backbone of the vine by connecting the nodes in a way 
that captures the most total dependency possible without creating 
redundant loops.

This "backbone" is critical because everything that happens in the 
later stages of the model—the conditional dependencies we discussed 
earlier—is built on top of these first-level links. If Tree 1 doesn't 
accurately capture the primary physical or economic connections in 
your data, the entire hierarchical model will be less effective.

```{r vine_deep_dive}
#| echo: true
#| fig-cap: "The First Tree: The Foundation of our Vine Model"

# Fit an R-vine (Regular Vine) to the 5 DAX stocks
# The algorithm automatically finds the best 'ordering' of variables
dax_vine <- vinecop(pobs(dax_sub), family_set = "all")

# Plot the first tree: This is the 'backbone' of our joint model
# Thick lines represent stronger dependencies
plot(dax_vine, tree = 1, var_names = "use")
```


#### Interpreting our DAX Risk Map

Looking at the plot generated above, we can see a clear "industrial 
logic" in how the algorithm has organized the German market. At the 
top of the tree, we typically see a very tight cluster between 
**BASF** and **Bayer**. These are both giant chemical/pharmaceutical 
conglomerates. The vine algorithm places them close together with 
a strong edge (often a Student-t or Gumbel copula) because their 
business models and market sensitivities are highly correlated. If 
the global chemical sector takes a hit, these two are almost 
guaranteed to move in lockstep.

Moving down the tree, we see how the "contagion" spreads. **Allianz**, 
a massive financial and insurance provider, often acts as a central 
hub or a major bridge in these market vines. This makes sense 
economically: as a major institutional investor, Allianz has its 
fingers in every other sector of the DAX. Its positioning in the 
center of the map suggests that it is a primary "transmitter" of 
systemic risk across the portfolio.

Finally, notice the automotive cluster at the bottom. **Daimler** 
and **BMW** are linked together, but they are often connected to the 
rest of the market through a single bridge (like Allianz or BASF). 
This tells us that while the car manufacturers are highly dependent 
on each other, their relationship with the broader market is somewhat 
more "filtered." A shock that hits the car industry specifically 
might crash BMW and Daimler together, but it has to pass through 
these intermediate "bridge" stocks before it severely impacts the 
chemical or financial sectors.


#### Why does this "Map" matter?

The vine structure tells us how shocks propagate through the system:

1.  **Direct Contagion**: If two assets are connected in Tree 1, they 
    have a direct, strong relationship. If one crashes, the other is 
    highly likely to follow immediately.
2.  **Indirect Contagion**: If two assets are only connected through 
    other nodes (e.g., A $\rightarrow$ B $\rightarrow$ C), then the 
    shock from A must pass through the "filter" of B before reaching C.
3.  **Stability**: By capturing the strongest dependencies in Tree 1, 
    we ensure the model remains stable even if we "truncate" (ignore) 
    the very weak, complex dependencies in higher-order trees.

### Structure Comparison: Automated vs. Fixed

Is a simple hub-and-spoke (C-Vine) enough, or do we need the complex 
automated search? We can compare models using AIC—the lower the AIC, 
the better the model captures the underlying "shape" of the risk.

```{r vine_structure_comp}
#| echo: true

# Compare a fixed C-vine (one pilot variable) vs the automated R-vine
# We'll use ALV.DE (Allianz) as a hypothetical pilot variable
fit_cvine <- vinecop(pobs(dax_sub), structure = cvine_structure(1:5))

vine_comparison <- tibble(
  Structure = c("C-Vine (Fixed Hub)", "R-Vine (Automated Search)"),
  LogLik    = c(logLik(fit_cvine), logLik(dax_vine)),
  AIC       = c(AIC(fit_cvine), AIC(dax_vine))
)

knitr::kable(vine_comparison, caption = "Vine Structure Comparison")
```

### Vine Truncation: Managing High Dimensions

For very high-dimensional problems (hundreds of variables), fitting a full 
vine can be computationally intensive and lead to overfitting. **Truncation** 
is a technique where we limit the "depth" of the vine, assuming that 
higher-order conditional dependencies (e.g., $(X_1, X_5 | X_2, X_3, X_4)$) 
become less important. A "1-truncated" vine only models unconditional 
bivariate dependencies.

### Stress-Testing and Simulation: Practical Risk Measures

The ultimate goal of copula modeling is to generate realistic "stress 
scenarios." Once we have a fitted vine, we can simulate thousands of 
years of joint behavior to calculate risk measures like **Value-at-Risk 
(VaR)** and **Expected Shortfall (ES)**.

#### Conditional Simulation: What if the Worst Happens?

A key advantage of copulas is the ability to perform **conditional 
simulations**. For instance, what if one specific market (e.g., Germany) 
experiences a 1% drop? What is the expected impact on the rest of the 
portfolio, given this condition? This helps quantify systemic risk.

#### Portfolio Risk Simulation (VaR and ES)

Let's assume an equal-weighted portfolio of the 5 DAX components. We 
want to know the 1-day 99% VaR—the loss we only expect to see once 
every 100 days. The Expected Shortfall then tells us the average loss 
given that the VaR threshold has been breached.

```{r portfolio_sim}
#| echo: true

# 1. Simulate ranks from our fitted DAX vine
n_sim_risk <- 5000
u_sim_risk <- rvinecop(n_sim_risk, dax_vine)

# 2. Map back to 'returns' space (assuming Empirical Marginals for simplicity)
# In a real model, we would use GARCH or heavy-tailed marginals
sim_returns <- matrix(NA, nrow = n_sim_risk, ncol = 5)
for (j in 1:5) {
  sim_returns[, j] <- quantile(dax_sub[, j], probs = u_sim_risk[, j])
}

# 3. Calculate Portfolio Return (Equal Weighted)
portfolio_returns <- rowMeans(sim_returns)

# 4. Calculate Risk Measures
var_99 <- quantile(portfolio_returns, 0.01)
es_99  <- mean(portfolio_returns[portfolio_returns <= var_99])

risk_measures <- tibble(
  Measure = c("99% Value-at-Risk (VaR)", "99% Expected Shortfall (ES)"),
  Value   = c(var_99, es_99)
)

knitr::kable(risk_measures, digits = 4, caption = "Portfolio Risk: Vine Copula Simulation")
```

The **Expected Shortfall (ES)** tells us the average loss given that the 
VaR threshold has been breached. Copula models provide a much more 
realistic estimate of ES because they capture the "crashing together" 
behavior of assets, which simple linear correlation models often miss.

## Other Real-World Applications (Conceptual)

Beyond finance, copulas are invaluable in:

*   **Hydrology**: Modeling joint probabilities of rainfall and river flow 
    to assess flood risk, or correlating drought severity across regions.
*   **Insurance**: Analyzing multi-line insurance products where different 
    types of claims (e.g., property, casualty, life) might be dependent 
    during systemic events. For example, a major hurricane could cause 
    both property damage and life insurance claims.
*   **Engineering**: Assessing the reliability of multi-component systems 
    where component failures are not independent.
*   **Supply Chain Risk**: Modeling how disruptions in one part of a supply 
    chain might propagate and impact other parts.

## Challenges and Limitations

Despite their power, copulas are not a silver bullet. Key challenges include:

*   **Model Selection**: Choosing the "best" copula family and vine structure 
    can be complex, especially with high-dimensional data.
*   **Computational Burden**: Fitting and simulating high-dimensional vine 
    copulas can be computationally intensive.
*   **Marginal Specification**: Errors in modeling the marginal distributions 
    will propagate into the copula model.
*   **Interpretability**: Interpreting complex vine structures can be less 
    intuitive than simple correlation matrices.

# Summary of the Copula Workflow

1.  **PIT**: Transform data to ranks to see the hidden dependence.

2.  **Visualize**: Check the "corners" of the Copula Space for tail risk.

3.  **Select**: Use AIC/BIC to find the right family (Lego connector).

4.  **Simulate**: Generate extreme scenarios to Stress-test your risk models.

# R Environment

```{r session_info}
#| echo: true

sessioninfo::session_info()
```
