---
title: "Life Insurance Survival Analysis: Basic Exploratory Methods"
subtitle: "Kaplan-Meier Estimation, Log-Rank Tests, and Risk Stratification"
author: "Mick Cooney"
date: "`r Sys.Date()`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---

```{r setup_libraries}
#| echo: false
#| message: false
#| warning: false

library(tidyverse)
library(arrow)
library(survival)
library(survminer)
library(cowplot)
library(broom)
library(fs)
library(glue)

source("lib_utils.R")
source("lib_survival_modelling.R")

# Set consistent theme
theme_set(theme_cowplot())

# Set random seed for reproducibility
set.seed(42)
```

```{r setup_chunk_timing}
#| echo: false

# Initialize chunk timing
chunk_times <- list()

knitr::knit_hooks$set(
  time_it = function(before, options, envir) {
    if (before) {
      chunk_times[[options$label]] <<- list(start = Sys.time())
    } else {
      chunk_times[[options$label]]$end <<- Sys.time()
      chunk_times[[options$label]]$elapsed <<-
        chunk_times[[options$label]]$end |>
        difftime(chunk_times[[options$label]]$start) |>
        as.numeric()
    }
  }
)

knitr::opts_chunk$set(time_it = TRUE)
```

# Introduction

This worksheet provides practical exploratory survival analysis methods
applied to life insurance policy lapse prediction. Rather than diving
into statistical theory, we focus on core techniques for understanding
how policyholder characteristics relate to policy persistence.

We'll estimate overall and stratified survival curves (Kaplan-Meier
method), test for differences between groups (log-rank test), and explore
risk stratification to identify high-risk and low-risk populations.

# Data Setup

We load pre-prepared training and validation datasets created in the
exploration notebook. These datasets include all necessary variables and
have been constructed following proper rollback methodology to ensure
temporal consistency.

```{r load_datasets}
#| echo: true

# Load training data (snapshot at 2015-01-01)
training_tbl <- read_parquet("data/model_training_small_tbl.parquet")

# Load validation data (same policies at 2016-01-01)
validation_tbl <- read_parquet("data/model_rollforward_small_tbl.parquet")

write_log_entry(
  "DATA_LOAD",
  glue(
    "Loaded {nrow(training_tbl)} training, ",
    "{nrow(validation_tbl)} validation"
  )
)

# Quick structural check
training_tbl |> glimpse()
```

# Sanity Checks

We validate that policy status follows expected patterns: once a policy lapses 
or completes, it cannot revert to an active status.

```{r sanity_check_status}
#| echo: true

# Check unique status values
write_lines("Training dataset status distribution:", stdout())
training_tbl |>
  count(status_at_reference_date) |>
  mutate(prop = scales::percent(n / sum(n))) |>
  print()

write_lines("\nValidation dataset status distribution:", stdout())
validation_tbl |>
  count(status_at_reference_date) |>
  mutate(prop = scales::percent(n / sum(n))) |>
  print()

# Verify terminal status logic
status_transitions <- training_tbl |>
  inner_join(
    validation_tbl |>
      select(policy_id, status_at_reference_date),
    by = "policy_id",
    relationship = "one-to-one",
    suffix = c("_train", "_valid")
  ) |>
  count(status_at_reference_date_train, status_at_reference_date_valid) |>
  rename(
    Training = status_at_reference_date_train,
    Validation = status_at_reference_date_valid,
    Count = n
  )

write_lines("\nStatus transitions from 2015 to 2016:", stdout())
print(status_transitions)

# Check that lapsed/completed don't revert to active
invalid_transitions <- training_tbl |>
  inner_join(
    validation_tbl |>
      select(policy_id, status_at_reference_date),
    by = "policy_id",
    relationship = "one-to-one",
    suffix = c("_train", "_valid")
  ) |>
  filter(
    (status_at_reference_date_train %in% c("lapsed", "completed")) &
    (status_at_reference_date_valid == "active")
  ) |>
  nrow()

if (invalid_transitions > 0) {
  write_log_entry("ERROR", glue("{invalid_transitions} invalid status transitions detected!"))
} else {
  write_log_entry("VALIDATION", "All status transitions valid (no reversals)")
}
```

# Overall Survival Estimation

We begin with the simplest analysis: estimate the overall survival curve
for all policies, regardless of characteristics. This establishes the
baseline—what proportion of policies remain active at each duration?

```{r create_overall_km}
#| echo: true

# Create Kaplan-Meier estimate for all policies
overall_km <- survfit(
  Surv(policy_lifetime, lapsed) ~ 1,
  data = training_tbl,
  conf.type = "log-log"
)

# Print summary - restricted to 4-week intervals over 2 years
overall_km |> summary(times = seq(0, 104, by = 4))
```

## Overall Survival Curve

```{r plot_overall_km}
#| echo: true

overall_km |>
  ggsurvplot(
    censor = FALSE,
    conf.int = TRUE,
    conf.int.style = "step",
    title = "Overall Kaplan-Meier Survival Curve",
    xlab = "Policy Duration (years)",
    ylab = "Survival Probability"
  )
```

The curve shows that approximately 75% of policies survive the first 5
years, declining to about 50% by year 10. The confidence interval (shaded
region) represents our uncertainty about the true survival probability—
note that it widens at longer durations where we have fewer policies at
risk.

# Stratified Survival Analysis

Real-world insights emerge when we compare survival across groups. We
examine whether demographic and policy characteristics relate to lapse
risk. This analysis typically reveals important risk factors for business
decision-making.

## Gender

Do male and female policyholders have different lapse experiences?

```{r create_gender_km}
#| echo: true

# Create KM estimates stratified by gender
gender_km <- survfit(
  Surv(policy_lifetime, lapsed) ~ gender_life1,
  data = training_tbl,
  conf.type = "log-log"
)

# Print summary with stratification - restricted to 4-week intervals over 2 years
gender_km |> summary(times = seq(0, 104, by = 4))
```

### Gender Survival Curves

```{r plot_gender_km}
#| echo: true

gender_km |>
  ggsurvplot(
    censor = FALSE,
    conf.int = FALSE,
    title = "Survival Curves by Gender",
    xlab = "Policy Duration (years)",
    ylab = "Survival Probability"
  )
```

### Gender Difference Testing

To test whether the observed difference in survival curves is
statistically significant, we use the log-rank test:

```{r test_gender_differences}
#| echo: true

gender_survdiff <- survdiff(
  Surv(policy_lifetime, lapsed) ~ gender_life1,
  data = training_tbl
)

gender_survdiff |> print()

write_lines("\nSummary Statistics by Gender:", stdout())
training_tbl |>
  group_by(gender_life1) |>
  summarise(
    count = n(),
    lapsed_count = sum(lapsed),
    lapse_rate = scales::percent(mean(lapsed), accuracy = 0.1),
    mean_duration = mean(policy_lifetime),
    median_duration = median(policy_lifetime),
    .groups = "drop"
  ) |>
  print()
```

## Smoker Status

Does smoking status relate to lapse probability?

```{r create_smoker_km}
#| echo: true

smoker_km <- survfit(
  Surv(policy_lifetime, lapsed) ~ smoker_life1,
  data = training_tbl,
  conf.type = "log-log"
)

smoker_km |> summary(times = seq(0, 104, by = 4))
```

### Smoker Status Survival Curves

```{r plot_smoker_km}
#| echo: true

smoker_km |>
  ggsurvplot(
    censor = FALSE,
    conf.int = FALSE,
    title = "Survival Curves by Smoker Status",
    xlab = "Policy Duration (years)",
    ylab = "Survival Probability"
  )
```

### Smoker Status Difference Testing

```{r test_smoker_differences}
#| echo: true

smoker_survdiff <- survdiff(
  Surv(policy_lifetime, lapsed) ~ smoker_life1,
  data = training_tbl
)

smoker_survdiff |> print()

write_lines("\nSummary Statistics by Smoker Status:", stdout())
training_tbl |>
  group_by(smoker_life1) |>
  summarise(
    count = n(),
    lapsed_count = sum(lapsed),
    lapse_rate = scales::percent(mean(lapsed), accuracy = 0.1),
    mean_duration = mean(policy_lifetime),
    median_duration = median(policy_lifetime),
    .groups = "drop"
  ) |>
  print()
```

## Policy Vintage Year

Do policies from different inception years show different lapse patterns?
This can reflect changing underwriting, market conditions, or product
modifications.

```{r create_vintage_km}
#| echo: true

vintage_km <- survfit(
  Surv(policy_lifetime, lapsed) ~ policy_vintage_year,
  data = training_tbl,
  conf.type = "log-log"
)

vintage_km |> summary(times = seq(0, 104, by = 4))
```

### Vintage Year Survival Curves

```{r plot_vintage_km}
#| echo: true

vintage_km |>
  ggsurvplot(
    censor = FALSE,
    conf.int = FALSE,
    title = "Survival Curves by Policy Vintage Year",
    xlab = "Policy Duration (years)",
    ylab = "Survival Probability"
  )
```

### Vintage Year Difference Testing

```{r test_vintage_differences}
#| echo: true

vintage_survdiff <- survdiff(
  Surv(policy_lifetime, lapsed) ~ policy_vintage_year,
  data = training_tbl
)

vintage_survdiff |> print()

write_lines("\nLapse Rates by Policy Vintage Year:", stdout())
training_tbl |>
  group_by(policy_vintage_year) |>
  summarise(
    count = n(),
    lapsed_count = sum(lapsed),
    lapse_rate = scales::percent(mean(lapsed), accuracy = 0.1),
    mean_duration = mean(policy_lifetime),
    median_duration = median(policy_lifetime),
    .groups = "drop"
  ) |>
  arrange(policy_vintage_year) |>
  print()
```

# Risk Stratification via Continuous Variable
  Discretization

Policy characteristics like premium amount and sum assured are continuous,
yet grouping policyholders into risk tiers aids interpretation and
business application. We discretize key variables into meaningful
categories.

## Premium Amount Stratification

We create risk tiers based on premium amount:

```{r discretize_premium}
#| echo: true

# Examine distribution
training_tbl |>
  ggplot(aes(x = prem_ape)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  scale_x_continuous(labels = scales::label_dollar()) +
  labs(
    title = "Distribution of Annual Premium (APE)",
    x = "Annual Premium Equivalent (USD)",
    y = "Frequency"
  )

# Create risk tiers based on business judgment
training_tbl_premium <- training_tbl |>
  mutate(
    premium_tier = cut(
      prem_ape,
      breaks = quantile(
        prem_ape,
        probs = c(0, 0.33, 0.67, 1.0)
      ),
      labels = c("Low", "Medium", "High"),
      include.lowest = TRUE
    )
  )

# Verify distribution
training_tbl_premium |>
  count(premium_tier) |>
  mutate(pct = scales::percent(n / sum(n))) |>
  print()
```

### Premium Tier Survival

```{r create_premium_km}
#| echo: true

premium_km <- survfit(
  Surv(policy_lifetime, lapsed) ~ premium_tier,
  data = training_tbl_premium,
  conf.type = "log-log"
)

premium_km |> summary(times = seq(0, 104, by = 4))
```

```{r plot_premium_km}
#| echo: true

premium_km |>
  ggsurvplot(
    censor = FALSE,
    conf.int = TRUE,
    conf.int.style = "step",
    title = "Survival Curves by Premium Tier",
    xlab = "Policy Duration (years)",
    ylab = "Survival Probability"
  )
```

### Premium Tier Difference Testing

```{r test_premium_differences}
#| echo: true

premium_survdiff <- survdiff(
  Surv(policy_lifetime, lapsed) ~ premium_tier,
  data = training_tbl_premium
)

premium_survdiff |> print()

write_lines("\nSummary Statistics by Premium Tier:", stdout())
training_tbl_premium |>
  group_by(premium_tier) |>
  summarise(
    count = n(),
    lapsed_count = sum(lapsed),
    lapse_rate = scales::percent(mean(lapsed), accuracy = 0.1),
    mean_premium = scales::label_dollar()(mean(prem_ape)),
    mean_duration = mean(policy_lifetime),
    .groups = "drop"
  ) |>
  print()
```

## Policy Amount Stratification

```{r discretize_sum_assured}
#| echo: true

# Create tiers for sum assured
training_tbl_sa <- training_tbl |>
  mutate(
    sa_tier = cut(
      sum_assured,
      breaks = quantile(
        sum_assured,
        probs = c(0, 0.33, 0.67, 1.0)
      ),
      labels = c("Low", "Medium", "High"),
      include.lowest = TRUE
    )
  )

# Verify distribution
training_tbl_sa |>
  count(sa_tier) |>
  mutate(pct = scales::percent(n / sum(n))) |>
  print()
```

### Sum Assured Tier Survival

```{r create_sa_km}
#| echo: true

sa_km <- survfit(
  Surv(policy_lifetime, lapsed) ~ sa_tier,
  data = training_tbl_sa,
  conf.type = "log-log"
)

sa_km |> summary(times = seq(0, 104, by = 4))
```

```{r plot_sa_km}
#| echo: true

sa_km |>
  ggsurvplot(
    censor = FALSE,
    conf.int = TRUE,
    conf.int.style = "step",
    title = "Survival Curves by Sum Assured Tier",
    xlab = "Policy Duration (years)",
    ylab = "Survival Probability"
  )
```

### Sum Assured Tier Difference Testing

```{r test_sa_differences}
#| echo: true

sa_survdiff <- survdiff(
  Surv(policy_lifetime, lapsed) ~ sa_tier,
  data = training_tbl_sa
)

sa_survdiff |> print()

write_lines("\nSummary Statistics by Sum Assured Tier:", stdout())
training_tbl_sa |>
  group_by(sa_tier) |>
  summarise(
    count = n(),
    lapsed_count = sum(lapsed),
    lapse_rate = scales::percent(mean(lapsed), accuracy = 0.1),
    mean_sa = scales::label_dollar()(mean(sum_assured)),
    mean_duration = mean(policy_lifetime),
    .groups = "drop"
  ) |>
  print()
```

# Hazard Rate Estimation

The hazard rate measures the instantaneous risk of lapse at each
duration. While the Kaplan-Meier estimator focuses on survival
probability, hazard rates help us understand *when* lapses occur—whether
risk is concentrated early (early lapse risk) or spread across durations
(ongoing maintenance risk).

```{r compute_smoothed_hazards}
#| echo: true

# Estimate smoothed hazard rates using muhaz
library(muhaz)

# Overall hazard
overall_hazard <- kphaz.fit(
  training_tbl$policy_lifetime,
  training_tbl$lapsed,
  method = "nelson"
)

# Hazard by gender
male_data <- training_tbl |> filter(gender_life1 == "M")
female_data <- training_tbl |> filter(gender_life1 == "F")

male_hazard <- kphaz.fit(
  male_data$policy_lifetime,
  male_data$lapsed,
  method = "nelson"
)

female_hazard <- kphaz.fit(
  female_data$policy_lifetime,
  female_data$lapsed,
  method = "nelson"
)
```

### Overall Hazard Rate

```{r plot_overall_hazard}
#| echo: true

# Create tibble for plotting
overall_hazard_tbl <- tibble(
  time = overall_hazard$time,
  hazard = overall_hazard$haz
)

ggplot(overall_hazard_tbl) +
  geom_line(aes(x = time, y = hazard), size = 1, color = "steelblue") +
  labs(
    title = "Overall Hazard Rate for Policy Lapse",
    x = "Policy Duration (years)",
    y = "Instantaneous Lapse Rate"
  )
```

The hazard curve reveals when lapses are most likely to occur. Early-peaked 
curves suggest lapse risk concentrates shortly after purchase; declining curves 
suggest risk decreases with duration; U-shaped curves combine both effects.

### Hazard by Gender

```{r plot_gender_hazard}
#| echo: true

gender_hazard_tbl <- bind_rows(
  tibble(
    time = male_hazard$time,
    hazard = male_hazard$haz,
    gender = "Male"
  ),
  tibble(
    time = female_hazard$time,
    hazard = female_hazard$haz,
    gender = "Female"
  )
)

ggplot(gender_hazard_tbl) +
  geom_line(aes(x = time, y = hazard, color = gender), size = 1) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Hazard Rate by Gender",
    x = "Policy Duration (years)",
    y = "Instantaneous Lapse Rate"
  )
```

# Validation Period Comparison

After understanding survival patterns in the training period, we ask: do
these patterns hold in the validation period? Comparing training and
validation survival curves reveals whether our exploratory findings
generalize.

```{r create_validation_km}
#| echo: true

# Validation period survival
validation_km <- survfit(
  Surv(policy_lifetime, lapsed) ~ 1,
  data = validation_tbl,
  conf.type = "log-log"
)

# Extract curves for comparison using ggsurvplot outputs
training_km_obj <- ggsurvplot(
  survfit(Surv(policy_lifetime, lapsed) ~ 1, data = training_tbl),
  conf.int = TRUE,
  censor = FALSE
)

validation_km_obj <- ggsurvplot(
  survfit(Surv(policy_lifetime, lapsed) ~ 1, data = validation_tbl),
  conf.int = TRUE,
  censor = FALSE
)

training_km_data <- training_km_obj$plot$data |> mutate(period = "Training")
validation_km_data <- validation_km_obj$plot$data |>
  mutate(period = "Validation")
```

### Comparing Training and Validation Survival

```{r plot_train_validation_km}
#| echo: true

# Manual KM curve comparison
train_km <- survfit(Surv(policy_lifetime, lapsed) ~ 1, data = training_tbl)
valid_km <- survfit(Surv(policy_lifetime, lapsed) ~ 1, data = validation_tbl)

# Combine tidy data for plotting
combined_km_tbl <- bind_rows(
  train_km |> tidy() |> mutate(period = "Training"),
  valid_km |> tidy() |> mutate(period = "Validation")
)

ggplot(combined_km_tbl) +
  geom_line(aes(x = time, y = estimate, color = period), size = 1) +
  geom_line(
    aes(x = time, y = conf.high, color = period),
    linetype = "dashed",
    alpha = 0.5
  ) +
  geom_line(
    aes(x = time, y = conf.low, color = period),
    linetype = "dashed",
    alpha = 0.5
  ) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Survival Curves: Training vs. Validation Period",
    x = "Policy Duration (years)",
    y = "Survival Probability",
    color = "Period"
  ) +
  theme(legend.position = "bottom")
```

### Validation Gender Comparison

```{r compare_gender_validation}
#| echo: true

# Training period gender survival
train_gender_km <- survfit(
  Surv(policy_lifetime, lapsed) ~ gender_life1,
  data = training_tbl
)

# Validation period gender survival
valid_gender_km <- survfit(
  Surv(policy_lifetime, lapsed) ~ gender_life1,
  data = validation_tbl
)

# Compare via log-rank test
train_gender_diff <- survdiff(
  Surv(policy_lifetime, lapsed) ~ gender_life1,
  data = training_tbl
)

valid_gender_diff <- survdiff(
  Surv(policy_lifetime, lapsed) ~ gender_life1,
  data = validation_tbl
)

write_lines("Training Period Gender Comparison:", stdout())
train_gender_diff |> print()

write_lines("\n\nValidation Period Gender Comparison:", stdout())
valid_gender_diff |> print()
```

The persistence of gender differences across periods suggests the
pattern is robust to data change, increasing confidence in its business
relevance.

# Summary Insights

## Key Findings

Exploratory survival analysis reveals several consistent patterns:

  1. **Baseline persistence**: Approximately 75% of policies survive
    the first 5 years, with persistence declining to roughly 50% by year
    10, indicating a moderate lapse experience typical for life insurance
    portfolios
  1. **Gender differences**: Female policyholders show modestly better
    persistence than males across all durations, suggesting gender may be
    a protective factor in lapse behaviour
  1. **Smoker effects**: Smokers do not demonstrate noticeably higher lapse
    rates compared to non-smokers, though smokers deemed to have 'quit'
    smoking exhibit higher lapsing than the other two categories
  1. **Vintage patterns**: Policies issued in earlier years show different
    lapse trajectories, likely reflecting changes in underwriting criteria,
    product features, or policyholder behaviour across cohorts
  1. **Premium stratification**: Higher-premium policies tend to show better
    persistence, possibly reflecting stronger policyholder engagement or
    commitment for larger coverage amounts
  1. **Early lapse concentration**: Hazard rate analysis suggests lapse risk
    peaks in the early policy years (months 3-12), then declines, consistent
    with "lapse shock" behaviour where dissatisfied customers exit quickly

## Interpretation and Business Implications

These univariate findings establish baseline understanding but don't
account for confounding. For instance, gender differences might reflect
underlying age or health status differences rather than gender itself.
Multivariate modelling (Cox regression) is needed to isolate independent
effects of each variable while controlling for others.

## Next Steps

The next worksheet shifts to **Cox proportional hazards models**, which allow us to:

  1. Estimate independent effects of multiple variables simultaneously
  1. Incorporate continuous variables directly without discretization loss
  1. Generate individual-level risk predictions
  1. Validate model assumptions and fit

# R Environment

```{r show_session_info}
#| echo: false
#| message: true

sessioninfo::session_info()
```
