---
title: "Dublin Data Science Workshop on the Statistical Analysis of Networks"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Monday, March 26 2018"
output:
  html_document:
    fig_caption: yes
    number_sections: yes
    theme: cerulean
    toc: yes
  word_document:
    toc: yes
  pdf_document: default
---

```{r knit_opts, include = FALSE}
knitr::opts_chunk$set(tidy  = FALSE
                     ,cache = FALSE
                     ,message = FALSE
                     ,warning = FALSE
                     ,fig.height =  8
                     ,fig.width  = 11)

library(tidyverse)
library(scales)
library(cowplot)

library(ggnetwork)
library(igraph)
library(igraphdata)
library(sand)


options(width = 80L
       ,warn  = 1
        )

set.seed(42)
```



# Introduction to Network Data

## Example Networks

### Florentine Marriage Data

```{r load_florentine_data, echo=TRUE}
data(flo, package = 'network')

florence_igraph <- graph_from_adjacency_matrix(flo, mode = 'undirected')

plot(florence_igraph)
```

We can use the package `ggnetwork` to allow us to plot networks within `ggplot2`

```{r plot_florentine_ggnetwork, echo=TRUE}
ggplot(florence_igraph, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges() +
    geom_label(aes(label = vertex.names), size = 5) +
    theme_blank()
```

```{r show_graph_details, echo=TRUE}
### Show graph vertices
V(florence_igraph)


### Show graph edges
E(florence_igraph)
```


To help access the edgelist in a more usable form we convert the edgelist to a
matrix, showing the origin and destination nodes.

```{r show_florence_edgelist, echo=TRUE}
get.edgelist(florence_igraph)
```

We also want to look at the adjacency matrix for this network

```{ show_florence_adjacency, echo=TRUE}
get.adjacency(florence_igraph)
```

The elements of a graph can all have attributes:

```{r show_element_attributes, echo=TRUE}
igraph::list.graph.attributes(florence_igraph)

igraph::list.vertex.attributes(florence_igraph)

igraph::list.edge.attributes(florence_igraph)
```


Subgraphs are subsets of graphs that are part of the whole.

```{r show_florence_subgraph, echo=TRUE}
family_keep <- c('Medici', 'Barbadori', 'Ridolfi','Tornabuoni','Pazzi'
                ,'Salviati', 'Albizzi', 'Guadagni')

florence_subgraph <- induced_subgraph(florence_igraph, family_keep)

ggplot(florence_subgraph, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges() +
    geom_label(aes(label = vertex.names), size = 5) +
    theme_blank()
```

### US Airport Data

```{r load_usairport_data, echo=TRUE}
data(USairports)


### Show graph vertices
V(USairports)


### Show graph edges
E(USairports)
```

This is a much larger network, and visualising it is likely going to be a mess,
but we will try anyway.

```{r plot_usairport_data, echo=TRUE}
ggplot(USairports, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges() +
    geom_label(aes(label = vertex.names), size = 5) +
    theme_blank()
```

Yeah, it is a mess.

We will try again with a small subgraph, using just 5 nodes

```{r plot_usairport_subgraph, echo=TRUE}
usairport_subgraph <- induced_subgraph(USairports, 1:10)

ggplot(usairport_subgraph, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges() +
    geom_label(aes(label = vertex.names), size = 5) +
    theme_blank()
```

Now that we have seen this network, we look at what additional information is
here.

```{r show_element_attributes_usairports, echo=TRUE}
igraph::list.graph.attributes(USairports)

igraph::list.vertex.attributes(USairports)

igraph::list.edge.attributes(USairports)
```

We see that the edges in particular now have a number of attributes. We access
them through `edge_attr()`

```{r list_airport_edge_attributes, echo=TRUE}
edge_attr(USairports) %>%
    as_tibble()
```

### Lazega Lawyers Network

```{r load_lazega_data, echo=TRUE}
data(lazega)

lazega <- lazega %>% upgrade_graph()  # Data is in deprecated format.


### Show graph vertices
V(lazega)


### Show graph edges
E(lazega)
```


```{r plot_lazega_data, echo=TRUE}
ggplot(lazega, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges() +
    geom_label(aes(label = vertex.names), size = 3) +
    theme_blank()
```

```{r show_element_attributes_lazega, echo=TRUE}
igraph::list.graph.attributes(lazega)

igraph::list.vertex.attributes(lazega)

igraph::list.edge.attributes(lazega)
```

We see that the vertices have the attributes but the edges have none.

```{r show_lazega_edge_attr, echo=TRUE}
vertex_attr(lazega) %>%
    as_tibble()
```


### Exercises

  1. Read in the network data contained in the file `NetScience.net`.
  1. Visualise the network using both base `plot` and `ggplot2`.
  1. Investigate the vertex and edge attributes.
  1. Is this graph directed or undirected?
  1. How many edges and vertices does it have?
  1. Is it a simple graph?
 


## Visualising Network Data

Plotting network data is not automatic - mathematical concepts allow us to
convert network topology into a form amenable to plotting.

```{r plot_florence_graph, echo=TRUE}
plot_1 <- ggplot(florence_igraph, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges() +
    geom_label(aes(label = vertex.names), size = 3) +
    theme_blank()

plot_2 <- ggplot(florence_igraph, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges() +
    geom_label(aes(label = vertex.names), size = 3) +
    theme_blank()

plot_3 <- ggplot(florence_igraph, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges() +
    geom_label(aes(label = vertex.names), size = 3) +
    theme_blank()

plot_4 <- ggplot(florence_igraph, aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges() +
    geom_label(aes(label = vertex.names), size = 3) +
    theme_blank()


plot_grid(plot_1, plot_2, plot_3, plot_4, ncol = 2)
```

### Creating Layouts

To plot graphs visually, we need a way to transform the graphs into 2D
coordinates. A number of layout algorithms exist.

To show the differences, we plot the Florentine network using a number of
different layout algorithms.

```{r plot_florence_fr, echo=TRUE}
ggplot(ggnetwork(florence_igraph, layout = 'fruchtermanreingold')
      ,aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges() +
    geom_label(aes(label = vertex.names)) +
    ggtitle('The Florentine Network Using Fruchterman-Reingold Layout') +
    theme_blank()
```


```{r plot_florence_spring, echo=TRUE}
ggplot(ggnetwork(florence_igraph, layout = 'spring')
      ,aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges() +
    geom_label(aes(label = vertex.names)) +
    ggtitle('The Florentine Network Using Spring Layout') +
    theme_blank()
```


```{r plot_florence_mds, echo=TRUE}
ggplot(ggnetwork(florence_igraph, layout = 'mds')
      ,aes(x = x, y = y, xend = xend, yend = yend)) +
    geom_edges() +
    geom_label(aes(label = vertex.names)) +
    ggtitle('The Florentine Network Using MDS Layout') +
    theme_blank()
```



### Exercises

  1. Produce similar visualisations as appropriate for the other graphs.
  1. Investigate alternative layout algorithms for the Florentine network.
  1. Try those layouts on the other network data.
  
  

# Network Statistics

## Network Degree

### Degree of Vertex

The *degree* of a vertex is the count of connections from that vertex.

We now look at the distribution of vertex degree for the Florentine network:


```{r plot_florentine_degree_distribution, echo=TRUE}
ggplot() +
    geom_bar(aes(x = igraph::degree(florence_igraph))) +
    xlab("Vertex Degree") +
    ylab("Count of Degrees")
```

### Average Neighbour Degree

Another interesting quantity is the average degree of all the neighbours of
each vertex.

```{r plot_degree_knn, echo=TRUE}
flor_knn <- graph.knn(florence_igraph)$knn

flor_knn_tbl <- data_frame(family = names(flor_knn)
                          ,degree = igraph::degree(florence_igraph)
                          ,knn    = flor_knn
                           )


ggplot(flor_knn_tbl) +
    geom_text(aes(x = degree, y = knn, label = family)) +
    expand_limits(y = 0) +
    xlab("Vertex Degree") +
    ylab("KNN Degree")
```


### Exercises

  1. Produce the degree distribution for the other networks
  1. Calaculate the average neighbour degree distribution for the other networks
  1. What can we infer about the importance of a vertex from these values?




## Measures of Centrality

### Between-ness Centrality

Between-ness measures how often a vertex comes between two other vertices in
the graph.

$$
c_B(\nu) = \sum_{s \neq t \neq \nu \in V} \frac{\sigma(s, t | \nu)}{\sigma(s,t)}
$$

where $\sigma(s, t| \nu)$ is the count of shortest paths between $s$ and $t$
that goes through $\nu$ and $\sigma(s, t)$ is the total number of shortest
paths between $s$ and $t$.

We now calculate the between-ness centrality for each vertex in the Florence
marriage network

```{r florence_graph_betweenness, echo=TRUE}
florence_betweenness <- florence_igraph %>%
    (igraph::betweenness)() %>%
    sort(decreasing = TRUE)

ggplot() +
    geom_col(aes(x = names(florence_betweenness), y = florence_betweenness)) +
    xlab("Family") +
    ylab("Between-ness") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```


### Closeness Centrality

Closeness centrality is measure of the distance of the vertex from other
vertices in the graph.

$$
c_{Cl}(\nu) = \frac{1}{\sum_{u \in V} \text{dist}(u, \nu)}
$$



```{r florence_graph_closeness, echo=TRUE}
florence_closeness <- florence_igraph %>%
    (igraph::closeness)() %>%
    sort(decreasing = TRUE)

ggplot() +
    geom_col(aes(x = names(florence_closeness), y = florence_closeness)) +
    xlab("Family") +
    ylab("Closeness") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```



### Eigenvector Centrality

Eigenvector centrality is a class of centrality measures based on linear
systems derived from the graph.

The most common of these are the eigenvectors of the adjacency matrix produced
from the graph.

The key idea here is that vertices which are 'central' in the network are so
due to their neighbours being 'central'. This concept is inherently implicit
in nature and so is calculated via linear algebra means.


```{r florence_graph_eigenvector, echo=TRUE}
florence_eigencent <- florence_igraph %>%
    igraph::eigen_centrality() %>%
    .$vector %>%
    sort(decreasing = TRUE)

ggplot() +
    geom_col(aes(x = names(florence_eigencent), y = florence_eigencent)) +
    xlab("Family") +
    ylab("Eigenvector Centrality") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```


### Exercises

  1. Calculate the centrality measures for the other networks


## Network Cohesion

### Cliques

```{r plot_clique_distribution, echo=TRUE}
clique_size <- florence_igraph %>%
    cliques %>%
    map_int(length)

ggplot() +
    geom_bar(aes(x = clique_size)) +
    xlab("Clique Size") +
    ylab("Count")
```


### Clustering Co-efficient / Transitivity

The transitivity of the network is a measure of the 'density' of connections
in the network. It is the ratio of triangles in the network to possible
triangles.

Local transitivity does the same for all triangles contain the vertex.

```{r calculate_transitivities, echo=TRUE}
florence_igraph %>%
    transitivity()

florence_igraph %>%
    transitivity(type = 'local', vids = c('Strozzi', 'Guadagni', 'Medici'))
```

### Shortest Paths

The shortest path measures are measures of the size and connectivity of the
graph

```{r show_distance_stats, echo=TRUE}
florence_igraph %>% mean_distance()

florence_igraph %>% diameter()
```

## Graph Partitioning

```{r partition_florence_graph, echo=TRUE}
florence_partition <- cluster_fast_greedy(florence_igraph)

florence_partition %>% str()
```

## Assortativity




# Random Graph Models

## Erdos-Renyi Graph Models


## Generalised Random Graph Models


## Small World Models


## Preferential Attachment Models


# Statistical Models for Graphs

