---
title: "Introduction to Copulas (Part 1: Basics and Elliptical Copulas)"
author: "Mick Cooney <mcooney@describedata.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---

```{r import_libraries}
#| echo: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(furrr)
library(glue)
library(fs)
library(copula)
library(VineCopula)
library(GGally)
library(lattice)
library(scatterplot3d)
library(broom)
library(arrow)
library(qs2)

source("lib_utils.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "copula")
  )

options(
  width    = 80L,
  warn     = 1,
  mc.cores = parallelly::availableCores()
  )

set.seed(42)

theme_set(theme_cowplot())

plan(multisession)
```

# Introduction

In many real-world applications, we are interested in the joint behavior of multiple random variables. While marginal distributions describe the behavior of each variable individually, they do not capture how the variables move together.

**Copulas** are mathematical tools that allow us to separate the modeling of marginal distributions from the modeling of the dependence structure between them.

## Sklar's Theorem

Sklar's Theorem is the foundation of copula theory. It states that any multivariate cumulative distribution function (CDF) $H(x, y)$ can be expressed in terms of its marginal CDFs $F(x)$ and $G(y)$, and a copula $C$:

$$H(x, y) = C(F(x), G(y))$$

where $C$ is a joint distribution function with uniform marginals on $[0, 1]$.

Conversely, given a copula $C$ and marginals $F$ and $G$, $C(F(x), G(y))$ defines a valid bivariate distribution.

# Probability Integral Transform (PIT)

The core idea behind copulas is the Probability Integral Transform. If $X$ is a continuous random variable with CDF $F$, then $U = F(X)$ is uniformly distributed on $[0, 1]$.

Let's illustrate this with an example.

```{r pit_example}
# Generate some data from a Gamma distribution
n_sim <- 5000
shape <- 2
rate  <- 0.5

x <- rgamma(n_sim, shape = shape, rate = rate)
u <- pgamma(x, shape = shape, rate = rate)

data_tbl <- tibble(x = x, u = u)

plot_x <- ggplot(data_tbl, aes(x = x)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "steelblue", alpha = 0.7) +
  stat_function(fun = dgamma, args = list(shape = shape, rate = rate), color = "red", linewidth = 1) +
  labs(title = "Original Gamma Distribution", x = "X", y = "Density")

plot_u <- ggplot(data_tbl, aes(x = u)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "darkorange", alpha = 0.7) +
  geom_hline(yintercept = 1, color = "red", linetype = "dashed") +
  labs(title = "PIT-transformed Uniform Distribution", x = "U = F(X)", y = "Density")

plot_grid(plot_x, plot_u, ncol = 2)
```

# Visualizing Dependence

When we have two variables, we can look at their joint distribution. If we transform both to uniform marginals using PIT, we can see the "copula" structure.

## Independent Variables

```{r independence_viz}
x1 <- rnorm(n_sim)
x2 <- rnorm(n_sim)

u1 <- pnorm(x1)
u2 <- pnorm(x2)

indep_tbl <- tibble(u1 = u1, u2 = u2)

ggplot(indep_tbl, aes(x = u1, y = u2)) +
  geom_point(alpha = 0.2, size = 0.5) +
  labs(title = "Independence Copula (Scatter Plot on [0,1]^2)", x = "U1", y = "U2")
```

## Dependent Variables

```{r dependence_viz}
# Generate dependent variables using a multivariate normal
sigma <- matrix(c(1, 0.7, 0.7, 1), 2, 2)
z <- MASS::mvrnorm(n_sim, mu = c(0, 0), Sigma = sigma)

u1_dep <- pnorm(z[, 1])
u2_dep <- pnorm(z[, 2])

dep_tbl <- tibble(u1 = u1_dep, u2 = u2_dep)

ggplot(dep_tbl, aes(x = u1, y = u2)) +
  geom_point(alpha = 0.2, size = 0.5) +
  labs(title = "Gaussian Copula (rho = 0.7)", x = "U1", y = "U2")
```

# Elliptical Copulas

Elliptical copulas are derived from elliptical distributions (like Normal or t).

## Gaussian Copula

The Gaussian copula is the most common. It is defined by its correlation matrix.

```{r gaussian_copula_demo}
# Using the 'copula' package
gauss_cop <- normalCopula(param = 0.5, dim = 2)
persp(gauss_cop, dCopula, main = "Gaussian Copula Density")
contour(gauss_cop, dCopula, main = "Gaussian Copula Density Contours")
```

## t-Copula

The t-copula adds a degree of freedom parameter, allowing for "tail dependence" (more extreme events occurring together).

```{r t_copula_demo}
t_cop <- tCopula(param = 0.5, dim = 2, df = 2)
persp(t_cop, dCopula, main = "t-Copula Density (df = 2)")
contour(t_cop, dCopula, main = "t-Copula Density Contours (df = 2)")
```

# Exercise 1: Exploring Parameters

Try changing the correlation parameter `rho` and the degrees of freedom `df` for the t-copula. How does the scatter plot on the uniform square change?

```{r exercise_1}
# YOUR CODE HERE
# Hint: use tCopula(param = ..., df = ...) and rCopula(n, copula)
```

# Summary

In this part, we learned:
- Copulas separate marginals from dependence.
- PIT transforms any continuous variable to a Uniform(0,1).
- Elliptical copulas (Gaussian and t) are easy to use but have specific properties.

In the next part, we will look at Archimedean copulas and how to fit copulas to real data.
