---
title: "Copula Modelling (Part 2: Archimedean Copulas and Fitting)"
author: "Mick Cooney <mcooney@describedata.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---

```{r import_libraries}
#| echo: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(furrr)
library(glue)
library(fs)
library(copula)
library(VineCopula)
library(GGally)
library(broom)
library(arrow)
library(qs2)

source("lib_utils.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "copula")
  )

options(
  width    = 80L,
  warn     = 1,
  mc.cores = parallelly::availableCores()
  )

set.seed(42)

theme_set(theme_cowplot())

plan(multisession)
```

# Archimedean Copulas

Archimedean copulas are a class of copulas that are easy to construct and have diverse dependence properties, particularly regarding tail dependence.

## Clayton Copula (Lower Tail Dependence)

The Clayton copula is useful for modeling variables that are strongly dependent in the lower tail (e.g., stock market crashes).

```{r clayton_demo}
clayton_cop <- claytonCopula(param = 2, dim = 2)
persp(clayton_cop, dCopula, main = "Clayton Copula Density")

sim_clayton <- rCopula(2000, clayton_cop)
ggplot(as.data.frame(sim_clayton), aes(x = V1, y = V2)) +
  geom_point(alpha = 0.3) +
  labs(title = "Clayton Copula Samples (Lower Tail Dependence)")
```

## Gumbel Copula (Upper Tail Dependence)

The Gumbel copula models strong dependence in the upper tail.

```{r gumbel_demo}
gumbel_cop <- gumbelCopula(param = 2, dim = 2)
persp(gumbel_cop, dCopula, main = "Gumbel Copula Density")

sim_gumbel <- rCopula(2000, gumbel_cop)
ggplot(as.data.frame(sim_gumbel), aes(x = V1, y = V2)) +
  geom_point(alpha = 0.3) +
  labs(title = "Gumbel Copula Samples (Upper Tail Dependence)")
```

## Frank Copula (Symmetric)

The Frank copula is symmetric and does not have tail dependence.

```{r frank_demo}
frank_cop <- frankCopula(param = 5, dim = 2)
persp(frank_cop, dCopula, main = "Frank Copula Density")

sim_frank <- rCopula(2000, frank_cop)
ggplot(as.data.frame(sim_frank), aes(x = V1, y = V2)) +
  geom_point(alpha = 0.3) +
  labs(title = "Frank Copula Samples (Symmetric)")
```

# Modelling Process

The standard way to fit a copula model is:
1. Estimate marginal distributions for each variable.
2. Transform the data to the uniform scale $[0, 1]$ using the estimated CDFs (or empirical CDF).
3. Fit the copula to the transformed data.

## Example: Fitting to Synthetic Data

Let's create some data with Gamma and Lognormal marginals and a Gumbel dependence.

```{r synthetic_data}
n <- 1000
gumbel_cop <- gumbelCopula(param = 3, dim = 2)
u <- rCopula(n, gumbel_cop)

# Transform to marginals
x1 <- qgamma(u[, 1], shape = 2, rate = 1)
x2 <- qlnorm(u[, 2], meanlog = 0, sdlog = 0.5)

data_tbl <- tibble(x1 = x1, x2 = x2)

ggplot(data_tbl, aes(x = x1, y = x2)) +
  geom_point(alpha = 0.5) +
  scale_x_log10() + scale_y_log10() +
  labs(title = "Synthetic Data (Gumbel Dependence)")
```

## Step 1: Fitting Marginals (Empirical approach)

```{r fit_marginals}
u1_hat <- pobs(data_tbl$x1)
u2_hat <- pobs(data_tbl$x2)

u_hat <- cbind(u1_hat, u2_hat)

plot(u_hat, xlab = "U1 hat", ylab = "U2 hat", main = "Empirical Copula Data")
```

## Step 2: Selecting and Fitting a Copula

We can use `VineCopula::BiCopSelect` to automatically select the best copula family based on AIC/BIC.

```{r select_copula}
fit_cop <- BiCopSelect(u1_hat, u2_hat, familyset = NA) # NA means check all available
summary(fit_cop)
```

# Simulation from a Fitted Model

Once we have the marginals and the copula, we can simulate new data.

```{r simulation}
# Simulate from the fitted copula
sim_u <- BiCopSim(5000, fit_cop$family, fit_cop$par, fit_cop$par2)

# Transform back to original marginals (using empirical quantiles for simplicity here)
sim_x1 <- quantile(data_tbl$x1, sim_u[, 1])
sim_x2 <- quantile(data_tbl$x2, sim_u[, 2])

sim_tbl <- tibble(x1 = sim_x1, x2 = sim_x2)

ggplot(sim_tbl, aes(x = x1, y = x2)) +
  geom_point(alpha = 0.2, color = "blue") +
  scale_x_log10() + scale_y_log10() +
  labs(title = "Simulated Data from Fitted Copula Model")
```

# Exercise 2: Real Data Challenge

If you have a dataset with two related variables (e.g., height and weight, or two stock returns), try fitting a copula to them.

1. Transform to uniform scale using `pobs()`.
2. Use `BiCopSelect()` to find the best copula.
3. Compare the Gaussian copula fit with the selected one.

# Summary

- Archimedean copulas (Clayton, Gumbel, Frank) offer different tail behaviors.
- Fitting involves a two-stage process: Marginals then Copula.
- `VineCopula` is a powerful tool for selecting and fitting bivariate (and multivariate) copulas.
