---
title: "Workshop: Introduction to Copulas"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---

```{r import_libraries}
#| echo: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(furrr)
library(glue)
library(fs)
library(copula)
library(copulaData)
library(VineCopula)
library(rvinecopulib)
library(GGally)
library(lattice)
library(scatterplot3d)
library(broom)
library(arrow)
library(qs2)

source("lib_utils.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "copula", "rvinecopulib")
)

options(
  width    = 80L,
  warn     = 1,
  mc.cores = parallelly::availableCores()
)

set.seed(42)

theme_set(theme_cowplot())

plan(multisession)
```

# Foundations of Dependency: Beyond Simple Correlation

In many real-world applications, we are interested in how multiple random variables 
behave together. While we often have a good handle on individual variables—like 
the height of a wave or the number of insurance claims in a year—it is the 
**joint behavior** that usually creates the most risk.

## The "Aggregation Problem" and the Lego Analogy

Traditional statistics often forces us into "one-size-fits-all" multivariate 
distributions. For example, the Multivariate Normal distribution requires that 
every variable follows a Normal distribution. In reality, data is messy: you 
might have a Gamma-distributed frequency of events linked to a Lognormal-distributed 
severity of costs.

**Copulas** solve this through the **Separation Principle**. Think of your 
multivariate model like a set of Lego bricks:

1.  **The Bricks (Marginals)**: You choose the best "brick" (distribution) for 
    each individual variable.

2.  **The Connectors (Copula)**: The copula is the specialized "peg" that 
    describes exactly how those bricks snap together, independent of their 
    shape or color.

## The Engine: Probability Integral Transform (PIT)

The magic that allows us to separate the "bricks" from the "connectors" is the 
**Probability Integral Transform (PIT)**. It acts as a universal translator. 

If you take any continuous data and pass it through its own Cumulative 
Distribution Function (CDF), you end up with a value between 0 and 1—essentially 
its "percentile rank." By converting everything to this uniform scale, we can 
link variables that have completely different units (like "Degrees Celsius" and 
"Megawatts") without the units getting in the way.

The following plots demonstrate this "uniformization" and show what a high-correlation 
relationship looks like once we've moved into this "Copula Space."

```{r theory_viz}
#| echo: true
#| fig-cap: "PIT and Copula Visualisation"

n_sim <- 5000

# PIT Demonstration: Transforming Gamma to Uniform
x <- rgamma(n_sim, shape = 2, rate = 0.5)
u <- pgamma(x, shape = 2, rate = 0.5)

plot_pit <- ggplot(tibble(u = u), aes(x = u)) +
  geom_histogram(bins = 30, fill = "darkorange", alpha = 0.7) +
  geom_hline(yintercept = n_sim / 30, color = "red", linetype = "dashed") +
  labs(
    x     = "U = F(X) (Percentile Rank)",
    y     = "Count",
    title = "PIT: Gamma to Uniform"
  )

# Dependence Visualization: Gaussian Copula
sigma <- matrix(c(1, 0.8, 0.8, 1), 2, 2)
z     <- MASS::mvrnorm(n_sim, mu = c(0, 0), Sigma = sigma)
u_dep <- pnorm(z) |> as_tibble(.name_repair = "unique")

plot_dep <- ggplot(u_dep, aes(x = ...1, y = ...2)) +
  geom_point(alpha = 0.2, size = 0.5, color = "steelblue") +
  labs(
    x     = "U1 (Variable A Rank)",
    y     = "U2 (Variable B Rank)",
    title = "Dependency in Copula Space (rho = 0.8)"
  )

plot_grid(plot_pit, plot_dep, ncol = 2)
```

# The Copula Zoo: Families and their "Tail Signatures"

Not all dependencies are created equal. Correlation tells us how variables 
move *on average*, but in risk management, we care about the **extremes**. 
This is where different "families" of copulas come into play.

## Rank Correlation: Measuring the "Order" of Things

Traditional correlation (Pearson’s $r$) is obsessed with straight lines. If 
your relationship is curved, Pearson's $r$ will underestimate its strength. 
In copula modelling, we use **Rank Correlation** (like Kendall’s Tau). It 
doesn't care about the scale or the units; it only cares about the *order* of 
the values. It is invariant to transformations—whether you model loss in 
Dollars or Log-Dollars, the rank correlation stays the same.

```{r rank_correlation_viz}
#| echo: true
#| fig-cap: "Pearson vs. Kendall Correlation under Non-Linear Transformation"

# Exponential relationship
x_lin <- seq(1, 10, length.out = 100)
y_exp <- exp(x_lin) + rnorm(100, sd = 1000)

cor_p <- cor(x_lin, y_exp, method = "pearson")
cor_k <- cor(x_lin, y_exp, method = "kendall")

tibble(x = x_lin, y = y_exp) |>
  ggplot(aes(x = x, y = y)) +
  geom_point(color = "darkblue") +
  labs(
    title    = "The Limit of Linear Correlation",
    subtitle = glue("Pearson r = {round(cor_p, 3)} (Low) vs. Kendall tau = {round(cor_k, 3)} (High)"),
    caption  = "Pearson misses the strength of the curved relationship; Kendall sees the perfect order."
  )
```

## Tail Dependence: "Ride-or-Die" vs. "Fair-Weather Friends"

This is perhaps the most critical concept in copula theory. It describes how 
likely two variables are to "crash" or "boom" together.

*   **Fair-Weather Friends (Gaussian Copula)**: These variables look correlated 
    in normal times, but as you move to the extreme edges, they actually become 
    independent. This "asymptotic independence" is why Gaussian models 
    famously failed during the 2008 financial crisis—they assumed that 
    just because two markets hadn't crashed together *recently*, they 
    *couldn't* crash together in the extreme.

*   **Ride-or-Die (t-Copula, Gumbel, Clayton)**: These exhibit **Tail Dependence**. 
    If Variable A is in its worst 1%, Variable B is highly likely to be there 
    too. 

### Visualizing the "Corners" of Risk

The plots below show how different families concentrate their risk in different 
"corners" of the uniform space.

```{r archimedean_viz}
#| echo: true
#| fig-cap: "Tail Dependence Signatures in Archimedean Copulas"

n_plot <- 2000

# Helper to simulate and plot
plot_arch_comp <- function(family, param_val, title_text, color_val) {
  if (family == "clayton") {
    cop <- claytonCopula(param = param_val, dim = 2)
  } else if (family == "gumbel") {
    cop <- gumbelCopula(param = param_val, dim = 2)
  } else {
    cop <- frankCopula(param = param_val, dim = 2)
  }

  u_sim_tbl <- rCopula(n_plot, cop) |>
    as_tibble()

  ggplot(u_sim_tbl, aes(x = V1, y = V2)) +
    geom_point(alpha = 0.3, color = color_val, size = 0.5) +
    labs(
      x     = "Rank U1",
      y     = "Rank U2",
      title = title_text
    ) +
    theme_minimal()
}

# Clayton (Lower Tail - Droughts/Crashes)
p1 <- plot_arch_comp("clayton", 2, "Clayton (Lower Tail Concentration)", "darkgreen")
# Gumbel (Upper Tail - Heatwaves/Booms)
p2 <- plot_arch_comp("gumbel", 2, "Gumbel (Upper Tail Concentration)", "firebrick")
# Frank (Symmetric - No Tail dependence)
p3 <- plot_arch_comp("frank", 5, "Frank (Symmetric / No Tails)", "purple")
# Student's t (Both Tails - Heavy Tails)
cop_t <- tCopula(param = 0.5, df = 2, dim = 2)
u_t   <- rCopula(n_plot, cop_t) |> as_tibble()
p4    <- ggplot(u_t, aes(x = V1, y = V2)) +
  geom_point(alpha = 0.3, color = "blue", size = 0.5) +
  labs(title = "Student's t (Both Tails)", x = "U1", y = "U2") +
  theme_minimal()

plot_grid(p1, p2, p3, p4, ncol = 2)
```

## The Critical Role of Degrees of Freedom ($df$)

In the **Student's t-Copula**, the $df$ parameter is our "dial" for tail risk. 
Low $df$ means extreme events happen together very often. As $df$ increases, 
the t-copula slowly turns into a Gaussian copula, and that joint tail risk 
evaporates.

```{r t_copula_viz}
#| echo: true
#| fig-cap: "The 'Normalization' of the t-Copula as df Increases"

n_plot  <- 2000
rho_fix <- 0.5

# Helper to simulate and plot
plot_t_comp <- function(df_val, title_text, color_val) {
  if (is.infinite(df_val)) {
    cop <- normalCopula(param = rho_fix, dim = 2)
  } else {
    cop <- tCopula(param = rho_fix, df = df_val, dim = 2)
  }

  u_sim_tbl <- rCopula(n_plot, cop) |>
    as_tibble(.name_repair = "unique")

  ggplot(u_sim_tbl, aes(x = ...1, y = ...2)) +
    geom_point(alpha = 0.4, color = color_val, size = 0.6) +
    labs(x = "U1", y = "U2", title = title_text) +
    theme_minimal()
}

p1_t <- plot_t_comp(1, "t (df=1): Heavy Tails", "firebrick")
p2_t <- plot_t_comp(30, "t (df=30): Almost Gaussian", "darkblue")

plot_grid(p1_t, p2_t, ncol = 2)
```

# Real-World Scenarios: Modelling Compounding Risk

In risk management, **compounding risk** occurs when multiple "bad" things 
happen at once. Copulas allow us to model these dependencies explicitly, 
even when the types of data are very different.

## Case Study: Insurance - Frequency and Severity

Imagine a year with a high number of storm events (Frequency). If those 
storms are also individually more severe (Severity), the total annual loss 
balloons. This is compounding risk.

```{r compounding_risk_sim}
#| echo: true

simulate_total_costs <- function(n_years, rho, lambda, shp, rt) {
  cop      <- normalCopula(param = rho, dim = 2)
  u_sim    <- rCopula(n_years, cop)
  n_events <- qpois(u_sim[, 1], lambda = lambda)

  total_costs <- map_dbl(
    1:n_years,
    function(i) {
      n <- n_events[i]
      if (n == 0) return(0)
      sev_p    <- u_sim[i, 2]
      year_sev <- qgamma(sev_p, shape = shp, rate = rt)
      return(n * year_sev)
    }
  )
  return(total_costs)
}

# Run simulations across correlation levels
n_years     <- 10000
costs_indep <- simulate_total_costs(n_years, 0.0, 10, 2, 0.02)
costs_high  <- simulate_total_costs(n_years, 0.7, 10, 2, 0.02)

cost_data_tbl <- tibble(
  correlation = rep(c("0.0 (Indep)", "0.7 (High)"), each = n_years),
  total_cost  = c(costs_indep, costs_high)
)
```

### The Impact on the "Tail"

Observe how high correlation doesn't just change the average; it dramatically 
increases the probability of extreme, "bankrupting" years.

```{r cost_density}
#| echo: true
#| fig-cap: "Compounding Risk: Impact of Correlation on Annual Loss"

ggplot(cost_data_tbl, aes(x = total_cost, fill = correlation)) +
  geom_density(alpha = 0.4) +
  scale_x_log10(labels = dollar) +
  labs(x = "Total Annual Cost (Log Scale)", title = "Log Scale: Seeing the Tail Risk")
```

## Diverse Coupling Examples

Copulas can link any two metrics that share a common environmental or 
economic driver.

### Public Health: Correlating Discrete Events

High pharmacy sales of flu medication often coincide with high ER admissions. 
Modeling these as independent would underestimate the peak load on the city’s 
healthcare system.

```{r public_health_discrete}
#| echo: true
#| fig-cap: "Correlated Public Health Indicators (Discrete-Discrete)"

# Moderate-High Correlation
rho_health <- 0.65
cop_health <- normalCopula(param = rho_health, dim = 2)
u_health   <- rCopula(2000, cop_health)

# Map to Marginals (Negative Binomial and Poisson)
sales  <- qnbinom(u_health[, 1], mu = 200, size = 20)
er_adm <- qpois(u_health[, 2], lambda = 15)

health_data_tbl <- tibble(Sales = sales, ER = er_adm)

ggplot(health_data_tbl, aes(x = Sales, y = ER)) +
  geom_jitter(alpha = 0.3, color = "darkred") +
  labs(x = "Daily Pharmacy Sales", y = "ER Admissions", title = "Public Health Coupling")
```

### Agriculture: Discrete-Continuous Coupling (Drought Risk)

In agriculture, we might model the number of rainy days (Discrete) against 
the final crop yield percentage (Continuous). Low rainfall is strongly 
linked to low yields, a classic "lower tail" dependency.

```{r agriculture_coupling}
#| echo: true
#| fig-cap: "Rainfall Frequency vs. Crop Yield (Discrete-Continuous)"

# Clayton Copula captures 'Lower Tail Dependence'
cop_agri <- claytonCopula(param = 1.5, dim = 2)
u_agri   <- rCopula(2000, cop_agri)

rain_days  <- qpois(u_agri[, 1], lambda = 12)
crop_yield <- qbeta(u_agri[, 2], shape1 = 5, shape2 = 2)

agri_data_tbl <- tibble(Rain = rain_days, Yield = crop_yield * 100)

ggplot(agri_data_tbl, aes(x = Rain, y = Yield)) +
  geom_jitter(alpha = 0.3, color = "darkgreen") +
  labs(x = "Rainy Days per Month", y = "Crop Yield (% of Max)", title = "Agriculture Coupling: Drought Risk")
```

### Energy: Continuous-Continuous Coupling (Heatwaves)

Extreme heatwaves drive non-linear spikes in electricity demand as air 
conditioning units work harder. This relationship is concentrated in the 
"upper tail."

```{r energy_coupling}
#| echo: true
#| fig-cap: "Energy Coupling: Heatwaves and Power Demand"

# Gumbel Copula for 'Upper Tail Dependence' (Heat -> Demand)
cop_energy <- gumbelCopula(param = 2.5, dim = 2)
u_energy   <- rCopula(2000, cop_energy)

temp   <- qgamma(u_energy[, 1], shape = 20, rate = 0.8)
demand <- qlnorm(u_energy[, 2], meanlog = 5, sdlog = 0.3)

energy_data_tbl <- tibble(Temp = temp, Demand = demand)

ggplot(energy_data_tbl, aes(x = Temp, y = Demand)) +
  geom_point(alpha = 0.2, color = "orange") +
  labs(x = "Peak Temp (°C)", y = "Demand (MW)", title = "Physical Coupling: Heat vs. Demand")
```

# The Empirical Detective: Fitting Copulas to Data

In practice, we don't start with the copula; we start with raw data. 
Finding the right dependency structure is a three-step detective process 
called **Inference Functions for Margins (IFM)**.

## The IFM Workflow

1.  **Marginal Estimation**: Fit the best distributions to each variable 
    independently (the "bricks").

2.  **Transformation (PIT)**: Use the fitted CDFs to move the data into 
    uniform "Copula Space."

3.  **Selection and Fitting**: Choose the copula family that best describes 
    the "corners" of the data.

## A "Recovery Test": Proving the Method

Let's see if we can "recover" the true dependency from our earlier 
Energy example. We know the truth was a **Gumbel Copula**.

```{r recovery_pit}
#| echo: true
#| fig-cap: "The Copula Space: Revealing the Hidden Structure"

# Transform raw data back to uniform scale (pseudo-observations)
u_energy_tbl <- pobs(energy_data_tbl) |>
  as_tibble() |>
  set_colnames(c("U1", "U2"))

ggplot(u_energy_tbl, aes(x = U1, y = U2)) +
  geom_point(alpha = 0.2, color = "darkblue") +
  labs(x = "Temp Rank", y = "Demand Rank", title = "Copula Space: Pure Dependency")
```

### Automated Selection (AIC/BIC)

Using `rvinecopulib`, we can automatically test all families and select 
the best one.

```{r recovery_fitting}
#| echo: true

# Select best bivariate copula
fit_energy <- bicop(data = as.matrix(u_energy_tbl), family_set = "all")
print(fit_energy)
```

The algorithm correctly identifies the Gumbel family, proving we can 
extract dependency structures from raw observations.

## The Danger of Choosing the Wrong Model

What if we chose a Gaussian model when the truth was Gumbel? We would 
severely underestimate how often extreme heat and extreme demand happen 
together.

```{r model_comparison}
#| echo: true

# Fit various families to the same data
fit_gauss   <- bicop(as.matrix(u_energy_tbl), family_set = "gaussian")
fit_frank   <- bicop(as.matrix(u_energy_tbl), family_set = "frank")

# Compare AIC (Lower is better)
model_comparison_tbl <- tibble(
  Model  = c("Gumbel (True)", "Gaussian", "Frank"),
  AIC    = c(AIC(fit_energy), AIC(fit_gauss), AIC(fit_frank))
) |>
  arrange(AIC)

knitr::kable(model_comparison_tbl, digits = 1, caption = "Model Selection via AIC")
```

# Scaling Complexity: Multivariate and Vine Copulas

The real world is rarely bivariate. In finance, engineering, and 
environmental science, we often need to model the joint behavior of 
dozens of variables. This introduces the **Curse of Dimensionality**: 
as the number of variables grows, the complexity of capturing every 
pairwise and joint interaction increases exponentially.

## Multivariate Continuous Data: The Uranium Case Study

A classic high-dimensional example is the **Uranium exploration dataset** 
(Cook & Johnson, 1986). It contains concentrations of 7 chemical elements 
in 655 water samples. Capturing the "fingerprint" of these relationships 
allows geologists to identify anomalous patterns indicative of mineral 
deposits.

```{r uranium_analysis}
#| echo: true
#| fig-cap: "Pairwise Dependencies in 7D Uranium Data"

data(uranium, package = "copula")

# Focus on a few elements for clarity: Uranium, Lithium, Cobalt, Titanium
uranium_sub <- uranium |>
  as_tibble() |>
  select(U, Li, Co, Ti)

# Visualize the dependencies using ranks
ggpairs(
  pobs(uranium_sub),
  lower = list(continuous = wrap("points", alpha = 0.3, size = 0.5, color = "darkblue")),
  title = "Copula Space (Ranks) for Uranium Data"
)
```

Notice how different pairs exhibit different structures—some are tight and 
symmetric, others are sparse. A standard multivariate Gaussian copula 
would force a single, symmetric "average" structure on all these pairs, 
potentially missing the physical constraints of the geochemistry.

## Financial Contagion: The DAX Index

In finance, we care about **contagion**—the phenomenon where assets that 
are uncorrelated during normal times suddenly crash together during a 
crisis. The `daxreturns` dataset (daily log-returns of major DAX 
components) from the `VineCopula` package illustrates this.

```{r dax_analysis}
#| echo: true

data(daxreturns, package = "VineCopula")

# Focus on first few assets
dax_sub <- daxreturns[, 1:5]

# Fit a multivariate Student-t copula vs. Gaussian using rvinecopulib
# The Student-t captures the 'Ride-or-Die' crashing together
fit_gauss_multi <- bicop(pobs(dax_sub[, 1:2]), family_set = "gaussian")
fit_t_multi     <- bicop(pobs(dax_sub[, 1:2]), family_set = "t")

model_comp_dax <- tibble(
  Model = c("Gaussian", "Student-t"),
  AIC   = c(AIC(fit_gauss_multi), AIC(fit_t_multi))
)

knitr::kable(model_comp_dax, caption = "Contagion Check: Gaussian vs. t-Copula (DAX)")
```

The superior fit of the Student-t model confirms that market components 
exhibit significant tail dependence—they crash together far more often 
than a Gaussian model would predict.

## Vine Copulas: The Modular Solution

How do we scale this to 50 variables without losing the unique 
"tail signatures" of each pair? We use **Vine Copulas**.

Vine Copulas break down a high-dimensional distribution into a 
hierarchy of **Bivariate Copulas**. 

1.  **Trees**: We organize variables into "trees." The first tree links 
    the strongest pairs.
2.  **Conditioning**: The second tree links variables *conditional* on 
    the relationships in the first tree.
3.  **Flexibility**: You can use a Clayton connector for one pair and a 
    Gumbel for another.

```{r vine_deep_dive}
#| echo: true
#| fig-cap: "Visualizing a Fitted Vine Structure"

# Fit an R-vine (Regular Vine) to the DAX data
dax_vine <- vinecop(pobs(dax_sub), family_set = "all")

# Plot the first tree of the vine (the most important dependencies)
plot(dax_vine, tree = 1)
```

The vine structure automatically prioritizes the most important physical 
or economic connections, providing a "map" of how risk propagates 
through the system.

# Summary of the Copula Workflow

1.  **PIT**: Transform data to ranks to see the hidden dependence.

2.  **Visualize**: Check the "corners" of the Copula Space for tail risk.

3.  **Select**: Use AIC/BIC to find the right family (Lego connector).

4.  **Simulate**: Generate extreme scenarios to Stress-test your risk models.

# R Environment

```{r session_info}
#| echo: true

sessioninfo::session_info()
```
