---
title: "Perform Basic Classic Survival Models"
author: "Mick Cooney <mcooney@describedata.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8

---


```{r import_libraries}
#| echo: false
#| message: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(arrow)
library(fs)
library(purrr)
library(furrr)
library(glue)
library(survival)
library(survminer)
library(muhaz)
library(broom)
library(ggsurvfit)
library(gtsummary)


source("lib_utils.R")


conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "MASS",
    "fitdistrplus")
  )

options(
  width    = 80L,
  warn     = 1,
  mc.cores = parallelly::availableCores()
  )

Sys.setenv(OPENBLAS_NUM_THREADS = 8)

set.seed(42)
stanfit_seed <- 4000

theme_set(theme_cowplot())

plan(multisession)
```


# Load Policy Data

We first need to load the policy data so we might visualise some of this data.

```{r load_insample_policy_data}
#| echo: true

model_training_tbl <- read_parquet("data/model_training_small_tbl.parquet")

model_training_tbl |> glimpse()
```






# Perform Initial Survival Modelling Explorations

## Kaplan-Meier Estimates

### Unsplit

We start with some Kaplan-Meier estimates for the whole dataset, with no splits
in our data.

```{r calculate_basic_km_estimates}
#| echo: true

lapse1_km <- survfit2(
  Surv(policy_lifetime, lapsed) ~ 1,
  data      = model_training_tbl,
  conf.type = 'log-log'
  )

ggsurvfit(lapse1_km) +
  expand_limits(y = 0) +
  add_confidence_interval() +
  add_risktable()
```

We now want to see the overall estimates for the 1, 2 and 5-year rates.

```{r summary_lapse1}
#| echo: true

lapse1_km |>
  summary(times = 52 * c(1, 2, 5))
```


### Split by Gender

We then split our data by the value on `gender_life` and construct KM estimates
on that.

```{r calculate_km_estimates_gender}
#| echo: true

lapse2_km <- survfit2(
  Surv(policy_lifetime, lapsed) ~ gender_life1,
  data      = model_training_tbl,
  conf.type = 'log-log'
  )

lapse2_survdiff <- survdiff(
  Surv(policy_lifetime, lapsed) ~ gender_life1,
  data      = model_training_tbl
  )

print(lapse2_survdiff)

ggsurvfit(lapse2_km) +
  expand_limits(y = 0) +
  add_confidence_interval() +
  add_risktable()
```

We also want to estimate the 1, 2 and 5 years lapse.

```{r summary_lapse2}
#| echo: true

lapse2_km |>
  summary(times = 52 * c(1, 2, 5))
```



### Split by Smoker

Finally, we look at splitting by `smoker_life1`

```{r calculate_km_estimates_smoker}
#| echo: true

lapse3_km <- survfit(
  Surv(policy_lifetime, lapsed) ~ smoker_life1,
  data      = model_training_tbl,
  conf.type = 'log-log'
  )

lapse3_survdiff <- survdiff(
  Surv(policy_lifetime, lapsed) ~ smoker_life1,
  data      = model_training_tbl
  )

print(lapse3_survdiff)

ggsurvfit(lapse3_km) +
  expand_limits(y = 0) +
  add_confidence_interval() +
  add_risktable()
```


We want to estimate the 1, 2 and 5 years lapse.

```{r summary_lapse3}
#| echo: true

lapse3_km |>
  summary(times = 52 * c(1, 2, 5))
```


## Use Alternative Descriptive Statistics



```{r summarise_model_data_by_gender}
#| echo: true

model_training_tbl |>
  tbl_summary(
    by           = c(gender_life1),
    include      = c(policy_lifetime, smoker_life1, cluster_id),
    statistic    = list(
      all_continuous()  ~ "{mean} ({sd})",
      all_categorical() ~ "{n} / {N} ({p}%)"
      ),
    digits       = all_continuous() ~ 2,
    missing_text = "(Missing)"
    )
```



# Fit First Cox-PH Models for Lapses

We now want to fit our Cox Proportional Hazards models to this data.


## First Model

We start with a simple model using both `gender_life1` and `smoker_life1` as
input variables, and we look to add more as things progress.


```{r fit_coxph1_model}
#| echo: true

lapse1_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~ smoker_life1 + gender_life1,
  data = model_training_tbl,
  x    = TRUE,
  y    = TRUE
  )

lapse1_coxph |> summary()
```

We now want to calculate the concordance of the first Cox-PH model.

```{r glance_coxph1_model}
#| echo: true

lapse1_coxph |>
  glance() |>
  select(n, nevent, r.squared, concordance, std.error.concordance)
```


## Model Diagnostics for Cox-PH Model

### Test Proportional Hazards Assumption

The Cox-PH model assumes that hazards are proportional over time. We test this
assumption using the Schoenfeld residuals test. A significant p-value indicates
violation of the proportional hazards assumption.

```{r test_ph_assumption_coxph1}
#| echo: true

lapse1_zph <- cox.zph(lapse1_coxph)

print(lapse1_zph)
```

We can visualize the scaled Schoenfeld residuals over time to assess whether
the proportional hazards assumption holds for each covariate.

```{r plot_schoenfeld_residuals}
#| echo: true

lapse1_zph |> ggcoxzph(resid = FALSE)
```


### Visualize Hazard Ratios

We create a forest plot to visualize the hazard ratios and their confidence
intervals for the covariates in our model.

```{r plot_hazard_ratios}
#| echo: true

### Need to use as.data.frame() to prevent errors
ggforest(lapse1_coxph, data = as.data.frame(model_training_tbl))
```


### Residual Diagnostics

We examine various types of residuals to assess model fit and identify
influential observations.

#### Martingale Residuals

Martingale residuals help identify the functional form of covariates and detect
outliers. Values should be randomly distributed around zero.

```{r plot_martingale_residuals}
#| echo: true

ggcoxdiagnostics(
    lapse1_coxph,
    type               = "martingale",
    linear.predictions = TRUE
    ) +
  labs(
    title    = "Martingale Residuals",
    subtitle = "Should be randomly distributed around zero"
    )
```


#### Deviance Residuals

Deviance residuals are normalized transformations of martingale residuals and
should be symmetrically distributed around zero with no patterns.

```{r plot_deviance_residuals}
#| echo: true

ggcoxdiagnostics(
    lapse1_coxph,
    type               = "deviance",
    linear.predictions = TRUE
    ) +
  labs(
    title    = "Deviance Residuals",
    subtitle = "Should be symmetrically distributed around zero"
    )
```


#### Dfbeta Residuals

Dfbeta residuals measure the influence of each observation on the coefficient
estimates. Large values indicate influential observations.

```{r plot_dfbeta_residuals}
#| echo: true

ggcoxdiagnostics(
    lapse1_coxph,
    type               = "dfbeta",
    linear.predictions = FALSE
    ) +
  labs(
    title    = "Dfbeta Residuals",
    subtitle = "Measures influence of observations on coefficients"
    )
```


### Influential Observations

We identify potentially influential observations using dfbetas values.

```{r identify_influential_observations}
#| echo: true

lapse1_dfbetas <- residuals(lapse1_coxph, type = "dfbetas")

influential_threshold <- 2 / sqrt(nrow(model_training_tbl))

influential_obs <- which(
  apply(abs(lapse1_dfbetas), 1, max) > influential_threshold
  )

write_lines(glue(
  "Number of influential observations: {length(influential_obs)}
  Threshold: {round(influential_threshold, 5)}"
  ), stdout())
```


# Extended Cox-PH Model

We now fit an extended model including additional covariates to improve model
performance.

```{r fit_coxph2_model}
#| echo: true

lapse2_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~ smoker_life1 + gender_life1 + cluster_id + sum_assured,
  data = model_training_tbl,
  x    = TRUE,
  y    = TRUE
  )

lapse2_coxph |> summary()
```





## Diagnostics for Extended Model

Test proportional hazards assumption for the extended model.

```{r test_ph_assumption_lapse2}
#| echo: true

lapse2_zph <- cox.zph(lapse2_coxph)

print(lapse2_zph)
```

We can visualize the scaled Schoenfeld residuals over time to assess whether
the proportional hazards assumption holds for each covariate.

```{r plot_schoenfeld_residuals_lapse2}
#| echo: true

lapse2_zph |> ggcoxzph(resid = FALSE)
```



## Hazard Ratios for Extended Model

```{r plot_hazard_ratios_model2}
#| echo: true

ggforest(lapse2_coxph, data = as.data.frame(model_training_tbl))
```


## Predicted Survival Curves

We can generate predicted survival curves for different covariate profiles to
understand how the model predicts survival for different groups.

```{r create_prediction_profiles}
#| echo: true

# Create profiles for prediction
sum_assured_tbl <- model_training_tbl |>
  pull(sum_assured) |>
  quantile(c(0.10, 0.25, 0.50, 0.75, 0.90)) |>
  enframe(
    name  = NULL,
    value = "sum_assured"
    )

prediction_profiles_tbl <- model_training_tbl |>
  select(smoker_life1, gender_life1, cluster_id) |>
  distinct() |>
  complete() |>
  expand_grid(sum_assured_tbl) |>
  arrange(smoker_life1, gender_life1, cluster_id, sum_assured)

prediction_profiles_tbl |> glimpse()
```


```{r plot_predicted_survival}
#| echo: true

sample_profiles_tbl <- prediction_profiles_tbl |>
  slice_sample(n = 5)

lapse2_survfit <- survfit(
  lapse2_coxph,
  newdata = sample_profiles_tbl
  )

lapse2_survfit |> summary(times = 52 * c(1, 2, 5))

ggsurvplot(
  lapse2_survfit,
  data         = sample_profiles_tbl,
  conf.int     = TRUE,
  censor       = FALSE,
  legend.title = "Profile",
  xlab         = "Time (weeks)",
  ylab         = "Survival Probability",
  title        = "Predicted Survival Curves by Profile"
  )
```


## Cumulative Hazard Function

We examine the cumulative hazard function to understand the risk accumulation
over time.

```{r plot_cumulative_hazard}
#| echo: true

ggsurvplot(
  lapse2_survfit,
  data         = sample_profiles_tbl,
  fun          = "cumhaz",
  censor       = FALSE,
  conf.int     = TRUE,
  legend.title = "Profile",
  xlab         = "Time (weeks)",
  ylab         = "Cumulative Hazard",
  title        = "Cumulative Hazard by Profile"
  )
```


## Model Comparison

We compare the two models using likelihood ratio test and AIC.

```{r compare_models}
#| echo: true

# Likelihood ratio test
anova(lapse1_coxph, lapse2_coxph, test = "LRT")

# Compare AIC values
aic1 <- AIC(lapse1_coxph)
aic2 <- AIC(lapse2_coxph)
aic_diff <- aic1 - aic2

write_lines(glue(
  "\nAIC Comparison:
  Model 1 AIC: {aic1}
  Model 2 AIC: {aic2}
  AIC Difference: {aic_diff}"
  ), stdout())

# Compare concordance
model_comparison_tbl <- bind_rows(
    lapse1_coxph |> glance() |> mutate(model = "Model 1"),
    lapse2_coxph |> glance() |> mutate(model = "Model 2")
    ) |>
  select(model, n, nevent, concordance, std.error.concordance, AIC, BIC)

print(model_comparison_tbl)
```







