---
title: "Dublin Data Science Workshop on Survival Analysis (Part 2: Cox Regression & Model Validation)"
author: "Mick Cooney <mcooney@describedata.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8

---

**Note:** This worksheet builds on the exploratory analysis from Part 1
(initial_survival_models.qmd). Ensure you have completed that notebook before
running this one, or manually load the precomputed objects from the
`precompute/` directory.

```{r import_libraries}
#| echo: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(furrr)
library(glue)
library(fs)
library(survival)
library(survminer)
library(broom)
library(arrow)
library(qs2)

source("lib_utils.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "MASS",
    "fitdistrplus")
  )

options(
  width    = 80L,
  warn     = 1,
  mc.cores = parallelly::availableCores()
  )

Sys.setenv(OPENBLAS_NUM_THREADS = 8)

set.seed(42)

theme_set(theme_cowplot())

plan(multisession)
```

```{r setup_chunk_timing}
#| echo: false
#| message: false

# Enable chunk timing for all chunks
knitr::opts_chunk$set(time_it = TRUE)

# Create a list to store chunk times
chunk_times <- list()

# Create a hook to capture chunk execution times
knitr::knit_hooks$set(time_it = function(before, options) {
  if (before) {
    assign(".chunk_start_time", Sys.time(), envir = .GlobalEnv)
  } else {
    elapsed <- difftime(
        Sys.time(), get(".chunk_start_time", envir = .GlobalEnv),
        units = "secs"
        ) |>
      as.numeric()
    
    chunk_times <- get("chunk_times", envir = .GlobalEnv)
    chunk_times[[options$label]] <- elapsed
    assign("chunk_times", chunk_times, envir = .GlobalEnv)
    
    message(glue("⏱️  Chunk '{options$label}': {round(elapsed, 2)} seconds"))
  }
})
```

## Load Data from Part 1

This worksheet builds directly on the exploratory analysis completed in
initial_survival_models.qmd. We load the processed data and initial models:

```{r load_part1_data}
#| echo: true

# Load discretized telco data (includes calls_cat)
telco_churn_cat_tbl <- read_parquet("precompute/telco_churn_cat.parquet")

# Load pre-fitted models
telco_model00_coxph <- qs_read("precompute/model00_null.qs")
telco_model01_coxph <- qs_read("precompute/model01_vmailplan.qs")
telco_model03_coxph <- qs_read("precompute/model03_combined.qs")

# For clarity: reload raw data to ensure all variables are available
telco_cols <- cols(areacode = col_character())
telco_churn_tbl <- read_csv("data/telcochurn.csv", col_types = telco_cols)

write_lines(
  glue(
    "Data loaded: {nrow(telco_churn_cat_tbl)} observations, ",
    "{ncol(telco_churn_cat_tbl)} variables"
    ),
  stdout()
  )
```




# The Cox Proportional Hazards Model

Up to this point we have focused on empirical and non-parametric approaches
to estimating survival and hazard functions. The Kaplan-Meier estimator
provides survival curves for different groups, and the log-rank test lets us
compare those curves. These methods are powerful for exploratory analysis and
simple group comparisons, but they have fundamental limitations when dealing
with real-world complexity.

## Understanding Cox Regression: Motivation and Theory

The group comparison approach encounters several practical challenges. Real-
world outcomes rarely depend on a single factor—customer churn might be
influenced by voicemail plan, international plan, customer service calls,
account age, and usage patterns all at once. We could split the data by all
these variables, but this creates an explosion of subgroups. With just 5
binary variables, we'd have 32 groups; many would be too small for reliable
estimation. Continuous predictors pose additional challenges, requiring
arbitrary discretization that loses information. Perhaps most importantly,
group-based approaches cannot disentangle confounding—differences between
groups on one variable might actually be driven by correlated factors.

The **Cox proportional hazards model** addresses these needs while respecting
the unique structure of survival data—censoring, time-to-event outcomes, and
non-constant hazard rates. It provides individual-level risk scores, handles
multiple predictors simultaneously, quantifies independent effects while
adjusting for confounding, and accommodates both continuous and categorical
variables naturally. It has become the standard regression approach in
survival analysis because it balances flexibility, interpretability, and
computational tractability.

### Why Specialized Methods for Survival Data?

Before diving into Cox models specifically, it's worth understanding why
survival data requires specialized regression rather than standard linear or
logistic regression. Linear regression assumes normally distributed outcomes
that can take any value; survival times are strictly positive and often
heavily right-skewed. More critically, **censored observations** don't fit
the framework—if a customer is still active at day 180, we know their true
survival time is "at least 180 days," not "exactly 180 days." Linear
regression has no mechanism to handle this partial information.

We could convert survival to binary (churned/not churned by day 180) and use
logistic regression, but this throws away timing information—a customer
churning on day 10 versus day 175 are treated identically. It also requires
choosing an arbitrary time cutoff and doesn't naturally handle cases censored
before that cutoff.

Survival regression models work directly with **hazard functions**—the
instantaneous risk of an event at each time point, conditional on survival
to that point. This respects the time-to-event structure and naturally
incorporates censored observations through likelihood functions that account
for partial information. The Cox model is the most widely used approach
because it makes minimal assumptions about the baseline hazard function (it's
"semi-parametric"), yet still provides interpretable coefficients and robust
predictions.

### The Proportional Hazards Assumption

The Cox model is built on a specific structural assumption about how
covariates affect the hazard function. Understanding this assumption is
critical for proper model use and interpretation.

**The Baseline Hazard**

Imagine a reference individual with all covariates set to zero (or reference
levels for categorical variables). This person's hazard function over time is
$h_0(t)$, called the **baseline hazard**. It represents the time-varying risk
pattern for this reference case.

Importantly, the Cox model does not require us to specify the shape of $h_0(t)$
—it could be constant (exponential distribution), increasing (Weibull), have
a bathtub shape, or any other pattern. This flexibility is a major advantage:
we don't need to assume a particular distribution for survival times.

**Proportionality: Constant Hazard Ratios**

The "proportional hazards" assumption states that every individual's hazard
function is a **constant multiple** of the baseline hazard:

$$
h(t | X) = \psi \, h_0(t)
$$

where $\psi$ is a positive constant that depends on the individual's covariate
values $X$, but **not on time**. This means:

  - If person A has twice the hazard of person B at time 10, they also have
    twice the hazard at time 100, time 500, etc.
  - The **hazard ratio** between any two individuals remains constant over time
  - All individuals experience the same time-varying pattern ($h_0(t)$), just
    scaled up or down by $\psi$

Graphically, this means survival curves for different groups should not cross
substantially—one group should consistently have better or worse survival
throughout the observation period.

**What Violates Proportional Hazards?**

The assumption fails when the effect of a covariate changes over time:

  - **Early vs. late effects**: A treatment might reduce hazard initially but
    lose effectiveness over time (curves eventually cross)
  - **Delayed effects**: A risk factor might not matter early but become
    important later
  - **Time-varying covariates**: If a covariate itself changes over time (e.g.,
    customer upgrades their plan), simple proportional hazards doesn't apply
    directly

Violations don't make the Cox model useless, but they require extensions like
time-varying coefficients or stratification (topics beyond this workshop).

We'll cover two sets of diagnostic tools later in this workshop:

  1. **Schoenfeld residuals** (Section "Assessing the Proportional Hazards
     Assumption"): Formal tests for whether coefficients vary over time
  1. **Residual-based variable exploration** (Section "Validating the
     Proportional Hazards Model"): Using martingale and deviance residuals to
     identify promising predictors and assess model fit

**Modeling the Hazard Multiplier**

The constant $\psi$ is modeled as an exponential linear function of covariates:

$$
\psi = \exp(\beta_1 x_1 + \beta_2 x_2 + \ldots + \beta_p x_p) = \exp(\beta^T X)
$$

Why the exponential? Several reasons:

  1. **Ensures positivity**: Hazards must be non-negative, and $\exp(\cdot)$ is
     always positive regardless of the linear predictor's value
  1. **Multiplicative effects**: $\exp(a + b) = \exp(a) \cdot \exp(b)$, so
     effects of different variables multiply rather than add
  1. **Interpretable coefficients**: $\exp(\beta_j)$ is the hazard ratio for a
     one-unit increase in $x_j$, holding other variables constant

Putting it together, the full Cox proportional hazards model is:

$$
h(t | X) = h_0(t) \exp(\beta^T X)
$$

The baseline hazard $h_0(t)$ captures the time-varying pattern, while
$\exp(\beta^T X)$ captures how individual characteristics scale that pattern
up or down.

### Interpreting Coefficients: Hazard Ratios

Cox model coefficients are interpreted through **hazard ratios** (HRs), which
are simply $\exp(\beta_j)$ for each coefficient $\beta_j$.

**For Binary Variables**

Suppose `intlplan` (international calling plan: yes/no) has a coefficient
$\beta = 0.69$. The hazard ratio is:

$$
HR = \exp(0.69) \approx 2.0
$$

**Interpretation**: Customers with an international plan have approximately
**twice the hazard** (twice the instantaneous churn risk) at any given time,
compared to customers without an international plan, holding all other
variables constant.

Equivalently, their risk is 100% higher. Generally:

  - $HR = 1.0$: No effect (hazard unchanged)
  - $HR > 1.0$: Increased hazard (worse survival)
  - $HR < 1.0$: Decreased hazard (better survival)
  - $HR = 2.0$: Hazard doubled (100% increase)
  - $HR = 0.5$: Hazard halved (50% reduction)

**For Continuous Variables**

Suppose `custservicecalls` has a coefficient $\beta = 0.15$. The hazard ratio
for a one-unit increase (one additional call) is:

$$
HR = \exp(0.15) \approx 1.16
$$

**Interpretation**: Each additional customer service call is associated with
a 16% increase in churn hazard, holding other variables constant.

For a 5-call increase, the hazard ratio is:

$$
HR = \exp(5 \times 0.15) = \exp(0.75) \approx 2.12
$$

So five additional calls more than double the hazard.

**Negative Coefficients**

If a coefficient is negative, $\exp(\beta) < 1$, indicating a protective
effect. For example, $\beta = -0.50$ gives $HR = \exp(-0.50) \approx 0.61$,
meaning a 39% reduction in hazard (better survival).

**The "Holding Other Variables Constant" Caveat**

Hazard ratios represent **independent associations** after adjusting for other
variables in the model. This is crucial: the effect of voicemail plan in a
model with multiple predictors might differ from its effect when modeled alone,
because other variables may confound or mediate the relationship.

### Estimation: Partial Likelihood

How do we estimate the coefficients $\beta$ in a Cox model? Standard maximum
likelihood doesn't work directly because we have censored observations and an
unspecified baseline hazard $h_0(t)$. The elegant solution is **partial
likelihood**, developed by Cox (1972).

**The Conceptual Idea**

Rather than modeling the probability of events happening at specific times
(which would require specifying $h_0(t)$), partial likelihood focuses on the
**relative risk** at each event time.

When an event occurs at time $t_i$, we ask: "Given that exactly one event
occurred at this time and these people were at risk, how likely is it that
this particular person experienced the event (versus someone else)?"

The probability that individual $j$ has the event, given the risk set at
time $t_i$, is:

$$
P(\text{individual } j \text{ fails at } t_i \mid \text{one failure at } t_i) = 
\frac{\exp(\beta^T X_j)}{\sum_{k \in R(t_i)} \exp(\beta^T X_k)}
$$

where $R(t_i)$ is the **risk set**—all individuals still under observation
(not yet failed or censored) just before time $t_i$.

Notice that $h_0(t_i)$ cancels out in this ratio—it appears in both numerator
and denominator, so we don't need to estimate it to calculate this probability.
This is the key insight that makes the Cox model work without distributional
assumptions.

**Constructing the Partial Likelihood**

We multiply these conditional probabilities across all observed event times:

$$
PL(\beta) = \prod_{i=1}^{D} \frac{\exp(\beta^T X_i)}{\sum_{k \in R(t_i)} \exp(\beta^T X_k)}
$$

where $D$ is the number of observed events. Maximizing this partial likelihood
(or equivalently, its log) yields estimates of $\beta$.

**Censored observations** contribute to the risk sets at event times before
their censoring time, but don't contribute an event term. Their information
isn't lost—they help define who was at risk when events occurred.

**After Estimating Coefficients**

Once we have $\hat{\beta}$, we can estimate the baseline hazard $h_0(t)$ using
the **Breslow estimator**, though this is often not needed for practical
applications. Most analyses focus on hazard ratios (which depend only on
$\beta$) or predicted survival probabilities (which can be calculated without
explicitly estimating $h_0(t)$).

**Computational Note**

Maximizing the partial likelihood is a standard numerical optimization problem,
solved by iterative algorithms like Newton-Raphson. The R `survival` package
implements this efficiently, handling tens of thousands of observations and
hundreds of covariates routinely.

### Advantages of the Cox Model

Before fitting models, it's worth summarizing why the Cox proportional hazards
model has become the workhorse of survival analysis:

  1. **Semi-parametric flexibility**: No distributional assumptions about
     survival times. The baseline hazard can have any shape.
  1. **Handles censoring naturally**: Partial likelihood incorporates censored
     observations without bias or information loss.
  1. **Interpretable coefficients**: Hazard ratios have clear, practical
     interpretations for stakeholders.
  1. **Multiple predictors**: Easily accommodates many covariates—continuous,
     categorical, interactions—in a single model.
  1. **Adjusts for confounding**: Estimates independent effects of each variable
     while controlling for others.
  1. **Robust and well-studied**: Decades of theoretical development, extensive
     software implementations, well-understood properties.
  1. **Extensions available**: Can be extended for time-varying coefficients,
     time-varying covariates, competing risks, frailty (random effects), and
     more.

The main limitation is the proportional hazards assumption, which needs
checking (we'll cover diagnostics later). When violated, alternative models or
extensions may be needed.

With this foundation, we're ready to fit Cox models to our data and interpret
the results.


# Continue from Initial Models

## Our First Cox-PH Model

Having developed the theoretical foundation, we now fit our first Cox
proportional hazards model to examine whether voicemail plan status affects
customer churn.

The `coxph()` function uses the same formula syntax as other R regression
functions: the left side is a `Surv()` object (time and event indicator),
and the right side specifies predictors. For a single binary predictor like
`vmailplan`, this gives us a simple baseline model.

```{r telco_model01_coxph_summary}
#| echo: true

telco_model01_coxph |> summary()
```

The `summary()` output provides several key pieces of information:

  - **Coefficient (coef)**: The estimated $\beta$ value from the model
  - **Hazard ratio (exp(coef))**: $\exp(\beta)$, the multiplicative effect on
    the hazard
  - **Standard error and z-score**: Quantify uncertainty in the coefficient
    estimate
  - **Concordance**: The c-index we'll discuss in the next section

For `vmailplan`, the negative coefficient means having a voicemail plan is
associated with *lower* hazard (better survival). The hazard ratio of
approximately 0.65 indicates that customers with voicemail plans have about
65% of the churn hazard compared to those without—equivalently, a 35%
reduction in churn risk.

This model was estimated using the partial likelihood framework described
earlier, comparing observed events to the risk set at each event time without
requiring specification of the baseline hazard shape.

## Model Assessment

Assessing Cox model fit differs from standard regression because we're modeling
a distribution (survival probabilities over time) rather than a single numeric
outcome. We cannot simply calculate residual sums of squares or standard $R^2$.
Instead, survival analysis uses adapted metrics that account for censoring and
time-to-event structure.

Two key measures help quantify how well a Cox model explains the data:
**pseudo-$R^2$** (variance explained) and the **concordance index**
(predictive discrimination). Both provide complementary views of model quality.

### R-Squared

The standard measure of model fit is $R^2$, the proportion of variance
explained by the model. This concept doesn't directly apply to Cox models
(which estimate hazard functions, not outcomes), but we can construct an
analogous measure based on likelihood ratios.

For Cox proportional hazards models, we use a **pseudo-$R^2$** defined as:

$$
R^2 = 1 - \left( \frac{l(0)}{l(\hat{\beta})} \right)^{\frac{2}{n}}
$$

where $l(\beta)$ is the log-partial-likelihood function, $l(0)$ is the
log-partial-likelihood of the null model (no predictors), and $n$ is the
sample size.

This measures improvement in model fit compared to the null model. Values
range from 0 (no improvement over null) to 1 (perfect fit), though in
practice survival models rarely achieve high $R^2$ values. A value of 0.10-0.20
might represent a reasonably useful model, as censoring and unmeasured factors
inherently limit explained variance.


### Concordance Index

The concordance index (c-index or c-statistic) measures **predictive
discrimination**: how well the model ranks individuals by risk. It answers
the question: "If we pick two random individuals, does our model correctly
predict which one will experience the event first?"

Formally, the c-index is the proportion of **concordant pairs** among all
evaluable pairs. A pair is evaluable if at least one individual experiences
the event (not both censored). The pair is concordant if:

  - The person predicted to have higher risk (higher linear predictor $\beta^T X$)
    actually experiences the event first, *or*
  - The person predicted to have lower risk is censored before the higher-risk
    person's event time

The c-index ranges from 0 to 1:

  - **0.5**: Random guessing (coin flip)—model has no discriminative ability
  - **0.6-0.7**: Reasonable discrimination—typical for good survival models
  - **0.7-0.8**: Strong discrimination—excellent model performance
  - **>0.8**: Exceptional (rare in practice due to censoring and noise)

Unlike classification accuracy, the c-index naturally handles censored
observations by excluding pairs where censoring prevents meaningful comparison.
It's analogous to the area under the ROC curve (AUC) in classification
problems.


### Assessing Our First Model

Let's examine our voicemail plan model using these metrics. We use `glance()`
to extract summary statistics and `tidy()` to view coefficient details:

```{r telco_model01_coxph_assessment}
#| echo: true

telco_model01_coxph |>
  glance() |>
  select(n, nevent, r.squared, concordance, std.error.concordance)

telco_model01_coxph |> tidy()
```

This simple model with one binary variable gives us a **c-index of 0.565**—
slightly better than random guessing (0.5), but only modestly so. The model
can distinguish high-risk from low-risk customers, but not strongly. This
makes sense: voicemail plan status alone cannot explain the full complexity
of customer churn behavior.

The **$R^2$ is essentially zero (0.011)**, indicating that voicemail plan
explains about 1% of the variance in churn hazard. Again, this is expected
for a single predictor in a complex process.

These low values aren't necessarily problematic—they establish a baseline.
As we add more predictive variables, we'll track how both metrics improve.
The question becomes: can we build a model with c-index approaching 0.70 and
$R^2$ above 0.10 by including additional risk factors?


## Adding International Plans

Before building more complex models, it's good practice to examine each
potential predictor individually. This helps us understand each variable's
independent relationship with churn and establishes individual baselines
before testing combined effects.

The combined model tests whether international plan status and voicemail
provide independent information about churn risk:

```{r telco_model03_coxph_summary}
#| echo: true

telco_model03_coxph |> summary()
```

The combined model shows both coefficients remain statistically meaningful,
suggesting each variable contributes independently.

Assessing the combined model:

```{r telco_model03_coxph_assessment}
#| echo: true

telco_model03_coxph |>
  glance() |>
  select(n, nevent, r.squared, concordance, std.error.concordance)

telco_model03_coxph |> tidy()
```

Model03 achieves a **c-index of 0.643**—a substantial jump from 0.565
(Model01). The combined model now discriminates between high-risk and
low-risk customers reasonably well, approaching the 0.65-0.70 range typical
of good survival models.

The **$R^2$ has improved to approximately 0.05** (5% of variance explained).
While this might seem low compared to linear regression standards, it's
reasonable for survival data where censoring and unmeasured factors inherently
limit explained variance.

The improvement from combining predictors confirms that voicemail plan and
international plan provide complementary information. This justifies including
both in the model and suggests that adding additional relevant predictors
might further improve performance.

Before we spend time iterating further, we check the proportional hazards
assumption to ensure our models are valid.


# Diagnostic Methods for Cox Models

Before building more complex models, we need a systematic approach to model
diagnostics and variable selection. This section introduces two fundamental
diagnostic tools for Cox proportional hazards models: **residual analysis**
for identifying promising predictors and detecting model inadequacies, and
**Schoenfeld residuals** for testing the proportional hazards assumption.

These diagnostic methods serve complementary purposes. Residual analysis helps
us understand which variables to include and how they should enter the model
(linear, non-linear, interactions). Proportional hazards diagnostics validate
the fundamental assumption underlying our coefficient interpretations—that
hazard ratios remain constant over time.

Together, these tools form the foundation for iterative model development:
fit a candidate model, check diagnostics, identify improvements, refit, and
repeat until we achieve adequate fit without assumption violations.


## Understanding Residuals in Survival Analysis

Before continuing to build more complex models, we need tools to answer two
key questions:

  1. **Model adequacy**: Does our model fit the data reasonably well? Are there
     systematic patterns in the residuals suggesting problems?
  1. **Variable selection**: Which additional variables might improve the model?
     Can we identify promising predictors before fitting dozens of models?

In linear regression, residual plots (observed minus predicted) help answer
both questions. For Cox models, residuals are more complex because we're
modeling hazards over time with censored observations, not simple numeric
outcomes. However, the conceptual approach remains the same: examine
differences between observed and expected values.

We'll use two types of residuals:

  - **Martingale residuals**: Measure differences between observed events and
    expected cumulative hazard. Useful for exploring functional forms and
    identifying influential observations.
  - **Deviance residuals**: Symmetric transformation of martingale residuals,
    better behaved for model diagnostics and goodness-of-fit assessment.

A particularly valuable application is using residuals from a **null model**
(no predictors) to explore which variables show relationships with the
outcome. Variables that show clear patterns when plotted against null-model
residuals are promising candidates for inclusion.

The intuition is simple: if a variable is associated with the outcome,
plotting it against null-model residuals should reveal a trend. The null
model doesn't account for any predictors, so all variation remains in the
residuals. A strong predictor will show up as a clear pattern (linear,
non-linear, or otherwise) when plotted against these residuals.

This approach is particularly powerful because it lets us screen many
candidate variables quickly without fitting dozens of models. We can visually
inspect residual plots and quantify relationships via correlation, focusing
our modeling efforts on the most promising predictors.

### Martingale Residuals from Null Model

Martingale residuals extend the concept of "observed minus expected" to the
survival context. For each individual $i$, the martingale residual is:

$$
m_i = \delta_i - \hat{H}_0(t_i) \exp(x_i \hat{\beta})
$$

where:

  - $\delta_i$ is the event indicator (1 if event occurred, 0 if censored)
  - $\hat{H}_0(t_i)$ is the estimated cumulative baseline hazard at time $t_i$
  - $\exp(x_i \hat{\beta})$ is the relative risk multiplier from covariates

The residual measures the difference between the **observed event status**
(did the event happen?) and the **expected cumulative hazard** (how much
risk has accumulated?). For the null model, this simplifies since
$\exp(x_i \hat{\beta}) = 1$ (no covariates).

**Interpretation:**

  - Positive residuals: Individual experienced event sooner than expected
  - Negative residuals: Individual survived longer than expected
  - Residuals sum to zero on average

```{r calculate_null_model_residuals}
#| echo: true

telco_model00_resid_m <- telco_model00_coxph |> residuals(type = "martingale")

telco_model00_resid_m |> summary()
```

### Residual-Based Variable Selection

By plotting residuals against potential predictors, we identify variables that
show relationships with the outcome. Variables with clear visual patterns are
strong candidates for inclusion in the model.

```{r systematic_residual_analysis_setup}
#| echo: true

# Add residuals to dataset
telco_null_assess_tbl <- telco_churn_cat_tbl |>
  mutate(resid_m = telco_model00_resid_m)

# Define continuous variables for systematic exploration
CONTINUOUS_VARS <- c(
  "daymins", "evemins", "nightmins", "intlmins",
  "daycalls", "evecalls", "nightcalls", "intlcalls",
  "vmailmessage", "custservicecalls"
  )
```

Function to generate residual plots:

```{r define_residual_plot_function}
#| echo: true

create_residual_plot <- function(data_tbl, var_name) {
  ggplot(data_tbl, aes(x = .data[[var_name]], y = resid_m)) +
    geom_point(size = 0.8, alpha = 0.15) +
    geom_smooth(
      method = "loess",
      se     = FALSE,
      colour = "firebrick",
      linewidth = 1.2
      ) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(
      x     = var_name,
      y     = "Martingale Residual",
      title = glue("Residuals vs {var_name}")
      ) +
    theme_minimal()
}
```

Generate systematic residual plots:

```{r generate_systematic_residual_plots}
#| echo: true
#| fig-width: 10
#| fig-height: 12

residual_plots <- map(
  CONTINUOUS_VARS,
  \(var) create_residual_plot(telco_null_assess_tbl, var)
  )

# Arrange plots in a grid
plot_grid(
  plotlist = residual_plots,
  ncol     = 2,
  labels   = "AUTO"
  )
```

### Quantifying Residual Relationships

Compute correlations between residuals and each variable:

```{r quantify_residual_correlations}
#| echo: true

residual_correlations_tbl <- tibble(
  variable    = CONTINUOUS_VARS,
  correlation = map_dbl(
    CONTINUOUS_VARS,
    \(var) cor(
      telco_null_assess_tbl$resid_m,
      telco_null_assess_tbl[[var]],
      use = "complete.obs"
      )
    )
  ) |>
  arrange(desc(abs(correlation)))

residual_correlations_tbl |> print()
```

**Key Finding**: `custservicecalls` shows the strongest relationship with
residuals (correlation ≈ 0.16-0.18). While this might seem modest, remember
that:

  - Survival data has inherent noise from censoring and unmeasured factors
  - We're looking at univariate relationships—the effect may be stronger when
    combined with other predictors
  - Even modest correlations can translate to meaningful improvements in model
    concordance and predictive power

The positive correlation indicates that higher customer service calls are
associated with higher martingale residuals, meaning these customers
experienced events sooner than expected under the null model. This confirms
the relationship we observed visually: more service calls predict earlier
churn.

Other variables show weaker or negligible correlations, suggesting they may
have limited predictive value (at least as linear main effects). This doesn't
mean they're useless—interactions or non-linear transformations might reveal
relationships—but `custservicecalls` is clearly the strongest candidate for
our next model iteration.

This variable will be our focus for extending the model beyond the simple
categorical predictors (voicemail and international plans) we've used so far.


## Testing Proportional Hazards: Schoenfeld Residuals

The proportional hazards assumption is fundamental to Cox regression: it
states that hazard ratios remain constant over time. If this assumption is
violated, our coefficient interpretations become unreliable—a hazard ratio
of 2.0 might be accurate early in the observation period but not later (or
vice versa).

**Schoenfeld residuals** provide a formal test of this assumption. These
residuals are computed for each covariate at each event time and should be
independent of time if the proportional hazards assumption holds. The
`cox.zph()` function performs a statistical test for each covariate,
testing whether the Schoenfeld residuals correlate with time.

**Interpreting the output:**

  - **p-value > 0.05**: No evidence against proportional hazards (assumption
    appears satisfied)
  - **p-value < 0.05**: Evidence of time-varying effects (assumption may be
    violated)
  - **GLOBAL test**: Overall test across all covariates simultaneously

The diagnostic plots show Schoenfeld residuals versus transformed time,
with a smoothed trend line. Flat trends (horizontal lines) suggest
proportional hazards; clear upward or downward trends indicate violations.

Let's check our initial models:

Test Model 01:

```{r testing_model01_ph_assumptions}
#| echo: true

telco_model01_coxzph <- telco_model01_coxph |> cox.zph(transform = "km")

telco_model01_coxzph |> print()
telco_model01_coxzph |> ggcoxzph()
```

The p-value for `vmailplan` should be well above 0.05, indicating no evidence
of time-varying effects. The diagnostic plot should show a relatively flat
trend with scattered residuals, consistent with proportional hazards.

### Testing Model 03 (Combined Predictors)

Test Model 03 (combined):

```{r testing_model03_ph_assumptions}
#| echo: true

telco_model03_coxzph <- telco_model03_coxph |> cox.zph(transform = "km")

telco_model03_coxzph |> print()
telco_model03_coxzph |> ggcoxzph()
```

Both predictors should pass the proportional hazards test, and the GLOBAL
test should show no overall evidence of violations. This validates that our
coefficient interpretations (e.g., "customers with international plans have
twice the hazard throughout the observation period") are appropriate.

The diagnostics suggest the proportional hazards assumption is reasonably
satisfied for both models. We can proceed with confidence that our hazard
ratio interpretations are valid.


# Building Models with Continuous Predictors

Having established a baseline with categorical predictors and identified
`custservicecalls` as a promising continuous variable through residual
analysis, we now extend our modeling to incorporate this variable. This
section explores three different approaches to including continuous predictors
in Cox models, each with distinct advantages and interpretive implications.

The key modeling question is: **how does the effect of customer service calls
on churn hazard behave?** Is it linear (each additional call has the same
proportional effect)? Is it stepped or threshold-based (low calls don't matter,
but high calls trigger churn)? Or is it smoothly non-linear (the effect
accelerates or decelerates)?

We'll test three specifications:

  1. **Linear**: Assumes constant proportional effect per unit increase
  1. **Discretised**: Groups calls into bins, allowing step-function effects
  1. **Splines**: Flexible smooth curves estimated from the data

Each approach has trade-offs. Linear models are parsimonious and
interpretable but may miss non-linear patterns. Discretisation is intuitive
and matches business logic (e.g., "customers with 6+ calls behave
differently") but requires choosing cutpoints and loses information within
bins. Splines are maximally flexible but harder to interpret and can overfit.

By comparing all three, we can understand the shape of the relationship and
choose the specification that best balances fit, interpretability, and
parsimony.


## Adding Continuous Variables to the Model

Having identified `custservicecalls` as a promising predictor through residual
analysis, we now add it to our Cox model. We start with the simplest
specification: a linear term.

### Model 04: Linear Effect of Customer Service Calls

We extend Model 03 (which included `vmailplan` and `intlplan`) by adding
`custservicecalls` as a continuous predictor with a linear effect:

```{r fit_model04_continuous}
#| echo: true

telco_model04_coxph <- coxph(
  Surv(accountdur, churned) ~ vmailplan + intlplan + custservicecalls,
  data = telco_churn_tbl
  )

telco_model04_coxph |> summary()
```

**Interpreting the output:**

The coefficient for `custservicecalls` represents the log-hazard ratio for a
one-unit increase (one additional call). Exponentiating this coefficient gives
the hazard ratio. For example, if the coefficient is 0.20, the hazard ratio
is exp(0.20) ≈ 1.22, meaning each additional customer service call increases
the churn hazard by approximately 22%.

The statistical significance (p-value) tells us whether this effect is
distinguishable from zero. Given the strong correlation we observed in
residual analysis, we expect a highly significant result.

Notice also the overall model statistics:

  - **Likelihood ratio test**: Compares this model to the null model (no
    predictors), testing overall model significance
  - **Wald test** and **Score test**: Alternative tests of overall significance,
    based on different statistical principles
  - **Concordance**: How well the model discriminates between patients (should
    improve substantially from Model 03)

### Assessing Model 04 Performance

Compare Model 04 to our previous models:

```{r assess_model04}
#| echo: true

telco_model04_coxph |>
  glance() |>
  select(n, nevent, r.squared, concordance, std.error.concordance)

telco_model04_coxph |> tidy()
```

**What to look for:**

  - **Concordance increase**: Model 03 achieved c ≈ 0.64. Model 04 should show
    a meaningful jump, likely to 0.68-0.72, indicating substantially better
    discrimination between high-risk and low-risk customers.
  - **R² increase**: While still modest by linear regression standards,
    adding a strong predictor should noticeably increase the proportion of
    variance explained.
  - **Coefficient significance**: All three predictors should remain
    statistically significant, confirming they provide independent information
    about churn risk.

The improvement from Model 03 to Model 04 demonstrates the value of
continuous predictors. While categorical variables like voicemail plan provide
useful baseline discrimination, quantitative measures like customer service
call frequency capture gradations of risk that binary indicators miss.

This is a general pattern in survival modeling: combining categorical risk
factors (which split the population into distinct subgroups) with continuous
measures (which provide fine-grained risk stratification within subgroups)
typically yields the best predictive performance.

### Checking Proportional Hazards for Model 04

```{r test_model04_ph_assumption}
#| echo: true

telco_model04_coxzph <- telco_model04_coxph |> cox.zph(transform = "km")

telco_model04_coxzph |> print()
telco_model04_coxzph |> ggcoxzph()
```

All three predictors should pass the test (p > 0.05), and the GLOBAL test
should show no evidence of violations. The proportional hazards assumption
appears satisfied for Model 04, validating our use of time-constant hazard
ratios for interpretation.


# Exploring Non-Linear Effects

The linear Model 04 assumes each additional customer service call has the same
proportional effect on hazard—the hazard ratio for going from 0 to 1 call is
the same as going from 5 to 6 calls. But is this realistic?

Behavioral intuition suggests non-linearity: a customer with 0-1 calls is
probably satisfied and unlikely to churn. A customer with 2-3 calls might be
frustrated but salvageable. A customer with 6+ calls is likely deeply
dissatisfied and at high risk. The incremental effect of each additional call
might accelerate rather than remain constant.

This section explores two approaches to modeling non-linear effects:
**discretisation** (grouping into bins) and **splines** (smooth flexible
curves). Both allow the hazard ratio to vary across the range of customer
service calls, but they differ in how they represent that variation.

Discretisation is intuitive, aligns with business thinking (e.g., "high-call
customers behave differently"), and provides easily interpretable hazard
ratios for each group. The downside is that it requires choosing cutpoints
(somewhat arbitrary) and treats customers within a bin as identical.

Splines are data-driven and maximally flexible, estimating a smooth curve
without imposing functional form. The downside is reduced interpretability—
instead of simple hazard ratios, we get a curve that must be visualized.

Both approaches can reveal whether the linear assumption is adequate or
whether the relationship is genuinely non-linear.


## Non-Linear Effects via Discretisation

The linear model assumes each additional customer service call has the same
proportional effect on hazard. We can test this by modeling non-linear effects
through discretisation, using the domain-driven bins created in Part 1.

### Model 05: Discretised Customer Service Calls

Instead of treating customer service calls as a continuous linear predictor,
we use the categorical variable `calls_cat` created in Part 1 with
domain-driven cutpoints: (0-1), (2), (3-5), (6+). This allows each group to
have its own hazard ratio relative to the reference group (0-1 calls).

```{r fit_model05_discretised}
#| echo: true

telco_model05_coxph <- coxph(
  Surv(accountdur, churned) ~ vmailplan + intlplan + calls_cat,
  data = telco_churn_cat_tbl
  )

telco_model05_coxph |> summary()
```

**Interpreting the output:**

The model produces three coefficients for `calls_cat` (since there are four
groups, one is the reference). Each coefficient represents the log-hazard
ratio for that group relative to the reference group (0-1 calls).

For example:
  - `calls_cat[2]`: hazard ratio for 2 calls vs 0-1 calls
  - `calls_cat[3-5]`: hazard ratio for 3-5 calls vs 0-1 calls
  - `calls_cat[6+]`: hazard ratio for 6+ calls vs 0-1 calls

If the coefficients increase monotonically (each group has a higher hazard
ratio than the previous), this confirms our intuition that more calls =
higher risk. If the hazard ratios are roughly equal to the linear model's
predictions (e.g., if 2 calls has HR ≈ linear_HR², 3-5 calls has HR ≈
linear_HR³⁻⁵), then the linear model was adequate.

But if we see non-linear patterns—for example, if 2 calls has nearly the
same risk as 0-1 calls, but 6+ calls has dramatically higher risk—then
discretisation captures important structure the linear model missed.

### Comparing Linear vs. Discretised Models

```{r compare_model04_model05}
#| echo: true

# Compare models
bind_rows(
  telco_model04_coxph |> glance() |> mutate(model = "Model04 (linear)"),
  telco_model05_coxph |> glance() |> mutate(model = "Model05 (discretised)")
  ) |>
  select(model, concordance, r.squared, AIC, BIC)
```

**Interpreting the comparison:**

  - **Concordance**: If Model 05 has noticeably higher concordance than Model 04,
    it suggests the discretisation captures non-linear patterns that improve
    risk discrimination.
  - **AIC/BIC**: Lower is better. AIC and BIC penalize model complexity
    (number of parameters), so even if Model 05 fits better, if it uses more
    parameters, the penalty-adjusted metrics might favor Model 04.
  - **Degrees of freedom**: Model 04 uses 3 parameters (vmail, intl, calls);
    Model 05 uses 5 parameters (vmail, intl, calls_cat[2], calls_cat[3-5],
    calls_cat[6+]). The question is whether the extra 2 parameters justify
    their complexity.

If AIC strongly favors Model 05 despite the parameter penalty, this is
compelling evidence for non-linearity. If Model 04 has similar or better
AIC/BIC, the linear model is adequate (simpler is better).

### Checking Proportional Hazards for Model 05

```{r test_model05_ph_assumption}
#| echo: true

telco_model05_coxzph <- telco_model05_coxph |> cox.zph(transform = "km")

telco_model05_coxzph |> print()
```

Each level of `calls_cat` should pass the test individually, and the GLOBAL
test should show no violations. If all tests pass, the discretised
specification satisfies proportional hazards just as the linear version did.


## Non-Linear Effects via Splines

Penalized splines (P-splines) provide a flexible, data-driven approach to
modeling smooth non-linear relationships without requiring arbitrary cutpoints.
Unlike discretisation, which creates step functions, splines fit smooth curves
that can capture gradual acceleration or deceleration in effects.

The `pspline()` function in the survival package uses penalized regression
splines, which balance fit (flexible curves that follow the data) against
smoothness (penalizing overly wiggly functions). The penalty is chosen
automatically via cross-validation, making splines largely automated and
objective.

Splines are particularly useful when we suspect non-linearity but don't have
strong domain knowledge about where the cutpoints should be or whether the
relationship has sudden jumps versus smooth transitions.

The main downside is interpretation: instead of a single hazard ratio (linear
model) or a set of group-specific hazard ratios (discretised model), we get
a curve that must be visualized to understand. This is acceptable for
exploratory analysis and prediction but can be harder to communicate to
stakeholders who want simple "risk increases by X% per call" statements.

### Model 06: Spline for Customer Service Calls

```{r fit_model06_spline}
#| echo: true

telco_model06_coxph <- coxph(
  Surv(accountdur, churned) ~ vmailplan + intlplan + pspline(custservicecalls),
  data = telco_churn_tbl
  )

telco_model06_coxph |> summary()
```

**Interpreting the output:**

The spline appears as `pspline(custservicecalls)` in the output with multiple
degrees of freedom (typically 4-6). Unlike the linear model (1 df) or
discretised model (3 df for 4 groups), the spline uses several basis functions
to represent the smooth curve.

The summary provides:

  - **Linear component**: A test of whether the linear term is significant
  - **Non-linear component**: A test of whether the non-linear deviations from
    linearity are significant

If the non-linear component is significant (p < 0.05), this is strong
evidence that the relationship is genuinely non-linear and the spline
captures structure the linear model missed.

The overall degrees of freedom consumed by the spline are shown in the
output. More degrees of freedom = more flexible curve, but also more
parameters to estimate.

### Visualising the Spline Effect

The best way to understand a spline is to visualize it. The `termplot()`
function shows the estimated curve across the range of customer service
calls:

```{r visualise_spline_effect}
#| echo: true
#| fig-width: 8
#| fig-height: 5

termplot(
  telco_model06_coxph,
  term   = 3,
  se     = TRUE,
  col.term = "firebrick",
  col.se   = "grey70",
  xlab   = "Customer Service Calls",
  ylab   = "Log Hazard (centered)",
  main   = "Spline Effect of Customer Service Calls"
  )
```

**Interpreting the plot:**

  - **X-axis**: Number of customer service calls
  - **Y-axis**: Log-hazard (relative to the mean), centered at zero
  - **Red curve**: Estimated effect of calls on log-hazard
  - **Grey band**: Confidence interval (uncertainty in the curve)

Look for:

  - **Shape**: Is the curve roughly linear (straight line)? Or does it curve
    (accelerate or decelerate)?
  - **Confidence width**: Narrow bands = precise estimates; wide bands =
    high uncertainty
  - **Outlier behavior**: Does the curve behave erratically at extreme values
    (very high calls)? This suggests overfitting or sparse data.

If the curve is approximately linear, the linear model (Model 04) was adequate.
If the curve clearly bends—especially if it accelerates (gets steeper) at
higher values—this confirms non-linearity and justifies the added complexity.

### Checking Proportional Hazards for Model 06

```{r test_model06_ph_assumption}
#| echo: true

telco_model06_coxzph <- telco_model06_coxph |> cox.zph(transform = "km")

telco_model06_coxzph |> print()
```

The spline components should all pass the proportional hazards test. If
violations occur, it may indicate that the non-linear effect itself varies
over time, requiring more advanced methods beyond this workshop's scope.


# Model Comparison and Selection

We've now built six models of increasing complexity, from a simple null model
to models incorporating categorical predictors, continuous linear effects,
discretised non-linear effects, and flexible spline effects. The question now
is: **which model should we choose?**

Model selection involves balancing multiple criteria:

  - **Fit quality**: How well does the model discriminate and explain variation?
  - **Parsimony**: Simpler models are easier to interpret and less prone to
    overfitting
  - **Assumption validity**: Does the model satisfy proportional hazards?
  - **Interpretability**: Can we explain the model to stakeholders?
  - **Predictive accuracy**: How well will it generalize to new data?

There's rarely a single "best" model. The choice depends on the analysis
goals: exploratory understanding, prediction, or communication to decision-
makers. An exploratory analysis might favor the most flexible model (splines);
a model for executive presentation might favor simpler discretised or linear
specifications.

We'll use several comparison frameworks:

  1. **Summary table**: Concordance, R², AIC, BIC across all models
  1. **Likelihood ratio tests**: Formal statistical tests comparing nested models
  1. **Degrees of freedom consideration**: How many parameters does each model use?

The summary table provides an overview, likelihood ratio tests offer formal
statistical comparisons, and degrees of freedom remind us of complexity costs.

## Summary Comparison Table

```{r create_model_comparison_table}
#| echo: true

model_comparison_tbl <- bind_rows(
  telco_model01_coxph |> glance() |> mutate(model = "Model01: vmailplan"),
  telco_model03_coxph |> glance() |> mutate(model = "Model03: vmail + intl"),
  telco_model04_coxph |> glance() |> mutate(model = "Model04: + csc (linear)"),
  telco_model05_coxph |> glance() |> mutate(model = "Model05: + csc (factor)"),
  telco_model06_coxph |> glance() |> mutate(model = "Model06: + csc (spline)")
  ) |>
  select(model, n, nevent, concordance, r.squared, AIC, BIC) |>
  arrange(AIC)

model_comparison_tbl |> print()
```

**Interpreting the table:**

  - **Concordance**: Measures discrimination. Model 01 (vmailplan only) likely
    achieves c ≈ 0.565. Model 04-06 (with custservicecalls) likely achieve
    c ≈ 0.68-0.72. Look for diminishing returns—if Model 05 and 06 don't
    meaningfully improve on Model 04, the added complexity isn't justified.
  - **R²**: Variance explained. Expect low values (0.05-0.15) due to censoring
    and unmeasured factors. Watch for incremental gains.
  - **AIC/BIC**: Lower is better. These metrics penalize model complexity, so
    they balance fit against parsimony. AIC penalizes complexity less than BIC,
    so BIC favors simpler models more strongly.
  - **Order**: Table sorted by AIC. The top model (lowest AIC) represents the
    best balance of fit and complexity according to this criterion.

Key questions:

  - Does adding custservicecalls (Model 04) substantially improve on Model 03?
    (Almost certainly yes.)
  - Do non-linear specifications (Model 05/06) improve on the linear Model 04?
    (Check if AIC drops by more than ~2-3 points.)
  - Is there a "sweet spot" model that achieves most of the predictive gain
    without excessive complexity?

## Likelihood Ratio Tests

The comparison table shows overall metrics, but doesn't tell us whether
differences are *statistically significant*. Likelihood ratio tests (LRTs)
provide formal hypothesis testing for nested models.

**What is a nested model?** Model A is nested in Model B if A is a special case
of B. For example:

  - Model 03 (vmailplan + intlplan) is nested in Model 04 (vmailplan +
    intlplan + custservicecalls)
  - Model 04 (linear custservicecalls) is nested in Model 05 (discretised
    custservicecalls)—the factor specification allows different slopes within
    bins, whereas linear forces a single slope

The LRT compares the log-likelihoods of two nested models:

  - **Null hypothesis (H₀)**: The simpler model is adequate
  - **Alternative (H₁)**: The complex model provides significantly better fit
  - **Test statistic**: -2 × (log-likelihood of simple model - log-likelihood
    of complex model)
  - **Distribution**: Chi-squared with degrees of freedom = difference in
    parameter count
  - **Interpretation**: p < 0.05 suggests the complex model improves fit beyond
    chance

This is particularly useful when information criteria (AIC/BIC) give ambiguous
results or when we need to justify model choice to stakeholders who want
statistical rigor.

```{r likelihood_ratio_tests}
#| echo: true

write_lines("\n--- Testing Model03 vs Model04 (adding linear csc) ---", stdout())
anova(telco_model03_coxph, telco_model04_coxph, test = "LRT")

write_lines("\n--- Testing Model04 vs Model05 (linear vs factor) ---", stdout())
anova(telco_model04_coxph, telco_model05_coxph, test = "LRT")
```

**Interpreting the results:**

  - **Model 03 vs Model 04**: Adding custservicecalls (one parameter) should
    yield p ≪ 0.001, strongly rejecting the null. This confirms what we saw in
    the coefficient estimates: customer service calls are highly predictive.
  - **Model 04 vs Model 05**: This tests linear vs discretised. If p < 0.05,
    the binned specification captures non-linearity the linear model misses. If
    p > 0.05, the simpler linear model is sufficient. Check the degrees of
    freedom: Model 05 adds 3 parameters (4 bins - 1 baseline = 3 coefficients)
    versus Model 04's single slope.

**Degrees of freedom note**: Model 05 uses 3 additional parameters compared to
Model 04 (custservicecalls_cat has 4 levels, requiring 3 dummy variables).
Model 06 (spline with 4 knots) uses similar degrees of freedom. The LRT
accounts for this when computing the chi-squared statistic.

**Why we don't test Model 05 vs Model 06**: These models aren't nested. Both
capture non-linearity but in different ways (bins vs smooth splines). To
compare them, we rely on AIC/BIC and domain considerations (interpretability,
stakeholder preferences).


# Validating Model Improvement with Residuals

We've built models and assessed them with summary statistics, but how do we
*confirm* that adding predictors genuinely improved model fit? One powerful
diagnostic is comparing residual patterns before and after adding covariates.

**The logic**: If a covariate captures meaningful variation in churn hazard,
including it in the model should "absorb" that variation, leaving residuals
that are less patterned and closer to noise. If residuals still show strong
patterns after including the covariate, either:

  1. The functional form is wrong (e.g., linear when relationship is non-linear)
  2. The covariate isn't predictive
  3. There are other unmeasured confounders

We'll compare martingale residuals from the null model (Model 00) to residuals
from the fitted model (Model 05). This is particularly revealing for
custservicecalls, which we explored exhaustively in earlier sections.

## Residuals from Model 05

```{r calculate_fitted_model_residuals}
#| echo: true

telco_model05_resid_m <- telco_model05_coxph |> residuals(type = "martingale")

telco_fitted_assess_tbl <- telco_churn_cat_tbl |>
  mutate(
    resid_null  = telco_model00_resid_m,
    resid_model = telco_model05_resid_m
    )
```

### Comparing Residual Patterns

```{r compare_residual_patterns}
#| echo: true
#| fig-width: 10
#| fig-height: 5

null_resid_plot <- ggplot(
    telco_fitted_assess_tbl,
    aes(x = custservicecalls, y = resid_null)
    ) +
  geom_point(size = 0.8, alpha = 0.15) +
  geom_smooth(method = "loess", se = FALSE, colour = "firebrick") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Customer Service Calls",
    y     = "Martingale Residual",
    title = "Null Model Residuals"
    ) +
  theme_minimal()

fitted_resid_plot <- ggplot(
    telco_fitted_assess_tbl,
    aes(x = custservicecalls, y = resid_model)
    ) +
  geom_point(size = 0.8, alpha = 0.15) +
  geom_smooth(method = "loess", se = FALSE, colour = "firebrick") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Customer Service Calls",
    y     = "Martingale Residual",
    title = "Fitted Model Residuals"
    ) +
  theme_minimal()

plot_grid(
  null_resid_plot,
  fitted_resid_plot,
  ncol   = 2,
  labels = c("A", "B")
  )
```

**What to look for:**

  - **Panel A (Null Model)**: The LOESS smoother should show a clear positive
    trend. Higher custservicecalls → more positive residuals → model
    under-predicts hazard. This is the pattern we identified earlier.
  - **Panel B (Model 05)**: The LOESS should be much flatter, hovering near
    zero. This indicates Model 05 has successfully captured the relationship
    between custservicecalls and churn hazard. Any remaining trend would
    suggest functional form misspecification.

The visual comparison provides intuitive confirmation: if Panel B looks like
random scatter around zero, the model has done its job.

## Checking for Remaining Patterns

Having confirmed that Model 05 captures custservicecalls, we should check
whether residuals reveal *additional* predictive patterns in other covariates.
If they do, those covariates might be candidates for future model expansion.

We'll examine two variables not currently in the model: daytime minutes and
international minutes. These usage metrics might correlate with customer
engagement or satisfaction.

```{r check_remaining_patterns}
#| echo: true
#| fig-width: 10
#| fig-height: 5

daymins_resid_plot <- ggplot(
    telco_fitted_assess_tbl,
    aes(x = daymins, y = resid_model)
    ) +
  geom_point(size = 0.8, alpha = 0.15) +
  geom_smooth(method = "loess", se = FALSE, colour = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "Daytime Minutes",
    y     = "Martingale Residual",
    title = "Residuals vs Daytime Minutes"
    ) +
  theme_minimal()

intlmins_resid_plot <- ggplot(
    telco_fitted_assess_tbl,
    aes(x = intlmins, y = resid_model)
    ) +
  geom_point(size = 0.8, alpha = 0.15) +
  geom_smooth(method = "loess", se = FALSE, colour = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    x     = "International Minutes",
    y     = "Martingale Residual",
    title = "Residuals vs International Minutes"
    ) +
  theme_minimal()

plot_grid(
  daymins_resid_plot,
  intlmins_resid_plot,
  ncol   = 2,
  labels = c("A", "B")
  )
```

**Interpreting these plots:**

  - **Flat LOESS**: No relationship between the covariate and residuals. The
    current model adequately captures all systematic variation. Including this
    variable likely wouldn't improve predictions.
  - **Trending LOESS**: A non-zero trend indicates the covariate contains
    information the model hasn't used. This suggests adding the variable might
    improve fit.

**What you'll likely see**: For this dataset, daytime minutes and
international minutes probably show weak or no trends. Customer service calls
dominate the predictive signal. However, residual plots are always worth
checking—they can reveal unexpected patterns that summary statistics miss.

**Future modeling directions**: If either plot shows a strong pattern, consider:

  - Adding the variable to the model
  - Discretising or spline transformation if the relationship is non-linear
  - Interaction terms (e.g., does the effect of daytime minutes depend on
    having a voicemail plan?)

## Summary of Residual Validation

Through systematic residual analysis, we:

  1. Identified `custservicecalls` as the strongest continuous predictor
  1. Tested three approaches to modeling its effect (linear, discretised, spline)
  1. Demonstrated substantial model improvement through concordance and AIC
  1. Confirmed that residual patterns are reduced after including the predictor

This workflow—residual exploration, variable selection, model fitting,
and diagnostic checking—provides a principled approach to building
robust Cox proportional hazards models.


## Conclusions and Recommendations

  - **Preferred specification**: If the LRT indicates significant improvement
    and AIC/BIC decrease, choose Model 05 (discretised `custservicecalls`) with
    `vmailplan` and `intlplan`. It balances performance and interpretability.
  - **Simpler alternative**: If non-linearity is not statistically supported,
    Model 04 (linear `custservicecalls`) is sufficient and easier to explain.
  - **Exploratory flexibility**: Use Model 06 (splines) to explore nuanced
    non-linear effects. When presenting results, include plots and clear
    explanations as spline coefficients are not directly interpretable.
  - **Always validate PH**: Confirm proportional hazards via `cox.zph()` for
    each model and review residual plots to ensure no remaining structure.
  - **Communicate clearly**: Report hazard ratios with confidence intervals and
    a readable baseline (e.g., 0–1 calls) versus higher-call bins; avoid
    jargon when presenting to non-technical stakeholders.
  - **Next steps**: Consider interactions (e.g., service calls × vmailplan),
    additional covariates indicated by residual patterns, and calibration
    assessments if the goal is prediction.


# R Environment

```{r show_session_info}
#| echo: true
#| message: true

sessioninfo::session_info()
```
