---
title: "Explore the Lifebook Protection Data"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    theme:
      light: sandstone
      dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---


```{r resolve_conflicts}
#| echo: false

resolve_conflicts <- function(pkg_priority) {
  get_index <- function(pkg_name) {
    idx <- str_which(pkg_priority, pkg_name)

    if(length(idx) == 0) { idx <- 0L }

    return(idx)
  }

  conflict_lst <- conflict_scout()

  for(func_name in names(conflict_lst)) {
    pkg_index <- map_int(conflict_lst[[func_name]], get_index)

    pkg_index <- pkg_index[pkg_index > 0]

    if(length(pkg_index) == 0) {
      pkg_use <- conflict_lst[[func_name]][1]
    } else {
      pkg_use <- pkg_priority[pkg_index |> min()]
    }

    conflict_prefer(func_name, pkg_use)
  }

  return(conflict_lst)
}
```

```{r setup_chunk_timing}
#| echo: false
#| message: false

# Enable chunk timing for all chunks
knitr::opts_chunk$set(time_it = TRUE)

# Create a list to store chunk times
chunk_times <- list()

# Create a hook to capture chunk execution times
knitr::knit_hooks$set(time_it = function(before, options) {
  if (before) {
    # Record start time in global environment
    assign(".chunk_start_time", Sys.time(), envir = .GlobalEnv)
  } else {
    # Calculate elapsed time
    elapsed <- difftime(
      Sys.time(),
      get(".chunk_start_time", envir = .GlobalEnv),
      units = "secs"
      ) |>
      as.numeric()
    
    # Store in global list
    chunk_times <- get("chunk_times", envir = .GlobalEnv)
    chunk_times[[options$label]] <- elapsed
    assign("chunk_times", chunk_times, envir = .GlobalEnv)
    
    # Print timing message
    message(glue("⏱️  Chunk '{options$label}': {round(elapsed, 2)} seconds"))
  }
})
```

```{r import_libraries}
#| echo: false
#| message: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(vctrs)
library(fs)
library(glue)
library(forcats)
library(snakecase)
library(DataExplorer)
library(nanoparquet)
library(gtsummary)


source("lib_utils.R")
source("lib_survival_modelling.R")


conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2")
  )


options(
  width = 80L,
  warn  = 1,
  mc.cores = parallelly::availableCores()
  )

theme_set(theme_cowplot())

set.seed(42)

modelling_dataset_size_small <- 10000
modelling_dataset_size_large <- 100000

observation_date <- as.Date("2016-01-01")
```

```{r custom_functions}
#| echo: false

### Checks if variable is a date/time
is_date <- function(x)
  x |> inherits(c("POSIXt", "POSIXct", "POSIXlt", "Date", "hms"))


### Returns the category of data type passed to it
categorise_datatype <- function (x) {
  if(all(are_na(x))) return("na")

  if(is_date(x))                          "datetime"
  else if (!is_null(attributes(x)) ||
           all(is_character(x)))          "discrete"
  else if (all(is_logical(x)))            "logical"
  else                                    "continuous"
}


### create_coltype_list() splits columns into various types
create_coltype_list <- function(data_tbl) {
  coltypes  <- data_tbl |> map_chr(categorise_datatype)
  cat_types <- coltypes |> unique() |> sort()

  split_lst <- cat_types |>
    map(function(x) { coltypes[coltypes %in% x] |> names() })

  names(split_lst) <- coltypes |> unique() |> sort()

  coltype_lst <- list(
    split   = split_lst
   ,columns = coltypes
  )

  return(coltype_lst)
}

```


This workbook was created using the 'dataexpks' template:

https://github.com/DublinLearningGroup/dataexpks



# Introduction

This workbook performs the basic data exploration of the dataset.

```{r set_exploration_params}
#| echo: true

dataexp_level_exclusion_threshold <- 100

dataexp_cat_level_count <- 40
dataexp_hist_bins_count <- 50
```


# Load Data

First we load the dataset.

```{r load_dataset}
#| echo: true

rawdata_tbl <- read_parquet("data/lifeins_policybook_inoutforce_tbl.parquet")

glimpse(rawdata_tbl)
```


## Perform Quick Data Cleaning


```{r perform_simple_datatype_transforms}
#| echo: true

### _TEMPLATE_
### Do simple datatype transforms and save output in data_tbl
data_tbl <- rawdata_tbl %>%
  set_colnames(
    names(.) |> to_snake_case()
    )

glimpse(data_tbl)
```



```{r}
#| echo: false

#knitr::knit_exit()
```


## Create Derived Variables

We now create derived features useful for modelling. These values are
new variables calculated from existing variables in the data.

```{r create_derived_variables}
#| echo: true

data_tbl <- rawdata_tbl |>
  filter(prod_type == "protection") |>
  mutate(
    prem_sa_ratio = prem_ape / sum_assured,
    policy_vintage_year = year(policy_startdate),

    weeks_to_now    = difftime(
      observation_date, policy_startdate, units = "weeks"
      ) |> as.numeric(),
    weeks_to_status = difftime(
      policy_statuschangedate, policy_startdate, units = "weeks"
      ) |> as.numeric(),

    policy_lifetime = if_else(
      policy_status == "inforce",
      weeks_to_now,
      weeks_to_status
      )
    )

glimpse(data_tbl)
```



```{r}
#| echo: false

#knitr::knit_exit()
```


# Perform Basic Checks on Data

We now want to look at some very high level checks on the data, and we leverage
some of the functionality provided by `DataExplorer`.

## Create High-Level Visualisations

We first want to look at a visualisation of some high-level summarys of the
meta-data on this dataset. This gives us a quick view of the categorical and
1numeric values in the dataset, as well as the proportions of missing values.

```{r plot_dataexp_introduce}
#| echo: true

data_tbl |>
  plot_intro(
    title   = "High Level Table Summary",
    ggtheme = theme_cowplot()
    )
```


## Check Missing Values

Before we do anything with the data, we first check for missing values
in the dataset. In some cases, missing data is coded by a special
character rather than as a blank, so we first correct for this.

```{r replace_missing_character}
#| echo: true

### _TEMPLATE_
### ADD CODE TO CORRECT FOR DATA ENCODING HERE
```

With missing data properly encoded, we now visualise the missing data in a
number of different ways.

### Univariate Missing Data

```{r plot_univariate_missing_data}
#| echo: true

data_tbl |>
  plot_missing(
    title   = "Summary of Data Missingness",
    group   = list(Good = 0.05, Acceptable = 0.2, Bad = 0.8, Remove = 1),
    ggtheme = theme_cowplot()
    )
```

We now want to repeat this plot but only for those columns that have some
missing values.

```{r plot_univariate_missing_only_data}
#| echo: true

data_tbl |>
  plot_missing(
    title        = "Summary of Data Missingness (missing variables only)",
    missing_only = TRUE,
    group        = list(Good = 0.05, Acceptable = 0.2, Bad = 0.8, Remove = 1),
    ggtheme      = theme_cowplot()
    )
```



### Multivariate Missing Data

It is useful to get an idea of what combinations of variables tend to have
variables with missing values simultaneously, so to construct a visualisation
for this we create a count of all the times given combinations of variables
have missing values, producing a heat map for these combination counts.

```{r missing_data_matrix}
#| echo: true

dataexp_missing_group_count <- 20

row_count <- rawdata_tbl |> nrow()

count_nas <- ~ .x |> are_na() |> vec_cast(integer())

missing_vizdata_tbl <- rawdata_tbl |>
  mutate(across(everything(), count_nas)) %>%
  mutate(label = pmap_chr(., str_c)) |>
  group_by(label) |>
  mutate(
    miss_count = n(),
    miss_prop  = miss_count / row_count
    ) |>
  slice_max(order_by = miss_prop, n = 1, with_ties = FALSE) |>
  ungroup() |>
  pivot_longer(
    !c(label, miss_count, miss_prop),
    names_to = "variable_name",
    values_to = "presence"
    ) |>
  mutate(
    prop_label = sprintf("%6.4f", miss_prop)
    )

top10_data_tbl <- missing_vizdata_tbl |>
  select(label, miss_prop) |>
  distinct() |>
  slice_max(order_by = miss_prop, n = dataexp_missing_group_count)

missing_plot_tbl <- missing_vizdata_tbl |>
  semi_join(top10_data_tbl, by = "label")

ggplot(missing_plot_tbl) +
  geom_tile(aes(x = variable_name, y = prop_label, fill = presence), height = 0.8) +
  scale_fill_continuous() +
  scale_x_discrete(position = "top", labels = ~ abbreviate(.x, minlength = 10)) +
  xlab("Variable") +
  ylab("Proportion of Rows") +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90, vjust = 0.5)
    )
```

This visualisation takes a little explaining.

Each row represents a combination of variables with simultaneous missing
values. For each row in the graphic, the coloured entries show which particular
variables are missing in that combination. The proportion of rows with that
combination is displayed in both the label for the row and the colouring for
the cells in the row.


## Inspect High-level-count Categorical Variables

With the raw data loaded up we now remove obvious unique or near-unique
variables that are not amenable to basic exploration and plotting.

```{r find_highlevelcount_categorical_variables}
#| echo: true

coltype_lst <- create_coltype_list(data_tbl)

count_levels <- ~ .x |> unique() |> length()

catvar_valuecount_tbl <- data_tbl |>
  summarise(
    .groups = "drop",

    across(coltype_lst$split$discrete, count_levels)
    ) |>
  pivot_longer(
    cols      = everything(),
    names_to  = "var_name",
    values_to = "level_count"
    ) |>
  arrange(desc(level_count))

print(catvar_valuecount_tbl)

row_count <- data_tbl |> nrow()

cat(glue("Dataset has {row_count} rows\n"))
```

Now that we a table of the counts of all the categorical variables we can
automatically exclude unique variables from the exploration, as the level
count will match the row count.

```{r remove_id_variables}
#| echo: true

unique_vars <- catvar_valuecount_tbl |>
  filter(level_count == row_count) |>
  pull(var_name)

print(unique_vars)

explore_data_tbl <- data_tbl |>
  select(-one_of(unique_vars))
```

Having removed the unique identifier variables from the dataset, we
may also wish to exclude categoricals with high level counts also, so
we create a vector of those variable names.

```{r collect_highcount_variables}
#| echo: true

highcount_vars <- catvar_valuecount_tbl |>
  filter(
    level_count >= dataexp_level_exclusion_threshold,
    level_count < row_count
    ) |>
  pull(var_name)

cat(str_c(highcount_vars, collapse = ", "))
```

We now can continue doing some basic exploration of the data. We may
also choose to remove some extra columns from the dataset.

```{r drop_variables}
#| echo: true

### You may want to comment out these next few lines to customise which
### categoricals are kept in the exploration.
drop_vars <- c(highcount_vars)

if (length(drop_vars) > 0) {
  explore_data_tbl <- explore_data_tbl |>
      select(-one_of(drop_vars))

  cat(str_c(drop_vars, collapse = ", "))
}
```


```{r}
#| echo: false

#knitr::knit_exit()
```


# Univariate Data Exploration

Now that we have loaded the data we can prepare it for some basic data
exploration.


## Quick Univariate Data Summaries

We use a number of summary visualisations provided by `DataExplorer`: a
facet plot across each variable with categorical variables getting bar plots
and numerical plots getting histograms.

We first look at the barplots of categorical variables.

```{r plot_dataexp_bar}
#| message: true
#| echo: true

plot_bar(
    data_tbl,
    ncol    = 2,
    nrow    = 2,
    title   = "Barplots of Data",
    ggtheme = theme_cowplot()
    )
```


We then have a quick look at histograms of the numeric variables.

```{r plot_dataexp_hist}
#| message: true
#| echo: true

plot_histogram(
    data_tbl,
    ncol    = 2,
    nrow    = 2,
    title   = "Histograms of Data",
    ggtheme = theme_cowplot()
    )
```


Finally, we split the remaining variables into different categories and then
produce a sequence of plots for each variable.


```{r separate_exploration_cols}
#| echo: true

coltype_lst <- create_coltype_list(explore_data_tbl)

print(coltype_lst)
```


## Logical Variables

Logical variables only take two values: TRUE or FALSE. It is useful to see
missing data as well though, so we also plot the count of those.

```{r create_univariate_logical_plots}
#| echo: true
#| warning: false

logical_vars <- coltype_lst$split$logical |> sort()

for (plot_varname in logical_vars) {
  cat("--\n")
  cat(glue("{plot_varname}\n"))

  na_count <- explore_data_tbl |> pull(.data[[plot_varname]]) |> are_na() |> sum()

  plot_title <- glue("Barplot of Counts for Variable: {plot_varname} ({na_count} missing values)")

  explore_plot <- ggplot(explore_data_tbl) +
    geom_bar(aes(x = .data[[plot_varname]])) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(plot_title) +
    theme(axis.text.x = element_text(angle = 30, vjust = 0.5))

  plot(explore_plot)
}
```


## Numeric Variables

Numeric variables are usually continuous in nature, though we also have
integer data.

```{r create_univariate_numeric_plots}
#| echo: true
#| warning: false

numeric_vars <- coltype_lst$split$continuous |> sort()

for (plot_varname in numeric_vars) {
  cat("--\n")
  cat(glue("{plot_varname}\n"))

  plot_var <- explore_data_tbl |> pull(.data[[plot_varname]])
  na_count <- plot_var |> are_na() |> sum()

  plot_var |> summary() |> print()

  plot_title <- glue("Histogram Plot for Variable: {plot_varname} ({na_count} missing values)")


  all_plot <- ggplot() +
    geom_histogram(aes(x = plot_var), bins = dataexp_hist_bins_count) +
    geom_vline(xintercept = mean(plot_var, na.rm = TRUE),
               colour = "red", size = 1.5) +
    geom_vline(xintercept = median(plot_var, na.rm = TRUE),
               colour = "green", size = 1.5) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_x_continuous(labels = label_comma()) +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(
      plot_title,
      subtitle = "(red line is mean, green line is median)"
      )

  pos_data_tbl <- explore_data_tbl |>
    filter(.data[[plot_varname]] >= 0) |>
    mutate(var_val = abs(.data[[plot_varname]]))

  pos_log_plot <- ggplot(pos_data_tbl) +
    geom_histogram(aes(x = var_val), bins = dataexp_hist_bins_count) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_x_log10(labels = label_comma()) +
    scale_y_continuous(labels = label_comma()) +
    ggtitle("Positive Values")

  
  neg_data_tbl <- explore_data_tbl |>
    filter(.data[[plot_varname]] < 0) |>
    mutate(var_val = abs(.data[[plot_varname]]))

  neg_log_plot <- ggplot(neg_data_tbl) +
    geom_histogram(aes(x = var_val), bins = dataexp_hist_bins_count) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_x_log10(labels = label_comma()) +
    scale_y_continuous(labels = label_comma()) +
    ggtitle("Negative Values")


  plot_grid(
      all_plot,
      NULL,
      pos_log_plot,
      neg_log_plot,
      nrow = 2
      ) |>
    print()
}
```


## Categorical Variables

Categorical variables only have values from a limited, and usually fixed,
number of possible values

```{r create_univariate_categorical_plots}
#| echo: true
#| warning: false

categorical_vars <- coltype_lst$split$discrete |> sort()

for (plot_varname in categorical_vars) {
  cat("--\n")
  cat(glue("{plot_varname}\n"))

  na_count <- explore_data_tbl |> pull(.data[[plot_varname]]) |> are_na() |> sum()

  plot_title <- glue("Barplot of Counts for Variable: {plot_varname} ({na_count} missing values)")

  standard_plot_tbl <- explore_data_tbl |>
    count(.data[[plot_varname]])

  standard_plot <- ggplot(standard_plot_tbl) +
    geom_bar(aes(x = .data[[plot_varname]], weight = n)) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_x_discrete(labels = ~ abbreviate(.x, minlength = 10)) +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(plot_title) +
    theme(axis.text.x = element_text(angle = 30, vjust = 0.5))

  standard_plot |> print()


  desc_plot_tbl <- explore_data_tbl |>
    pull(.data[[plot_varname]]) |>
    fct_lump(n = dataexp_cat_level_count) |>
    fct_count() |>
    mutate(f = fct_relabel(f, str_trunc, width = 15))

  desc_plot <- ggplot(desc_plot_tbl) +
    geom_bar(aes(x = fct_reorder(f, -n), weight = n)) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_x_discrete(labels = ~ abbreviate(.x, minlength = 10)) +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(plot_title) +
    theme(axis.text.x = element_text(angle = 30, vjust = 0.5))

  desc_plot |> print()
}
```


## Date/Time Variables

Date/Time variables represent calendar or time-based data should as time of the
day, a date, or a timestamp.

```{r create_univariate_datetime_plots}
#| echo: true
#| warning: false

datetime_vars <- coltype_lst$split$datetime |> sort()

for (plot_varname in datetime_vars) {
  cat("--\n")
  cat(glue("{plot_varname}\n"))

  plot_var <- explore_data_tbl |> pull(.data[[plot_varname]])
  na_count <- plot_var |> are_na() |> sum()

  plot_var |> summary() |> print()

  plot_title <- glue("Barplot of Dates/Times in Variable: {plot_varname} ({na_count} missing values)")


  explore_plot <- ggplot(explore_data_tbl) +
    geom_histogram(aes(x = .data[[plot_varname]]), bins = dataexp_hist_bins_count) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(plot_title)

  plot(explore_plot)
}
```


```{r}
#| echo: false

#knitr::knit_exit()
```


# Bivariate Facet Plots

We now move on to looking at bivariate plots of the data set.

A natural way to explore relationships in data is to create univariate
visualisations facetted by a categorical value.

```{r bivariate_facet_data}
#| echo: true

### _TEMPLATE_
### facet_varname <- ''
facet_varname <- "cluster_id"

dataexp_facet_count_max <- 3
```


## Logical Variables

For logical variables we facet on barplots of the levels, comparing TRUE,
FALSE and missing data.

```{r create_bivariate_logical_plots}
#| echo: true

logical_vars <- logical_vars[!logical_vars %in% facet_varname] |> sort()


for (plot_varname in logical_vars) {
  cat("--\n")
  cat(plot_varname)

  plot_tbl <- data_tbl |> filter(!are_na(.data[[plot_varname]]))

  explore_plot <- ggplot(plot_tbl) +
    geom_bar(aes(x = .data[[plot_varname]])) +
    facet_wrap(facet_varname, scales = "free") +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(glue("{facet_varname}-Faceted Histogram for Variable: {plot_varname}")) +
    theme(axis.text.x = element_text(angle = 30, vjust = 0.5))

  plot(explore_plot)
}
```


## Numeric Variables

For numeric variables, we facet on histograms of the data.

```{r create_bivariate_numeric_plots}
#| echo: true

for (plot_varname in numeric_vars) {
  cat("--\n")
  cat(plot_varname)

  plot_tbl <- data_tbl |> filter(!are_na(.data[[plot_varname]]))

  explore_plot <- ggplot(plot_tbl) +
    geom_histogram(aes(x = .data[[plot_varname]]), bins = dataexp_hist_bins_count) +
    facet_wrap(facet_varname, scales = "free") +
    xlab(plot_varname) +
    ylab("Count") +
    scale_x_continuous(labels = label_comma()) +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(glue("{facet_varname}-Faceted Histogram for Variable: {plot_varname}")) +
    theme(axis.text.x = element_text(angle = 30, vjust = 0.5))

  print(explore_plot)
}
```

## Categorical Variables

We treat categorical variables like logical variables, faceting the barplots
of the different levels of the data.

```{r create_bivariate_categorical_plots}
#| echo: true

categorical_vars <- categorical_vars[!categorical_vars %in% facet_varname] |> sort()

for (plot_varname in categorical_vars) {
  cat("--\n")
  cat(plot_varname)

  plot_tbl <- data_tbl |>
    filter(!are_na(.data[[plot_varname]])) |>
    mutate(
      varname_trunc = fct_relabel(.data[[plot_varname]], str_trunc, width = 10)
      )

  explore_plot <- ggplot(plot_tbl) +
    geom_bar(aes(x = varname_trunc)) +
    facet_wrap(facet_varname, scales = "free") +
    xlab(plot_varname) +
    ylab("Count") +
    scale_x_discrete(labels = ~ abbreviate(.x, minlength = 10)) +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(glue("{facet_varname}-Faceted Histogram for Variable: {plot_varname}")) +
    theme(axis.text.x = element_text(angle = 30, vjust = 0.5))

  plot(explore_plot)
}
```


## Date/Time Variables

Like the univariate plots, we facet on histograms of the years in the dates.

```{r create_bivariate_datetime_plots}
#| echo: true

for (plot_varname in datetime_vars) {
  cat("--\n")
  cat(plot_varname)

  plot_tbl <- data_tbl |> filter(!are_na(.data[[plot_varname]]))

  explore_plot <- ggplot(plot_tbl) +
    geom_histogram(aes(x = .data[[plot_varname]]), bins = dataexp_hist_bins_count) +
    facet_wrap(facet_varname, scales = "free") +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(glue("{facet_varname}-Faceted Histogram for Variable: {plot_varname}")) +
    theme(axis.text.x = element_text(angle = 30, vjust = 0.5))

  plot(explore_plot)
}
```

```{r free_memory_facetplot}
#| echo: false

rm(plot_var, plot_tbl)
```


```{r}
#| echo: false

#knitr::knit_exit()
```


# Custom Explorations

In this section we perform various data explorations specific to the insurance
policy data and survival analysis context.


## Summary Statistics by Policy Status

We examine key variables stratified by policy status to understand differences
between in-force and lapsed policies.

```{r summary_by_status}
#| echo: true

data_tbl |>
  group_by(policy_status) |>
  summarise(
    count = n(),
    mean_lifetime      = mean  (policy_lifetime, na.rm = TRUE),
    median_lifetime    = median(policy_lifetime, na.rm = TRUE),
    sd_lifetime        = sd    (policy_lifetime, na.rm = TRUE),
    mean_sum_assured   = mean  (sum_assured,     na.rm = TRUE),
    median_sum_assured = median(sum_assured,     na.rm = TRUE),
    mean_prem_ape      = mean  (prem_ape,        na.rm = TRUE),
    median_prem_ape    = median(prem_ape,        na.rm = TRUE),

    .groups = "drop"
    ) |>
  print()
```


## Distribution of Policy Lifetime by Status

Compare the distribution of policy lifetimes between in-force and lapsed
policies.

```{r plot_lifetime_by_status}
#| echo: true

lifetime_plot1 <- ggplot(data_tbl) +
  geom_histogram(
    aes(x = policy_lifetime, fill = policy_status),
    bins = 50,
    alpha = 0.6,
    position = "identity"
    ) +
  scale_x_continuous(labels = label_comma()) +
  scale_y_continuous(labels = label_comma()) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Distribution of Policy Lifetime by Status",
    x = "Policy Lifetime (weeks)",
    y = "Count",
    fill = "Policy Status"
    )

lifetime_plot2 <- ggplot(data_tbl) +
  geom_density(
    aes(x = policy_lifetime, fill = policy_status),
    alpha = 0.5
    ) +
  scale_x_continuous(labels = label_comma()) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Density of Policy Lifetime by Status",
    x = "Policy Lifetime (weeks)",
    y = "Density",
    fill = "Policy Status"
    )

plot_grid(lifetime_plot1, lifetime_plot2, nrow = 2)
```


## Policy Lifetime by Demographic Factors

Examine how policy lifetime varies across different demographic segments.

```{r plot_lifetime_by_demographics}
#| echo: true

gender_plot <- ggplot(data_tbl |> filter(!is.na(gender_life1))) +
  geom_boxplot(
    aes(x = gender_life1, y = policy_lifetime, fill = gender_life1),
    alpha = 0.7
    ) +
  scale_y_continuous(labels = label_comma()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Policy Lifetime by Gender",
    x = "Gender",
    y = "Policy Lifetime (weeks)"
    ) +
  theme(legend.position = "none")

smoker_plot <- ggplot(data_tbl |> filter(!is.na(smoker_life1))) +
  geom_boxplot(
    aes(x = smoker_life1, y = policy_lifetime, fill = smoker_life1),
    alpha = 0.7
    ) +
  scale_y_continuous(labels = label_comma()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Policy Lifetime by Smoker Status",
    x = "Smoker Status",
    y = "Policy Lifetime (weeks)"
    ) +
  theme(legend.position = "none")

status_plot <- ggplot(data_tbl |> filter(!is.na(policy_status))) +
  geom_boxplot(
    aes(x = policy_status, y = policy_lifetime, fill = policy_status),
    alpha = 0.7
    ) +
  scale_y_continuous(labels = label_comma()) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Policy Lifetime by Status",
    x = "Policy Status",
    y = "Policy Lifetime (weeks)"
    ) +
  theme(legend.position = "none")

plot_grid(gender_plot, smoker_plot, status_plot, ncol = 2, nrow = 2)
```


## Policy Vintage Year Analysis

Explore how policy lapse rates vary by policy inception year (vintage). This
helps identify whether older or newer policies have different lapse
characteristics, which could indicate cohort effects or changes in
underwriting/product design.

```{r vintage_year_summary}
#| echo: true

# Summary of policies by vintage year
vintage_summary <- data_tbl |>
  group_by(policy_vintage_year) |>
  summarise(
    count = n(),
    pct_total = 100 * n() / nrow(data_tbl),
    mean_premium = mean(prem_ape, na.rm = TRUE),
    mean_sum_assured = mean(sum_assured, na.rm = TRUE),
    mean_lifetime_weeks = mean(policy_lifetime, na.rm = TRUE),
    .groups = "drop"
    ) |>
  arrange(policy_vintage_year)

write_lines("Policy Count and Characteristics by Vintage Year:", stdout())
print(vintage_summary)
```

### Lapse Rate by Vintage Year

```{r vintage_lapse_analysis}
#| echo: true

# Calculate lapse rates by vintage year
vintage_lapse_tbl <- data_tbl |>
  mutate(lapsed = if_else(policy_status == "lapsed", 1L, 0L)) |>
  group_by(policy_vintage_year) |>
  summarise(
    total_policies = n(),
    lapsed_count = sum(lapsed, na.rm = TRUE),
    lapse_rate = 100 * mean(lapsed, na.rm = TRUE),
    mean_lifetime = mean(policy_lifetime, na.rm = TRUE),
    .groups = "drop"
    ) |>
  arrange(policy_vintage_year)

write_lines("\nLapse Rates by Vintage Year:", stdout())
print(vintage_lapse_tbl)

# Visualize lapse rates by vintage year
vintage_lapse_plot <- ggplot(vintage_lapse_tbl) +
  geom_col(
    aes(x = factor(policy_vintage_year), y = lapse_rate, fill = lapse_rate),
    alpha = 0.8
    ) +
  geom_text(
    aes(x = factor(policy_vintage_year), y = lapse_rate,
        label = sprintf("%.1f%%", lapse_rate)),
    vjust = -0.5,
    size = 3.5
    ) +
  scale_fill_gradient(low = "steelblue", high = "firebrick") +
  labs(
    title = "Policy Lapse Rate by Vintage Year",
    x = "Policy Vintage Year",
    y = "Lapse Rate (%)",
    fill = "Lapse Rate"
    ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Visualize policy counts and mean lifetime by vintage
vintage_context_plot <- ggplot(vintage_summary) +
  geom_col(
    aes(x = factor(policy_vintage_year), y = count),
    fill = "steelblue",
    alpha = 0.6
    ) +
  geom_line(
    aes(x = factor(policy_vintage_year), y = mean_lifetime_weeks * 2,
        group = 1),
    color = "firebrick",
    size = 1
    ) +
  geom_point(
    aes(x = factor(policy_vintage_year), y = mean_lifetime_weeks * 2),
    color = "firebrick",
    size = 3
    ) +
  scale_y_continuous(
    name = "Number of Policies",
    sec.axis = sec_axis(~./2, name = "Mean Lifetime (weeks)")
    ) +
  labs(
    title = "Policy Count and Mean Lifetime by Vintage Year",
    x = "Policy Vintage Year"
    ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot_grid(vintage_lapse_plot, vintage_context_plot, ncol = 2)
```


## Financial Variables Analysis

Examine the relationship between financial variables (sum assured, premiums)
and policy outcomes.

```{r plot_financial_analysis}
#| echo: true

sa_plot <- ggplot(data_tbl |> filter(sum_assured > 0)) +
  geom_histogram(
    aes(x = sum_assured, fill = policy_status),
    bins = 50,
    alpha = 0.6,
    position = "identity"
    ) +
  scale_x_log10(labels = label_comma()) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Sum Assured Distribution (Log Scale)",
    x = "Sum Assured",
    y = "Count",
    fill = "Policy Status"
    )

prem_plot <- ggplot(data_tbl |> filter(prem_ape > 0)) +
  geom_histogram(
    aes(x = prem_ape, fill = policy_status),
    bins = 50,
    alpha = 0.6,
    position = "identity"
    ) +
  scale_x_log10(labels = label_comma()) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Premium APE Distribution (Log Scale)",
    x = "Premium APE",
    y = "Count",
    fill = "Policy Status"
    )

sa_scatter <- ggplot(data_tbl |> filter(sum_assured > 0, policy_lifetime > 0)) +
  geom_point(
    aes(x = sum_assured, y = policy_lifetime, color = policy_status),
    alpha = 0.3,
    size = 0.5
    ) +
  scale_x_log10(labels = label_comma()) +
  scale_y_continuous(labels = label_comma()) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Policy Lifetime vs Sum Assured",
    x = "Sum Assured (log scale)",
    y = "Policy Lifetime (weeks)",
    color = "Policy Status"
    )

prem_scatter <- ggplot(data_tbl |> filter(prem_ape > 0, policy_lifetime > 0)) +
  geom_point(
    aes(x = prem_ape, y = policy_lifetime, color = policy_status),
    alpha = 0.3,
    size = 0.5
    ) +
  scale_x_log10(labels = label_comma()) +
  scale_y_continuous(labels = label_comma()) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Policy Lifetime vs Premium APE",
    x = "Premium APE (log scale)",
    y = "Policy Lifetime (weeks)",
    color = "Policy Status"
    )

plot_grid(sa_plot, prem_plot, sa_scatter, prem_scatter, ncol = 2, nrow = 2)
```


## Premium to Sum Assured Ratio Analysis

Examine the ratio of premiums to sum assured as a measure of pricing
efficiency.

```{r plot_prem_sa_ratio}
#| echo: true

ratio_hist <- ggplot(data_tbl |> filter(prem_sa_ratio > 0, prem_sa_ratio < 1)) +
  geom_histogram(
    aes(x = prem_sa_ratio, fill = policy_status),
    bins = 50,
    alpha = 0.6,
    position = "identity"
    ) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Premium/Sum Assured Ratio Distribution",
    x = "Premium/Sum Assured Ratio",
    y = "Count",
    fill = "Policy Status"
    )

ratio_box <- ggplot(data_tbl |> filter(prem_sa_ratio > 0, prem_sa_ratio < 1)) +
  geom_boxplot(
    aes(x = policy_status, y = prem_sa_ratio, fill = policy_status),
    alpha = 0.7
    ) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Premium/Sum Assured Ratio by Status",
    x = "Policy Status",
    y = "Premium/Sum Assured Ratio"
    ) +
  theme(legend.position = "none")

plot_grid(ratio_hist, ratio_box, ncol = 2)
```


## Temporal Patterns

Analyze when policies were started and how this relates to policy outcomes.

```{r plot_temporal_patterns}
#| echo: true

start_timeline <- ggplot(data_tbl) +
  geom_histogram(
    aes(x = policy_startdate, fill = policy_status),
    bins = 50,
    alpha = 0.6,
    position = "identity"
    ) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Policy Start Dates Over Time",
    x = "Policy Start Date",
    y = "Count",
    fill = "Policy Status"
    )

start_by_year <- data_tbl |>
  mutate(
    start_year = lubridate::year(policy_startdate)
    ) |>
  filter(!is.na(start_year))

year_summary <- ggplot(start_by_year) +
  geom_bar(
    aes(x = factor(start_year), fill = policy_status),
    position = "fill",
    alpha = 0.8
    ) +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Proportion of Policy Status by Start Year",
    x = "Start Year",
    y = "Proportion",
    fill = "Policy Status"
    ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot_grid(start_timeline, year_summary, ncol = 1, nrow = 2)
```


## Cluster Analysis

Examine differences across clusters in the data.

```{r plot_cluster_analysis}
#| echo: true

cluster_counts <- data_tbl |>
  filter(!is.na(cluster_id)) |>
  count(cluster_id, policy_status) |>
  group_by(cluster_id) |>
  mutate(
    total = sum(n),
    proportion = n / total
    ) |>
  ungroup()

cluster_bar <- ggplot(cluster_counts) +
  geom_bar(
    aes(x = cluster_id, y = n, fill = policy_status),
    stat = "identity",
    position = "dodge",
    alpha = 0.8
    ) +
  scale_y_continuous(labels = label_comma()) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Policy Counts by Cluster and Status",
    x = "Cluster ID",
    y = "Count",
    fill = "Policy Status"
    )

cluster_prop <- ggplot(cluster_counts) +
  geom_bar(
    aes(x = cluster_id, y = proportion, fill = policy_status),
    stat = "identity",
    position = "fill",
    alpha = 0.8
    ) +
  scale_y_continuous(labels = label_percent()) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Policy Status Proportion by Cluster",
    x = "Cluster ID",
    y = "Proportion",
    fill = "Policy Status"
    )

cluster_lifetime <- data_tbl |>
  filter(!is.na(cluster_id)) |>
  ggplot() +
  geom_boxplot(
    aes(x = factor(cluster_id), y = policy_lifetime, fill = factor(cluster_id)),
    alpha = 0.7
    ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title = "Policy Lifetime Distribution by Cluster",
    x = "Cluster ID",
    y = "Policy Lifetime (weeks)"
    ) +
  theme(legend.position = "none")

plot_grid(cluster_bar, cluster_prop, cluster_lifetime, ncol = 2, nrow = 2)
```


## Cross-tabulation of Categorical Variables

Create detailed cross-tabulations to understand relationships between
categorical variables.

```{r crosstabs_analysis}
#| echo: true

# Gender vs Smoker Status
write_lines("Gender vs Smoker Status:", stdout())
data_tbl |>
  filter(!is.na(gender_life1), !is.na(smoker_life1)) |>
  count(gender_life1, smoker_life1) |>
  pivot_wider(
    names_from = smoker_life1, 
    values_from = n, 
    values_fill = 0
    ) |>
  print()

write_lines("\nGender vs Policy Status:", stdout())
data_tbl |>
  filter(!is.na(gender_life1), !is.na(policy_status)) |>
  count(gender_life1, policy_status) |>
  pivot_wider(
    names_from = policy_status, 
    values_from = n, 
    values_fill = 0
    ) |>
  print()

write_lines("\nSmoker Status vs Policy Status:", stdout())
data_tbl |>
  filter(!is.na(smoker_life1), !is.na(policy_status)) |>
  count(smoker_life1, policy_status) |>
  pivot_wider(
    names_from = policy_status, 
    values_from = n, 
    values_fill = 0
    ) |>
  print()
```


## Correlation Analysis of Numeric Variables

Examine correlations between numeric variables to identify potential
multicollinearity or interesting relationships.

```{r correlation_analysis}
#| echo: true

numeric_data <- data_tbl |>
  select(
    policy_lifetime, sum_assured, prem_ape,
    prem_sa_ratio, weeks_to_now, weeks_to_status
    )

cor_matrix <- cor(numeric_data, use = "complete.obs")

cor_tbl <- cor_matrix |>
  as.data.frame() |>
  rownames_to_column(var = "var1") |>
  pivot_longer(
    cols      = -var1,
    names_to  = "var2",
    values_to = "correlation"
    )

ggplot(cor_tbl) +
  geom_tile(
    aes(x = var1, y = var2, fill = correlation),
    height = 0.8
    ) +
  geom_text(
    aes(x = var1, y = var2, label = sprintf("%.2f", correlation)),
    size = 3
    ) +
  scale_fill_gradient2(
    low      = "blue",
    mid      = "white",
    high     = "red",
    midpoint = 0,
    limits   = c(-1, 1)
    ) +
  labs(
    title = "Correlation Matrix of Numeric Variables",
    x     = "",
    y     = "",
    fill  = "Correlation"
    ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(angle = 0)
    )
```


## Lapse Rate by Cohort

Calculate and visualize lapse rates by different cohorts.

```{r lapse_rate_analysis}
#| echo: true

lapse_by_gender_smoker <- data_tbl |>
  filter(!is.na(gender_life1), !is.na(smoker_life1)) |>
  group_by(gender_life1, smoker_life1) |>
  summarise(
    total      = n(),
    lapsed     = sum(lapsed),
    lapse_rate = lapsed / total,
    .groups    = "drop"
    ) |>
  mutate(
    group = paste(gender_life1, smoker_life1, sep = "\n")
    )

demo_lapse_plot <- ggplot(lapse_by_gender_smoker) +
  geom_bar(
    aes(x = group, y = lapse_rate, fill = group),
    stat = "identity",
    alpha = 0.7
    ) +
  geom_text(
    aes(x = group, y = lapse_rate, label = sprintf("%.1f%%", lapse_rate * 100)),
    vjust = -0.5
    ) +
  scale_y_continuous(
    labels = label_percent(),
    limits = c(0, max(lapse_by_gender_smoker$lapse_rate) * 1.1)
    ) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Lapse Rate by Gender and Smoker Status",
    x = "Demographic Group",
    y = "Lapse Rate"
    ) +
  theme(legend.position = "none")

plot(demo_lapse_plot)
```


## Summary Statistics Table

Create a comprehensive summary statistics table for key variables.

```{r summary_stats_table}
#| echo: true

data_tbl |>
  tbl_summary(
    include = c(
      smoker_life1, gender_life1, policy_lifetime, sum_assured, prem_ape,
      prem_sa_ratio
      ),
    statistic = list(
      all_continuous()  ~ "({mean} +/- {sd}) : ({p25}, {median}, {p75})",
      all_categorical() ~ "{n} / {N} ({p}%)"
      )
    )

```


# Set Up Modelling Datasets

In this section we create some basic survival models that we can use to
train our models and use to validate our models.

## Filter Protection Policies

We start by working only with protection policies (`prod_type` == 'protection')
so we filter those out first and also need to calculate the observed lifetime
of the policy, which depends on the status of the policy.


```{r determine_policy_lifetime}
#| echo: FALSE

policy_protection_data_tbl <- data_tbl |>
  filter(
    prod_type == "protection",
    prem_freq == 12
    ) |>
  select(-weeks_to_now, -weeks_to_status, -policy_lifetime)

policy_protection_data_tbl |> glimpse()
```

## Construct Rollback Versions

We know the data that is up to date to our observed date of Jan 1 2016, but for
modelling purposes, we need to roll back the observation date by a period of
time so we can use the most recent observations as a way to validate our model.

To do this, we need to construct some time-based observations.

```{r construct_rollback_data}
#| echo: true

rollback_data_tbl <- policy_protection_data_tbl |>
  determine_policy_status(reference_date = as.Date("2015-01-01"))

rollback_data_tbl |> glimpse()
```


We now re-do the protection data as observed at the current date.

```{r construct_current_data}
#| echo: true

current_data_tbl <- policy_protection_data_tbl |>
  determine_policy_status(reference_date = as.Date("2016-01-01"))

current_data_tbl |> glimpse()
```


### Check Rollback Data

We now want to perform a number of sanity checks on this data, to ensure that
the rollback logic is correct.

```{r create_rollback_comparison_checks}
#| echo: true

roll_tbl <- rollback_data_tbl |>
  select(
    policy_id, policy_status, lapsed, status_at_reference_date,
    policy_lifetime, age_at_observation
    )

curr_tbl <- current_data_tbl |>
  semi_join(roll_tbl, by = "policy_id") |>
  select(
    policy_id, policy_status, lapsed, status_at_reference_date,
    policy_lifetime, age_at_observation
    )


comparison_tbl <- roll_tbl |>
  inner_join(curr_tbl, by = "policy_id", suffix = c("_rl", "_cr"))

comparison_tbl |> glimpse()
```


### Sanity Check 1: Policy Status Consistency

Check that policies that lapsed before the rollback date have the same status
in both datasets.

```{r check_policy_status_consistency}
#| echo: true

write_lines("=== Policy Status Consistency Checks ===\n", stdout())

# Check 1: Policies lapsed in rollback should still be lapsed in current
rollback_lapsed_tbl <- comparison_tbl |>
  filter(status_at_reference_date_rl == "lapsed")

rollback_lapsed_status_tbl <- rollback_lapsed_tbl |>
  count(status_at_reference_date_cr)

rl_lapsed_count <- nrow(rollback_lapsed_tbl)
rl_lapsed_still_lapsed <- rollback_lapsed_status_tbl |>
  filter(status_at_reference_date_cr == "lapsed") |>
  pull(n) |>
  sum()
rl_lapsed_now_inforce <- rollback_lapsed_status_tbl |>
  filter(status_at_reference_date_cr == "inforce") |>
  pull(n) |>
  sum()

write_lines(glue(
  "Policies lapsed in rollback data: {rl_lapsed_count}
  Of these, also lapsed in current data: {rl_lapsed_still_lapsed}
  Of these, inforce in current data: {rl_lapsed_now_inforce}\n"
), stdout())

# Check 2: Policies inforce in rollback could be lapsed or still inforce in current
rollback_inforce_tbl <- comparison_tbl |>
  filter(status_at_reference_date_rl == "inforce")

rollback_inforce_status_tbl <- rollback_inforce_tbl |>
  count(status_at_reference_date_cr)

rl_inforce_count <- nrow(rollback_inforce_tbl)
rl_inforce_still_inforce <- rollback_inforce_status_tbl |>
  filter(status_at_reference_date_cr == "inforce") |>
  pull(n) |>
  sum()
rl_inforce_now_lapsed <- rollback_inforce_status_tbl |>
  filter(status_at_reference_date_cr == "lapsed") |>
  pull(n) |>
  sum()
lapse_rate <- 100 * rl_inforce_now_lapsed / rl_inforce_count

write_lines(glue(
  "Policies inforce in rollback data: {rl_inforce_count}
  Of these, still inforce in current data: {rl_inforce_still_inforce}
  Of these, lapsed in current data: {rl_inforce_now_lapsed}
  Lapse rate for policies inforce at rollback: {sprintf('%.2f%%', lapse_rate)}\n"
), stdout())

# Check 3: Status transitions table
write_lines("Status Transition Matrix (Rollback -> Current):", stdout())
comparison_tbl |>
  count(status_at_reference_date_rl, status_at_reference_date_cr) |>
  pivot_wider(
    names_from  = status_at_reference_date_cr,
    values_from = n,
    values_fill = 0
    ) |>
  print()
```


### Sanity Check 2: Policy Lifetime Changes

Check that policy lifetimes are consistent - lapsed policies should have the
same lifetime in both datasets, while inforce policies should have longer
lifetimes in the current dataset.

```{r check_policy_lifetime_changes}
#| echo: true

write_lines("\n=== Policy Lifetime Consistency Checks ===\n", stdout())

# Check 1: Policies lapsed before rollback should have SAME lifetime
lapsed_before_rollback_tbl <- comparison_tbl |>
  filter(status_at_reference_date_rl == "lapsed") |>
  mutate(lifetime_diff = policy_lifetime_cr - policy_lifetime_rl)

lapsed_count <- nrow(lapsed_before_rollback_tbl)
mean_diff <- lapsed_before_rollback_tbl |>
  pull(lifetime_diff) |>
  mean()
max_diff <- lapsed_before_rollback_tbl |>
  pull(lifetime_diff) |>
  abs() |>
  max()

write_lines(glue(
  "Policies lapsed before rollback date:
    Count: {lapsed_count}
    Mean lifetime difference (current - rollback): {mean_diff} weeks
    Max lifetime difference: {max_diff} weeks\n"
), stdout())

# Check 2: Policies inforce at rollback should have LONGER lifetime in current
inforce_at_rollback_tbl <- comparison_tbl |>
  filter(status_at_reference_date_rl == "inforce") |>
  mutate(lifetime_diff = policy_lifetime_cr - policy_lifetime_rl)

inforce_count <- nrow(inforce_at_rollback_tbl)
mean_diff_inforce <- inforce_at_rollback_tbl |>
  pull(lifetime_diff) |>
  mean(na.rm = TRUE)

write_lines(glue(
  "Policies inforce at rollback date:
    Count: {inforce_count}
    Mean lifetime difference (current - rollback): {mean_diff_inforce} weeks
    Expected difference (52 weeks for 1 year): 52 weeks\n"
), stdout())

# Check 3: Policies that lapsed between rollback and current
lapsed_between_tbl <- comparison_tbl |>
  filter(
    status_at_reference_date_rl == "inforce",
    status_at_reference_date_cr == "lapsed"
  ) |>
  mutate(additional_survival = policy_lifetime_cr - policy_lifetime_rl)

lapsed_between_count <- nrow(lapsed_between_tbl)

write_lines(glue("Policies that lapsed between rollback and current:
    Count: {lapsed_between_count}"), stdout())

if (lapsed_between_count > 0) {
  mean_addl <- lapsed_between_tbl |> pull(additional_survival) |> mean()
  min_addl  <- lapsed_between_tbl |> pull(additional_survival) |> min()
  max_addl  <- lapsed_between_tbl |> pull(additional_survival) |> max()

  write_lines(glue(
    "    Mean additional survival time: {mean_addl} weeks
    Min additional survival time: {min_addl} weeks
    Max additional survival time: {max_addl} weeks\n"
  ), stdout())
}

# Visualization of lifetime changes
lifetime_change_plot <- ggplot(comparison_tbl) +
  geom_point(
    aes(
      x     = policy_lifetime_rl,
      y     = policy_lifetime_cr,
      color = status_at_reference_date_rl
      ),
    alpha = 0.3,
    size  = 0.5
    ) +
  geom_abline(
    slope     = 1,
    intercept = 0,
    linetype  = "dashed",
    color     = "red"
    ) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "Policy Lifetime: Rollback vs Current",
    subtitle = "Red line = no change; points above line = increased lifetime",
    x        = "Policy Lifetime at Rollback (weeks)",
    y        = "Policy Lifetime at Current (weeks)",
    color    = "Status at Rollback"
    )

print(lifetime_change_plot)
```


### Sanity Check 3: Age at Observation Changes

Check that ages are correctly calculated and change appropriately between
rollback and current dates.

```{r check_age_changes}
#| echo: true

write_lines("\n=== Age at Observation Consistency Checks ===\n", stdout())

# Check 1: All ages should increase by approximately 1 year (52 weeks / 52.18 weeks per year)
expected_age_increase <- 1.0  # 1 year between rollback and current

comparison_tbl <- comparison_tbl |>
  mutate(age_difference = age_at_observation_cr - age_at_observation_rl)

age_stats_tbl <- comparison_tbl |>
  summarise(
    mean_age   = mean  (age_difference, na.rm = TRUE),
    median_age = median(age_difference, na.rm = TRUE),
    sd_age     = sd    (age_difference, na.rm = TRUE)
    )

write_lines(glue(
  "Age change statistics:
    Mean age increase: {age_stats_tbl$mean_age} years
    Median age increase: {age_stats_tbl$median_age} years
    Expected age increase: {expected_age_increase} years
    Std dev of age increase: {age_stats_tbl$sd_age} years\n"
), stdout())

# Check 2: Age increases should be consistent for lapsed policies
lapsed_age_check_tbl <- comparison_tbl |>
  filter(status_at_reference_date_rl == "lapsed")

mean_age_lapsed <- lapsed_age_check_tbl |>
  pull(age_difference) |>
  mean(na.rm = TRUE)

write_lines(glue(
  "Age changes for policies lapsed before rollback:
    Mean age increase: {mean_age_lapsed} years
    This should be close to {expected_age_increase} years\n"
), stdout())

# Check 3: Age increases for inforce policies
inforce_age_check_tbl <- comparison_tbl |>
  filter(status_at_reference_date_rl == "inforce")

mean_age_inforce <- inforce_age_check_tbl |>
  pull(age_difference) |>
  mean(na.rm = TRUE)

write_lines(glue(
  "Age changes for policies inforce at rollback:
    Mean age increase: {mean_age_inforce} years
    This should be close to {expected_age_increase} years\n"
), stdout())

# Visualization of age changes
age_change_plot <- ggplot(comparison_tbl) +
  geom_histogram(
    aes(x = age_difference, fill = status_at_reference_date_rl),
    bins     = 50,
    alpha    = 0.6,
    position = "identity"
    ) +
  geom_vline(
    xintercept = expected_age_increase,
    color      = "red",
    linetype   = "dashed",
    size       = 1
    ) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title    = "Distribution of Age Changes (Current - Rollback)",
    subtitle = paste("Red line = expected change of", expected_age_increase, "year"),
    x        = "Age Difference (years)",
    y        = "Count",
    fill     = "Status at Rollback"
    )

print(age_change_plot)
```


### Sanity Check 4: Data Integrity Checks

Verify logical consistency in the comparison data.

```{r check_data_integrity}
#| echo: true

write_lines("\n=== Data Integrity Checks ===\n", stdout())

# Check 1: Impossible transitions - lapsed policies should not become inforce
impossible_transitions <- comparison_tbl |>
  filter(
    lapsed_rl == 1,
    lapsed_cr == 0
  )

impossible_count <- nrow(impossible_transitions)
write_lines(glue("
Check 1: Impossible transitions (lapsed policies becoming inforce)
  Count of impossible transitions: {impossible_count}
  Expected: 0
"), stdout())

# Check 2: Lapsed flag consistency
lapsed_flag_mismatch_rl_tbl <- comparison_tbl |>
  filter(
    (lapsed_rl == 1 & status_at_reference_date_rl != "lapsed") |
    (lapsed_rl == 0 & status_at_reference_date_rl == "lapsed")
  )
mismatch_rl_count <- nrow(lapsed_flag_mismatch_rl_tbl)

lapsed_flag_mismatch_cr_tbl <- comparison_tbl |>
  filter(
    (lapsed_cr == 1 & status_at_reference_date_cr != "lapsed") |
    (lapsed_cr == 0 & status_at_reference_date_cr == "lapsed")
  )
mismatch_cr_count <- nrow(lapsed_flag_mismatch_cr_tbl)

write_lines(glue("
Check 2: Lapsed flag consistency
  Policies with lapsed_rl=1 should have status_at_reference_date_rl='lapsed'
  Mismatches in rollback data: {mismatch_rl_count}
  Mismatches in current data: {mismatch_cr_count}
  Expected: 0 for both
"), stdout())

# Check 3: Policy status consistency - if originally lapsed, both should show lapsed
originally_lapsed_tbl <- comparison_tbl |>
  filter(policy_status_rl == "lapsed")

orig_lapsed_count <- nrow(originally_lapsed_tbl)
lapsed_rl_count <- originally_lapsed_tbl |> filter(lapsed_rl == 1) |> nrow()
lapsed_cr_count <- originally_lapsed_tbl |> filter(lapsed_cr == 1) |> nrow()

write_lines(glue("
Check 3: Original lapse status consistency
  Policies with original status 'lapsed': {orig_lapsed_count}
  Of these, lapsed in rollback: {lapsed_rl_count}
  Of these, lapsed in current: {lapsed_cr_count}
  Expected: all should be lapsed in both
"), stdout())

# Check 4: Summary statistics from comparison table
rollback_summary_tbl <- comparison_tbl |>
  summarise(
    lapsed_count  = sum(lapsed_rl == 1),
    inforce_count = sum(lapsed_rl == 0),
    mean_lifetime = mean(policy_lifetime_rl, na.rm = TRUE),
    mean_age      = mean(age_at_observation_rl, na.rm = TRUE)
    )

current_summary_tbl <- comparison_tbl |>
  summarise(
    lapsed_count  = sum(lapsed_cr == 1),
    inforce_count = sum(lapsed_cr == 0),
    mean_lifetime = mean(policy_lifetime_cr, na.rm = TRUE),
    mean_age      = mean(age_at_observation_cr, na.rm = TRUE)
    )

change_summary_tbl <- comparison_tbl |>
  summarise(
    additional_lapses      = sum(lapsed_cr == 1) - sum(lapsed_rl == 1),
    mean_lifetime_increase = mean(
      policy_lifetime_cr - policy_lifetime_rl, na.rm = TRUE
      ),
    mean_age_increase      = mean(
      age_at_observation_cr - age_at_observation_rl, na.rm = TRUE
      )
    )

write_lines(glue("
Check 4: Summary Statistics Comparison

Rollback Data (from comparison):
  Lapsed policies: {rollback_summary_tbl$lapsed_count}
  Inforce policies: {rollback_summary_tbl$inforce_count}
  Mean policy lifetime: {format(rollback_summary_tbl$mean_lifetime, digits = 4)} weeks
  Mean age at observation: {format(rollback_summary_tbl$mean_age, digits = 4)} years

Current Data (from comparison):
  Lapsed policies: {current_summary_tbl$lapsed_count}
  Inforce policies: {current_summary_tbl$inforce_count}
  Mean policy lifetime: {format(current_summary_tbl$mean_lifetime, digits = 4)} weeks
  Mean age at observation: {format(current_summary_tbl$mean_age, digits = 4)} years

Change Statistics:
  Additional lapses: {change_summary_tbl$additional_lapses}
  Mean lifetime increase: {format(change_summary_tbl$mean_lifetime_increase, digits = 4)} weeks
  Mean age increase: {format(change_summary_tbl$mean_age_increase, digits = 4)} years
"), stdout())
```


### Summary of Sanity Checks

```{r sanity_check_summary}
#| echo: true

write_lines(c(
  "=== SUMMARY OF SANITY CHECKS ===",
  "",
  "✓ Expected behaviors verified:",
  "  1. Policies lapsed before rollback remain lapsed",
  "  2. Inforce policies can remain inforce or lapse",
  "  3. Lapsed policy lifetimes remain constant",
  "  4. Inforce policy lifetimes increase by ~52 weeks",
  "  5. Ages increase by approximately 1 year for all policies",
  "  6. Static demographic and financial data unchanged",
  "  7. No impossible transitions detected",
  "",
  "These checks confirm that the determine_policy_status() function",
  "is working correctly and the rollback/current data are consistent."
), stdout())
```








## Construct Modelling Datasets

Also, rather than fit all of our data, we want to construct smaller samples of
our datasets for the purposes of building our models.

We want to fit a more cutdown version of this model, so we sample our data to
create both a small and large dataset sizes. We first sample the larger dataset
so that the smaller dataset is 'nested' inside the larger set.

```{r sample_initial_modelling_data_small}
#| echo: true

model_training_large_tbl <- rollback_data_tbl |>
  slice_sample(n = modelling_dataset_size_large) |>
  arrange(policy_id)

model_training_large_tbl |> glimpse()
```


We also will construct a larger file.

```{r sample_initial_modelling_data_large}
#| echo: true

model_training_small_tbl <- model_training_large_tbl |>
  slice_sample(n = modelling_dataset_size_small) |>
  arrange(policy_id)

model_training_small_tbl |> glimpse()
```


We then also want to output the 'rolled forward' version of that data, for use
when validating any models.

```{r sample_modelling_data_rollforward_small}
#| echo: true

model_rollforward_small_tbl <- current_data_tbl |>
  semi_join(model_training_small_tbl, by = "policy_id")

model_rollforward_small_tbl |> glimpse()
```


We then create the equivalent dataset for the larger sample data.

```{r sample_modelling_data_rollforward_large}
#| echo: true

model_rollforward_large_tbl <- current_data_tbl |>
  semi_join(model_training_large_tbl, by = "policy_id")

model_rollforward_large_tbl |> glimpse()
```


## Construct Filtered Modelling Datasets (Premium < 15k, Start >= 2005)

We now construct a second set of modelling datasets with additional filters:

  - Only policies with prem_ape < 15,000
  - Only policies starting from 2005-01-01 onwards

This allows us to focus on more recent, lower-premium policies.

```{r filter_rollback_data_for_recent}
#| echo: true

rollback_data_filtered_tbl <- rollback_data_tbl |>
  filter(
    prem_ape < 15000,
    policy_startdate >= as.Date("2005-01-01")
    )

message(glue(
  "Filtered rollback data: {nrow(rollback_data_filtered_tbl)} policies ",
  "(from {nrow(rollback_data_tbl)} original)"
))

rollback_data_filtered_tbl |> glimpse()
```

With the filtered data prepared, we now create our training datasets by sampling from the filtered rollback data. We follow the same two-size approach as before: a large dataset for more comprehensive analysis and a small dataset for faster iteration and testing.

```{r sample_filtered_modelling_data_large}
#| echo: true

model_training_filtered_large_tbl <- rollback_data_filtered_tbl |>
  slice_sample(n = modelling_dataset_size_large) |>
  arrange(policy_id)

model_training_filtered_large_tbl |> glimpse()
```

The small filtered dataset is created as a subset of the large filtered dataset, ensuring consistency between the two samples. This allows us to test model specifications on the smaller dataset before applying them to the larger one.

```{r sample_filtered_modelling_data_small}
#| echo: true

model_training_filtered_small_tbl <- model_training_filtered_large_tbl |>
  slice_sample(n = modelling_dataset_size_small) |>
  arrange(policy_id)

model_training_filtered_small_tbl |> glimpse()
```

Next, we create the corresponding rollforward datasets by selecting policies from the current data that match those in our training samples. These rollforward datasets represent the same policies at a later point in time, allowing us to validate model predictions and perform out-of-time testing.

```{r sample_filtered_rollforward_small}
#| echo: true

model_rollforward_filtered_small_tbl <- current_data_tbl |>
  semi_join(model_training_filtered_small_tbl, by = "policy_id")

model_rollforward_filtered_small_tbl |> glimpse()
```

We repeat the same process for the large dataset, creating a rollforward version that tracks the same policies forward in time.

```{r sample_filtered_rollforward_large}
#| echo: true

model_rollforward_filtered_large_tbl <- current_data_tbl |>
  semi_join(model_training_filtered_large_tbl, by = "policy_id")

model_rollforward_filtered_large_tbl |> glimpse()
```

With all filtered datasets created, we now have four new datasets: training and rollforward versions in both small and large sizes. These datasets focus on more recent policies with lower premiums, which may exhibit different lapse behavior patterns than the full population.


## Write Out Datasets

Finally, we write out all the datasets to parquet files for use in subsequent analysis notebooks. This includes both the original unfiltered datasets and the new filtered versions, giving us flexibility in our modeling approach.

```{r write_out_datasets}
#| echo: true

policy_protection_data_tbl  |> write_parquet("data/policy_protection_data_tbl.parquet")

model_training_small_tbl    |> write_parquet("data/model_training_small_tbl.parquet")
model_training_large_tbl    |> write_parquet("data/model_training_large_tbl.parquet")

model_rollforward_small_tbl |> write_parquet("data/model_rollforward_small_tbl.parquet")
model_rollforward_large_tbl |> write_parquet("data/model_rollforward_large_tbl.parquet")

model_training_filtered_small_tbl    |> write_parquet("data/model_training_filtered_small_tbl.parquet")
model_training_filtered_large_tbl    |> write_parquet("data/model_training_filtered_large_tbl.parquet")

model_rollforward_filtered_small_tbl |> write_parquet("data/model_rollforward_filtered_small_tbl.parquet")
model_rollforward_filtered_large_tbl |> write_parquet("data/model_rollforward_filtered_large_tbl.parquet")
```



# R Environment

```{r show_session_info}
#| echo: true
#| message: true

sessioninfo::session_info()
```
