---
title: "Life Insurance Policy Lapse Survival Analysis"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    theme:
      light: sandstone
      dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---


```{r import_libraries}
#| echo: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(glue)
library(survival)
library(survminer)
library(broom)
library(arrow)
library(muhaz)

source("lib_utils.R")
source("lib_survival_modelling.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2")
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallelly::availableCores()
  )

theme_set(theme_cowplot())

set.seed(42)

# Enable chunk timing for all chunks
knitr::opts_chunk$set(time_it = TRUE)

# Create a list to store chunk times
chunk_times <- list()

# Create a hook to capture chunk execution times
knitr::knit_hooks$set(time_it = function(before, options) {
  if (before) {
    # Record start time in global environment
    assign(".chunk_start_time", Sys.time(), envir = .GlobalEnv)
  } else {
    # Calculate elapsed time
    elapsed <- difftime(
      Sys.time(),
      get(".chunk_start_time", envir = .GlobalEnv),
      units = "secs"
      ) |>
      as.numeric()
    
    # Store in global list
    chunk_times <- get("chunk_times", envir = .GlobalEnv)
    chunk_times[[options$label]] <- elapsed
    assign("chunk_times", chunk_times, envir = .GlobalEnv)
    
    # Print timing message
    message(glue("⏱️  Chunk '{options$label}': {round(elapsed, 2)} seconds"))
  }
})
```

```{r setup_constants}
#| echo: true

# Define observation dates for rollback validation
TRAINING_DATE   <- as.Date("2015-01-01")
VALIDATION_DATE <- as.Date("2016-01-01")

write_lines(
  glue(
    "Training date:   {TRAINING_DATE}\n",
    "Validation date: {VALIDATION_DATE}\n",
    "Prediction horizon: 52 weeks (1 year)"
    ),
  stdout()
  )
```


# Introduction

This workbook performs survival analysis on life insurance policy lapse data
using a rollback validation framework. Unlike traditional model validation that
splits data randomly, rollback validation respects the temporal nature of
insurance data by training on historical snapshots and validating on subsequent
periods.

## Objective

Our goal is to build Cox proportional hazards models that predict policy lapse
behavior and quantify premium at risk. Specifically, we will:

  1. Train models on the policy book as it stood on **2015-01-01**
  2. Use those models to predict lapses in the subsequent year
  3. Compare predictions to actual outcomes observed by **2016-01-01**
  4. Assess both discrimination (how well we rank risk) and calibration (how
     accurate our probabilities are)

## Rollback Validation Methodology

Traditional cross-validation randomly splits data, but this can leak future
information into training. For time-dependent data like insurance policies,
we need **rollback validation**:

  - **Training data**: Policy book reconstructed as of 2015-01-01
    - Policies lapsed before 2015 are marked as events
    - Policies inforce as of 2015 are censored at that date
    - Policies not yet incepted are excluded

  - **Validation data**: Policy book as of 2016-01-01
    - Same policies tracked forward one year
    - New lapses between 2015-2016 identified
    - Policies incepted after 2015 excluded (model didn't know about them)

This approach mimics real-world usage: we fit a model with information
available at time $t$, then test how well it predicts outcomes from time $t$
to $t+1$.

## Business Context

Policy lapse (voluntary termination by the policyholder) represents a key risk
for life insurers:

  - Lost premium revenue
  - Disruption to cash flow planning
  - Potential anti-selection if healthy lives lapse more frequently

By identifying drivers of lapse behavior and quantifying premium at risk, we
can inform retention strategies, product design, and capital planning.


# Load Prepared Datasets

We load the pre-prepared modeling datasets from the exploration workbook. All
data manipulation, rollback construction, and filtering has been performed in
`exploration_lifebook_data.qmd`.

## Training Data (2015-01-01)

Load the training dataset, which represents the policy book as of 2015-01-01.

```{r load_training_data}
#| echo: true

training_tbl <- read_parquet("data/model_training_small_tbl.parquet")

write_lines(
  glue(
    "Training data: {nrow(training_tbl)} policies as of {TRAINING_DATE}"
    ),
  stdout()
  )

glimpse(training_tbl)

# Summary of status at training date
training_tbl |>
  count(status_at_reference_date) |>
  mutate(proportion = n / sum(n)) |>
  print()

# Summary of lapse indicator
training_tbl |>
  count(lapsed) |>
  mutate(
    status = if_else(lapsed == 1, "Event (lapsed)", "Censored (inforce/completed)"),
    proportion = n / sum(n)
    ) |>
  select(status, n, proportion) |>
  print()
```


## Validation Data (2016-01-01)

Load the rollforward dataset for the same policies as of 2016-01-01.

```{r load_validation_data}
#| echo: true

validation_tbl <- read_parquet("data/model_rollforward_small_tbl.parquet")

write_lines(
  glue(
    "Validation data: {nrow(validation_tbl)} policies as of {VALIDATION_DATE}"
    ),
  stdout()
  )

glimpse(validation_tbl)

# Summary of status at validation date
validation_tbl |>
  count(status_at_reference_date) |>
  mutate(proportion = n / sum(n)) |>
  print()
```


## Verify Data Alignment

The training and validation datasets contain the same policies at different
points in time.

```{r filter_common_policies}
#| echo: true

common_policies <- training_tbl$policy_id

training_tbl <- training_tbl |>
  filter(policy_id %in% common_policies)

validation_tbl <- validation_tbl |>
  filter(policy_id %in% common_policies)

write_lines(
  glue(
    "Common policies: {length(common_policies)}\n",
    "Training dataset:   {nrow(training_tbl)} policies\n",
    "Validation dataset: {nrow(validation_tbl)} policies"
    ),
  stdout()
  )
```


## Sanity Checks

Validate that rollback logic and status transitions are sensible.

```{r rollback_sanity_checks}
#| echo: true

# Join training and validation data to check status consistency
check_tbl <- training_tbl |>
  inner_join(
    validation_tbl |> select(
      policy_id, status_2016 = status_at_reference_date
      ),
    by = "policy_id",
    relationship = "one-to-one"
    )

write_lines("Status transitions (2015 -> 2016):", stdout())
transition_tbl <- check_tbl |>
  count(status_at_reference_date, status_2016) |>
  pivot_wider(
    names_from = status_2016,
    values_from = n,
    values_fill = 0
    )

print(transition_tbl)

# Validate that transitions are sensible (note: no status should regress)
# Expected valid transitions:
#   - inforce (2015) can only go to: inforce, lapsed, completed (2016)
#   - lapsed (2015) must stay lapsed (2016) - cannot reverse
#   - completed (2015) must stay completed (2016) - cannot reverse
invalid_transitions <- check_tbl |>
  filter(
    # Lapsed in 2015 but not lapsed in 2016
    (status_at_reference_date == "lapsed" & status_2016 != "lapsed") |
    # Completed in 2015 but not completed in 2016
    (status_at_reference_date == "completed" & status_2016 != "completed")
    )

if (nrow(invalid_transitions) > 0) {
  write_lines(
    glue(
      "\n⚠ WARNING: {nrow(invalid_transitions)} policies have impossible status transitions:",
      "\n  These may indicate data quality issues in the source data."
      ),
    stdout()
    )
  print(invalid_transitions |> select(policy_id, status_at_reference_date, status_2016))
} else {
  write_lines("\n✓ All status transitions are valid", stdout())
}

# Check 2: Policy lifetime should increase for inforce policies
lifetime_check_tbl <- training_tbl |>
  filter(status_at_reference_date == "inforce") |>
  inner_join(
    validation_tbl |> select(
      policy_id, lifetime_2016 = policy_lifetime
      ),
    by = "policy_id",
    relationship = "one-to-one"
    ) |>
  mutate(lifetime_diff = lifetime_2016 - policy_lifetime)

write_lines(
  glue(
    "\nLifetime change for inforce policies:\n",
    "  Mean:   {sprintf('%.1f', mean(lifetime_check_tbl$lifetime_diff))} weeks\n",
    "  Median: {sprintf('%.1f', median(lifetime_check_tbl$lifetime_diff))} weeks\n",
    "  Expected: ~52 weeks (1 year)"
    ),
  stdout()
  )
```


# Exploratory Data Analysis

Before modeling, we explore the training data to understand lapse patterns and
identify potential predictors.

## Summary Statistics by Status

```{r summary_by_status}
#| echo: true

training_tbl |>
  group_by(status_at_reference_date) |>
  summarise(
    count              = n(),
    mean_lifetime      = mean(policy_lifetime, na.rm = TRUE),
    median_lifetime    = median(policy_lifetime, na.rm = TRUE),
    mean_sum_assured   = mean(sum_assured, na.rm = TRUE),
    mean_prem_ape      = mean(prem_ape, na.rm = TRUE),
    mean_prem_sa_ratio = mean(prem_sa_ratio, na.rm = TRUE),
    .groups = "drop"
    ) |>
  print()
```


## Kaplan-Meier Survival Curves

### Overall Survival

```{r km_overall}
#| echo: true

# Fit KM estimator for entire population
km_overall <- survfit(
  Surv(policy_lifetime, lapsed) ~ 1,
  data = training_tbl,
  conf.type = "log-log"
  )

# Plot using base survminer (avoids ggplot2 ribbon issue)
plot(
  km_overall,
  conf.int = TRUE,
  xlab = "Policy Lifetime (weeks)",
  ylab = "Survival Probability",
  main = "Overall Survival Curve (Training Data)"
  )
```


### Survival by Gender

```{r km_gender}
#| echo: true

km_gender <- survfit(
  Surv(policy_lifetime, lapsed) ~ gender_life1,
  data = training_tbl,
  conf.type = "log-log"
  )

# Plot using base graphics
plot(
  km_gender,
  conf.int = TRUE,
  col = c("blue", "red"),
  xlab = "Policy Lifetime (weeks)",
  ylab = "Survival Probability",
  main = "Survival Curves by Gender"
  )
legend(
  "topright",
  legend = names(km_gender$strata),
  col = c("blue", "red"),
  lty = 1,
  bty = "n"
  )

# Log-rank test
survdiff_gender <- survdiff(
  Surv(policy_lifetime, lapsed) ~ gender_life1,
  data = training_tbl
  )
print(survdiff_gender)
```


### Survival by Smoker Status

```{r km_smoker}
#| echo: true

km_smoker <- survfit(
  Surv(policy_lifetime, lapsed) ~ smoker_life1,
  data = training_tbl,
  conf.type = "log-log"
  )

# Plot using base graphics
plot(
  km_smoker,
  conf.int = TRUE,
  col = c("darkgreen", "orange"),
  xlab = "Policy Lifetime (weeks)",
  ylab = "Survival Probability",
  main = "Survival Curves by Smoker Status"
  )
legend(
  "topright",
  legend = names(km_smoker$strata),
  col = c("darkgreen", "orange"),
  lty = 1,
  bty = "n"
  )

# Log-rank test
survdiff_smoker <- survdiff(
  Surv(policy_lifetime, lapsed) ~ smoker_life1,
  data = training_tbl
  )
print(survdiff_smoker)
```


## Distribution of Key Variables

```{r plot_key_distributions}
#| echo: true

# Policy lifetime distribution
lifetime_plot <- ggplot(training_tbl) +
  geom_histogram(
    aes(x = policy_lifetime, fill = status_at_reference_date),
    bins = 50,
    alpha = 0.6,
    position = "identity"
    ) +
  scale_x_continuous(labels = label_comma()) +
  scale_fill_brewer(palette = "Set1") +
  labs(
    title = "Distribution of Policy Lifetime",
    x = "Policy Lifetime (weeks)",
    y = "Count",
    fill = "Status"
    )

# Premium APE distribution
prem_plot <- ggplot(training_tbl |> filter(prem_ape > 0, prem_ape < 15000)) +
  geom_histogram(
    aes(x = prem_ape, fill = factor(lapsed)),
    bins = 50,
    alpha = 0.6,
    position = "identity"
    ) +
  scale_x_log10(labels = label_comma()) +
  scale_fill_manual(
    values = c("0" = "steelblue", "1" = "firebrick"),
    labels = c("0" = "Censored", "1" = "Lapsed")
    ) +
  labs(
    title = "Premium APE Distribution (Log Scale)",
    x = "Premium APE",
    y = "Count",
    fill = "Event"
    )

# Sum assured distribution
sa_plot <- ggplot(training_tbl |> filter(sum_assured > 0)) +
  geom_histogram(
    aes(x = sum_assured, fill = factor(lapsed)),
    bins = 50,
    alpha = 0.6,
    position = "identity"
    ) +
  scale_x_log10(labels = label_comma()) +
  scale_fill_manual(
    values = c("0" = "steelblue", "1" = "firebrick"),
    labels = c("0" = "Censored", "1" = "Lapsed")
    ) +
  labs(
    title = "Sum Assured Distribution (Log Scale)",
    x = "Sum Assured",
    y = "Count",
    fill = "Event"
    )

plot_grid(lifetime_plot, prem_plot, sa_plot, ncol = 2, nrow = 2)
```


## Lapse Rates by Demographics

```{r lapse_rates_demographics}
#| echo: true

training_tbl |>
  filter(!is.na(gender_life1), !is.na(smoker_life1)) |>
  group_by(gender_life1, smoker_life1) |>
  summarise(
    total      = n(),
    lapsed     = sum(lapsed),
    lapse_rate = lapsed / total,
    .groups    = "drop"
    ) |>
  mutate(
    group = paste(gender_life1, smoker_life1, sep = " / ")
    ) |>
  ggplot() +
  geom_bar(
    aes(x = group, y = lapse_rate, fill = group),
    stat = "identity",
    alpha = 0.7
    ) +
  geom_text(
    aes(x = group, y = lapse_rate, label = sprintf("%.1f%%", lapse_rate * 100)),
    vjust = -0.5
    ) +
  scale_y_continuous(
    labels = label_percent(),
    limits = c(0, NA)
    ) +
  labs(
    title = "Lapse Rate by Gender and Smoker Status",
    x = "Demographic Group",
    y = "Lapse Rate"
    ) +
  theme(legend.position = "none")
```


# Residual-Based Variable Selection

Before building complex models, we use martingale residuals from a null model
to identify promising continuous predictors.

## Fit Null Model

```{r fit_null_model}
#| echo: true

lapse_model00_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~ 1,
  data = training_tbl
  )

lapse_model00_coxph |> summary()
```


## Extract Martingale Residuals

```{r extract_residuals}
#| echo: true

training_tbl <- training_tbl |>
  mutate(resid_m = residuals(lapse_model00_coxph, type = "martingale"))

write_lines(
  glue(
    "Martingale residuals: mean = {sprintf('%.3f', mean(training_tbl$resid_m))}, ",
    "range = [{sprintf('%.3f', min(training_tbl$resid_m))}, ",
    "{sprintf('%.3f', max(training_tbl$resid_m))}]"
    ),
  stdout()
  )
```


## Plot Residuals vs Continuous Variables

```{r plot_residuals_continuous}
#| echo: true

# Helper function for residual plots
create_residual_plot <- function(data_tbl, var_name, use_log = FALSE) {
  plot_data <- data_tbl |>
    filter(!is.na(.data[[var_name]]), .data[[var_name]] > 0)

  if (use_log) {
    p <- ggplot(plot_data, aes(x = log(.data[[var_name]]), y = resid_m))
    x_label <- glue("log({var_name})")
  } else {
    p <- ggplot(plot_data, aes(x = .data[[var_name]], y = resid_m))
    x_label <- var_name
    }

  p <- p +
    geom_point(size = 0.8, alpha = 0.15) +
    geom_smooth(method = "loess", se = FALSE, colour = "firebrick") +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(
      title = glue("Martingale Residuals vs {var_name}"),
      x = x_label,
      y = "Martingale Residual"
      )

  return(p)
}

# Plot residuals vs key financial variables
prem_resid_plot <- create_residual_plot(training_tbl, "prem_ape", use_log = TRUE)
sa_resid_plot   <- create_residual_plot(training_tbl, "sum_assured", use_log = TRUE)
ratio_resid_plot <- create_residual_plot(training_tbl, "prem_sa_ratio", use_log = FALSE)

plot_grid(prem_resid_plot, sa_resid_plot, ratio_resid_plot, ncol = 2, nrow = 2)
```


## Calculate Correlations

```{r calculate_correlations}
#| echo: true

# Calculate correlations between residuals and continuous variables
variables_vec <- c("prem_ape", "sum_assured", "prem_sa_ratio")

# Build correlation table by mapping over variables
cor_values <- map_dbl(
  variables_vec,
  ~cor(
    training_tbl$resid_m,
    training_tbl[[.x]],
    use = "complete.obs"
    )
  )

log_cor_values <- map_dbl(
  variables_vec,
  ~{
    valid_data <- training_tbl |> filter(.data[[.x]] > 0)
    cor(
      valid_data$resid_m,
      log(valid_data[[.x]]),
      use = "complete.obs"
      )
    }
  )

correlation_tbl <- tibble(
  variable = variables_vec,
  correlation = cor_values,
  log_correlation = log_cor_values
  ) |>
  mutate(abs_correlation = abs(correlation)) |>
  arrange(desc(abs_correlation)) |>
  select(variable, correlation, log_correlation)

print(correlation_tbl)
```

The residual plots and correlations suggest that log transformations of
financial variables may be appropriate, and that `prem_ape` and `sum_assured`
are promising predictors.


# Incremental Model Building

We build a sequence of increasingly complex Cox models, assessing each before
moving to the next.

## Model 00: Null Model (Baseline)

Already fitted above. This establishes the baseline with no predictors.

```{r model00_assessment}
#| echo: true

# Baseline metrics
model00_metrics <- lapse_model00_coxph |>
  glance() |>
  select(n, nevent, logLik, AIC, BIC)

print(model00_metrics)
```


## Model 01: Gender Only

```{r model01_gender}
#| echo: true

lapse_model01_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~ gender_life1,
  data = training_tbl
  )

# Summary
lapse_model01_coxph |> summary()

# Coefficients
lapse_model01_coxph |> tidy() |> print()

# Fit metrics
lapse_model01_coxph |>
  glance() |>
  select(n, nevent, concordance, std.error.concordance, r.squared, logLik, AIC, BIC) |>
  print()

# Check proportional hazards
cox_zph_01 <- lapse_model01_coxph |> cox.zph(transform = "km")
print(cox_zph_01)
```


## Model 02: Demographics (Gender + Smoker)

```{r model02_demographics}
#| echo: true

lapse_model02_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~ gender_life1 + smoker_life1,
  data = training_tbl
  )

# Summary
lapse_model02_coxph |> summary()

# Coefficients with confidence intervals
lapse_model02_coxph |> tidy(conf.int = TRUE) |> print()

# Fit metrics
lapse_model02_coxph |>
  glance() |>
  select(n, nevent, concordance, std.error.concordance, r.squared, logLik, AIC, BIC) |>
  print()

# Check proportional hazards
cox_zph_02 <- lapse_model02_coxph |> cox.zph(transform = "km")
print(cox_zph_02)

# Likelihood ratio test vs Model 01
anova(lapse_model01_coxph, lapse_model02_coxph, test = "Chisq")
```


## Model 03: Add Premium (Linear)

```{r model03_premium_linear}
#| echo: true

lapse_model03_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~ gender_life1 + smoker_life1 + prem_ape,
  data = training_tbl
  )

# Summary
lapse_model03_coxph |> summary()

# Coefficients
lapse_model03_coxph |> tidy(conf.int = TRUE) |> print()

# Fit metrics
lapse_model03_coxph |>
  glance() |>
  select(n, nevent, concordance, std.error.concordance, r.squared, logLik, AIC, BIC) |>
  print()

# Check proportional hazards
cox_zph_03 <- lapse_model03_coxph |> cox.zph(transform = "km")
print(cox_zph_03)

# LRT vs Model 02
anova(lapse_model02_coxph, lapse_model03_coxph, test = "Chisq")
```


## Model 04: Premium with Log Transformation

```{r model04_log_premium}
#| echo: true

# Filter to positive premiums for log transformation
training_pos_prem_tbl <- training_tbl |>
  filter(prem_ape > 0, sum_assured > 0)

lapse_model04_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~ gender_life1 + smoker_life1 + log(prem_ape),
  data = training_pos_prem_tbl
  )

# Summary
lapse_model04_coxph |> summary()

# Coefficients
lapse_model04_coxph |> tidy(conf.int = TRUE) |> print()

# Fit metrics
lapse_model04_coxph |>
  glance() |>
  select(n, nevent, concordance, std.error.concordance, r.squared, logLik, AIC, BIC) |>
  print()

# Check proportional hazards
cox_zph_04 <- lapse_model04_coxph |> cox.zph(transform = "km")
print(cox_zph_04)
```


## Model 05: Add Sum Assured (Log)

```{r model05_add_sum_assured}
#| echo: true

lapse_model05_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~ gender_life1 + smoker_life1 +
    log(prem_ape) + log(sum_assured),
  data = training_pos_prem_tbl
  )

# Summary
lapse_model05_coxph |> summary()

# Coefficients
lapse_model05_coxph |> tidy(conf.int = TRUE) |> print()

# Fit metrics
lapse_model05_coxph |>
  glance() |>
  select(n, nevent, concordance, std.error.concordance, r.squared, logLik, AIC, BIC) |>
  print()

# Check proportional hazards
cox_zph_05 <- lapse_model05_coxph |> cox.zph(transform = "km")
print(cox_zph_05)

# LRT vs Model 04
anova(lapse_model04_coxph, lapse_model05_coxph, test = "Chisq")
```


## Model 06: Add Premium/SA Ratio with Spline

```{r model06_spline}
#| echo: true

lapse_model06_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~ gender_life1 + smoker_life1 +
    log(prem_ape) + log(sum_assured) + pspline(prem_sa_ratio),
  data = training_pos_prem_tbl
  )

# Summary
lapse_model06_coxph |> summary()

# Fit metrics
lapse_model06_coxph |>
  glance() |>
  select(n, nevent, concordance, std.error.concordance, r.squared, logLik, AIC, BIC) |>
  print()

# Check proportional hazards
cox_zph_06 <- lapse_model06_coxph |> cox.zph(transform = "km")
print(cox_zph_06)

# LRT vs Model 05
anova(lapse_model05_coxph, lapse_model06_coxph, test = "Chisq")

# Visualize spline effect
termplot(
  lapse_model06_coxph,
  term = 5,
  se = TRUE,
  col.term = "firebrick",
  col.se = "grey70",
  main = "Spline Effect of Premium/SA Ratio on Log Hazard"
  )
```


# Model Comparison

We compare all models systematically to select the best specification.

```{r model_comparison_table}
#| echo: true

# Extract metrics for all models
model_list <- list(
  lapse_model00_coxph,
  lapse_model01_coxph,
  lapse_model02_coxph,
  lapse_model03_coxph,
  lapse_model04_coxph,
  lapse_model05_coxph,
  lapse_model06_coxph
  )

model_comparison_tbl <- tibble(
  model = c("00", "01", "02", "03", "04", "05", "06"),
  description = c(
    "Null (no predictors)",
    "Gender only",
    "Gender + Smoker",
    "Demographics + prem_ape",
    "Demographics + log(prem_ape)",
    "Demographics + log(prem_ape) + log(sum_assured)",
    "Demographics + log(prem_ape) + log(sum_assured) + pspline(prem_sa_ratio)"
    ),
  n = map_dbl(model_list, ~glance(.x)$n),
  nevent = map_dbl(model_list, ~glance(.x)$nevent),
  concordance = map_dbl(model_list, ~glance(.x)$concordance),
  std_error = map_dbl(model_list, ~glance(.x)$std.error.concordance),
  r_squared = map_dbl(model_list, ~glance(.x)$r.squared),
  logLik = map_dbl(model_list, ~glance(.x)$logLik),
  AIC = map_dbl(model_list, ~glance(.x)$AIC),
  BIC = map_dbl(model_list, ~glance(.x)$BIC)
  )

print(model_comparison_tbl)

# Identify best model by AIC
best_model_idx <- which.min(model_comparison_tbl$AIC)
best_model_name <- model_comparison_tbl$model[best_model_idx]

write_lines(
  glue("Best model by AIC: Model {best_model_name}"),
  stdout()
  )
```


## Forest Plot for Best Model

```{r forest_plot}
#| echo: true

# Select best model (Model 06)
best_model <- lapse_model06_coxph

# Create forest plot
ggforest(best_model, data = as.data.frame(training_pos_prem_tbl))
```


# Validation and Cashflow Analysis

Now we test our best model on the validation data (2016-01-01) to assess
out-of-time performance.

## Generate Predictions

```{r generate_predictions}
#| echo: true

# Filter validation data to match training data
validation_pos_prem_tbl <- validation_tbl |>
  filter(
    policy_id %in% training_pos_prem_tbl$policy_id,
    prem_ape > 0,
    sum_assured > 0
    )

# Generate predictions for 52-week horizon
predictions <- calculate_cashflow_predictions(
  model_coxph = best_model,
  training_data_tbl = training_pos_prem_tbl,
  prediction_horizon_weeks = 52
  )

# Display aggregate summary
print(predictions$aggregate_summary)
```


## Compare to Actual Outcomes

```{r compare_to_actual}
#| echo: true

comparison <- compare_predicted_vs_actual(
  predictions_list = predictions,
  actual_data_tbl = validation_pos_prem_tbl
  )

# Summary metrics
print(comparison$summary_metrics)

# Calibration by decile
print(comparison$calibration_tbl)
```


## Visualize Results

```{r visualize_cashflow_results}
#| echo: true

# Generate all comparison plots
cashflow_plots <- create_cashflow_comparison_plots(comparison)

# Display plots
plot_grid(
  cashflow_plots$lapse_comparison_plot,
  cashflow_plots$premium_comparison_plot,
  cashflow_plots$calibration_plot,
  cashflow_plots$residual_plot,
  ncol = 2,
  nrow = 2
  )
```


# Results Interpretation

Based on the validation results:

```{r interpret_results}
#| echo: false

metrics <- comparison$summary_metrics

write_lines(
  glue(
    "\n=== MODEL PERFORMANCE SUMMARY ===\n\n",
    "Lapse Prediction:\n",
    "  Predicted: {sprintf('%.0f', metrics$predicted_lapses)} policies\n",
    "  Actual:    {sprintf('%.0f', metrics$actual_lapses)} policies\n",
    "  Error:     {sprintf('%.1f%%', metrics$lapse_error_pct)}\n\n",
    "Premium at Risk:\n",
    "  Predicted: £{sprintf('%.0f', metrics$predicted_premium_loss)}\n",
    "  Actual:    £{sprintf('%.0f', metrics$actual_premium_loss)}\n",
    "  Error:     {sprintf('%.1f%%', metrics$premium_error_pct)}\n\n",
    "Interpretation:\n"
    ),
  stdout()
  )

# Assess prediction accuracy
if (abs(metrics$lapse_error_pct) < 10) {
  write_lines("  ✓ Lapse predictions are well-calibrated (< 10% error)", stdout())
} else if (abs(metrics$lapse_error_pct) < 20) {
  write_lines("  ~ Lapse predictions are acceptable (10-20% error)", stdout())
} else {
  write_lines("  ✗ Lapse predictions show significant error (> 20%)", stdout())
}

if (abs(metrics$premium_error_pct) < 10) {
  write_lines("  ✓ Premium predictions are well-calibrated (< 10% error)", stdout())
} else if (abs(metrics$premium_error_pct) < 20) {
  write_lines("  ~ Premium predictions are acceptable (10-20% error)", stdout())
} else {
  write_lines("  ✗ Premium predictions show significant error (> 20%)", stdout())
}
```

**Key Findings:**

1. **Discrimination (C-index)**: Model 06 achieved a concordance of
   `r sprintf("%.3f", glance(best_model)$concordance)`, indicating
   `r if(glance(best_model)$concordance > 0.65) "good" else if(glance(best_model)$concordance > 0.60) "moderate" else "weak"`
   ability to rank risk.

2. **Calibration**: The calibration plot shows how well predicted probabilities
   match observed lapse rates across risk deciles. Points near the diagonal
   indicate good calibration.

3. **Premium at Risk**: The model predicts
   £`r sprintf("%.0f", metrics$predicted_premium_loss)` in annual premium at
   risk from lapses, compared to actual losses of
   £`r sprintf("%.0f", metrics$actual_premium_loss)`.

4. **Variable Effects**: Key drivers of lapse risk include:
   - Demographics (gender, smoker status)
   - Financial variables (premium level, sum assured)
   - Pricing efficiency (premium/sum assured ratio)


# Conclusions

This analysis demonstrates the application of Cox proportional hazards models
to life insurance lapse prediction with rollback validation. The approach
provides:

  - **Interpretable risk factors**: Clear hazard ratios for each covariate
  - **Individual risk scores**: Policy-level lapse probabilities
  - **Aggregate cashflow forecasts**: Total expected lapses and premium at risk
  - **Validation metrics**: Out-of-time performance assessment

## Business Recommendations

Based on these findings, insurers can:

  1. **Targeted retention**: Focus efforts on high-risk policies identified by
     the model
  1. **Pricing adjustments**: Revisit pricing for segments with high lapse rates
  1. **Product design**: Address structural factors (e.g., premium/SA ratio)
     that drive lapses
  1. **Capital planning**: Use calibrated cashflow forecasts for reserve
     calculations

## Model Limitations

  - Assumes proportional hazards (validated via cox.zph)
  - Does not capture time-varying covariates
  - Limited to observable policy characteristics
  - Retrospective validation; future lapse behavior may differ

## Future Work

  - Incorporate time-varying covariates (e.g., payment history)
  - Test stratified models by product type or vintage
  - Develop automated monitoring for model drift
  - Extend to multi-state models (lapse vs other exit modes)


# R Environment

```{r show_session_info}
#| echo: true
#| message: true

sessioninfo::session_info()
```
