---
title: "Life Insurance Lapse Survival Analysis with Rollback Validation"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    theme:
      light: sandstone
      dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---


```{r import_libraries}
#| echo: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(glue)
library(survival)
library(survminer)
library(broom)
library(arrow)
library(muhaz)
library(lubridate)

source("lib_utils.R")
source("lib_survival_modelling.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "lubridate")
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallelly::availableCores()
  )

theme_set(theme_cowplot())

set.seed(42)

# Enable chunk timing for all chunks
knitr::opts_chunk$set(time_it = TRUE)

# Create a list to store chunk times
chunk_times <- list()

# Create a hook to capture chunk execution times
knitr::knit_hooks$set(time_it = function(before, options) {
  if (before) {
    # Record start time in global environment
    assign(".chunk_start_time", Sys.time(), envir = .GlobalEnv)
  } else {
    # Calculate elapsed time
    elapsed <- difftime(
      Sys.time(),
      get(".chunk_start_time", envir = .GlobalEnv),
      units = "secs"
      ) |>
      as.numeric()
    
    # Store in global list
    chunk_times <- get("chunk_times", envir = .GlobalEnv)
    chunk_times[[options$label]] <- elapsed
    assign("chunk_times", chunk_times, envir = .GlobalEnv)
    
    # Print timing message
    message(glue("⏱️  Chunk '{options$label}': {round(elapsed, 2)} seconds"))
  }
})
```

```{r setup_constants}
#| echo: true

TRAINING_DATE   <- as.Date("2015-01-01")
VALIDATION_DATE <- as.Date("2016-01-01")

write_lines(
  glue(
    "Training date:   {TRAINING_DATE}\n",
    "Validation date: {VALIDATION_DATE}\n",
    "Prediction horizon: 52 weeks (1 year)"
    ),
  stdout()
  )
```


# Introduction

This worksheet demonstrates survival analysis applied to life insurance policy
lapse modeling using Cox proportional hazards regression with rollback
validation.

## Objectives

Build Cox PH models to predict policy lapse behavior:

  1. Train models on historical policy book snapshot (2015-01-01)
  2. Predict lapses for the subsequent 12-month period
  3. Validate predictions against actual outcomes (2016-01-01)
  4. Quantify premium at risk and assess model calibration

## Rollback Validation Methodology

Unlike random cross-validation, rollback validation respects temporal structure:

  - **Training**: Reconstruct policy book as of 2015-01-01
    - Lapses before 2015: events
    - Policies inforce at 2015: censored at that date
    - Policies not yet incepted: excluded

  - **Validation**: Same policies tracked forward to 2016-01-01
    - New lapses between 2015-2016 identified
    - Policies incepted after 2015: excluded (model had no info)

This mimics real-world deployment: fit model at time $t$, predict outcomes
from $t$ to $t+1$.

## Business Context

Policy lapse (voluntary termination) represents key risks:

  - Lost premium revenue
  - Disrupted cash flow planning
  - Potential anti-selection if healthier lives lapse more

Survival models help identify lapse drivers, quantify premium at risk, and
inform retention strategies.


# Load Prepared Datasets

Load pre-prepared modeling datasets from the exploration workbook. All data
manipulation, rollback construction, and filtering performed in
`exploration_lifebook_data.qmd`.

## Training Data (2015-01-01)

Load training dataset representing the policy book as of 2015-01-01.

```{r load_training}
#| echo: true

training_tbl <- read_parquet("data/model_training_small_tbl.parquet")

write_lines(
  glue("Training: {comma(nrow(training_tbl))} policies as of {TRAINING_DATE}"),
  stdout()
  )

glimpse(training_tbl)

training_tbl |>
  count(status_at_reference_date) |>
  mutate(proportion = n / sum(n)) |>
  print()

training_tbl |>
  count(lapsed) |>
  mutate(
    status = if_else(lapsed == 1, "Event (lapsed)", "Censored (inforce/completed)"),
    proportion = n / sum(n)
    ) |>
  select(status, n, proportion) |>
  print()
```


## Validation Data (2016-01-01)

Load rollforward dataset for the same policies as of 2016-01-01.

```{r load_validation}
#| echo: true

validation_tbl <- read_parquet("data/model_rollforward_small_tbl.parquet")

write_lines(
  glue("Validation: {comma(nrow(validation_tbl))} policies as of {VALIDATION_DATE}"),
  stdout()
  )

glimpse(validation_tbl)

validation_tbl |>
  count(status_at_reference_date) |>
  mutate(proportion = n / sum(n)) |>
  print()
```


## Verify Data Alignment

The training and validation datasets contain the same policies at different
time points.

```{r filter_common}
#| echo: true

common_policy_ids <- training_tbl$policy_id

training_tbl <- training_tbl |>
  filter(policy_id %in% common_policy_ids)

validation_tbl <- validation_tbl |>
  filter(policy_id %in% common_policy_ids)

write_lines(
  glue(
    "Common policies: {comma(length(common_policy_ids))}\n",
    "Training:   {comma(nrow(training_tbl))}\n",
    "Validation: {comma(nrow(validation_tbl))}"
    ),
  stdout()
  )
```


## Create Derived Features

Create age bands and other categorical predictors.

```{r create_derived_features}
#| echo: true

training_tbl <- training_tbl |>
  mutate(
    age_band = cut(
      age_at_observation,
      breaks = c(0, 30, 40, 50, 60, 70, Inf),
      labels = c("Under 30", "30-39", "40-49", "50-59", "60-69", "70+"),
      right = FALSE
      ),
    vintage_cohort = cut(
      policy_vintage_year,
      breaks = c(1900, 2000, 2005, 2010, 2015, Inf),
      labels = c("Pre-2000", "2000-2004", "2005-2009", "2010-2014", "2015+"),
      right = FALSE
      )
    )

validation_tbl <- validation_tbl |>
  mutate(
    age_band = cut(
      age_at_observation,
      breaks = c(0, 30, 40, 50, 60, 70, Inf),
      labels = c("Under 30", "30-39", "40-49", "50-59", "60-69", "70+"),
      right = FALSE
      ),
    vintage_cohort = cut(
      policy_vintage_year,
      breaks = c(1900, 2000, 2005, 2010, 2015, Inf),
      labels = c("Pre-2000", "2000-2004", "2005-2009", "2010-2014", "2015+"),
      right = FALSE
      )
    )

write_lines("Derived features created: age_band, vintage_cohort", stdout())
```


## Sanity Checks

Verify rollback logic is working correctly.

```{r sanity_checks}
#| echo: true

check_tbl <- training_tbl |>
  inner_join(
    validation_tbl |> select(policy_id, status_2016 = status_at_reference_date),
    by = "policy_id",
    relationship = "one-to-one"
    )

write_lines("Status transitions (2015 -> 2016):", stdout())
check_tbl |>
  count(status_at_reference_date, status_2016) |>
  pivot_wider(names_from = status_2016, values_from = n, values_fill = 0) |>
  print()

lifetime_check_tbl <- training_tbl |>
  filter(status_at_reference_date == "inforce") |>
  inner_join(
    validation_tbl |> select(policy_id, lifetime_2016 = policy_lifetime),
    by = "policy_id",
    relationship = "one-to-one"
    ) |>
  mutate(lifetime_diff = lifetime_2016 - policy_lifetime)

write_lines(
  glue(
    "\nLifetime change for inforce policies:\n",
    "  Mean:   {sprintf('%.1f', mean(lifetime_check_tbl$lifetime_diff))} weeks\n",
    "  Median: {sprintf('%.1f', median(lifetime_check_tbl$lifetime_diff))} weeks\n",
    "  Expected: ~52 weeks (1 year)"
    ),
  stdout()
  )
```


# Exploratory Data Analysis

Explore training data to understand lapse patterns.

## Summary Statistics

```{r summary_stats}
#| echo: true

training_tbl |>
  group_by(status_at_reference_date) |>
  summarise(
    count = n(),
    mean_lifetime_wks = mean(policy_lifetime, na.rm = TRUE),
    median_lifetime_wks = median(policy_lifetime, na.rm = TRUE),
    mean_sum_assured = mean(sum_assured, na.rm = TRUE),
    mean_prem_ape = mean(prem_ape, na.rm = TRUE),
    mean_age = mean(age_at_observation, na.rm = TRUE),
    .groups = "drop"
    ) |>
  print()
```


## Kaplan-Meier Survival Curves

### Overall Survival

```{r km_overall}
#| echo: true

km_overall <- survfit(
  Surv(policy_lifetime, lapsed) ~ 1,
  data = training_tbl,
  conf.type = "log-log"
  )

plot(
  km_overall,
  conf.int = TRUE,
  xlab = "Policy Lifetime (weeks)",
  ylab = "Survival Probability",
  main = "Overall Survival Curve"
  )
```


### By Gender

```{r km_gender}
#| echo: true

km_gender <- survfit(
  Surv(policy_lifetime, lapsed) ~ gender_life1,
  data = training_tbl,
  conf.type = "log-log"
  )

plot(
  km_gender,
  conf.int = TRUE,
  col = c("blue", "red"),
  xlab = "Policy Lifetime (weeks)",
  ylab = "Survival Probability",
  main = "Survival by Gender"
  )
legend("topright", legend = names(km_gender$strata), col = c("blue", "red"), lty = 1)

survdiff_gender <- survdiff(
  Surv(policy_lifetime, lapsed) ~ gender_life1,
  data = training_tbl
  )
print(survdiff_gender)
```


### By Smoker Status

```{r km_smoker}
#| echo: true

km_smoker <- survfit(
  Surv(policy_lifetime, lapsed) ~ smoker_life1,
  data = training_tbl,
  conf.type = "log-log"
  )

plot(
  km_smoker,
  conf.int = TRUE,
  col = c("darkgreen", "orange"),
  xlab = "Policy Lifetime (weeks)",
  ylab = "Survival Probability",
  main = "Survival by Smoker Status"
  )
legend("topright", legend = names(km_smoker$strata), col = c("darkgreen", "orange"), lty = 1)

survdiff_smoker <- survdiff(
  Surv(policy_lifetime, lapsed) ~ smoker_life1,
  data = training_tbl
  )
print(survdiff_smoker)
```


### By Cluster

```{r km_cluster}
#| echo: true

km_cluster <- survfit(
  Surv(policy_lifetime, lapsed) ~ cluster_id,
  data = training_tbl,
  conf.type = "log-log"
  )

plot(
  km_cluster,
  conf.int = FALSE,
  xlab = "Policy Lifetime (weeks)",
  ylab = "Survival Probability",
  main = "Survival by Cluster"
  )
legend("topright", legend = names(km_cluster$strata), lty = 1, cex = 0.7)

survdiff_cluster <- survdiff(
  Surv(policy_lifetime, lapsed) ~ cluster_id,
  data = training_tbl
  )
print(survdiff_cluster)
```


### By Age Band

```{r km_age}
#| echo: true

km_age <- survfit(
  Surv(policy_lifetime, lapsed) ~ age_band,
  data = training_tbl,
  conf.type = "log-log"
  )

plot(
  km_age,
  conf.int = FALSE,
  xlab = "Policy Lifetime (weeks)",
  ylab = "Survival Probability",
  main = "Survival by Age Band"
  )
legend("topright", legend = names(km_age$strata), lty = 1, cex = 0.8)

survdiff_age <- survdiff(
  Surv(policy_lifetime, lapsed) ~ age_band,
  data = training_tbl
  )
print(survdiff_age)
```


## Distribution Plots

```{r distribution_plots}
#| echo: true

lifetime_plot <- ggplot(training_tbl) +
  geom_histogram(
    aes(x = policy_lifetime, fill = factor(lapsed)),
    bins = 50, alpha = 0.6, position = "identity"
    ) +
  scale_fill_manual(
    values = c("0" = "steelblue", "1" = "firebrick"),
    labels = c("0" = "Censored", "1" = "Lapsed")
    ) +
  scale_x_continuous(labels = comma) +
  labs(
    title = "Policy Lifetime Distribution",
    x = "Weeks", y = "Count", fill = "Status"
    )

prem_plot <- ggplot(
  training_tbl |> filter(prem_ape > 0, prem_ape < 15000)
  ) +
  geom_histogram(
    aes(x = prem_ape, fill = factor(lapsed)),
    bins = 50, alpha = 0.6, position = "identity"
    ) +
  scale_x_log10(labels = comma) +
  scale_fill_manual(
    values = c("0" = "steelblue", "1" = "firebrick"),
    labels = c("0" = "Censored", "1" = "Lapsed")
    ) +
  labs(
    title = "Premium APE (Log Scale)",
    x = "Premium APE", y = "Count", fill = "Status"
    )

sa_plot <- ggplot(training_tbl |> filter(sum_assured > 0)) +
  geom_histogram(
    aes(x = sum_assured, fill = factor(lapsed)),
    bins = 50, alpha = 0.6, position = "identity"
    ) +
  scale_x_log10(labels = comma) +
  scale_fill_manual(
    values = c("0" = "steelblue", "1" = "firebrick"),
    labels = c("0" = "Censored", "1" = "Lapsed")
    ) +
  labs(
    title = "Sum Assured (Log Scale)",
    x = "Sum Assured", y = "Count", fill = "Status"
    )

age_plot <- ggplot(training_tbl) +
  geom_histogram(
    aes(x = age_at_observation, fill = factor(lapsed)),
    bins = 30, alpha = 0.6, position = "identity"
    ) +
  scale_fill_manual(
    values = c("0" = "steelblue", "1" = "firebrick"),
    labels = c("0" = "Censored", "1" = "Lapsed")
    ) +
  labs(
    title = "Age at Observation",
    x = "Age (years)", y = "Count", fill = "Status"
    )

plot_grid(lifetime_plot, prem_plot, sa_plot, age_plot, ncol = 2)
```


## Lapse Rates by Demographics

```{r lapse_rates_demo}
#| echo: true

training_tbl |>
  filter(!is.na(gender_life1), !is.na(smoker_life1)) |>
  group_by(gender_life1, smoker_life1) |>
  summarise(
    total = n(),
    lapsed = sum(lapsed),
    lapse_rate = lapsed / total,
    .groups = "drop"
    ) |>
  mutate(group = paste(gender_life1, smoker_life1, sep = " / ")) |>
  ggplot(aes(x = group, y = lapse_rate, fill = group)) +
  geom_bar(stat = "identity", alpha = 0.7) +
  geom_text(aes(label = sprintf("%.1f%%", lapse_rate * 100)), vjust = -0.5) +
  scale_y_continuous(labels = percent, limits = c(0, NA)) +
  labs(
    title = "Lapse Rate by Gender and Smoker Status",
    x = NULL, y = "Lapse Rate"
    ) +
  theme(legend.position = "none")
```


# Residual-Based Variable Selection

Use martingale residuals from null model to identify promising continuous
predictors.

## Fit Null Model

```{r null_model}
#| echo: true

lapse_model00_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~ 1,
  data = training_tbl
  )

lapse_model00_coxph |> summary()

lapse_model00_coxph |>
  glance() |>
  select(n, nevent, logLik, AIC, BIC) |>
  print()
```


## Extract Martingale Residuals

```{r extract_residuals}
#| echo: true

training_tbl <- training_tbl |>
  mutate(resid_m = residuals(lapse_model00_coxph, type = "martingale"))

write_lines(
  glue(
    "Martingale residuals: mean = {sprintf('%.3f', mean(training_tbl$resid_m))}, ",
    "range = [{sprintf('%.3f', min(training_tbl$resid_m))}, ",
    "{sprintf('%.3f', max(training_tbl$resid_m))}]"
    ),
  stdout()
  )
```


## Residual Plots

```{r residual_plots}
#| echo: true

create_residual_plot <- function(data_tbl, var_name, use_log = FALSE) {
  plot_data <- data_tbl |>
    filter(!is.na(.data[[var_name]]), .data[[var_name]] > 0)

  if (use_log) {
    p <- ggplot(plot_data, aes(x = log(.data[[var_name]]), y = resid_m))
    x_label <- glue("log({var_name})")
    } else {
    p <- ggplot(plot_data, aes(x = .data[[var_name]], y = resid_m))
    x_label <- var_name
    }

  p +
    geom_point(size = 0.8, alpha = 0.15) +
    geom_smooth(method = "loess", se = FALSE, colour = "firebrick") +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    labs(
      title = glue("Martingale Residuals vs {var_name}"),
      x = x_label,
      y = "Martingale Residual"
      )
}

prem_resid <- create_residual_plot(training_tbl, "prem_ape", use_log = TRUE)
sa_resid <- create_residual_plot(training_tbl, "sum_assured", use_log = TRUE)
ratio_resid <- create_residual_plot(training_tbl, "prem_sa_ratio", use_log = FALSE)
age_resid <- create_residual_plot(training_tbl, "age_at_observation", use_log = FALSE)

plot_grid(prem_resid, sa_resid, ratio_resid, age_resid, ncol = 2)
```


## Correlations

```{r correlations}
#| echo: true

continuous_vars <- c("prem_ape", "sum_assured", "prem_sa_ratio", "age_at_observation")

cor_values <- map_dbl(
  continuous_vars,
  ~cor(
    training_tbl$resid_m,
    training_tbl[[.x]],
    use = "complete.obs"
    )
  )

log_cor_values <- map_dbl(
  continuous_vars,
  ~{
    valid_data <- training_tbl |> filter(.data[[.x]] > 0)
    cor(
      valid_data$resid_m,
      log(valid_data[[.x]]),
      use = "complete.obs"
      )
    }
  )

correlation_tbl <- tibble(
  variable = continuous_vars,
  correlation = cor_values,
  log_correlation = log_cor_values
  ) |>
  arrange(desc(abs(correlation)))

print(correlation_tbl)
```


# Incremental Model Building

Build sequence of increasingly complex Cox models, assessing each step.

## Model 00: Null Model

Already fitted above as baseline.

```{r model00_metrics}
#| echo: true

model00_metrics <- lapse_model00_coxph |>
  glance() |>
  select(n, nevent, logLik, AIC, BIC)

print(model00_metrics)
```


## Model 01: Gender

```{r model01}
#| echo: true

lapse_model01_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~ gender_life1,
  data = training_tbl
  )

lapse_model01_coxph |> summary()
lapse_model01_coxph |> tidy(conf.int = TRUE) |> print()

lapse_model01_coxph |>
  glance() |>
  select(n, nevent, concordance, std.error.concordance, r.squared, AIC, BIC) |>
  print()

cox_zph_01 <- lapse_model01_coxph |> cox.zph(transform = "km")
print(cox_zph_01)
```


## Model 02: Demographics (Gender + Smoker)

```{r model02}
#| echo: true

lapse_model02_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~ gender_life1 + smoker_life1,
  data = training_tbl
  )

lapse_model02_coxph |> summary()
lapse_model02_coxph |> tidy(conf.int = TRUE) |> print()

lapse_model02_coxph |>
  glance() |>
  select(n, nevent, concordance, std.error.concordance, r.squared, AIC, BIC) |>
  print()

cox_zph_02 <- lapse_model02_coxph |> cox.zph(transform = "km")
print(cox_zph_02)

anova(lapse_model01_coxph, lapse_model02_coxph, test = "Chisq")
```


## Model 03: Add Log Premium

Filter to positive premiums and sum assured for log transformations.

```{r model03}
#| echo: true

training_pos_tbl <- training_tbl |>
  filter(prem_ape > 0, sum_assured > 0)

lapse_model03_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~ gender_life1 + smoker_life1 + log(prem_ape),
  data = training_pos_tbl
  )

lapse_model03_coxph |> summary()
lapse_model03_coxph |> tidy(conf.int = TRUE) |> print()

lapse_model03_coxph |>
  glance() |>
  select(n, nevent, concordance, std.error.concordance, r.squared, AIC, BIC) |>
  print()

cox_zph_03 <- lapse_model03_coxph |> cox.zph(transform = "km")
print(cox_zph_03)
```


## Model 04: Add Log Sum Assured

```{r model04}
#| echo: true

lapse_model04_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~
    gender_life1 + smoker_life1 + log(prem_ape) + log(sum_assured),
  data = training_pos_tbl
  )

lapse_model04_coxph |> summary()
lapse_model04_coxph |> tidy(conf.int = TRUE) |> print()

lapse_model04_coxph |>
  glance() |>
  select(n, nevent, concordance, std.error.concordance, r.squared, AIC, BIC) |>
  print()

cox_zph_04 <- lapse_model04_coxph |> cox.zph(transform = "km")
print(cox_zph_04)

anova(lapse_model03_coxph, lapse_model04_coxph, test = "Chisq")
```


## Model 05: Add Cluster

```{r model05}
#| echo: true

lapse_model05_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~
    gender_life1 + smoker_life1 + log(prem_ape) + log(sum_assured) + cluster_id,
  data = training_pos_tbl
  )

lapse_model05_coxph |> summary()

lapse_model05_coxph |>
  glance() |>
  select(n, nevent, concordance, std.error.concordance, r.squared, AIC, BIC) |>
  print()

cox_zph_05 <- lapse_model05_coxph |> cox.zph(transform = "km")
print(cox_zph_05)

anova(lapse_model04_coxph, lapse_model05_coxph, test = "Chisq")
```


## Model 06: Add Age (Linear)

```{r model06}
#| echo: true

lapse_model06_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~
    gender_life1 + smoker_life1 + log(prem_ape) + log(sum_assured) +
    cluster_id + age_at_observation,
  data = training_pos_tbl
  )

lapse_model06_coxph |> summary()
lapse_model06_coxph |> tidy(conf.int = TRUE) |> print()

lapse_model06_coxph |>
  glance() |>
  select(n, nevent, concordance, std.error.concordance, r.squared, AIC, BIC) |>
  print()

cox_zph_06 <- lapse_model06_coxph |> cox.zph(transform = "km")
print(cox_zph_06)

anova(lapse_model05_coxph, lapse_model06_coxph, test = "Chisq")
```


## Model 07: Add Policy Vintage

```{r model07}
#| echo: true

lapse_model07_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~
    gender_life1 + smoker_life1 + log(prem_ape) + log(sum_assured) +
    cluster_id + age_at_observation + policy_vintage_year,
  data = training_pos_tbl
  )

lapse_model07_coxph |> summary()

lapse_model07_coxph |>
  glance() |>
  select(n, nevent, concordance, std.error.concordance, r.squared, AIC, BIC) |>
  print()

cox_zph_07 <- lapse_model07_coxph |> cox.zph(transform = "km")
print(cox_zph_07)

anova(lapse_model06_coxph, lapse_model07_coxph, test = "Chisq")
```


## Model 08: Full Model with Splines

```{r model08}
#| echo: true

lapse_model08_coxph <- coxph(
  Surv(policy_lifetime, lapsed) ~
    gender_life1 + smoker_life1 + log(prem_ape) + log(sum_assured) +
    cluster_id + pspline(age_at_observation) + policy_vintage_year +
    pspline(prem_sa_ratio),
  data = training_pos_tbl
  )

lapse_model08_coxph |> summary()

lapse_model08_coxph |>
  glance() |>
  select(n, nevent, concordance, std.error.concordance, r.squared, AIC, BIC) |>
  print()

cox_zph_08 <- lapse_model08_coxph |> cox.zph(transform = "km")
print(cox_zph_08)

anova(lapse_model07_coxph, lapse_model08_coxph, test = "Chisq")

par(mfrow = c(2, 2))
termplot(lapse_model08_coxph, term = 6, se = TRUE, col.term = "firebrick",
         main = "Age Effect (Spline)")
termplot(lapse_model08_coxph, term = 8, se = TRUE, col.term = "firebrick",
         main = "Prem/SA Ratio Effect (Spline)")
par(mfrow = c(1, 1))
```


# Model Comparison

Compare all models systematically.

```{r model_comparison}
#| echo: true

model_list <- list(
  lapse_model00_coxph,
  lapse_model01_coxph,
  lapse_model02_coxph,
  lapse_model03_coxph,
  lapse_model04_coxph,
  lapse_model05_coxph,
  lapse_model06_coxph,
  lapse_model07_coxph,
  lapse_model08_coxph
  )

model_comparison_tbl <- tibble(
  model = sprintf("%02d", 0:8),
  description = c(
    "Null",
    "Gender",
    "Gender + Smoker",
    "Demographics + log(prem)",
    "Demo + log(prem) + log(SA)",
    "Model04 + cluster",
    "Model05 + age",
    "Model06 + vintage",
    "Full with splines"
    ),
  n = map_dbl(model_list, ~glance(.x)$n),
  nevent = map_dbl(model_list, ~glance(.x)$nevent),
  concordance = map_dbl(model_list, ~glance(.x)$concordance),
  std_error = map_dbl(model_list, ~glance(.x)$std.error.concordance),
  r_squared = map_dbl(model_list, ~glance(.x)$r.squared),
  logLik = map_dbl(model_list, ~glance(.x)$logLik),
  AIC = map_dbl(model_list, ~glance(.x)$AIC),
  BIC = map_dbl(model_list, ~glance(.x)$BIC)
  )

print(model_comparison_tbl)

best_model_idx <- which.min(model_comparison_tbl$AIC)
best_model_name <- model_comparison_tbl$model[best_model_idx]
best_model <- model_list[[best_model_idx]]

write_lines(glue("Best model by AIC: Model {best_model_name}"), stdout())
```


## Forest Plot for Best Model

```{r forest_plot}
#| echo: true
#| fig-height: 10

# Use model07 for forest plot (model08 has splines which ggforest can't handle)
# Model07 is the best non-spline model
forest_model <- lapse_model07_coxph

# ggforest requires data frame with unique row names
forest_data <- training_pos_tbl |>
  ungroup() |>
  as.data.frame()

rownames(forest_data) <- paste0("row_", seq_len(nrow(forest_data)))

# Create forest plot for non-spline model
ggforest(forest_model, data = forest_data)

write_lines(
  "\nNote: Forest plot uses Model 07 (best non-spline model).",
  stdout()
  )
write_lines(
  "Model 08 includes pspline terms which cannot be visualized in forest plots.",
  stdout()
  )
```


# Cashflow Validation

Test best model on validation data to assess out-of-time performance.

## Prepare Validation Data

Filter validation data to match training policies that were inforce at training
date.

```{r prepare_validation}
#| echo: true

training_inforce_ids <- training_pos_tbl |>
  filter(status_at_reference_date == "inforce") |>
  pull(policy_id)

validation_pos_tbl <- validation_tbl |>
  filter(
    policy_id %in% training_inforce_ids,
    prem_ape > 0,
    sum_assured > 0
    )

write_lines(
  glue(
    "Policies inforce at training date: {comma(length(training_inforce_ids))}\n",
    "Validation sample (matched): {comma(nrow(validation_pos_tbl))}"
    ),
  stdout()
  )
```


## Generate Predictions

```{r generate_predictions}
#| echo: true

predictions_lst <- calculate_cashflow_predictions(
  model_coxph = best_model,
  training_data_tbl = training_pos_tbl |>
    filter(status_at_reference_date == "inforce"),
  prediction_horizon_weeks = 52
  )

print(predictions_lst$aggregate_summary)
```


## Compare to Actual Outcomes

```{r compare_outcomes}
#| echo: true

comparison_lst <- compare_predicted_vs_actual(
  predictions_list = predictions_lst,
  actual_data_tbl = validation_pos_tbl
  )

write_lines("\n=== SUMMARY METRICS ===", stdout())
print(comparison_lst$summary_metrics)

write_lines("\n=== CALIBRATION BY DECILE ===", stdout())
print(comparison_lst$calibration_tbl)
```


## Visualization

```{r visualize_results}
#| echo: true

plots_lst <- create_cashflow_comparison_plots(comparison_lst)

plot_grid(
  plots_lst$lapse_comparison_plot,
  plots_lst$premium_comparison_plot,
  plots_lst$calibration_plot,
  plots_lst$residual_plot,
  ncol = 2
  )
```


# Results Interpretation

```{r interpret_results}
#| echo: false

metrics <- comparison_lst$summary_metrics

write_lines(
  glue(
    "\n=== MODEL PERFORMANCE SUMMARY ===\n\n",
    "Lapse Prediction:\n",
    "  Predicted: {comma(metrics$predicted_lapses, accuracy = 1)} policies\n",
    "  Actual:    {comma(metrics$actual_lapses, accuracy = 1)} policies\n",
    "  Error:     {sprintf('%.1f%%', metrics$lapse_error_pct)}\n\n",
    "Premium at Risk:\n",
    "  Predicted: £{comma(metrics$predicted_premium_loss, accuracy = 1)}\n",
    "  Actual:    £{comma(metrics$actual_premium_loss, accuracy = 1)}\n",
    "  Error:     {sprintf('%.1f%%', metrics$premium_error_pct)}\n"
    ),
  stdout()
  )

if (abs(metrics$lapse_error_pct) < 10) {
  write_lines("Lapse predictions: Well-calibrated (< 10% error)", stdout())
  } else if (abs(metrics$lapse_error_pct) < 20) {
  write_lines("Lapse predictions: Acceptable (10-20% error)", stdout())
  } else {
  write_lines("Lapse predictions: Significant error (> 20%)", stdout())
  }

if (abs(metrics$premium_error_pct) < 10) {
  write_lines("Premium predictions: Well-calibrated (< 10% error)", stdout())
  } else if (abs(metrics$premium_error_pct) < 20) {
  write_lines("Premium predictions: Acceptable (10-20% error)", stdout())
  } else {
  write_lines("Premium predictions: Significant error (> 20%)", stdout())
  }
```

## Key Findings

1. **Discrimination**: Model 08 achieved concordance of
   `r sprintf("%.3f", glance(best_model)$concordance)`, indicating
   `r if(glance(best_model)$concordance > 0.65) "good" else if(glance(best_model)$concordance > 0.60) "moderate" else "weak"`
   ability to rank lapse risk.

2. **Calibration**: The calibration plot shows how predicted probabilities match
   observed rates across risk deciles. Points near the diagonal indicate good
   calibration.

3. **Premium at Risk**: Model predicts
   £`r comma(metrics$predicted_premium_loss, accuracy = 1)` in annual
   premium at risk vs actual £`r comma(metrics$actual_premium_loss, accuracy = 1)`.

4. **Key Risk Factors**: Demographics (gender, smoker), financial variables
   (premium, sum assured), age, policy vintage, and cluster membership all
   contribute to lapse risk.


# Conclusions

This analysis demonstrates Cox PH modeling for life insurance lapse prediction
with rollback validation.

## Business Value

  - **Interpretable risk factors**: Clear hazard ratios for each covariate
  - **Individual risk scores**: Policy-level lapse probabilities
  - **Aggregate forecasts**: Total expected lapses and premium at risk
  - **Validated performance**: Out-of-time assessment mimics real deployment

## Recommendations

  1. **Targeted retention**: Focus on high-risk policies identified by model
  1. **Pricing review**: Examine segments with high lapse rates
  1. **Product design**: Address structural drivers (prem/SA ratio)
  1. **Capital planning**: Use calibrated forecasts for reserves

## Limitations

  - Assumes proportional hazards (tested via cox.zph)
  - No time-varying covariates
  - Limited to observable policy characteristics
  - Past lapse behavior may not predict future

## Future Work

  - Incorporate time-varying covariates (payment history)
  - Test stratified models by product/vintage
  - Automate model monitoring for drift
  - Extend to multi-state models (lapse vs other exits)


# R Environment

```{r session_info}
#| echo: true

sessioninfo::session_info()
```
