---
title: "Bond Portfolio"
author: "Mick Cooney"
date: "13 July 2016"
output:
  html_document:
    toc: true
    number_sections: true
    fig_caption: yes
    theme: cerulean
    css: styles.css
  pdf_document: default
---

<!--
(Title:) Bond Portfolio

Author: Mick Cooney

Date: 2016

Abstract: This is a quick exploratory analysis for a bond portfolio

Keywords: bond portfolio

-->

```{r knit_opts, include = FALSE}
rm(list = ls())

knitr::opts_chunk$set(tidy = FALSE, cache = FALSE)

library(lubridate)
library(data.table)
library(ggplot2)
library(scales)

library(readODS)


options(width            = 90)
options(stringsAsFactors = FALSE)

options(datatable.print.nrows      = 10L)
options(datatable.prettyprint.char = 80L)

source("custom_functions.R")
```

```{r set_variables, echo=TRUE}
portfolio_date    <- as.Date('2014-10-01')
portfolio_capital <- 3000000

```



---

# Load Old Portfolio

Just to test the proof of concept, I am loading an old speculative
portfolio from 2014 where I was taking the bond data and constructing
the portfolio based on the price and FX rate for USD. We can then use
the count of contracts (assuming a common face value of \$1,000
USD/CAD).

```{r load_data, echo=TRUE}
bond_portfolio_dt <- fread("data/highyield_bond_portfolio.csv")

bond_portfolio_dt[, bond_id       := .I]
bond_portfolio_dt[, issued_date   := as.Date(issued_date)]
bond_portfolio_dt[, maturity_date := as.Date(maturity_date)]
bond_portfolio_dt[, face_value    := 1000]

str(bond_portfolio_dt)

print(bond_portfolio_dt)

cadusd_dt <- fread("cadusd.csv")
cadusd_dt[, date := as.Date(Date, format = '%d/%m/%Y')]
fx_dt <- cadusd_dt[, .(symbol = 'CADUSD.FX', date, close = Rate)][order(symbol,date)]

print(fx_dt)
```

With the desired issues and FX data loaded we then calculate the
contract count for each issue.


```{r calculate_portfolio, echo=TRUE}
exchange_rate     <- fx_dt[date == portfolio_date, close]

port_dt <- calculate_portfolio_amounts(bond_portfolio_dt
                                      ,fund_capital  = portfolio_capital
                                      ,exchange_rate = exchange_rate)

print(port_dt)
```

We also want to look at the schedule of coupon payments.

```{r show_coupon_payments, echo=TRUE}
coupon_data_dt <- create_coupon_schedule(port_dt, portfolio_date)

schedule_dt <- merge(port_dt[, .(bond_id, issuer, currency)]
                    ,coupon_data_dt
                    ,by = 'bond_id')
```


# Load New Portfolio

We now move forward to July 2016, and we have a given portfolio as
exported out of Bloomberg's PRTU functionality. We have exported the
portfolio as an OpenDocument ODS file, and we will load that file
directly into our code.

```{r load_new_portfolio, echo=TRUE}
portfolio_date <- as.Date('2016-07-14')

portfolio_file <- "~/Dropbox/Shared/HTnR/Transfer/bond_portfolio_20160711.ods"

bondportfolio_dt <- load_portfolio_file(portfolio_file)

print(bondportfolio_dt)
```

With the data loaded and all the datatypes set to the correct types,
we can now start doing some simple analysis on the cashflows.

## Perform Data Cleaning Checks

We want to do some very simple data cleaning checks. For example, we
need to ensure that the given FX rate for a bond issue is not 1.00
only in the case when the bond is USD denominated.

```{r data_clean, echo=TRUE}
print(bondportfolio_dt[abs(fx_rate - 1) > 1e-4 & currency != 'USD'])
```


# Show Coupon and Maturity Cashflows

We now want to investigate combined cashflows from the bond portfolio
as a whole. In particular, USD cashflows will be hedged using futures
so we want to do that in a rigourous way and determine what to trade
in futures to help hedge the currency risk.

```{r show_coupon, echo=TRUE}
max_date <- bondportfolio_dt[, max(maturity_date)]

bondportfolio_coupon_dt <- create_coupon_schedule(bondportfolio_dt, portfolio_date)

bondportfolio_coupon_dt <- merge(bondportfolio_dt[, .(bond_id, currency, coupon, position)]
                                ,bondportfolio_coupon_dt
                                ,by = 'bond_id')

bondportfolio_coupon_dt[, coupon_month := format(coupon_date, '%Y%m')]


cashflow_dt <- dcast(bondportfolio_coupon_dt
                    ,bond_id + currency + coupon + position ~ coupon_month
                    ,value.var = 'cashflow'
                    ,fill = 0)

ym_values <- format(seq(portfolio_date, max_date, by = 'month'), '%Y%m')

ym_fill <- ym_values[!ym_values %in% names(cashflow_dt)]

zero_cols <- matrix(0, ncol = length(ym_fill), nrow = nrow(cashflow_dt))
colnames(zero_cols) <- ym_fill


cashflow_dt <- cbind(cashflow_dt, zero_cols)

ordered_cols <- sort(names(cashflow_dt)[grep("\\d{6}", names(cashflow_dt))])

cashflow_dt <- cashflow_dt[, c("bond_id", "currency", "coupon", "position"
                              ,sort(ordered_cols))
                           ,with = FALSE]

print(cashflow_dt[, lapply(.SD, sum), by = 'currency', .SDcols = ym_values])
```

# Calculate Yield Data

We often wish to calculate the *yield to maturity* - the discount rate
for the cashflows that gives us a Net Present Value for the bond to be
0. The yield-to-maturity is one way of quantifying the amount of
interest paid by that particular contract. As the price of the bond
changes, the yield on the bond will change.

```{r calculate_bond_ytm, echo=TRUE}
bonddata_dt <- merge(bondportfolio_dt[, .(bond_id, market_val)]
                    ,bondportfolio_coupon_dt
                    ,by = 'bond_id')

bond_ytm_dt <- bonddata_dt[, {
    bond_cost = market_val[1]

    days  = as.numeric(coupon_date - portfolio_date)

    duration_data = calculate_duration_data(days, cashflow, bond_cost)

    data.table(t(duration_data))
}, by = bond_id]

bondportfolio_dt <- merge(bondportfolio_dt
                         ,bond_ytm_dt
                         ,by = 'bond_id')

bondportfolio_dt[, maturity := as.numeric(maturity_date - portfolio_date) / 365]

print(bondportfolio_dt)
```



# Calculate Hedging Amounts

In this section we develop the ability to extract the coupon dates for
each of the bond issues and assign a cashflow to the next delivery
date for the FX futures. This allows us to see how many futures of
each contract size are required at each delivery.

We currently will assume contract sizes of 100,000 CAD and 10,000 CAD.

```{r calculate_hedging_data, echo=TRUE}
datedelivery_dt <- fread("data/delivery_dates.csv")
datedelivery_dt[, delivery := as.Date(delivery)]

datedelivery_dt[, mergedate := delivery]
setkey(datedelivery_dt, mergedate)

bondportfolio_coupon_dt[, mergedate := coupon_date]
setkey(bondportfolio_coupon_dt, mergedate)

bondportfolio_cashflow_dt <- datedelivery_dt[bondportfolio_coupon_dt, roll = -Inf]
bondportfolio_cashflow_dt <- bondportfolio_cashflow_dt[, .(bond_id, currency, coupon_date, delivery, cashflow)]
bondportfolio_cashflow_dt <- bondportfolio_cashflow_dt[order(bond_id, coupon_date)]

bondportfolio_delivery_dt <- bondportfolio_cashflow_dt[, .(total_cashflow = sum(cashflow))
                                                       ,keyby = .(currency, delivery)]

print(bondportfolio_delivery_dt)
```

We aggregate the cashflows by delivery and then display them in a
table, ordering by delivery date.

```{r show_delivery_currency, echo=TRUE}
bondportfolio_currency_dt <- dcast(bondportfolio_delivery_dt
                                  ,delivery ~ currency
                                  ,value.var = 'total_cashflow'
                                  ,fill = 0)

print(bondportfolio_currency_dt)
```

We also do the hedging calculations for the futures.


```{r show_hedge_trades, echo=TRUE}
calc_hedges <- create_hedging_function(c(100000, 10000))

bondportfolio_contracts_dt <- bondportfolio_currency_dt[, data.table(t(calc_hedges(USD))), by = delivery]

bondportfolio_hedge_dt <- merge(bondportfolio_currency_dt
                               ,bondportfolio_contracts_dt
                               ,keyby = 'delivery')

print(bondportfolio_hedge_dt)
```
