---
title: "Building Transactional Frequency Models"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: TRUE
    code_folding: hide
    fig_caption: TRUE

  html_document:
    fig_caption: yes
    theme: spacelab #sandstone #spacelab #flatly
    highlight: pygments
    number_sections: TRUE
    toc: TRUE
    toc_depth: 2
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---


```{r import_libraries, echo=FALSE, message=FALSE}
library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(directlabels)
library(magrittr)
library(rlang)
library(fs)
library(purrr)
library(furrr)
library(glue)
library(cmdstanr)
library(brms)
library(posterior)
library(bayesplot)
library(tidybayes)


source("lib_utils.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2")
  )


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width        = 80L,
  warn         = 1,
  brms.backend = "cmdstanr",
  mc.cores     = parallel::detectCores()
  )

theme_set(theme_cowplot())

set.seed(42)

plan(multisession)
```

In this workbook we investigate different ways to model the transaction
frequency of individual customers, with a view to expanding this approach into
a more traditional P/NBD model.

# Load and Configure Datasets

We first want to load some synthetic transaction data.

```{r load_synth_transaction_data, echo=TRUE}
customer_cohort_tbl <- read_rds("data/synthdata_singleyear_cohort_tbl.rds")
customer_cohort_tbl %>% glimpse()

customer_simparams_tbl <- read_rds("data/synthdata_singleyear_simparams_tbl.rds")
customer_simparams_tbl %>% glimpse()

customer_transactions_tbl <- read_rds("data/synthdata_singleyear_transactions_tbl.rds")
customer_transactions_tbl %>% glimpse()
```


## Construct Frequency Data

```{r construct_frequency_modelling_data, echo=TRUE}
customer_summarystats_tbl <- customer_transactions_tbl %>%
  group_by(customer_id) %>%
  summarise(
    .groups = "drop",
    
    tnx_count    = n(),
    first_tnx_ts = min(tnx_timestamp),
    last_tnx_ts  = max(tnx_timestamp),
    btyd_count   = tnx_count - 1,
    all_weeks    = 52,
    tnx_weeks    = difftime(last_tnx_ts, first_tnx_ts, units = "weeks") %>% as.numeric(),
    tnx_weeks    = if_else(btyd_count > 0, tnx_weeks, 52),    
    obs_freq     = btyd_count / tnx_weeks,
    emp_freq     = btyd_count / all_weeks
    )

customer_summarystats_tbl %>% glimpse()
```

We want to view the transactions to get a sense of how regular our transactions
are in general.

```{r plot_customer_transaction_times, echo=TRUE}
plot_tbl <- customer_transactions_tbl %>%
  group_nest(customer_id, .key = "cust_data") %>%
  filter(map_int(cust_data, nrow) > 3) %>%
  slice_sample(n = 30) %>%
  unnest(cust_data)

ggplot(plot_tbl, aes(x = tnx_timestamp, y = customer_id)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Date",
    y = "Customer ID",
    title = "Visualisation of Customer Transaction Times"
    ) +
  theme(axis.text.y = element_text(size = 10))
```


## Construct Data Subsets

For the purposes of saving on time and computation, we construct a subset of
the data first, and rather than constructing different samples each time we
instead construct sets of `customer_id` to draw upon, then using those list
of values when we want to take a subset.

We do this by randomly shuffling the data and then selecting the top $n$ values
of `customer_id`.

```{r construct_data_subset_ids, echo=TRUE}
shuffle_tbl <- customer_summarystats_tbl %>%
  slice_sample(n = nrow(.), replace = FALSE)

id_1000  <- shuffle_tbl %>% head(1000)  %>% pull(customer_id) %>% sort()
id_5000  <- shuffle_tbl %>% head(5000)  %>% pull(customer_id) %>% sort()
id_10000 <- shuffle_tbl %>% head(10000) %>% pull(customer_id) %>% sort()
```

We now have a list of `customer_id` values we use to subset the data.


# Construct Flat Frequency Model

For first frequency model we fit each customer's transaction frequency
independent of the other customers. We start with the estimation that the
average customer frequency is about one transaction every four weeks (once a
month) so this value determines our initial prior distribution for the
transaction frequency.

For our initial fit, we use only 1,000 eligible customers to reduce computation
time and construct a dataset used to fit these models.

```{r construct_fit_1000_data, echo=TRUE}
fit_1000_data_tbl <- customer_summarystats_tbl %>%
  filter(
    customer_id %in% id_1000
    )

fit_1000_data_tbl %>% glimpse()
```

We do the same for the 10k customers.

```{r construct_fit_10000_data, echo=TRUE}
fit_10000_data_tbl <- customer_summarystats_tbl %>%
  filter(
    customer_id %in% id_10000
    )

fit_10000_data_tbl %>% glimpse()
```


## Fit Stan Model

We start by fitting our first Stan model, fitting a Poisson rate for each
individual customer.

```{r display_freqmodel_flat_stancode, echo=FALSE}
read_lines("stan_code/freqmodel_flat.stan") %>% cat(sep = "\n")
```

We now compile this model using `CmdStanR`.

```{r compile_freqmodel_flat_stanmodel, echo=TRUE}
freqmodel_flat_stanmodel <- cmdstan_model(
  "stan_code/freqmodel_flat.stan",
  include_paths = "stan_code/",
  pedantic      =  TRUE,
  dir           = "stan_models/"
  )
```

We then use this compiled model with our data to produce a fit of the data.

```{r fit_freqmodel_flat_stanmodel, echo=TRUE}
stan_data_lst <- fit_1000_data_tbl %>%
  select(customer_id, btyd_count, tnx_weeks = all_weeks) %>%
  compose_data(
    r     =  1,
    alpha =  4
    )

freqmodel_flat_stanfit <- freqmodel_flat_stanmodel$sample(
  data            =    stan_data_lst,
  chains          =                4,
  iter_warmup     =              500,
  iter_sampling   =              500,
  seed            =              421,
  output_dir      =    "stan_models",
  output_basename = "freqmodel_flat"
  )

freqmodel_flat_stanfit$summary()
```

We first want to check the HMC diagnostics.

```{r calculate_freqmodel_flat_hmc_diagnostics, echo=TRUE}
freqmodel_flat_stanfit$cmdstan_diagnose()
```


## Check Model Fit

We now need to check the parameters of this fit against the data to see how
effective our model is at capturing the data. In this case we have the benefit
of knowing the 'true' data, and so we compare our model output against the
input parameters.

```{r construct_freqmodel_flat_validation, echo=TRUE}
freqmodel_flat_validation_tbl <- freqmodel_flat_stanfit %>%
  recover_types(fit_1000_data_tbl) %>%
  spread_draws(lambda[customer_id]) %>%
  ungroup() %>%
  inner_join(customer_simparams_tbl, by = "customer_id") %>%
  select(customer_id, post_lambda = lambda, customer_lambda)

freqmodel_flat_validation_tbl %>% glimpse()
```

Having constructed the validation data we now want to check the quantile of
each 'true' value in the posterior distribution for the parameter. If our
model is valid, this distribution will be uniform on $[0, 1]$.

```{r calculate_freqmodel_flat_qvalues, echo=TRUE}
plotvalid_tbl <- freqmodel_flat_validation_tbl %>%
  calculate_distribution_qvals(post_lambda, customer_lambda)

ggplot(plotvalid_tbl) +
  geom_histogram(aes(x = q_val), bins = 50) +
  geom_hline(aes(yintercept = plotvalid_tbl %>% nrow() %>% divide_by(50)), colour = "red") +
  labs(
    x = "Quantile",
    y = "Count",
    title = "Quantile Plot of the q-Values for the Posterior Distribution"
    )

ggplot(plotvalid_tbl) +
  geom_point(aes(x = q_val, y = customer_lambda)) +
  labs(
    x = "Quantile",
    y = "Customer Lambda",
    title = "Scatterplot of q-Value against Lambda"
    )
```



## Fit Larger Model

We now repeat this model with a larger dataset.

```{r fit_10k_freqmodel_flat_stanmodel, echo=TRUE}
stan_data_lst <- fit_10000_data_tbl %>%
  select(customer_id, btyd_count, tnx_weeks = all_weeks) %>%
  compose_data(
    r     =  1,
    alpha =  4
    )

freqmodel_flat_10k_stanfit <- freqmodel_flat_stanmodel$sample(
  data            =        stan_data_lst,
  chains          =                    4,
  iter_warmup     =                  500,
  iter_sampling   =                  500,
  seed            =                  422,
  output_dir      =        "stan_models",
  output_basename = "freqmodel_flat_10k"
  )

freqmodel_flat_10k_stanfit$summary()
```

We first want to check the HMC diagnostics.

```{r calculate_freqmodel_flat_10k_hmc_diagnostics, echo=TRUE}
freqmodel_flat_10k_stanfit$cmdstan_diagnose()
```


## Check Larger Model Fit

We now repeat the validation exercise once more, extracting the posterior
customer $\lambda$ values and comparing them to our input values.

```{r construct_freqmodel_flat_10k_validation, echo=TRUE}
freqmodel_flat_10k_validation_tbl <- freqmodel_flat_10k_stanfit %>%
  recover_types(fit_10000_data_tbl) %>%
  spread_draws(lambda[customer_id]) %>%
  ungroup() %>%
  inner_join(customer_simparams_tbl, by = "customer_id") %>%
  select(customer_id, post_lambda = lambda, customer_lambda)

freqmodel_flat_10k_validation_tbl %>% glimpse()
```

We also construct validation plots for this.

```{r calculate_freqmodel_flat_10k_qvalues, echo=TRUE}
plotvalid_tbl <- freqmodel_flat_10k_validation_tbl %>%
  calculate_distribution_qvals(post_lambda, customer_lambda)

ggplot(plotvalid_tbl) +
  geom_histogram(aes(x = q_val), bins = 50) +
  geom_hline(aes(yintercept = plotvalid_tbl %>% nrow() %>% divide_by(50)), colour = "red") +
  labs(
    x = "Quantile",
    y = "Count",
    title = "Quantile Plot of the q-Values for the Posterior Distribution"
    )

ggplot(plotvalid_tbl) +
  geom_point(aes(x = q_val, y = customer_lambda), alpha = 0.2) +
  labs(
    x = "Quantile",
    y = "Customer Lambda",
    title = "Scatterplot of q-Value against Lambda"
    )
```


## Construct Observed-Time Frequency Model

We now want to update our approach to use the observed duration of the
transactions rather than using a constant time period of 52 weeks.

```{r fit_observed_freqmodel_flat_stanmodel, echo=TRUE}
stan_data_lst <- fit_1000_data_tbl %>%
  transmute(customer_id, btyd_count, tnx_weeks) %>%
  compose_data(
    r     =  1,
    alpha =  4
    )

freqmodel_flat_obs_stanfit <- freqmodel_flat_stanmodel$sample(
  data            =        stan_data_lst,
  chains          =                    4,
  iter_warmup     =                  500,
  iter_sampling   =                  500,
  seed            =                  423,
  output_dir      =        "stan_models",
  output_basename = "freqmodel_flat_obs"
  )

freqmodel_flat_obs_stanfit$summary()
```

Now that we have fit our models, we repeat our validation model much like we
did the last time. We expect this model to deviate from our first model however
as we have a less-reliable estimate of the observation time for each
customer - as it is now calculated as ending at the last observation.

```{r calculate_validation_observed_data, echo=TRUE}
freqmodel_flat_obs_validation_tbl <- freqmodel_flat_obs_stanfit %>%
  recover_types(fit_1000_data_tbl) %>%
  spread_draws(lambda[customer_id]) %>%
  ungroup() %>%
  inner_join(customer_simparams_tbl, by = "customer_id") %>%
  select(customer_id, post_lambda = lambda, customer_lambda)

freqmodel_flat_obs_validation_tbl %>% glimpse()
```


We also construct validation plots for this.

```{r calculate_freqmodel_flat_obs_qvalues, echo=TRUE}
plotvalid_tbl <- freqmodel_flat_obs_validation_tbl %>%
  calculate_distribution_qvals(post_lambda, customer_lambda)

ggplot(plotvalid_tbl) +
  geom_histogram(aes(x = q_val), bins = 50) +
  geom_hline(aes(yintercept = plotvalid_tbl %>% nrow() %>% divide_by(50)), colour = "red") +
  labs(
    x = "Quantile",
    y = "Count",
    title = "Quantile Plot of the q-Values for the Posterior Distribution"
    )

ggplot(plotvalid_tbl) +
  geom_point(aes(x = q_val, y = customer_lambda), alpha = 0.2) +
  labs(
    x = "Quantile",
    y = "Customer Lambda",
    title = "Scatterplot of q-Value against Lambda"
    )
```


## Construct Posterior Distribution Comparisons

We now combine the outputted posterior distributions and compare them together.

```{r combine_flat_posterior_distribs, echo=TRUE}
comparison_distrib_tbl <- list(
    `Fixed Time`    = freqmodel_flat_validation_tbl,
    `Observed Time` = freqmodel_flat_obs_validation_tbl
    ) %>%
  bind_rows(.id = "label")

comparison_distrib_tbl %>% glimpse()
```

Now that we have this data we can construct a number of visualisations to
help us compare these two distributions on a per-customer basis.

```{r plot_distribution_posterior_lambda, echo=TRUE}
comparison_summary_tbl <- comparison_distrib_tbl %>%
  group_by(label, customer_id) %>%
  summarise(
    .groups = "drop",
    
    labels = c("q10", "q25", "q50", "q75", "q90", "mean", "customer_lambda"),
    vals   = c(quantile(post_lambda, probs = c(0.1, 0.25, 0.50, 0.75, 0.90)),
               mean(post_lambda),
               unique(customer_lambda))
    ) %>%
  pivot_wider(
    names_from  = labels,
    values_from = vals
    )

plot_tbl <- comparison_summary_tbl %>%
  group_nest(customer_id) %>%
  slice_sample(n = 30) %>%
  unnest(data)

ggplot(plot_tbl) +
  geom_errorbar(
    aes(x = customer_id, ymin = q10, ymax = q90, colour = label),
    position = position_dodge(width = 0.75), width = 0, size = 1) +
  geom_errorbar(
    aes(x = customer_id, ymin = q25, ymax = q75, colour = label),
    position = position_dodge(width = 0.75), width = 0, size = 2) +
  geom_point(aes(x = customer_id, y = customer_lambda)) +
  labs(
    x = "Customer ID",
    y = "Frequency Rate",
    colour = "Distribution"
    ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```






# Construct Hierarchical Frequency Model

We now need to


















# Fit brms Model

To start with we fit our model using `brms` as this provides a large amount
of infrastructure and additional functionality that may be of use.

```{r fit_flat_brms_frequency_model, echo=TRUE}
# freqmodel_flat_brmfit <- brm(
#   formula = btyd_count ~ offset(log(all_weeks)) + customer_id,
#   data    = fit_1000_data_tbl,
#   family  = poisson(),
#   prior   = c(
#     set_prior("normal(-1.96, 1.25)", class = "b", lb = 0)
#     ),
#   warmup   = 500,
#   iter     = 1000,
#   chains   = 4,
#   seed     = 421
#   )
```


Now that we have fit the model we want to see how well our posterior
distribution is covering the 'correct' values.

```{r construct_flat_freq_model_check, echo=TRUE}
# freqmodel_flat_postcheck_tbl <- freqmodel_flat_brmfit %>%
#   tidy_draws() %>%
#   pivot_longer(
#     !c(.chain, .iteration, .draw),
#     names_prefix = "b_customer_id",
#     names_to     = "customer_id",
#     values_to    = "post_freqmean"
#     ) %>%
#   inner_join(customer_simparams_tbl, by = "customer_id") %>%
#   select(.draw, customer_id, customer_lambda, post_freqmean)
# 
# 
# freqmodel_flat_postcheck_qvals_tbl <- freqmodel_flat_postcheck_tbl %>%
#   calculate_distribution_qvals(post_freqmean, customer_lambda, customer_id)
# 
# freqmodel_flat_postcheck_qvals_tbl %>% glimpse()
```




# R Environment

```{r show_session_info, echo=TRUE, message=TRUE}
options(width = 120L)
sessioninfo::session_info()
options(width = 80L)
```
