---
title: "Using Buy-Till-You-Die (BTYD) Models the Online Retail Dataset"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: yes
    css: styles.css

  html_document:
    fig_caption: yes
    theme: spacelab #sandstone #spacelab #flatly
    highlight: pygments
    number_sections: TRUE
    toc: TRUE
    toc_depth: 2
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---


```{r import_libraries, echo=FALSE, message=FALSE}
library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(directlabels)
library(magrittr)
library(rlang)
library(fs)
library(purrr)
library(furrr)
library(tidyquant)
library(glue)
library(BTYD)
library(BTYDplus)
library(CLVTools)


source("lib_utils.R")

resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2")
  )


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallel::detectCores()
  )

theme_set(theme_cowplot())

set.seed(42)

plan(multisession)
```


# Load Data

We first want to load our datasets and perform some basic manipulation similar
to those done for the association rules mining.

```{r load_transaction_data, echo=TRUE}
tnx_data_tbl <- read_rds("data/retail_data_cleaned_tbl.rds")

tnx_data_tbl %>% glimpse()
```


We model the purchase data first, then combine this to create an individual
customer/invoice pairing with the total amount spent as an additional column.

```{r prepare_purchase_data, echo=TRUE}
tnx_purchase_tbl <- tnx_data_tbl %>%
  filter(
    quantity > 0,
    exclude == FALSE
    ) %>%
  select(
    invoice_date, invoice_id, stock_code, customer_id, description,
    quantity, price, stock_value
    )

tnx_purchase_tbl %>% glimpse()
```

Use of BTYD models assumes a total spend over a period of day and those
differences between the times. This is calculated internally by the various
BTYD routines so we keep just the per-invoice spend.

```{r calculate_customer_daily_spend, echo=TRUE}
daily_spend_invoice_tbl <- tnx_purchase_tbl %>%
  drop_na(customer_id) %>%
  group_by(invoice_date, customer_id, invoice_id) %>%
  summarise(
    .groups = "drop",
    
    invoice_spend = sum(stock_value)
    )

daily_spend_invoice_tbl %>% glimpse()


daily_spend_tbl <- daily_spend_invoice_tbl %>%
  group_by(invoice_date, customer_id) %>%
  summarise(
    .groups = "drop",
    
    total_spend = sum(invoice_spend),
    tnx_count   = n()
    )

daily_spend_tbl %>% glimpse()
```


Finally, we want to set a date to make the end of the "fitting" data, which
will be used in the various estimation routines to fit parameters of our models.

```{r set_training_data_date, echo=TRUE}
training_data_date <- as.Date("2011-03-31")
```



# Background Theory

Before we start working on fitting and using the various Buy-Till-You-Die
models, we first need to discuss the basic underlying theory and model.

In this model, we assume a customer becomes 'alive' to the business at the
first purchase and then makes purchases stochastically but at a steady-rate
for a period of time, and then 'dies' - i.e. becomes inactive to the business -
hence the use of "Buy-Till-You-Die".

Thus, at a high level these models decompose into modelling the transaction
events using distributions such as the Poisson or Negative Binomial, and then
modelling the 'dropout' process using some other method.

A number of BTYD models exist and for this workshop we will focus on the
BG/NBD model - the Beta-Geometric Negative Binomial Distribution model (though
we will discuss the P/NBD model also).

These models require only two pieces of information about each customer's
purchasing history: the "recency" (when the last transaction occurred) and
"frequency" (the count of transactions made by that customer in a specified
time period).

The notation used to represent this information is

$$
X = (x, t_x, T),
$$
where $x$ is the number of transactions observed in the time period $T$ and
$t_x$ was the time of the last transaction.

From this summary data we can fit both P/NBD and BG/NBD models.


## Beta-Geometric/Negative-Binomial Distribution (BG/NBD) Model

This model relies on a number of base assumptions:

  1. While active, the number of transactions made by a customer follows a
  Poisson process with transaction rate $\lambda$.
  1. Heterogeneity in $\lambda$ follows a Gamma distribution
  $\Gamma(\lambda \, | \, \alpha, r)$ with parameters shape $r$ and rate
  $\alpha$. 
  1. After any transaction, a customer becomes inactive with probability $p$.
  1. Heterogeneity in $p$ follows a Beta distribution $B(p \, | \, a, b)$ with
  shape parameters $a$ and $b$.
  1. The transaction rate $\lambda$ and the dropout probability $p$ vary
  independently across customers.


Note that it follows from the above assumptions that the probability of a
customer being 'alive' after any transaction is given by the Geometric
distribution, and hence the Beta-Geometric in the name.


## Pareto/Negative-Binomial Distribution (P/NBD) Model

Similar to the BG/NBD model, the P/NBD model relies on five assumptions:

  1. While active, the number of transactions made by a customer follows a
  Poisson process with transaction rate $\lambda$.
  1. Heterogeneity in $\lambda$ follows a Gamma distribution
  $\Gamma(\lambda \, | \, \alpha, r)$ with shape $r$ and rate $\alpha$. 
  1. Each customer has an unobserved 'lifetime' of length $\tau$. This point at
  which the customer becomes inactive is distributed as an exponential with
  dropout rate $\mu$.
  1. Heterogeneity in dropout rates across customers follows a Gamma
  distribution $\Gamma(\mu \, | \, s, \beta)$ with shape parameter $s$ and
  rate parameter $\beta$.
  1. The transaction rate $\lambda$ and the dropout rate $\mu$ vary
  independently across customers.


## Gamma/Gamma Spending Model

The final piece of this approach is to construct a spending model for the
transactions. Similar to transaction rates, spending patterns are often
skewed, so a Gamma distribution is used.

For the Gamma/Gamma model, we use the following assumptions:

  1. Each transaction amount, $v_i$, is distributed according to a Gamma
  distribution $\Gamma(v_i \, | \, p, \nu)$ with shape parameter $p$ and
  rate parameter $\nu$.
  2. Heterogeneity in $p$ follows a Gamma distribution
  $\Gamma(p \, | \, q, \gamma)$





# Data Preparation

```{r}
daily_spend_tbl %>%
  clvdata(
    date.format = "%Y-%m-%d",
    time.unit = "weeks",
    estimation.split = training_data_date,
    name.id          = "customer_id",
    name.date        = "invoice_date",
    name.price       = "total_spend"
    )
```


# R Environment

```{r show_session_info, echo=TRUE, message=TRUE}
sessioninfo::session_info()
```
