---
title: "Building Initial Monetary Models"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: TRUE
    code_folding: hide
    fig_caption: TRUE

  html_document:
    fig_caption: yes
    theme: spacelab #sandstone #spacelab #flatly
    highlight: pygments
    number_sections: TRUE
    toc: TRUE
    toc_depth: 2
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---


```{r import_libraries, echo=FALSE, message=FALSE}
library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(directlabels)
library(magrittr)
library(rlang)
library(fs)
library(purrr)
library(furrr)
library(glue)
library(cmdstanr)
library(brms)
library(posterior)
library(bayesplot)
library(tidybayes)


source("lib_utils.R")
source("lib_btyd.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2")
  )


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width        = 80L,
  warn         = 1,
  brms.backend = "cmdstanr",
  mc.cores     = parallel::detectCores()
  )

theme_set(theme_cowplot())

set.seed(42)

plan(multisession)
```

In this workbook we investigate different ways to model the transaction
frequency of individual customers, with a view to expanding this approach into
a more traditional P/NBD model.

# Load and Configure Datasets

We first want to load some synthetic transaction data.

```{r load_synth_transaction_data, echo=TRUE}
customer_cohort_tbl <- read_rds("data/synthdata_allfixed_cohort_tbl.rds")
customer_cohort_tbl %>% glimpse()

customer_simparams_tbl <- read_rds("data/synthdata_allfixed_simparams_tbl.rds")
customer_simparams_tbl %>% glimpse()

customer_transactions_tbl <- read_rds("data/synthdata_allfixed_transactions_tbl.rds")
customer_transactions_tbl %>% glimpse()
```

Our transaction data is our main input data for this work, so we will show the
first few rows of this.

```{r show_transaction_data, echo=TRUE}
customer_transactions_tbl %>% arrange(tnx_timestamp) %>% head(10)
```

We also want to set up a number of parameters for use in this workbook

```{r setup_workbook_parameters, echo=TRUE}
stan_modeldir <- "stan_models"
stan_codedir  <-   "stan_code"
```


## Construct Visualisation Data

```{r construct_frequency_modelling_data, echo=TRUE}
customer_summarystats_tbl <- customer_transactions_tbl %>%
  calculate_transaction_summary_stats()

customer_summarystats_tbl %>% glimpse()
```

We want to view the transactions to get a sense of how regular our transactions
are in general.

```{r plot_customer_transaction_times, echo=TRUE}
plot_tbl <- customer_transactions_tbl %>%
  group_nest(customer_id, .key = "cust_data") %>%
  filter(map_int(cust_data, nrow) > 3) %>%
  slice_sample(n = 30) %>%
  unnest(cust_data)

ggplot(plot_tbl, aes(x = tnx_timestamp, y = customer_id)) +
  geom_line() +
  geom_point() +
  labs(
    x = "Date",
    y = "Customer ID",
    title = "Visualisation of Customer Transaction Times"
    ) +
  theme(axis.text.y = element_text(size = 10))
```


## Construct Data Subsets

For the purposes of saving on time and computation, we construct a subset of
the data first, and rather than constructing different samples each time we
instead construct sets of `customer_id` to draw upon, then using those list
of values when we want to take a subset.

We do this by randomly shuffling the data and then selecting the top $n$ values
of `customer_id`.

```{r construct_data_subset_ids, echo=TRUE}
shuffle_tbl <- customer_summarystats_tbl %>%
  slice_sample(prop = 1, replace = FALSE)

id_1000  <- shuffle_tbl %>% head(1000)  %>% pull(customer_id) %>% sort()
id_5000  <- shuffle_tbl %>% head(5000)  %>% pull(customer_id) %>% sort()
id_10000 <- shuffle_tbl %>% head(10000) %>% pull(customer_id) %>% sort()
```

We now have a list of `customer_id` values we use to subset the data.

# Fit Transaction Amount Models

We first want to use these samples of the dataset to look at it.


```{r construct_fit_data_subsets, echo=TRUE}
fit_1000_data_tbl <- customer_transactions_tbl %>%
  filter(customer_id %in% id_1000)

fit_1000_data_tbl %>% glimpse()


fit_10000_data_tbl <- customer_transactions_tbl %>%
  filter(customer_id %in% id_10000)
  
fit_10000_data_tbl %>% glimpse()
```



## Fit Stan Model

We start by fitting our first Stan model, fitting a Poisson rate for each
individual customer.

```{r display_amtmodel_flat_stancode, echo=FALSE}
read_lines("stan_code/amtmodel_flat.stan") %>% cat(sep = "\n")
```

We now compile this model using `CmdStanR`.

```{r compile_amtmodel_flat_stanmodel, echo=TRUE, results="hide"}
amtmodel_flat_stanmodel <- cmdstan_model(
  "stan_code/amtmodel_flat.stan",
  include_paths =   stan_codedir,
  pedantic      =           TRUE,
  dir           =  stan_modeldir
  )
```

We then use this compiled model with our data to produce a fit of the data.

```{r fit_amtmodel_flat_stanmodel, echo=TRUE, cache=TRUE}
stan_modelname <- "amtmodel_flat"
stanfit_prefix <- str_c("fit_", stan_modelname) 

stan_data_lst <- fit_1000_data_tbl %>%
  select(customer_id, tnx_amt = tnx_amount) %>%
  compose_data(
    p_p1  = 100,
    p_p2  =   1,
    nu_p1 =   1,
    nu_p2 =   1
    )

amtmodel_flat_stanfit <- amtmodel_flat_stanmodel$sample(
  data            =                stan_data_lst,
  chains          =                            4,
  iter_warmup     =                          500,
  iter_sampling   =                          500,
  seed            =                         4201,
  save_warmup     =                         TRUE,
  output_dir      =                stan_modeldir,
  output_basename =               stanfit_prefix,
  )

amtmodel_flat_stanfit$summary()
```

We first want to check the HMC diagnostics.

```{r calculate_amtmodel_flat_hmc_diagnostics, echo=TRUE}
amtmodel_flat_stanfit$cmdstan_diagnose()
```


### Visual Diagnostics of the Sample Validity

Now that we have a sample from the posterior distribution we need to create a
few different visualisations of the diagnostics.

```{r plot_p_nu_traceplots_warmup, echo=TRUE}
parameter_subset <- c(
  "p[1]", "nu[1]", "p[2]", "nu[2]", "p[3]", "nu[3]"
  )

amtmodel_flat_stanfit$draws(inc_warmup = TRUE) %>%
  mcmc_trace(
    pars     = parameter_subset,
    n_warmup = 500
    ) +
  ggtitle("Full Traceplots of Some Posterior Values")
```


As the warmup is skewing the y-axis somewhat, we repeat this process without
the warmup.

```{r plot_traceplots_nowarmup, echo=TRUE}
amtmodel_flat_stanfit$draws(inc_warmup = FALSE) %>%
  mcmc_trace(
    pars     = parameter_subset
    ) +
  expand_limits(y = 0) +
  labs(
    x = "Iteration",
    y = "Value",
    title = "Traceplot of Sample of Posterior Values"
    ) +
  theme(axis.text.x = element_text(size = 10))
```

A common MCMC diagnostic is $\hat{R}$ - which is a measure of the 'similarity'
of the chains.

```{r plot_amtmodel_flat_parameter_rhat, echo=TRUE}
amtmodel_flat_stanfit %>%
  rhat(pars = c("p", "nu")) %>%
  mcmc_rhat() +
    ggtitle("Plot of Parameter R-hat Values")
```

Related to this quantity is the concept of *effective sample size*, $N_{eff}$,
an estimate of the size of the sample from a statistical information point of
view.


```{r plot_amtmodel_flat_parameter_neffratio, echo=TRUE}
amtmodel_flat_stanfit %>%
  neff_ratio(pars = c("p", "nu")) %>%
  mcmc_neff() +
    ggtitle("Plot of Parameter Effective Sample Sizes")
```

Finally, we also want to look at autocorrelation in the chains for each
parameter.

```{r plot_amtmodel_flat_parameter_acf, echo=TRUE}
amtmodel_flat_stanfit$draws() %>%
  mcmc_acf(pars = parameter_subset) +
    ggtitle("Autocorrelation Plot of Sample Posterior Values")
```

In practice we do not always require a comprehensive exploration of the
diagnostics, but it is good practice to run through the various visualisations
when we fit a model to ensure our sample is valid.



## Check Model Fit

We now need to check the parameters of this fit against the data to see how
effective our model is at capturing the data. In this case we have the benefit
of knowing the 'true' data, and so we compare our model output against the
input parameters.

```{r construct_amtmodel_flat_validation, echo=TRUE}
amtmodel_flat_validation_tbl <- amtmodel_flat_stanfit %>%
  recover_types(fit_1000_data_tbl) %>%
  spread_draws(p[customer_id], nu[customer_id]) %>%
  ungroup() %>%
  inner_join(
    customer_summarystats_tbl %>% select(customer_id, tnx_count),
    by = "customer_id"
    ) %>%
  inner_join(customer_simparams_tbl, by = "customer_id") %>%
  select(
    customer_id, draw_id = .draw, tnx_count,
    post_p = p, post_nu = nu, customer_p, customer_nu
    )

amtmodel_flat_validation_tbl %>% glimpse()
```

Having constructed the validation data we now want to check the quantile of
each 'true' value in the posterior distribution for the parameter. If our
model is valid, this distribution will be uniform on $[0, 1]$.

```{r calculate_amtmodel_flat_qvalues, echo=TRUE}
plotvalid_tbl <- amtmodel_flat_validation_tbl %>%
  calculate_distribution_qvals(post_p, customer_p, customer_id)

ref_value <- plotvalid_tbl %>% nrow() %>% divide_by(50)

ggplot(plotvalid_tbl) +
  geom_histogram(aes(x = q_val), bins = 50) +
  geom_hline(aes(yintercept = ref_value), colour = "red") +
  labs(
    x = "Quantile",
    y = "Count",
    title = "Quantile Plot of the q-Values for the Posterior Distribution"
    )

ggplot(plotvalid_tbl) +
  geom_point(aes(x = q_val, y = customer_p)) +
  labs(
    x = "Quantile",
    y = "Customer p",
    title = "Scatterplot of q-Value against p"
    )
```





# R Environment

```{r show_session_info, echo=TRUE, message=TRUE}
options(width = 120L)
sessioninfo::session_info()
options(width = 80L)
```
