---
title: "Building the Customer and Product Modelling"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: TRUE
    code_folding: hide
    fig_caption: TRUE

  html_document:
    fig_caption: yes
    theme: spacelab #sandstone #spacelab #flatly
    highlight: pygments
    number_sections: TRUE
    toc: TRUE
    toc_depth: 3
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---


```{r import_libraries, echo=FALSE, message=FALSE}
library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(stringr)
library(glue)
library(purrr)
library(furrr)
library(zoo)
library(arules)
library(arulesViz)
library(DT)
library(tidygraph)
library(rfm)
library(FactoMineR)
library(factoextra)
library(ggpubr)
library(tidytext)
library(ggwordcloud)
library(wordcloud2)


source("lib_utils.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "arules",
    "Matrix", "DT", "zoo")
  )


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallel::detectCores()
  )

theme_set(theme_cowplot())

set.seed(42)

plan(multisession)
```


# Load Data

We first want to load our datasets and prepare them for some simple association
rules mining.

## Load Transaction Data

```{r load_transaction_data, echo=TRUE}
tnx_data_tbl <- read_rds("data/retail_data_cleaned_tbl.rds")

tnx_data_tbl %>% glimpse()
```

To use our rules mining we just need the invoice data and the stock code, so
we can ignore the rest. Also, we ignore the issue of returns and just look at
purchases.

```{r prepare_data_arules, echo=TRUE}
tnx_purchase_tbl <- tnx_data_tbl %>%
  filter(
    quantity > 0,
    price > 0,
    exclude == FALSE
    ) %>%
  drop_na(customer_id) %>%
  select(
    invoice_id, invoice_date, stock_code, customer_id, quantity, price,
    stock_value, description
    )

tnx_purchase_tbl %>% glimpse()
```


To build the association rules, we need to load the transactions in the format
required for the `arules` package. We set a date for our dataset before which
we wish to train our data and use the remainder as our model validation.


```{r set_training_data_date, echo=TRUE}
training_data_date <- as.Date("2011-03-31")
```


We now combine all this data to construct our association rules.


```{r setup_arules_structure, echo=TRUE}
tnx_purchase_tbl %>%
  filter(invoice_date <= training_data_date) %>%
  select(invoice_id, stock_code) %>%
  write_csv("data/tnx_arules_input.csv")

basket_tnxdata <- read.transactions(
    file   = "data/tnx_arules_input.csv",
    format = "single",
    sep    = ",",
    header = TRUE,
    cols   = c("invoice_id", "stock_code")
    )

basket_tnxdata %>% glimpse()
```


## Load Customer Data

We also want to load the various data about the customers such as their cohort
and so on.

```{r load_customer_cohort_data, echo=TRUE}
customer_cohort_tbl <- read_rds("data/customer_cohort_tbl.rds")

customer_cohort_tbl %>% glimpse()
```


## Load Product Data

We also want to load the free-text description of the various stock items as
this will help will interpretation of the various rules we have found.

```{r load_product_data, echo=TRUE}
product_data_tbl <- read_rds("data/stock_description_tbl.rds")

product_data_tbl %>% glimpse()
```





# Build Association Rules Model

We now build our association rules based on the lower support data.


The idea is to repeat some of the initial association rules analysis: we use
the APRIORI algorithm to mine the rules, and then convert the discovered rules
to produce a graph of the products and the rules.

With this graph, we then use the disjoint components of this graph to cluster
the products, and take the largest subgraph and cluster that one according
to some standard clustering.





## Construct Association Rules

Having loaded the individual transaction data we now construct our basket data
and use the APRIORI algorithm to discover our rules.


```{r construct_association_rules, echo=TRUE}
basket_arules <- apriori(
    basket_tnxdata,
    parameter = list(supp = 0.005, conf = 0.01)
  )

basket_arules_tbl <- basket_arules %>%
  as("data.frame") %>%
  as_tibble() %>%
  arrange(desc(lift))

basket_arules_tbl %>% glimpse()
```


Having constructed the main association rules, we then convert the discovered
rules into a graph.


```{r convert_rules_to_graph, echo=TRUE}
apriori_rules_igraph <- basket_arules %>%
  plot(
    measure = "support",
    method  = "graph",
    engine  = "igraph",
    control = list(max = 20000)
    )

apriori_rules_igraph %>% summary()
```

Having constructed the graph, we now want to visualise it.

```{r plot_interactive_rules_graph, echo=TRUE}
basket_arules %>%
  head(n = 500, by = "support") %>%
  plot(
    measure  = "lift",
    method   = "graph",
    engine   = "htmlwidget"
    )
```


## Determine Graph Clusters

With the constructed graph we now want to label the elements that are part
of the disjoint components of the graph.


```{r create_component_labels, echo=TRUE}
apriori_rules_tblgraph <- apriori_rules_igraph %>%
  igraph::as.undirected(mode = "collapse") %>%
  as_tbl_graph() %>%
  mutate(
    component_id = group_components()
    ) %>%
  group_by(component_id) %>%
  mutate(
    component_size = n()
    ) %>%
  ungroup()

apriori_rules_tblgraph %>% print()
```

From the graph, we extract the nodes that correspond to the products (as
opposed to the nodes corresponding to the mined association rules). These are
identified as the various numeric values attached to the rules are blank.

We also wish to add an additional column that is the size of the group, so
it is easier to identify outsized subgraphs suitable for further partitioning.


```{r combine_connected_products, echo=TRUE}
product_cluster_disjoint_tbl <- apriori_rules_tblgraph %>%
  activate(nodes) %>%
  as_tibble() %>%
  filter(are_na(support)) %>%
  group_by(component_id) %>%
  mutate(
    cluster_size = n()
    ) %>%
  ungroup() %>%
  arrange(desc(cluster_size), label) %>%
  group_by(component_id) %>%
  mutate(
    product_group_id = sprintf("AR_DISJOINT_%03d", cur_group_id()),
    cluster_size,
    stock_code       = label
    ) %>%
  ungroup() %>%
  select(product_group_id, cluster_size, stock_code) %>%
  arrange(product_group_id, stock_code)

product_cluster_disjoint_tbl %>% glimpse()
```

We now segment up the largest disjoint subgraph using alternative clustering
techniques.

We try a few different types - inspecting the output of the various algorithms
to see which clustering may be the 



```{r create_largest_subgraph_clusters, echo=TRUE, cache=TRUE}
run_subgraph_clusters <- function(graph_cluster_func, rules_tblgraph, ...) {
  subgraph_clusters_tbl <- rules_tblgraph %>%
    to_subgraph(component_size == max(component_size)) %>%
    use_series(subgraph) %>%
    morph(to_undirected) %>%
    mutate(
      sub_id = graph_cluster_func(...)
      ) %>%
    unmorph() %>%
    activate(nodes) %>%
    as_tibble() %>%
    filter(are_na(support)) %>%
    count(sub_id, name = "cluster_size", sort = TRUE) %>%
    mutate(
      sub_id = factor(1:n(), levels = 1:n())
    )
  
  return(subgraph_clusters_tbl)
}

cluster_func <- c(
    "group_fast_greedy",
    "group_infomap",
    "group_label_prop",
    "group_louvain",
    "group_spinglass"
    )

cluster_data_tbl <- tibble(cluster_func_name = cluster_func) %>%
  mutate(
    cluster_func = map(cluster_func_name, get),
    clustered    = map(cluster_func, run_subgraph_clusters,
                       rules_tblgraph = apriori_rules_tblgraph)
    ) %>%
  select(cluster_func_name, clustered) %>%
  unnest(clustered)

cluster_data_tbl %>% glimpse()
```

Having split this largest component into various splits, we now visualise the
count and size of each cluster and use this to determine which clustering
splits the data into a smaller number of larger clusters.

```{r visualise_cluster_count, echo=TRUE}
ggplot(cluster_data_tbl) +
  geom_col(aes(x = sub_id, y = cluster_size)) +
  geom_hline(aes(yintercept = 5), colour = "red") +
  facet_wrap(vars(cluster_func_name), scales = "free") +
  labs(
    x = "ID",
    y = "Cluster Size"
    ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size = 8))
```




```{r plot_distribution_cluster_sizes, echo=TRUE}
plot_tbl <- cluster_data_tbl %>%
  group_by(cluster_func_name) %>%
  count(cluster_size, name = "cluster_count", sort = TRUE) %>%
  ungroup() %>%
  mutate(cluster_size = as.factor(cluster_size))

ggplot(plot_tbl) +
  geom_col(aes(x = cluster_size, y = cluster_count, group = cluster_size)) +
  facet_wrap(vars(cluster_func_name), scales = "free") +
  labs(
    x = "Cluster Size",
    y = "Community Count",
    title = "Visualisation of Spread of Cluster Sizes"
    ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```


From this, it appears that `louvain` is the method of choice.

Thus, we re-run the clustering for this larger component using the chosen
algorithm and use this to create our various product groups.


```{r construct_fast_greedy_clusters, echo=TRUE}
subgraph_groups_tbl <- apriori_rules_tblgraph %>%
  to_subgraph(component_size == max(component_size)) %>%
  use_series(subgraph) %>%
  morph(to_undirected) %>%
  mutate(
    sub_id = group_louvain()
    ) %>%
  unmorph() %>%
  activate(nodes) %>%
  as_tibble() %>%
  filter(are_na(support)) %>%
  group_by(sub_id) %>%
  mutate(
    cluster_size = n()
    ) %>%
  ungroup() %>%
  arrange(desc(cluster_size), label) %>%
  group_by(sub_id) %>%
  mutate(
    product_group_id = sprintf("AR_LARGE_%03d", cur_group_id()),
    cluster_size,
    stock_code       = label
    ) %>%
  ungroup() %>%
  select(product_group_id, cluster_size, stock_code) %>%
  arrange(product_group_id, stock_code)
  

subgraph_groups_tbl %>% glimpse()
```

We now combine both these lists of groupings and combine them.

```{r combine_product_cluster, echo=TRUE}
product_cluster_tbl <- list(
    product_cluster_disjoint_tbl,
    subgraph_groups_tbl
    ) %>%
  bind_rows() %>%
  filter(product_group_id != "AR_DISJOINT_001")

product_cluster_tbl %>% glimpse()
```



## Assign Products to Groups

We now want to look at our complete list of products and then assign them to
each of our product groups. In terms of coverage, we need to check to see if
all the products appearing in the most invoices.


We also want to look at the most commonly purchased items (in terms of
appearance in baskets as opposed to quantity sold).

```{r construct_popular_product_data, echo=TRUE}
product_popular_tbl <- tnx_purchase_tbl %>%
  mutate(
    stock_code = str_to_upper(stock_code)
    ) %>%
  count(stock_code, name = "invoice_count", sort = TRUE)

product_popular_tbl %>% glimpse()
```


We now combine this data to construct a product dataset containing the
relevant summary data about each product.

```{r construct_product_dataset, echo=TRUE}
product_data_full_tbl <- product_data_tbl %>%
  left_join(product_cluster_tbl, by = "stock_code") %>%
  left_join(product_popular_tbl, by = "stock_code") %>%
  replace_na(
    list(product_group_id = "none", cluster_size = "0")
    ) %>%
  arrange(desc(invoice_count)) %>%
  mutate(ranking = 1:n()) %>%
  semi_join(tnx_purchase_tbl, by = "stock_code") %>%
  arrange(stock_code)

product_data_full_tbl %>% glimpse()
```

First, let us export the table to help us inspect the data.

```{r show_product_data_dt, echo=TRUE}
product_data_full_tbl %>% datatable()
```

To make it more obvious, we look at the products unassigned to a group and
see how they rank in terms of invoice count.

```{r show_unassigned_products, echo=TRUE}
product_data_full_tbl %>% filter(product_group_id == "none") %>% datatable()
```


# Construct Transaction-Based Graph Clusters

We can treat the transaction data as a graph, turning both invoices and stock
items into nodes on the graph, and create an edge between stock and invoices
when the item occurs on the invoice.

We then cluster the graph to create groupings for the different stock codes.


```{r construct_basket_graph, echo=TRUE}
stock_nodes_tbl <- tnx_purchase_tbl %>%
  select(stock_code) %>%
  distinct() %>%
  transmute(node_label = stock_code, node_type = "stock")

invoice_nodes_tbl <- tnx_purchase_tbl %>%
  select(invoice_id) %>%
  distinct() %>%
  transmute(node_label = invoice_id, node_type = "invoice")

nodes_tbl <- list(stock_nodes_tbl, invoice_nodes_tbl) %>%
  bind_rows()

edges_tbl <- tnx_purchase_tbl %>%
  select(stock_code, invoice_id, quantity, price)


basket_tblgraph <- tbl_graph(
  nodes    = nodes_tbl,
  edges    = edges_tbl,
  directed = FALSE,
  node_key = "node_label"
)
```



## Check Graph Clustering Approaches

First we perform our basic clustering by splitting off the different disjoint
components of the graph.

```{r create_disjoint_component_labels, echo=TRUE}
basket_tblgraph <- basket_tblgraph %>%
  mutate(
    component_id = group_components()
    ) %>%
  group_by(component_id) %>%
  mutate(
    component_size = n()
    ) %>%
  ungroup()

basket_tblgraph %>% print()
```

We now want to check the sizes of the disjoint components of this graph.

```{r display_main_component_sizes, echo=TRUE}
basket_tblgraph %>%
  as_tibble() %>%
  filter(node_type == "stock") %>%
  count(component_id, name = "stock_count", sort = TRUE)
```

We see that almost all the stock codes are contained in that one large
component and so confine the rest of this analysis to that one large component.

```{r run_subgraph_clusters, echo=TRUE}
run_subgraph_clusters <- function(graph_cluster_func, labelling, input_tblgraph, ...) {
  message(glue("Clustering the graph using {labelling}..."))
  
  subgraph_clusters_tbl <- input_tblgraph %>%
    mutate(
      cluster_id = graph_cluster_func(...)
      ) %>%
    activate(nodes) %>%
    as_tibble() %>%
    filter(node_type == "stock") %>%
    count(cluster_id, name = "cluster_size", sort = TRUE) %>%
    mutate(
      cluster_id = factor(1:n(), levels = 1:n())
    )
  
  return(subgraph_clusters_tbl)
}
```


```{r test_subgraph_cluster_sizes, echo=TRUE, cache=TRUE}
cluster_func <- c(
    "group_fast_greedy",
    "group_infomap",
    "group_leading_eigen",
    "group_louvain"
    )

largecomp_tblgraph <- basket_tblgraph %>%
  to_subgraph(component_size == max(component_size)) %>%
  use_series(subgraph)

cluster_data_tbl <- tibble(cluster_func_name = cluster_func) %>%
  mutate(
    cluster_func = map(cluster_func_name, get),
    clustered    = map2(
      cluster_func, cluster_func_name,
      run_subgraph_clusters,
      input_tblgraph = largecomp_tblgraph
      )
    ) %>%
  select(cluster_func_name, clustered) %>%
  unnest(clustered)

cluster_data_tbl %>% glimpse()
```

Having created a summary of the data splits, we now want to construct a
visualisation of how the various cluster routines split the data.

To do this, we turn the size of each cluster into a 'label' and then count how
many clusters of that size there are. We then use this summary data to
construct barplots of the size.

```{r visualise_community_splits, echo=TRUE}
plot_tbl <- cluster_data_tbl %>%
  group_by(cluster_func_name) %>%
  count(cluster_size, name = "cluster_count", sort = TRUE) %>%
  ungroup() %>%
  mutate(cluster_size = as.factor(cluster_size))


ggplot(plot_tbl) +
  geom_col(aes(x = cluster_size, y = cluster_count, group = cluster_size)) +
  facet_wrap(vars(cluster_func_name), scales = "free") +
  labs(
    x = "Cluster Size",
    y = "Community Count",
    title = "Visualisation of Spread of Cluster Sizes"
    ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

From this graphic, we see that we want to use `group_louvain` gives us the
most even split across the data - though the sizes are still hugely unequal.


## Create Cluster-Based Allocation

We now use this algorithm to cluster this large component in the graph, and
this gives us an alternative allocation of the each `stock_code` to a product
group.

```{r cluster_largest_component_louvain, echo=TRUE}
largecomp_clustered_tbl <- largecomp_tblgraph %>%
  mutate(
    cluster_id = group_louvain()
    ) %>%
  activate(nodes) %>%
  as_tibble() %>%
  filter(node_type == "stock") %>%
  mutate(
    cluster_group = sprintf("TNX_%03d", cluster_id)
    ) %>%
  select(stock_code = node_label, cluster_group)

largecomp_clustered_tbl %>% glimpse()
```




## Combine Clustering Data

We now want to combine this data to construct our stock code allocations.

```{r combine_clustering_data, echo=TRUE}
other_tbl <- basket_tblgraph %>%
  activate(nodes) %>%
  as_tibble() %>%
  filter(
    node_type == "stock",
    component_size != max(component_size)
    ) %>%
  transmute(
    stock_code = node_label, cluster_group = "TNX_010"
    )

product_group_tnxgroups_tbl <- list(
    largecomp_clustered_tbl,
    other_tbl
    ) %>%
  bind_rows() %>%
  arrange(stock_code) %>%
  inner_join(product_data_tbl, by = "stock_code") %>%
  select(stock_code, product_group = cluster_group, desc)

product_group_tnxgroups_tbl %>% glimpse()
```



# Construct RFM Customer Segments

We now wish to repeat our RFM analysis, and then we reassign the customer base
to each of these groupings.


```{r construct_customer_segments, echo=TRUE}
segment_names <- c(
  "Champions", "Loyal Customers", "Potential Loyalist", "New Customers",
  "Promising", "Need Attention", "About To Sleep", "At Risk",
  "Can't Lose Them", "Lost"
  )

recency_lower   <- c(4, 2, 3, 4, 3, 2, 2, 1, 1, 1)
recency_upper   <- c(5, 5, 5, 5, 4, 3, 3, 2, 1, 2)
frequency_lower <- c(4, 3, 1, 1, 1, 2, 1, 2, 4, 1)
frequency_upper <- c(5, 5, 3, 1, 1, 3, 2, 5, 5, 2)
monetary_lower  <- c(4, 3, 1, 1, 1, 2, 1, 2, 4, 1)
monetary_upper  <- c(5, 5, 3, 1, 1, 3, 2, 5, 5, 2)

segment_defs_tbl <- tibble(
  segment_names,
  recency_lower,
  recency_upper,
  frequency_lower,
  frequency_upper,
  monetary_lower,
  monetary_upper
  )

segment_defs_tbl %>% glimpse()
```

We first visually inspect these segment definitions and the bands.

```{r display_customer_segment_definitions, echo=TRUE}
segments_show_tbl <- segment_defs_tbl %>%
  mutate(
    recency   = glue("{recency_lower}-{recency_upper}")     %>% as.character(),
    frequency = glue("{frequency_lower}-{frequency_upper}") %>% as.character(),
    monetary  = glue("{monetary_lower}-{monetary_upper}")   %>% as.character()
    ) %>%
  select(
    segment_names, recency, frequency, monetary
    )

segments_show_tbl %>%
  datatable(
    colnames = c("Segment", "R", "F", "M"),
    options = list(
      columnDefs = list(list(className = 'dt-left', targets = 0:4))
      )
    )
```

We now construct the RFM data from the purchase data and assign each of the
customers to a segment based on their RFM score.

There is a reasonable number of transactions with a missing `customer_id`, so
we exclude this from the analysis.

```{r construct_basic_rfm_structures, echo=TRUE}
customer_rfmdata <- tnx_purchase_tbl %>%
  filter(
    !are_na(customer_id),
    invoice_date <= training_data_date
    ) %>%
  group_by(invoice_date, customer_id) %>%
  summarise(
    .groups = "drop",
    
    total_spend = sum(stock_value)
    ) %>%
  rfm_table_order(
    customer_id   = customer_id,
    order_date    = invoice_date,
    revenue       = total_spend,
    analysis_date = training_data_date
    )

customer_rfmdata %>% print()
```

## Visualise RFM Data

As we explored earlier, the `rfm` package provides a number of inbuilt
descriptive visualisations.

First we look at the count of customers at each order count:

```{r rfm_order_count_barplot, echo=TRUE}
customer_rfmdata %>%
  rfm_order_dist(print_plot = FALSE)
```


We also have a few summary plots - showing the histograms of the recency,
frequency and monetary measures.

```{r rfm_histograms, echo=TRUE}
customer_rfmdata %>%
  rfm_histograms(print_plot = FALSE) +
    scale_x_continuous(labels = label_comma()) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

Finally, we look at each of the three bivariate plots to explore the
relationship between the three quantities.

```{r plot_bivariate_visualisation, echo=TRUE}
customer_rfmdata %>%
  rfm_rm_plot(print_plot = FALSE) +
    scale_x_log10(labels = label_comma()) +
    scale_y_log10(labels = label_comma())

customer_rfmdata %>%
  rfm_rf_plot(print_plot = FALSE) +
    scale_x_log10(labels = label_comma()) +
    scale_y_log10(labels = label_comma())

customer_rfmdata %>%
  rfm_fm_plot(print_plot = FALSE) +
    scale_x_log10(labels = label_comma()) +
    scale_y_log10(labels = label_comma())
```




## Assign Customer Segments

We now assign each customer to a segment and this allows us to analyse each
of the segments.

```{r segment_customer_base, echo=TRUE}
customer_segments_tbl <- customer_rfmdata %>%
  rfm_segment(
    segment_names   = segment_names,
    recency_lower   = recency_lower,
    recency_upper   = recency_upper,
    frequency_lower = frequency_lower,
    frequency_upper = frequency_upper,
    monetary_lower  = monetary_lower,
    monetary_upper  = monetary_upper
    )

customer_segments_tbl %>% glimpse()
```

We want to plot the count of each of the customer segments, before we calculate
the various summary statistics.

```{r plot_segment_count, echo=TRUE}
customer_segment_count_tbl <- customer_segments_tbl %>%
  count(segment, name = "count", sort = TRUE)

ggplot(customer_segment_count_tbl) +
  geom_col(aes(x = segment, y = count, fill = segment)) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  labs(
    x = "Segment",
    y = "Count"
    ) +
  theme(
    axis.text.x = element_text(angle = 20, vjust = 0.5),
    legend.position = "none"
    )
```

Again, `rfm` provides a number of inbuilt plots of the segments, but as they
create very simple summary plots we create these plots ourselves and this
allows us to summarise the segments however we wish.


```{r create_customer_segment_summary_plots, echo=TRUE}
plot_tbl <- customer_segments_tbl %>%
  select(
    customer_id, segment,
    Transactions  = transaction_count,
    Recency       = recency_days,
    `Total Spend` = amount
    ) %>%
  pivot_longer(
    !c(customer_id, segment),
    names_to = "quantity",
    values_to = "value"
    ) %>%
  mutate(
    value = pmax(0.1, value)
    )

ggplot(plot_tbl) +
  geom_boxplot(aes(x = segment, y = value, fill = segment)) +
  expand_limits(y = 0.1) +
  facet_wrap(vars(quantity), ncol = 2, scales = "free_y") +
  scale_y_log10(labels = label_comma()) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  labs(
    x = "Customer Segment",
    y = "Value"
    ) +
  theme(
    axis.text.x = element_text(angle = 20, vjust = 0.5, size = 8),
    legend.position = "none"
    )
```


## Inspect Segment Validation

Now that we have assigned customers to segments, we use these segments to
assess the transactions made after the cutoff date `training_data_date`.

```{r construct_post_training_data, echo=TRUE}
segments_alloc_tbl <- customer_segments_tbl %>%
  select(customer_id, segment)

daily_spend_tbl <- tnx_purchase_tbl %>%
  filter(invoice_date > training_data_date) %>%
  group_by(invoice_date, customer_id) %>%
  summarise(
    .groups = "drop",
    
    daily_spend = sum(stock_value)
    ) %>%
  left_join(segments_alloc_tbl, by = "customer_id") %>%
  replace_na(list(segment = "New Customer"))

daily_spend_tbl %>% glimpse()
```

Having constructed this data, we now calculate some per-customer summary
statistics.

```{r construct_post_date_stats, echo=TRUE}
postdate_customer_stats_tbl <- daily_spend_tbl %>%
  group_by(customer_id, segment) %>%
  summarise(
    .groups = "drop",
    
    total_transactions = n(),
    total_spend        = sum(daily_spend)
    )

postdate_customer_stats_tbl %>% glimpse()
```


First we compare the segment counts from both pre- and post- dates. The first
metric to check is the proportion of the segment in the dataset, though we
exclude newly arrived customers in the post-date dataset to enable a direct
comparison.

```{r compare_pre_post_segment_data, echo=TRUE}
pre_data_tbl <- customer_segment_count_tbl %>%
  mutate(prop = count / sum(count))

post_data_tbl <- postdate_customer_stats_tbl %>%
  filter(segment != "New Customer") %>%
  count(segment, name = "count", sort = TRUE) %>%
  mutate(prop = count / sum(count))

comparison_tbl <- list(
    Training   = pre_data_tbl,
    Validation = post_data_tbl
    ) %>%
  bind_rows(.id = "data")

ggplot(comparison_tbl) +
  geom_col(aes(x = segment, y = prop, fill = data), position = "dodge") +
  labs(
    x = "Segment",
    y = "Segment Proportion",
    fill = "Dataset"
    ) +
  scale_fill_brewer(type = "qual", palette = "Dark2") +
  theme(axis.text.x = element_text(angle = 20, vjust = 0.5))
```


Next, we check our RFM stats according the validation data as of the final
date in the dataset.

First we construct the data and then construct boxplots of each segment.

```{r calculate_validation_rfm_metrics, echo=TRUE}
validation_date <- daily_spend_tbl %>% pull(invoice_date) %>% max()

validation_rfm_data_tbl <- daily_spend_tbl %>%
  group_by(segment, customer_id) %>%
  summarise(
    .groups = "drop",
    
    transaction_count  = n(),
    recency_days       = (validation_date - max(invoice_date)) %>% as.numeric(),
    amount             = sum(daily_spend)
    )

validation_rfm_data_tbl %>% glimpse()
```

Having constructed the table - we view the data as a JS datatable.

```{r show_validation_rfm_data_dt, echo=TRUE}
validation_rfm_data_tbl %>% datatable()
```

We now produce boxplots of the three metrics using these segments, and also
look at the new customers as a separate category.

```{r create_validation_segment_boxplots, echo=TRUE}
plot_tbl <- validation_rfm_data_tbl %>%
  select(
    customer_id, segment,
    Transactions  = transaction_count,
    Recency       = recency_days,
    `Total Spend` = amount
    ) %>%
  pivot_longer(
    !c(customer_id, segment),
    names_to  = "quantity",
    values_to = "value"
    ) %>%
  mutate(
    value = pmax(0.1, value)
    )

ggplot(plot_tbl) +
  geom_boxplot(aes(x = segment, y = value, fill = segment)) +
  expand_limits(y = 0.1) +
  facet_wrap(vars(quantity), ncol = 2, scales = "free_y") +
  scale_y_log10(labels = label_comma()) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  labs(
    x = "Customer Segment",
    y = "Value"
    ) +
  theme(
    axis.text.x = element_text(angle = 20, vjust = 0.5, size = 8),
    legend.position = "none"
    )
```

As an alternative plot, we also look at the violin plots rather boxplots.

```{r create_validation_segment_violin_plots, echo=TRUE}
ggplot(plot_tbl) +
  geom_violin(aes(x = segment, y = value, fill = segment)) +
  expand_limits(y = 0.1) +
  facet_wrap(vars(quantity), ncol = 2, scales = "free_y") +
  scale_y_log10(labels = label_comma()) +
  scale_fill_brewer(type = "qual", palette = "Set1") +
  labs(
    x = "Customer Segment",
    y = "Value"
    ) +
  theme(
    axis.text.x = element_text(angle = 20, vjust = 0.5, size = 8),
    legend.position = "none"
    )
```

We certainly see that our "Champions" are the 'best' segment in terms of the
three metrics we measure. That said, our new customers also appear to have
desirable metrics on aggregate.




## Simple Product Description Analysis

With each of the products classified into different groups we now inspect the
product descriptions of those products.

We split the product descriptions by `product_group` and construct a word cloud
for each one. Word clouds are a simple starting point to help visualise common
words in a group.

```{r construct_description_word_cloud, echo=TRUE}
create_wordcloud <- function(tokens_tbl, seed = 42) {
  cloud_plot <- tokens_tbl %>%
    count(word, name = "freq", sort = TRUE) %>%
    slice_max(order_by = freq, n = 100) %>%
    ggwordcloud2(seed = seed)
  
  return(cloud_plot)
}

product_group_tokens_tbl <- product_group_tnxgroups_tbl %>%
  group_by(product_group) %>%
  mutate(product_count = n()) %>%
  mutate(label = glue("{product_group} ({product_count})")) %>%
  ungroup() %>%
  unnest_tokens(word, desc)

w_tbl <- product_group_tokens_tbl %>%
  filter(product_group != "TNX_010") %>%
  group_nest(label, product_count, .key = "token_data") %>%
  mutate(
    cloud = map2(token_data, product_count, create_wordcloud)
    )

plot_grid(
    plotlist = w_tbl$cloud,
    labels   = w_tbl$label,
    ncol     = 3
    )
```

We now repeat this exercise but using a slightly lower-level use of the
word cloud data.

```{r create_custom_facetted_word_cloud, echo=TRUE}
wc_input_tbl <- product_group_tokens_tbl %>%
  count(label, word, name = "freq", sort = TRUE) %>%
  group_by(label) %>%
  slice_max(order_by = freq, n = 50) %>%
  ungroup()


ggplot(wc_input_tbl %>% filter(label != "TNX_010 (6)")) +
  geom_text_wordcloud_area(
    aes(label = word, size = freq),
    shape = "square", seed = 42, rm_outside = TRUE) +
  facet_wrap(vars(label), scales = "free", ncol = 3) +
  scale_size_area(max_size = 15) +
  theme_minimal()
```



# Analyse Categorical Co-occurrences

We now need to perform a correspondence analysis (CA) for the co-occurence
of customer grouping and product grouping to see what types of products
are associated with the various groups.


## Construct Customer and Product Allocations

We need to construct a lookup table for all customers in our dataset. Those
customers which first appeared in our 'validation' data are assigned the
grouping "New Customer".


```{r construct_comprehensive_segment_allocation, echo=TRUE}
customer_segment_allocation_tbl <- tnx_purchase_tbl %>%
  drop_na(customer_id) %>%
  select(customer_id) %>%
  distinct() %>%
  left_join(segments_alloc_tbl, by = "customer_id") %>%
  replace_na(list(segment = "New Customer")) %>%
  arrange(customer_id)

customer_segment_allocation_tbl %>% glimpse()
```


We also construct a lookup table allocating each `stock_code` to a
`product_group_id`.

```{r construct_product_group_allocation_table, echo=TRUE}
product_group_allocation_tbl <- product_data_full_tbl %>%
  select(stock_code, product_group_id) %>%
  arrange(stock_code)

product_group_allocation_tbl %>% glimpse()
```


## Construct Co-occurence Analysis

We now wish to add the various product and customer groupings to our
transaction data to construct a table we can later us for some correspondence
analysis. Prior to pursuing library routines to perform this analysis, we first
look at some of the co-occurence frequencies.


```{r construct_correspondence_data, echo=TRUE}
tnx_correspondence_tbl <- tnx_purchase_tbl %>%
  select(
    invoice_id, invoice_date, stock_code, customer_id
    ) %>%
  mutate(
    tnx_ym    = invoice_date %>% format("%Y%m"),
    tnx_dow   = invoice_date %>% format("%A"),
    tnx_month = invoice_date %>% format("%B"),
    tnx_qtr   = invoice_date %>% as.yearqtr() %>% as.character()
    ) %>%
  inner_join(customer_cohort_tbl, by = "customer_id") %>%
  inner_join(customer_segment_allocation_tbl, by = "customer_id") %>%
  inner_join(product_group_tnxgroups_tbl, by = "stock_code")

tnx_correspondence_tbl %>% glimpse()
```

We now want to do some basic analysis based on the relative frequencies of 
the values of both the product groups and the customer segments.

```{r calculate_segment_group_frequency, echo=TRUE}
segment_group_contingency_tbl <- tnx_correspondence_tbl %>%
  filter(
    invoice_date <= training_data_date
    ) %>%
  count(segment, product_group, name = "group_count") %>%
  mutate(
    obs_prop = group_count / sum(group_count)
    )

segment_group_contingency_tbl %>% glimpse()
```


```{r calculate_independence_proportions, echo=TRUE}
segment_props_tbl <- segment_group_contingency_tbl %>%
  count(segment, wt = group_count, name = "segment_count") %>%
  mutate(
    segment_prop = segment_count / sum(segment_count)
    )

prodgroup_props_tbl <- segment_group_contingency_tbl %>%
  count(product_group, wt = group_count, name = "prodgroup_count") %>%
  mutate(
    prodgroup_prop = prodgroup_count / sum(prodgroup_count)
    )

segment_group_independ_tbl <- expand_grid(
    segment_props_tbl,
    prodgroup_props_tbl
    ) %>%
  transmute(
    segment, product_group, segment_count, prodgroup_count,
    independ_prop = segment_prop * prodgroup_prop
    )

segment_group_independ_tbl %>% glimpse()
```

We now compare the observed proportions of these combinations against the
"theoretical" proportions under the assumptions of independence.

```{r construct_combined_proportion_data, echo=TRUE}
segment_group_combined_tbl <- segment_group_independ_tbl %>%
  left_join(segment_group_contingency_tbl, by = c("segment", "product_group")) %>%
  replace_na(list(group_count = 0, obs_prop = 0)) %>%
  mutate(
    ignore = (group_count == 0) | (independ_prop <= 1e-4),
    diff = (obs_prop - independ_prop),
    perc = if_else(ignore == TRUE, 0, (obs_prop / independ_prop) - 1)
    )

segment_group_combined_tbl %>% glimpse()
```

Now that we have combined this data, we construct a heatmap of the proportions.

```{r plot_proportion_differences_heatmap, echo=TRUE}
ggplot(segment_group_combined_tbl) +
  geom_tile(aes(x = segment, y = product_group, fill = diff)) +
  geom_text(aes(x = segment, y = product_group,
                label = label_comma(accuracy = 1)(group_count))) +
  labs(
    x = "Customer Segment",
    y = "Product Group",
    fill = "Diff"
    ) +
  scale_fill_gradient2(low = "blue", high = "red", midpoint = 0) +
  ggtitle("Heatmap of Differences Between Observed and Theoretical Proportions") +
  theme(axis.text.x = element_text(angle = 10, vjust = 0.5))
```

We now make a similar plot, but plotting the relative percentage difference
between the theoretical proportion under assumptions and independence and what
was observed.

```{r plot_proportion_percentage_diffs_heatmap, echo=TRUE}
ggplot(segment_group_combined_tbl) +
  geom_tile(aes(x = segment, y = product_group, fill = perc)) +
  geom_text(aes(x = segment, y = product_group,
                label = label_comma(accuracy = 1)(group_count))) +
  labs(
    x = "Customer Segment",
    y = "Product Group",
    fill = "Perc"
    ) +
  scale_fill_gradient2(low = "blue", high = "red", midpoint = 0) +
  ggtitle("Heatmap of Percentage Differences Between Observed and Theoretical Proportions") +
  theme(axis.text.x = element_text(angle = 10, vjust = 0.5))
```

## Construct Balloon Plot

Another way to visualise this co-occurence of values is to construct a balloon
plot, which shows the counts of each combination of the categorical values
within our dataset.


```{r construct_segment_product_balloonplot, echo=TRUE}
plot_tbl <- tnx_correspondence_tbl %>%
  count(segment, product_group, name = "freq_count")

ggballoonplot(
    plot_tbl,
    show.label = TRUE,
    ggtheme = theme_cowplot()
    ) +
  labs(
    x = "Customer Segment",
    y = "Product Group",
    size = "Count"
    ) +
  theme(axis.text.x = element_text(angle = 20))
```



## Perform Basic Correspondence Analysis

To run the various routines to calculate the correspondence analysis we need
to convert our data into a matrix format required by the libraries.

To do this efficiently, we construct a quick function to convert these data
tables into a matrix with the first column becoming the row name for the
matrix.


```{r make_df_matrix, echo=TRUE}
make_df_matrix <- function(data_tbl) {
  seg_names <- data_tbl %>% pull(1)
  
  data_mat <- data_tbl %>%
    select(-1) %>%
    as.matrix() %>%
    set_rownames(seg_names)
  
  return(data_mat)
}
```


```{r create_segment_group_frequency_data, echo=TRUE}
segment_group_freq_tbl <- tnx_correspondence_tbl %>%
  count(segment, product_group, name = "freq_count") %>%
  pivot_wider(
    id_cols     = segment,
    names_from  = product_group,
    values_from = freq_count,
    values_fill = 0
    )

segment_group_mat <- segment_group_freq_tbl %>%
  make_df_matrix()

segment_group_mat %>% glimpse()
```

Now that we have the frequency matrix we can run the `CA` routine and visualise
the outputs.

```{r calculate_segment_group_ca, echo=TRUE}
segment_group_ca <- segment_group_mat %>% CA(graph = FALSE)
```

First we visualise the variance determined from the eigenvalues of the matrix.
This gives us an indication of the 'true' underlying dimensionality of the
data.

```{r visualise_ca_eigenvalues_variance, echo=TRUE}
segment_group_ca %>% fviz_eig()
```

From this plot, we see that we can capture almost 94% of the variance with
just two dimensions.

We then construct the biplot - which is a way of visualising the co-occurrences
of each of the categorical values.

```{r visualise_ca_biplots, echo=TRUE}
segment_group_ca %>%
  fviz_ca_biplot(
    repel = TRUE,
    title = "CA Biplot of Customer Segment Against Product Group"
    )
```


According to the biplots, there is a suggested relationship between customers
in the "Champions" category and those products in grouping "TNX_009" - we
ignore "TNX_010" as it is very small.

As before, let us look at a wordcloud on the types of items in that.

```{r plot_tnx_009_word_cloud, echo=TRUE}
wc_009_tbl <- product_group_tokens_tbl %>%
  filter(product_group == "TNX_009") %>%
  count(word, name = "freq") %>%
  slice_max(order_by = freq, n = 100)

wc_plot <- ggwordcloud2(
    data    = wc_009_tbl,
    shuffle = FALSE,
    size    = 4,
    seed    = 42421
    )

wc_plot %>% plot()
```

We also see that customers designated "Potential Loyalist" are connected to
"TNX_005". So, we show a word cloud for this group also.

```{r plot_tnx_005_word_cloud, echo=TRUE}
wc_005_tbl <- product_group_tokens_tbl %>%
  filter(product_group == "TNX_005") %>%
  count(word, name = "freq") %>%
  slice_max(order_by = freq, n = 100)

wc_plot <- ggwordcloud2(
    data    = wc_005_tbl,
    shuffle = FALSE,
    size    = 4,
    seed    = 42422
    )

wc_plot %>% plot()
```


# R Environment
 
```{r show_session_info, echo=TRUE, message=TRUE}
sessioninfo::session_info()
```
