---
title: "Alternative Explorations of the Online Retail Dataset"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: yes

  html_document:
    fig_caption: yes
    theme: spacelab #sandstone #spacelab #flatly
    highlight: pygments
    number_sections: TRUE
    toc: TRUE
    toc_depth: 2
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---


```{r import_libraries, echo=FALSE, message=FALSE}
library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(vctrs)
library(fs)
library(glue)
library(forcats)
library(snakecase)
library(lubridate)
library(evir)
library(DT)
library(tidygraph)



source("custom_functions.R")

resolve_conflicts(c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2"))


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallel::detectCores()
  )

theme_set(theme_cowplot())

set.seed(42)
```


# Basket Analysis

We want to do some basic basket analysis before we investigate using full
'association rules' libraries for our modelling.

Rather than repeating all the logic, it might be worth looking at building
some basic code to construct all the item pairs in each basket.


```{r construct_basket_pairs, echo=TRUE}
construct_pairwise_items <- function(stock_codes) {
  pairs_tbl <- expand_grid(
      one = stock_codes,
      two = stock_codes
      ) %>%
    filter(one < two)
  
  return(pairs_tbl)
}

basket_data_tbl <- data_tbl %>%
  group_by(invoice_id) %>%
  summarise(
    .groups = "drop",
    
    stock_codes = list(stock_code)
    ) %>%
  mutate(
    basket_data = future_map(stock_codes, construct_pairwise_items,
                             .progress = TRUE
                             )
    )

basket_data_tbl %>% glimpse()
```


# Graph Analysis

We can treat this data as a graph, turning both invoices and stock items into
nodes on the graph, and create a connection between stock and invoices when the
item occurs on the invoice.

This graph will get large, but it is a starting point.


```{r construct_basket_graph, echo=TRUE}
stock_nodes_tbl <- data_tbl %>%
  select(stock_code) %>%
  distinct() %>%
  transmute(node_label = stock_code, node_type = "stock")

invoice_nodes_tbl <- data_tbl %>%
  select(invoice_id) %>%
  distinct() %>%
  transmute(node_label = invoice_id, node_type = "invoice")

nodes_tbl <- list(stock_nodes_tbl, invoice_nodes_tbl) %>%
  bind_rows()

edges_tbl <- data_tbl %>%
  select(stock_code, invoice_id, quantity, price)


basket_graph <- tbl_graph(
  nodes    = nodes_tbl,
  edges    = edges_tbl,
  directed = FALSE
)
```





# R Environment

```{r show_session_info, echo=TRUE, message=TRUE}
sessioninfo::session_info()
```
