---
title: "Exploring the Online Retail Dataset"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: yes

  html_document:
    fig_caption: yes
    theme: spacelab #sandstone #spacelab #flatly
    highlight: pygments
    number_sections: TRUE
    toc: TRUE
    toc_depth: 2
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---


```{r import_libraries, echo=FALSE, message=FALSE}
library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(furrr)
library(vctrs)
library(fs)
library(glue)
library(forcats)
library(snakecase)
library(lubridate)
library(evir)
library(DT)
library(tidygraph)



source("custom_functions.R")

resolve_conflicts(c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2"))


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallel::detectCores()
  )

theme_set(theme_cowplot())

set.seed(42)

plan(multisession)
```


```{r custom_functions, echo=FALSE}
### Checks if variable is a date/time
is_date <- function(x)
  x %>% inherits(c("POSIXt", "POSIXct", "POSIXlt", "Date", "hms"))


### Returns the category of data type passed to it
categorise_datatype <- function(x) {
  if (all(are_na(x))) return("na")

  if (is_date(x))                          "datetime"
  else if (!is_null(attributes(x)) ||
           all(is_character(x)))          "discrete"
  else if (all(is_logical(x)))            "logical"
  else                                    "continuous"
}


### create_coltype_list() splits columns into various types
create_coltype_list <- function(data_tbl) {
  coltypes  <- data_tbl %>% map_chr(categorise_datatype)
  cat_types <- coltypes %>% unique() %>% sort()

  split_lst <- cat_types %>% map(~ coltypes[coltypes %in% .x] %>% names())

  names(split_lst) <- coltypes %>% unique() %>% sort()

  coltype_lst <- list(
    split   = split_lst,
    columns = coltypes
  )

  return(coltype_lst)
}

```


This workbook was created using the "dataexpks" template:

https://github.com/DublinLearningGroup/dataexpks



# Introduction

This workbook performs the basic data exploration of the dataset.

```{r set_exploration_params, echo=TRUE}
level_exclusion_threshold <- 100

cat_level_count <- 40
hist_bins_count <- 50
```


# Load Data

First we load the dataset as well as some support datasets.

```{r load_dataset, echo=TRUE}
rawdata_tbl <- read_rds("data/retail_data_raw_tbl.rds")

rawdata_tbl %>% glimpse()
```


## Perform Quick Data Cleaning

Some of the dates provided in the dataset are in an irregular format.

```{r clean_names, echo=TRUE}
data_tbl <- rawdata_tbl %>% set_colnames(names(.) %>% to_snake_case())

data_tbl %>% glimpse()
```


```{r, echo=FALSE}
#knitr::knit_exit()
```


## Create Derived Variables

We now create derived features useful for modelling. These values are
new variables calculated from existing variables in the data.

```{r construct_derived_values, echo=FALSE}
data_tbl <- data_tbl %>%
  rename(invoice_id = invoice) %>%
  mutate(
    row_id = sprintf("ROW%07d", 1:n()),
    
    .before = 1
    ) %>%
  mutate(
    stock_code_upr = stock_code %>% str_to_upper(),
    
    cancellation   = str_detect(invoice_id, "^C"),
    invoice_dttm   = invoice_date,
    invoice_date   = invoice_date %>% as.Date(),
    invoice_month  = format(invoice_dttm, "%B"),
    invoice_dow    = format(invoice_dttm, "%A"),
    invoice_dom    = format(invoice_dttm, "%d"),
    invoice_hour   = format(invoice_dttm, "%H"),
    invoice_minute = format(invoice_dttm, "%M"),
    invoice_woy    = format(invoice_dttm, "%V"),
    invoice_ym     = format(invoice_dttm, "%Y%m"),

    stock_amount   = price * quantity
    ) %>%
  group_by(invoice_ym) %>%
  mutate(
    invoice_monthprop = as.numeric(invoice_dom) / max(as.numeric(invoice_dom))
    ) %>%
  ungroup() %>%
  arrange(invoice_dttm)

data_tbl %>% glimpse()
```



# Perform Basic Checks on Data

## Check/Remove Duplicated Data

Now that we have data across multiple sheets in the Excel file we need to see
if the data is duplicated across the sheets.

```{r check_duplicated}
data_tbl %>%
  group_by(excel_sheet) %>%
  summarise(
    .groups = "drop",
    
    min_date = min(invoice_date),
    max_date = max(invoice_date)
    ) %>%
  print()
```

There appears to be some overlap for the two sheets for the first 10 days or so
of December. It is possible this data is not duplicated though, so we will
check that.


```{r check_duplicated_entries, echo=TRUE}
data_tbl %>%
  group_by(excel_sheet, invoice_date, invoice_id, stock_code) %>%
  mutate(row_id = 1:n()) %>%
  ungroup() %>%
  filter(row_id > 1) %>%
  arrange(invoice_date, invoice_id, invoice_dttm, excel_sheet)
```

It looks like there are about 24,000 rows of data that are duplicated, so we
can remove them.

```{r remove_duplicated_entries, echo=TRUE}
data_tbl <- data_tbl %>%
  group_by(invoice_date, invoice_id, stock_code) %>%
  mutate(dup_id = 1:n()) %>%
  ungroup() %>%
  filter(dup_id == 1) %>%
  select(-dup_id)

data_tbl %>% glimpse()
```

## Check Dates

We now want to check for missing dates in this dataset.

```{r check_missing_dates, echo=TRUE}
data_tbl %>%
  select(invoice_date, invoice_dow) %>%
  distinct() %>%
  mutate(date_diff = invoice_date - lag(invoice_date)) %>%
  filter(date_diff > 1) %>%
  datatable()
```

It appears that there are no missing dates in this dataset, though we may need
to account for the fact that the business is not open all the time as there
are gaps in dates for one day every weekend, as well as being closed over the
Easter and Christmas holidays.

For our time series analysis on a daily basis we may be able to ignore this
wrinkle in our dataset.



## Check Missing Values

Before we do anything with the data, we first check for missing values
in the dataset. In some cases, missing data is coded by a special
character rather than as a blank, so we first correct for this.

```{r replace_missing_character, echo=TRUE}
### _TEMPLATE_
### ADD CODE TO CORRECT FOR DATA ENCODING HERE
```

With missing data properly encoded, we now visualise the missing data in a
number of different ways.

### Univariate Missing Data

We first examine a simple univariate count of all the missing data:

```{r missing_data_univariate_count, echo=TRUE}
row_count <- data_tbl %>% nrow()

missing_univariate_tbl <- data_tbl %>%
  summarise_all(list(~sum(are_na(.)))) %>%
  gather("variable", "missing_count") %>%
  mutate(missing_prop = missing_count / row_count)

ggplot(missing_univariate_tbl) +
  geom_bar(aes(x = fct_reorder(variable, -missing_prop),
               weight = missing_prop)) +
  xlab("Variable") +
  ylab("Missing Value Proportion") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

We remove all variables where all of the entries are missing

```{r remove_entirely_missing_vars, echo=TRUE}
remove_vars <- missing_univariate_tbl %>%
  filter(missing_count == row_count) %>%
  pull(variable)

lessmiss_data_tbl <- data_tbl %>%
  select(-one_of(remove_vars))
```

With these columns removed, we repeat the exercise.

```{r missing_data_univariate_count_redux, echo=TRUE}
missing_univariate_tbl <- lessmiss_data_tbl %>%
  summarise_all(list(~sum(are_na(.)))) %>%
  gather("variable", "missing_count") %>%
  mutate(missing_prop = missing_count / row_count)

ggplot(missing_univariate_tbl) +
  geom_bar(aes(x = fct_reorder(variable, -missing_prop),
               weight = missing_prop)) +
  xlab("Variable") +
  ylab("Missing Value Proportion") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```


To reduce the scale of this plot, we look at the top twenty missing data
counts.

```{r missing_data_univariate_top10_count, echo=TRUE}
missing_univariate_top_tbl <- missing_univariate_tbl %>%
  arrange(desc(missing_count)) %>%
  top_n(n = 50, wt = missing_count)

ggplot(missing_univariate_top_tbl) +
  geom_bar(aes(x = fct_reorder(variable, -missing_prop),
               weight = missing_prop)) +
  xlab("Variable") +
  ylab("Missing Value Proportion") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```



### Multivariate Missing Data

It is useful to get an idea of what combinations of variables tend to have
variables with missing values simultaneously, so to construct a visualisation
for this we create a count of all the times given combinations of variables
have missing values, producing a heat map for these combination counts.

```{r missing_data_matrix, echo=TRUE}
row_count <- rawdata_tbl %>% nrow()

count_nas <- ~ .x %>% are_na() %>% vec_cast(integer())

missing_plot_tbl <- rawdata_tbl %>%
  mutate_all(count_nas) %>%
  mutate(label = pmap_chr(., str_c)) %>%
  group_by(label) %>%
  summarise_all(list(sum)) %>%
  arrange(desc(label)) %>%
  select(-label) %>%
  mutate(label_count = pmap_int(., pmax)) %>%
  gather("col", "count", -label_count) %>%
  mutate(miss_prop   = count / row_count,
         group_label = sprintf("%6.4f", round(label_count / row_count, 4))
        )

ggplot(missing_plot_tbl) +
  geom_tile(aes(x = col, y = group_label, fill = miss_prop), height = 0.8) +
  scale_fill_continuous() +
  scale_x_discrete(position = "top") +
  xlab("Variable") +
  ylab("Missing Value Proportion") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```

This visualisation takes a little explaining.

Each row represents a combination of variables with simultaneous missing
values. For each row in the graphic, the coloured entries show which particular
variables are missing in that combination. The proportion of rows with that
combination is displayed in both the label for the row and the colouring for
the cells in the row.

## Inspect High-level-count Categorical Variables

With the raw data loaded up we now remove obvious unique or near-unique
variables that are not amenable to basic exploration and plotting.

```{r find_highlevelcount_categorical_variables, echo=TRUE}
coltype_lst <- create_coltype_list(data_tbl)

count_levels <- ~ .x %>% unique() %>% length()

catvar_valuecount_tbl <- data_tbl %>%
  summarise_at(coltype_lst$split$discrete, count_levels) %>%
  gather("var_name", "level_count") %>%
  arrange(-level_count)

print(catvar_valuecount_tbl)

row_count <- data_tbl %>% nrow()

cat(glue("Dataset has {row_count} rows\n"))
```

Now that we a table of the counts of all the categorical variables we can
automatically exclude unique variables from the exploration, as the level
count will match the row count.

```{r remove_id_variables, echo=TRUE}
unique_vars <- catvar_valuecount_tbl %>%
  filter(level_count == row_count) %>%
  pull(var_name)

print(unique_vars)

explore_data_tbl <- data_tbl %>%
  select(-one_of(unique_vars))
```

Having removed the unique identifier variables from the dataset, we
may also wish to exclude categoricals with high level counts also, so
we create a vector of those variable names.

```{r collect_highcount_variables, echo=TRUE}
highcount_vars <- catvar_valuecount_tbl %>%
  filter(level_count >= level_exclusion_threshold,
         level_count < row_count) %>%
  pull(var_name)

cat(str_c(highcount_vars, collapse = ", "))
```

We now can continue doing some basic exploration of the data. We may
also choose to remove some extra columns from the dataset.

```{r drop_variables, echo=TRUE}
### You may want to comment out these next few lines to customise which
### categoricals are kept in the exploration.
drop_vars <- c(highcount_vars)

if (length(drop_vars) > 0) {
  explore_data_tbl <- explore_data_tbl %>%
      select(-one_of(drop_vars))

  cat(str_c(drop_vars, collapse = ", "))
}
```


```{r write_to_disk, echo=TRUE}
data_tbl %>% write_rds("data/retail_data_tbl.rds")
```

```{r, echo=FALSE}
#knitr::knit_exit()
```



# Univariate Data Exploration

Now that we have loaded the data we can prepare it for some basic data
exploration. We first exclude the variables that are unique
identifiers or similar, and tehen split the remaining variables out
into various categories to help with the systematic data exploration.


```{r separate_exploration_cols, echo=TRUE}
coltype_lst <- create_coltype_list(explore_data_tbl)

print(coltype_lst)
```


## Logical Variables

Logical variables only take two values: TRUE or FALSE. It is useful to see
missing data as well though, so we also plot the count of those.

```{r create_univariate_logical_plots, echo=TRUE, warning=FALSE}
logical_vars <- coltype_lst$split$logical %>% sort()

for (plot_varname in logical_vars) {
  cat("--\n")
  cat(glue("{plot_varname}\n"))

  na_count <- explore_data_tbl %>% pull(!! plot_varname) %>% are_na() %>% sum()

  explore_plot <- ggplot(explore_data_tbl) +
    geom_bar(aes(x = !! sym(plot_varname))) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(str_c("Barplot of Counts for Variable: ", plot_varname,
                  " (", na_count, " missing values)")) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

  plot(explore_plot)
}
```


## Numeric Variables

Numeric variables are usually continuous in nature, though we also have
integer data.

```{r create_univariate_numeric_plots, echo=TRUE, warning=FALSE}
numeric_vars <- coltype_lst$split$continuous %>% sort()

for (plot_varname in numeric_vars) {
  cat("--\n")
  cat(glue("{plot_varname}\n"))

  plot_var <- explore_data_tbl %>% pull(!! plot_varname)
  na_count <- plot_var %>% are_na() %>% sum()

  plot_var %>% summary() %>% print()

  neg_data_tbl <- explore_data_tbl %>%
    filter(!! sym(plot_varname) < 0) %>%
    mutate(var_val = abs(!! sym(plot_varname)))
  
  pos_data_tbl <- explore_data_tbl %>%
    filter(!! sym(plot_varname) >= 0) %>%
    mutate(var_val = abs(!! sym(plot_varname)))

  all_plot <- ggplot(explore_data_tbl) +
    geom_histogram(aes(x = !! sym(plot_varname)), bins = hist_bins_count) +
    geom_vline(xintercept = mean(plot_var, na.rm = TRUE),
               colour = "red",   size = 1.5) +
    geom_vline(xintercept = median(plot_var, na.rm = TRUE),
               colour = "green", size = 1.5) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(glue("Histogram Plot for Variable: {plot_varname} ({na_count} missing values)"),
            subtitle = "(red line is mean, green line is median)")


  pos_log_plot <- ggplot(pos_data_tbl) +
    geom_histogram(aes(x = var_val), bins = hist_bins_count) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_x_log10(labels = label_comma()) +
    scale_y_continuous(labels = label_comma()) +
    ggtitle("Positive Values")

  neg_log_plot <- ggplot(neg_data_tbl) +
    geom_histogram(aes(x = var_val), bins = hist_bins_count) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_x_log10(labels = label_comma()) +
    scale_y_continuous(labels = label_comma()) +
    ggtitle("Negative Values")


  plot_grid(
      all_plot,
      NULL,
      pos_log_plot,
      neg_log_plot,
      nrow = 2
      ) %>%
    print()
}
```

## Categorical Variables

Categorical variables only have values from a limited, and usually fixed,
number of possible values

```{r create_univariate_categorical_plots, echo=TRUE, warning=FALSE}
categorical_vars <- coltype_lst$split$discrete %>% sort()

for (plot_varname in categorical_vars) {
  cat("--\n")
  cat(str_c(plot_varname, "\n"))

  na_count <- explore_data_tbl %>% pull(!! plot_varname) %>% are_na() %>% sum()

  plot_tbl <- explore_data_tbl %>%
    pull(!! plot_varname) %>%
    fct_lump(n = cat_level_count) %>%
    fct_count() %>%
    mutate(f = fct_relabel(f, str_trunc, width = 15))

  explore_plot <- ggplot(plot_tbl) +
    geom_bar(aes(x = f, weight = n)) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(str_c("Barplot of Counts for Variable: ", plot_varname,
                  " (", na_count, " missing values)")) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

  plot(explore_plot)
}
```


## Date/Time Variables

Date/Time variables represent calendar or time-based data should as time of the
day, a date, or a timestamp.

```{r create_univariate_datetime_plots, echo=TRUE, warning=FALSE}
datetime_vars <- coltype_lst$split$datetime %>% sort()

for (plot_varname in datetime_vars) {
  cat("--\n")
  cat(str_c(plot_varname, "\n"))

  plot_var <- explore_data_tbl %>% pull(!! plot_varname)
  na_count <- plot_var %>% are_na() %>% sum()

  plot_var %>% summary() %>% print()

  explore_plot <- ggplot(explore_data_tbl) +
    geom_histogram(aes(x = !! sym(plot_varname)),
                   bins = hist_bins_count) +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(str_c("Barplot of Dates/Times in Variable: ", plot_varname,
                  " (", na_count, " missing values)"))

  plot(explore_plot)
}
```


```{r, echo=FALSE}
#knitr::knit_exit()
```


# Bivariate Data Exploration

We now move on to looking at bivariate plots of the data set.

## Facet Plots on Variables

A natural way to explore relationships in data is to create univariate
visualisations facetted by a categorical value.

```{r bivariate_facet_data, echo=TRUE}
facet_varname <- "excel_sheet"

facet_count_max <- 3
```


### Logical Variables

For logical variables we facet on barplots of the levels, comparing TRUE,
FALSE and missing data.

```{r create_bivariate_logical_plots, echo=TRUE}
logical_vars <- logical_vars[!logical_vars %in% facet_varname] %>% sort()


for (plot_varname in logical_vars) {
  cat("--\n")
  cat(str_c(plot_varname, "\n"))

  plot_tbl <- data_tbl %>% filter(!are_na(!! plot_varname))

  explore_plot <- ggplot(plot_tbl) +
    geom_bar(aes(x = !! sym(plot_varname))) +
    facet_wrap(facet_varname, scales = "free") +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(str_c(facet_varname, "-Faceted Barplots for Variable: ",
                  plot_varname)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

  plot(explore_plot)
}
```


### Numeric Variables

For numeric variables, we facet on histograms of the data.

```{r create_bivariate_numeric_plots, echo=TRUE}
for (plot_varname in numeric_vars) {
  cat("--\n")
  cat(str_c(plot_varname, "\n"))

  plot_tbl <- data_tbl %>% filter(!are_na(!! plot_varname))

  explore_plot <- ggplot(plot_tbl) +
    geom_histogram(aes(x = !! sym(plot_varname)),
                   bins = hist_bins_count) +
    facet_wrap(facet_varname, scales = "free") +
    xlab(plot_varname) +
    ylab("Count") +
    scale_x_continuous(labels = label_comma()) +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(str_c(facet_varname, "-Faceted Histogram for Variable: ",
                  plot_varname)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

  print(explore_plot)
}
```

### Categorical Variables

We treat categorical variables like logical variables, faceting the barplots
of the different levels of the data.

```{r create_bivariate_categorical_plots, echo=TRUE}
categorical_vars <- categorical_vars[!categorical_vars %in% facet_varname] %>% sort()

for (plot_varname in categorical_vars) {
  cat("--\n")
  cat(str_c(plot_varname, "\n"))

  plot_tbl <- data_tbl %>%
    filter(!are_na(!! plot_varname)) %>%
    mutate(
      varname_trunc = fct_relabel(!! sym(plot_varname), str_trunc, width = 10)
      )
  
  explore_plot <- ggplot(plot_tbl) +
    geom_bar(aes(x = varname_trunc)) +
    facet_wrap(facet_varname, scales = "free") +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(str_c(facet_varname, "-Faceted Histogram for Variable: ",
                  plot_varname)) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5))

  plot(explore_plot)
}
```


### Date/Time Variables

Like the univariate plots, we facet on histograms of the years in the dates.

```{r create_bivariate_datetime_plots, echo=TRUE}
for (plot_varname in datetime_vars) {
  cat("--\n")
  cat(str_c(plot_varname, "\n"))

  plot_tbl <- data_tbl %>% filter(!are_na(!! plot_varname))

  explore_plot <- ggplot(plot_tbl) +
    geom_histogram(aes(x = !! sym(plot_varname)),
                   bins = hist_bins_count) +
    facet_wrap(facet_varname, scales = "free") +
    xlab(plot_varname) +
    ylab("Count") +
    scale_y_continuous(labels = label_comma()) +
    ggtitle(str_c(facet_varname, "-Faceted Histogram for Variable: ",
                  plot_varname))

  plot(explore_plot)
}
```

```{r free_memory_facetplot, echo=FALSE}
rm(plot_var, plot_tbl)
```


```{r, echo=FALSE}
#knitr::knit_exit()
```


# Custom Explorations

In this section we perform various data explorations.


## Custom Checks for Data Integrity

We want to check the transaction data for consistency, so we create a table
of all distinct 

```{r check_stock_codes, echo=TRUE}
stock_codes_lookup_tbl <- data_tbl %>%
  select(stock_code_upr, description) %>%
  distinct() %>%
  arrange(stock_code_upr, description) %>%
  drop_na(description)

stock_codes_lookup_tbl %>% glimpse()
```

We now take a look at the first 50 rows of this table to get a sense of any
possible duplication of `stock_code`.

```{r plot_stock_codes_distinct_50, echo=TRUE}
stock_codes_lookup_tbl %>% datatable()
```


### Items per Transactions

As another check on the data, we want to look at how many different objects
are included in 

```{r plot_histogram_distinct_items, echo=TRUE}
plot_tbl <- data_tbl %>%
  filter(quantity > 0) %>%
  count(invoice_id, name = "n_items")

ggplot(plot_tbl) +
  geom_histogram(aes(x = n_items), bins = 40) +
  scale_x_log10(labels = label_comma()) +
  scale_y_continuous(labels = label_comma()) +
  xlab("Number of Items") +
  ylab("Transaction Count") +
  ggtitle("Histogram of Item Counts per Transactions")
```


## Explore Aggregate Amounts

We now turn our focus to aggregating the data set in various ways and inspect
how these aggregate totals are distributed.


### Invoice-Level Amounts

We first aggregate the data at the invoice level, and inspect how those amounts
are distributed.

```{r explore_invoice_amounts, echo=TRUE}
invoice_data_tbl <- data_tbl %>%
  group_by(invoice_id) %>%
  summarise(
    .groups = "drop",
    
    invoice_amount = sum(price * quantity) %>% round(2)
  )

invoice_mean   <- invoice_data_tbl %>% pull(invoice_amount) %>% mean()   %>% round(2)
invoice_median <- invoice_data_tbl %>% pull(invoice_amount) %>% median() %>% round(2)

ggplot(invoice_data_tbl) +
  geom_histogram(aes(x = invoice_amount), bins = 50) +
  geom_vline(aes(xintercept = invoice_mean),   colour = "black") +
  geom_vline(aes(xintercept = invoice_median), colour = "red") +
  xlab("Invoice Amount") +
  ylab("Count") +
  scale_x_log10(labels = label_comma()) +
  scale_y_continuous(labels = label_comma()) +
  ggtitle(
    label    = "Histogram Plot for Invoice Amount",
    subtitle = glue("Mean is {invoice_mean}, Median is {invoice_median}")
    )
```

We see there is a broad range of different invoice totals, with mean and
median being a few hundred pounds.



### Customer-Level Amounts


```{r explore_customer_amounts, echo=TRUE}
customer_data_tbl <- data_tbl %>%
  group_by(customer_id) %>%
  summarise(
    .groups = "drop",
    
    customer_spend = sum(price * quantity) %>% round(2)
  )
  
ggplot(customer_data_tbl) +
  geom_histogram(aes(x = customer_spend), bins = 50) +
  xlab("Customer Spend") +
  ylab("Count") +
  scale_x_log10(labels = label_comma()) +
  scale_y_continuous(labels = label_comma()) +
  ggtitle("Histogram Plot for Customer Spend")
```


```{r customer_spend_hill_plot, echo=TRUE}
customer_data_tbl %>% pull(customer_spend) %>% hill()
```


### Stock Level Data

```{r plot_stock_price_counts, echo=TRUE}
stock_price_counts_tbl <- data_tbl %>%
  group_by(stock_code) %>%
  summarise(
    .groups = "drop",
    
    n_prices    = n(),
    min_price   = min(price),
    p25_price   = quantile(price, 0.25),
    mean_price  = mean(price) %>% round(2),
    p50_price   = median(price),
    p75_price   = quantile(price, 0.75),
    max_price   = max(price),
    range_price = ((max_price - min_price) / mean_price) %>% round(4)
    )

stock_price_counts_tbl %>% datatable()
  
stock_distinct_price_counts_tbl <- data_tbl %>%
  select(stock_code, price) %>%
  distinct() %>%
  group_by(stock_code) %>%
  summarise(
    .groups = "drop",
    
    n_prices    = n(),
    min_price   = min(price),
    p25_price   = quantile(price, 0.25),
    mean_price  = mean(price) %>% round(2),
    p50_price   = median(price),
    p75_price   = quantile(price, 0.75),
    max_price   = max(price),
    range_price = (max_price - min_price) / mean_price
    )

stock_distinct_price_counts_tbl %>% datatable()
```





## Construct Time-Series / Date-Based Data

Another way to look at this data is to combine all the invoice values by
various time period such as daily, weekly and monthly to see how it looks.

As we are going to do a number of aggregations based on various aspects of the
date, we construct a function that takes a table of data and adds a number of
derived columns based on that date: things like day of week, calendar month and
so on.

```{r append_calendar_columns, echo=TRUE}
append_calendar_columns <- function(data_tbl, date_col) {
  updated_data_tbl <- data_tbl %>%
    mutate(
      invoice_date   = {{date_col}} %>% as.Date(),
      
      invoice_month  = {{date_col}} %>% format("%B"),
      invoice_dow    = {{date_col}} %>% format("%A"),
      invoice_dom    = {{date_col}} %>% format("%d"),
      invoice_hour   = {{date_col}} %>% format("%H"),
      invoice_minute = {{date_col}} %>% format("%M"),
      invoice_woy    = {{date_col}} %>% format("%V"),
      invoice_ym     = {{date_col}} %>% format("%Y%m"),
      
      .after = {{date_col}}
      )
  
  return(updated_data_tbl)
}

data_tbl %>%
  select(excel_sheet, invoice_id, invoice_date, stock_amount) %>%
  append_calendar_columns(invoice_date) %>%
  glimpse()


```


### Create Univariate Time-Series of Amounts

```{r construct_timeseries_data, echo=TRUE}
ts_data_tbl <- data_tbl %>%
  mutate(
    ts_week  = format(invoice_dttm, "%Y-%U"),
    ts_month = format(invoice_dttm, "%Y-%m")
    )

ts_daily_tbl <- ts_data_tbl %>%
  group_by(label = invoice_date %>% format("%Y-%m-%d")) %>%
  summarise(
    .groups = "drop",

    period_date = min(invoice_date),
    total_spend = sum(price * quantity) %>% round(2)
    )

ggplot(ts_daily_tbl) +
  geom_line(aes(x = period_date, y = total_spend)) +
  expand_limits(y = 0) +
  scale_y_continuous(labels = label_comma()) +
  xlab("Date") +
  ylab("Total Spend") +
  ggtitle("Lineplot of Total Spend by Day")


ts_weekly_tbl <- ts_data_tbl %>%
  group_by(label = ts_week) %>%
  summarise(
    .groups = "drop",

    period_date = min(invoice_date),
    total_spend = sum(price * quantity)
    )

ggplot(ts_weekly_tbl) +
  geom_line(aes(x = period_date, y = total_spend)) +
  expand_limits(y = 0) +
  scale_y_continuous(labels = label_comma()) +
  xlab("Date") +
  ylab("Total Spend") +
  ggtitle("Lineplot of Total Spend by Week")


ts_monthly_tbl <- ts_data_tbl %>%
  group_by(label = ts_month) %>%
  summarise(
    .groups = "drop",

    period_date = min(invoice_date),
    total_spend = sum(price * quantity) %>% round(2)
    )

ggplot(ts_monthly_tbl) +
  geom_line(aes(x = period_date, y = total_spend)) +
  expand_limits(y = 0) +
  scale_y_continuous(labels = label_comma()) +
  xlab("Date") +
  ylab("Total Spend") +
  ggtitle("Lineplot of Total Spend by Month")
```

To avoid dealing with multiple files for the time series, we combine them into
a single object.

```{r combine_ts_tbl, echo=TRUE}
ts_data_tbl <- list(
    daily   = ts_daily_tbl,
    weekly  = ts_weekly_tbl,
    monthly = ts_monthly_tbl
    ) %>%
  bind_rows(.id = "series") %>%
  arrange(series, period_date)

ts_data_tbl %>% write_rds("data/retail_timeseries_tbl.rds")
```


### Calendar-Based Boxplots

We have aggregated our data across time periods, but it is also worth looking
at both transaction-level and invoice-level amount over time.

```{r lineitem_calendar_boxplot, echo=TRUE}
ggplot(data_tbl) +
  geom_boxplot(aes(x = invoice_woy, y = stock_amount, group = invoice_woy)) +
  scale_y_log10(labels = label_comma()) +
  xlab("Week of Year") +
  ylab("Transaction Amount") +
  ggtitle("Boxplot of Transaction Sizes by Week of Year") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```



```{r invoice_calendar_boxplot, echo=TRUE}
plot_tbl <- data_tbl %>%
  group_by(invoice_woy, invoice_id) %>%
  summarise(
    .groups = "drop",
    
    invoice_amount = sum(stock_amount) %>% round(2)
  )

ggplot(plot_tbl) +
  geom_boxplot(aes(x = invoice_woy, y = invoice_amount, group = invoice_woy)) +
  scale_y_log10(labels = label_comma()) +
  xlab("Week of Year") +
  ylab("Invoice Amount") +
  ggtitle("Boxplot of Invoice Amounts by Week of Year") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5))
```


## Check Distribution of Daily Purchases

We now look at individual invoice amounts, and look at how they are distributed
on a daily basis.

```{r plot_daily_amount_distributions, echo=TRUE}
daily_distribution_tbl <- data_tbl %>%
  group_by(invoice_date, invoice_id) %>%
  summarise(
    .groups = "drop",
    
    total_spend = sum(stock_amount)
    )

daily_distribution_tbl %>% glimpse()
```

We now produce a boxplot for each day so we have a sense of how the
distribution of transaction amounts changes and evolves over time.

```{r plot_daily_distribution_transactions, echo=TRUE}
plot_tbl <- daily_distribution_tbl %>%
  filter(total_spend > 0) %>%
  group_by(invoice_date) %>%
  summarise(
    .groups = "drop",
    
    qn   = c("total_spend", "p10", "p25", "p50", "p75", "p90"),
    vals = c(
      sum(total_spend),
      quantile(total_spend, probs = c(0.10, 0.25, 0.50, 0.75, 0.90))
      )
    ) %>%
  pivot_wider(
    names_from = qn,
    values_from = vals
  )

ggplot(plot_tbl) +
  geom_errorbar(aes(x = invoice_date, ymin = p25, ymax = p75, colour = total_spend),
                width = 0) +
  scale_y_log10(labels = label_comma()) +
  xlab("Date") +
  ylab("Transaction Amount") +
  ggtitle("P25-P75 Quartile of Transaction Amounts by Date")
```

It does not look there is much of a change over time or any particular patterns
that stand out, so we also want to look at the refunds.

```{r plot_daily_distribution_refunds, echo=TRUE}
plot_tbl <- daily_distribution_tbl %>%
  filter(total_spend < 0) %>%
  mutate(total_refund = abs(total_spend)) %>%
  group_by(invoice_date) %>%
  summarise(
    .groups = "drop",
    
    qn   = c("total_refund", "p10", "p25", "p50", "p75", "p90"),
    vals = c(
      sum(total_refund),
      quantile(total_refund, probs = c(0.10, 0.25, 0.50, 0.75, 0.90))
      )
    ) %>%
  pivot_wider(
    names_from = qn,
    values_from = vals
  )

ggplot(plot_tbl) +
  geom_errorbar(aes(x = invoice_date, ymin = p25, ymax = p75, colour = total_refund),
                width = 0) +
  scale_y_log10(labels = label_comma()) +
  xlab("Date") +
  ylab("Transaction Amount") +
  ggtitle("P25-P75 Quartile of Refund Amounts by Date")
```


## Investigate Refunds

This transactions data also includes returns and refunds - adding a level of
complexity to this analysis as we need to account for this when assessing the
data.

One option is to ignore the refunds, at least initially, but this may add
a large source of bias to our analysis and we need to get a feel for this.

To do this, we will look at each line-item entry with a negative quantity and
we will look back in time to the previous time such an time was sold, and we
match on the `customer_id`, `stock_code` and `price`. This is strong evidence
that the previous sale was then returned, and so we can label these entries.

Due to the time snapshot nature of this data, this also means there are likely
to be some returns in our data that do not have a corresponding sale, so we
can also ignore those.


```{r create_matching_returns_data, echo=TRUE}
filter_returns_transactions <- function(use_stock, use_customer,  use_dttm,
                                        data_tbl) {
  matches_tbl <- data_tbl %>%
    filter(stock_code   == use_stock,
           customer_id  == use_customer,
           invoice_dttm <= use_dttm,
           quantity     >  0)
  
  return(matches_tbl)
}

returns_data_tbl <- data_tbl %>%
  filter(quantity < 0) %>%
  select(row_id, stock_code, customer_id, quantity, invoice_dttm) %>%
  mutate(
    prev_tnx_data = pmap(
      list(use_stock    = stock_code,
           use_customer = customer_id,
           use_dttm     = invoice_dttm),
      filter_returns_transactions,
      data_tbl = data_tbl
      ),
    match_data = map2(prev_tnx_data, quantity,
                      ~ .x %>% filter(quantity == abs(.y))),
    last_match = map(match_data,
                     ~ .x %>% arrange(desc(invoice_dttm)) %>% head(1))
    )

returns_data_tbl %>% glimpse()
```

We can then use this data to get an idea of how often items are returned, and
how much time tends to pass between the purchase and the return.

```{r construct_returns_lookups, echo=TRUE}
returns_lookup_tbl <- returns_data_tbl %>%
  transmute(
    return_id = row_id,
    return_dttm = invoice_dttm,
    data        = map(last_match, ~ .x %>% select(row_id, invoice_dttm))
    ) %>%
  unnest(data) %>%
  select(row_id, return_id)

returns_lookup_tbl %>% glimpse()
```



We now want to look at the distribution of times between the purchase and the
item being returned.

```{r plot_return_times_histogram, echo=TRUE}
return_times_tbl <- returns_lookup_tbl %>%
  left_join(data_tbl %>% select(row_id, invoice_dttm), by = "row_id") %>%
  left_join(data_tbl %>% select(return_id = row_id, return_dttm = invoice_dttm), by = "return_id") %>%
  mutate(
    return_time = return_dttm - invoice_dttm,
    return_days  = as.numeric(return_time) / (24 * 60 * 60)
    )

norm_plot <- ggplot(return_times_tbl) +
  geom_histogram(aes(x = return_days), bins = 50) +
  geom_vline(aes(xintercept = 90), colour = "red") +
  scale_x_continuous(labels = label_comma()) +
  xlab("Days") +
  ylab("Count") +
  ggtitle("Histogram of Days Between Purchase and Return per Item Transaction")

log_plot <- norm_plot +
  scale_x_log10(labels = label_comma())

plot_grid(norm_plot, log_plot, nrow = 2)
```


# R Environment

```{r show_session_info, echo=TRUE, message=TRUE}
sessioninfo::session_info()
```
