---
title: "Using Association Rules of the Online Retail Dataset"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: yes

  html_document:
    fig_caption: yes
    theme: spacelab #sandstone #spacelab #flatly
    highlight: pygments
    number_sections: TRUE
    toc: TRUE
    toc_depth: 2
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---


```{r import_libraries, echo=FALSE, message=FALSE}
library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(arules)
library(arulesViz)
library(DT)
library(tidygraph)



source("custom_functions.R")

resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "arules", "Matrix")
  )


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallel::detectCores()
  )

theme_set(theme_cowplot())

set.seed(42)
```

# Load Data

We first want to load our datasets and prepare them for some simple association
rules mining.

```{r load_transaction_data, echo=TRUE}
tnx_data_tbl <- read_rds("data/retail_data_cleaned_tbl.rds")

tnx_data_tbl %>% glimpse()
```

To use our rules mining we just need the invoice data and the stock code, so
we can ignore the rest. Also, we ignore the issue of returns and just look at
purchases.

```{r prepare_data_arules, echo=TRUE}
tnx_purchase_tbl <- tnx_data_tbl %>%
  filter(
    quantity > 0,
    exclude == FALSE
    ) %>%
  select(invoice_id, stock_code, customer_id, quantity, price, stock_value, description)

tnx_purchase_tbl %>% glimpse()
```

We now write this data out as a CSV so `arules` can read it in and process it.

```{r read_transactions, echo=TRUE}
tnx_purchase_tbl %>% write_csv("data/tnx_purchase_tbl.csv")
```



# Basket Analysis with Association Rules

We now want to do some basic basket analysis using association rules, which
tries to determine which items get bought together, similar to taking a graph
approachin many ways.

```{r setup_arules_structures, echo=TRUE}
basket_arules <- read.transactions(
    file   = "data/tnx_purchase_tbl.csv",
    format = "single",
    sep    = ",",
    header = TRUE,
    cols   = c("invoice_id", "stock_code")
    )

basket_arules %>% glimpse()
```

Now that we have this data we can look at some basic plots much like we
produced before. For example, we can look at the relative frequency of the
different items.

```{r plot_arules_item_frequency, echo=TRUE}
itemFrequencyPlot(basket_arules, topN = 20)

itemFrequencyPlot(basket_arules, topN = 20, type = "absolute")
```

The stock codes do not mean a huge amount to us, so we also want to look at
the description field for these items.

```{r show_frequency_items, echo=TRUE}
freq_codes <- itemFrequency(basket_arules) %>%
  sort(decreasing = TRUE) %>%
  head(20) %>%
  names()

tnx_purchase_tbl %>%
  select(stock_code, description) %>%
  filter(stock_code %in% freq_codes) %>%
  distinct() %>%
  drop_na(description) %>%
  group_by(stock_code) %>%
  summarise(
    .groups = "drop",
    desc = str_c(description, collapse = " : ")
    ) %>%
  arrange(stock_code) %>%
  datatable()
```


## Basic Concepts

The basic ideas of association rule mining and basket analysis draws on basic
ideas from probability theory.

We speak in terms of the *itemset*: that is, a collection of one or more items
that co-occur in a transaction.

For example, suppose we have a list of transactions as follows:


| ID | Items               |
|----|---------------------|
| 1  | milk, bread         |
| 2  | bread, butter       |
| 3  | beer                |
| 4  | milk, bread, butter |
| 5  | bread, butter       |


Using the above set of transactions, and itemset may be "milk" or
"bread, butter".

The support of an itemset $X$, $\text{Supp}(X)$, is defined as the proportion of
transactions in the dataset which contain the itemset.

In the above example:

$$
\text{Supp}(\text{\{milk, bread\}}) = \frac{2}{5} = 0.40.
$$


A *rule*, $X \Rightarrow Y$, between two itemsets $X$ and $Y$ is a directed
relationship of the itemset $X$ showing the presence of $Y$. Not that the rule
is not symmetric: $X \Rightarrow Y$ and $Y \Rightarrow X$ are not the same.

The *confidence* for the rule $X \implies Y$, $\text{Conf}(X \Rightarrow Y)$ is
defined by

$$
\text{Conf}(X \Rightarrow Y) = \frac{\text{Supp}(X \cup Y)}{\text{Supp}(X)}.
$$

So, to calculate the confidence for a rule:

$$
\text{Supp}(\text{\{milk, bread\}} \Rightarrow \text{\{butter\}}) = \frac{0.2}{0.4} = 0.5.
$$

To illustrate how rules are not symmetric:

$$
\text{Supp}(\text{\{butter\}} \Rightarrow \text{\{milk, bread\}}) = \frac{0.2}{0.6} = 0.33.
$$


Finally, we want a measure of the strength of the relationship between the 
itemsets $X$ and $Y$. That is, measuring the effect of the presence of $X$ on
the presence of $Y$. We measure this by defining the *lift* of a rule as

$$
\text{Lift}(X \Rightarrow Y) = \frac{\text{Supp}(X \cup Y)}{\text{Supp}(X) \text{Supp}(Y)}.
$$

Again, we repeat our calculations for our rule.

$$
\text{Lift}(\text{\{bread, milk\}} \Rightarrow \text{\{butter\}}) = \frac{0.2}{(0.4)(0.6)} = \frac{0.2}{0.24} = 0.8333
$$

Lift values greater than 1 implies the presence of $X$ increases the
probability of $Y$ being present when compared to the unconditional
probability.


Now that we have these metrics and concepts, we can turn our attention to
trying to find rules in a given dataset, using these metrics to rank them.

Rather than using brute-force approaches to discovering these rules, we use a
number of different algorithms to find associations within the dataset.

The two main algorithms for discovering some rules are the `apriori` and the
`eclat` algorithms.


## Construct apriori Rules

We now want to construct the association rules using the `apriori` algorithm.
To do this, we need to set parameters such as the minimum support and the
minimum confidence level.

This gives us a set of association rules, along with the support and lift.


```{r construct_apriori_rules, echo=TRUE}
basket_apriori <- apriori(
    basket_arules,
    parameter = list(supp = 0.005, conf = 0.8)
    )

basket_apriori_tbl <- basket_apriori %>%
  as("data.frame") %>%
  as_tibble() %>%
  arrange(desc(lift))

basket_apriori_tbl %>% glimpse()
```

We now want to inspect this table using `datatable`

```{r inspect_apriori_rules, echo=TRUE}
basket_apriori %>% inspectDT()
```

To help visualise these rules, we can produce a basic scatterplot of the
metrics.

```{r plot_arules_metrics, echo=TRUE}
ggplot(basket_apriori_tbl) +
  geom_point(aes(x = confidence, y = lift), alpha = 0.2) +
  xlab("Rule Confidence") +
  ylab("Rule Lift") +
  ggtitle("Scatterplot of Association Rule Metrics")
```


## Construct eclat Rules

An alternative method of constructing association rules is to use the `eclat`
algorithm. The code for doing this is slightly different, but gives us similar
outputs.

```{r construct_eclat_rules, echo=TRUE}
basket_eclat <- eclat(
    basket_arules,
    parameter = list(support = 0.005)
    ) %>%
  ruleInduction(
    basket_arules,
    confidence = 0.8
    )


basket_eclat_tbl <- basket_eclat %>%
  as("data.frame") %>%
  as_tibble() %>%
  arrange(desc(lift))

basket_eclat_tbl %>% glimpse()
```

Once again, we inspect the data.

```{r inspect_eclat_rules, echo=TRUE}
basket_eclat %>% inspectDT()
```


## Compare Algorithms

We now want to compare the outputs of both algorithms in terms of association
rules and how they compare.

```{r compare_arules_algorithms, echo=TRUE}
basket_ap_tbl <- basket_apriori_tbl %>%
  select(rules, support, confidence, lift)

basket_ec_tbl <- basket_eclat_tbl %>%
  select(rules, support, confidence, lift)

rules_comparison_tbl <- basket_ap_tbl %>%
  full_join(basket_ec_tbl, by = "rules", suffix = c("_a", "_e"))

rules_comparison_tbl %>% glimpse()
```


## Reducing Minimum Confidence

While high confidence rules are useful, they are more likely to find rules that
are "obvious" as the probabilities are such that co-occuring basket items will
be noticed as being together - or possibly be natural complements: butter, milk
and bread is a good example.

Instead, we are also interested in less obvious rules, and so we reduce our
confidence threshold and see how many additional rules are discovered.

```{r determine_lower_confidence_rules, echo=TRUE}
basket_lower_rules <- apriori(
    basket_arules,
    parameter = list(supp = 0.005, conf = 0.4)
    )

basket_lower_rules_tbl <- basket_lower_rules %>%
  as("data.frame") %>%
  as_tibble() %>%
  arrange(desc(lift))


```

```{r plot_lower_arules_metrics, echo=TRUE}
ggplot(basket_lower_rules_tbl) +
  geom_point(aes(x = confidence, y = lift), alpha = 0.2) +
  xlab("Rule Confidence") +
  ylab("Rule Lift") +
  ggtitle("Scatterplot of Association Rule Metrics")
```










# Converting Rules to Graphs

We also have the ability to convert these rules to a graph representation,
where each node is either a `stock_code` or a rule, with the edges of the
graph representing that item being contained in the rule.


```{r convert_rules_to_graph, echo=TRUE}
apriori_rules_igraph <- basket_apriori %>%
  plot(
    measure = "support",
    method  = "graph",
    control = list(max = 1000)
    ) %>%
  as("igraph")

apriori_rules_igraph %>% glimpse()
```

We should first visualise this graph, using the top 30 rules in the dataset,
as measured by the support of the rule.

```{r plot_interactive_rules_graph, echo=TRUE}
basket_apriori %>%
  head(n = 30, by = "support") %>%
  plot(
    measure  = "lift",
    method   = "graph",
    engine   = "htmlwidget"
    )

```







# R Environment

```{r show_session_info, echo=TRUE, message=TRUE}
sessioninfo::session_info()
```
