---
title: "Using RFM Models the Online Retail Dataset"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: TRUE
    code_folding: hide
    css: styles.css
    fig_caption: TRUE

  html_document:
    fig_caption: yes
    theme: spacelab #sandstone #spacelab #flatly
    highlight: pygments
    number_sections: TRUE
    toc: TRUE
    toc_depth: 2
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---


```{r import_libraries, echo=FALSE, message=FALSE}
library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(directlabels)
library(magrittr)
library(rlang)
library(fs)
library(purrr)
library(furrr)
library(glue)
library(tidyquant)
library(rsample)
library(rfm)
library(xplorerr)
library(DT)


source("lib_utils.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2")
  )


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallel::detectCores()
  )

theme_set(theme_cowplot())

set.seed(42)

plan(multisession)
```


# Load Data

We first want to load our dataset and prepare them as before.


```{r load_transaction_data, echo=TRUE}
tnx_data_tbl <- read_rds("data/retail_data_cleaned_tbl.rds")

tnx_data_tbl %>% glimpse()
```

Having loaded the transaction data we also want to load the daily summary
datasets.

```{r load_daily_spend_data, echo=TRUE}
daily_spend_invoice_tbl <- read_rds("data/daily_spend_invoice_tbl.rds")
daily_spend_invoice_tbl %>% glimpse()

daily_spend_tbl <- read_rds("data/daily_spend_tbl.rds")
daily_spend_tbl %>% glimpse()
```



Finally, we want to set a date to make the end of the "fitting" data, which
will be used in the various estimation routines to fit parameters of our models.

```{r set_training_data_date, echo=TRUE}
training_data_date <- as.Date("2011-03-31")
```


# Basic RFM Analysis

We have some basic transaction data so we now add this to routines provided by
the `rfm` package.


```{r construct_basic_rfm_structures, echo=TRUE}
customer_rfmdata <- daily_spend_invoice_tbl %>%
  filter(invoice_date <= training_data_date) %>%
  rfm_table_order(
    customer_id   = customer_id,
    order_date    = invoice_date,
    revenue       = invoice_spend,
    analysis_date = training_data_date
    )

customer_rfmdata %>% print()
```

## RFM Visualisations

We now want to look at a few provided visualisations of this data.

```{r plot_rfm_heatmap_data, echo=TRUE}
customer_rfmdata %>% rfm_heatmap()

customer_rfmdata %>% rfm_bar_chart()

customer_rfmdata %>% rfm_histograms()

customer_rfmdata %>% rfm_rm_plot()

customer_rfmdata %>% rfm_fm_plot()

customer_rfmdata %>% rfm_rf_plot()
```


# Customer Segments

We can use the RFM measures to segment our customer base, and to do so we need
to first define each of the segments.

```{r construct_customer_segments, echo=TRUE}
segment_names <- c(
  "Champions", "Loyal Customers", "Potential Loyalist", "New Customers",
  "Promising", "Need Attention", "About To Sleep", "At Risk",
  "Can't Lose Them", "Lost"
  )

recency_lower   <- c(4, 2, 3, 4, 3, 2, 2, 1, 1, 1)
recency_upper   <- c(5, 5, 5, 5, 4, 3, 3, 2, 1, 2)
frequency_lower <- c(4, 3, 1, 1, 1, 2, 1, 2, 4, 1)
frequency_upper <- c(5, 5, 3, 1, 1, 3, 2, 5, 5, 2)
monetary_lower  <- c(4, 3, 1, 1, 1, 2, 1, 2, 4, 1)
monetary_upper  <- c(5, 5, 3, 1, 1, 3, 2, 5, 5, 2)

segment_defs_tbl <- tibble(
  segment_names,
  recency_lower,
  recency_upper,
  frequency_lower,
  frequency_upper,
  monetary_lower,
  monetary_upper
  )

segment_defs_tbl %>% glimpse()
```

We first visually inspect these segment definitions and the bands.

```{r display_customer_segment_definitions, echo=TRUE}
segments_show_tbl <- segment_defs_tbl %>%
  mutate(
    recency   = glue("{recency_lower}-{recency_upper}")     %>% as.character(),
    frequency = glue("{frequency_lower}-{frequency_upper}") %>% as.character(),
    monetary  = glue("{monetary_lower}-{monetary_upper}")   %>% as.character()
    ) %>%
  select(
    segment_names, recency, frequency, monetary
    )

segments_show_tbl %>%
  datatable(
    colnames = c("Segment", "R", "F", "M"),
    options = list(
      columnDefs = list(list(className = 'dt-left', targets = 0:4))
      )
    )
```


We now want to check all the different possibilities for customers to check if
it is possible for some combinations to fit in multiple categories, as this
can help with segment definition.


```{r validate_segment_definitions, echo=TRUE}
select_customer_segment <- function(rc, fr, mn, data_tbl) {
  selection_tbl <- data_tbl %>%
    filter(
      recency_lower   <= rc, rc <= recency_upper,
      frequency_lower <= fr, fr <= frequency_upper,
      monetary_lower  <= mn, mn <= monetary_upper,
      )
  
  return(selection_tbl)
}

test_customer_tbl <- expand_grid(
    recency   = 1:5,
    frequency = 1:5,
    monetary  = 1:5
    ) %>%
  mutate(
    segment_id = 1:n(),
    
    .before = 1
    ) %>%
  mutate(
    select_data = pmap(
      list(rc = recency, fr = frequency, mn = monetary),
      select_customer_segment,
      data_tbl = segment_defs_tbl
      )
    )
  
test_customer_tbl %>% glimpse()
```

Now we check these definitions against our customers and in particular we 
need to focus on customers assigned to none of the segments, and customers
assigned to multiple segments.

```{r find_missing_customer_segments, echo=TRUE}
missing_customer_segments_tbl <- test_customer_tbl %>%
  filter(map_int(select_data, nrow) == 0) %>%
  select(-select_data)

missing_customer_segments_tbl %>% datatable()
```


## Analysing the Customer Segments

Now that we have checked our definitions, we categorise each of the customers
into one of these segments.

```{r segment_customer_base, echo=TRUE}
customer_segments_tbl <- customer_rfmdata %>%
  rfm_segment(
    segment_names   = segment_names,
    recency_lower   = recency_lower,
    recency_upper   = recency_upper,
    frequency_lower = frequency_lower,
    frequency_upper = frequency_upper,
    monetary_lower  = monetary_lower,
    monetary_upper  = monetary_upper
    )

customer_segments_tbl %>% glimpse()
```

Having assigned the customers to each segment we now can now use this data to
run some summary information.

```{r plot_segment_counts, echo=TRUE}
plot_tbl <- customer_segments_tbl %>% count(segment)

ggplot(plot_tbl) +
  geom_col(aes(x = segment, y = n, fill = segment), alpha = 0.50) +
  geom_text(aes(x = segment, y = n, label = n)) +
  labs(
    x = "Segment",
    y = "Count",
    title = "Plot of Size of Customer Segments"
    ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10, angle = 20, vjust = 0.5)
    )
```


Finally, we produce boxplots of the customer segments by the various metrics
we have calculated.

```{r construct_segment_boxplots, echo=TRUE}
ggplot(customer_segments_tbl) +
  geom_boxplot(aes(x = segment, y = recency_days)) +
  labs(
    x = "Customer Segment",
    y = "Days Since Last Transaction",
    title = "Boxplot of Transaction Recency by Customer Segment"
    ) +
  theme(
    axis.text.x = element_text(size = 10, angle = 20, vjust = 0.5)
    )


ggplot(customer_segments_tbl) +
  geom_boxplot(aes(x = segment, y = transaction_count)) +
  scale_y_log10(labels = label_comma()) +
  expand_limits(y = 0) +
  labs(
    x = "Customer Segment",
    y = "Count of Transactions",
    title = "Boxplot of Transaction Count by Customer Segment"
    ) +
  theme(
    axis.text.x = element_text(size = 10, angle = 20, vjust = 0.5)
    )


ggplot(customer_segments_tbl) +
  geom_boxplot(aes(x = segment, y = amount)) +
  scale_y_log10(labels = label_comma()) +
  expand_limits(y = 0) +
  labs(
    x = "Customer Segment",
    y = "Total Revenue",
    title = "Boxplot of Customer Total Revenue by Customer Segment"
    ) +
  theme(
    axis.text.x = element_text(size = 10, angle = 20, vjust = 0.5)
    )

```


## Out-of-Sample Segment Validation

We can check the validity or usefulness of these segments by using the segment
classification on the data from after our cutoff date. If our segments are
useful we should see some discrimination between them in the data.

```{r check_post_cutoff_segments_data, echo=TRUE}
customer_segment_classification_tbl <- customer_segments_tbl %>%
  select(customer_id, segment)

post_cutoff_data_tbl <- daily_spend_invoice_tbl %>%
  filter(invoice_date > training_data_date)

customer_summary_tbl <- post_cutoff_data_tbl %>%
  group_by(customer_id) %>%
  summarise(
    most_recent_purchase = max(invoice_date),
    transaction_count    = n(),
    total_purchase       = sum(invoice_spend)
    )

customer_summary_tbl %>% glimpse()
```


```{r create_segment_summaries, echo=TRUE}
segment_check_tbl <- customer_segment_classification_tbl %>%
  left_join(customer_summary_tbl, by = "customer_id") %>%
  group_by(segment) %>%
  mutate(
    segment_count = n(),
    
    .after = "segment"
    ) %>%
  ungroup()

plot_tbl <- segment_check_tbl %>%
  group_by(segment) %>%
  summarise(
    count_total = n(),
    count_dead  = sum(are_na(transaction_count))
    ) %>%
  mutate(
    count_alive = count_total - count_dead,
    
    dropped_prop = count_dead / count_total,
    dropped_p10  = qbeta(0.10, count_dead + 1, count_alive + 1),
    dropped_p25  = qbeta(0.25, count_dead + 1, count_alive + 1),
    dropped_p50  = qbeta(0.50, count_dead + 1, count_alive + 1),
    dropped_p75  = qbeta(0.75, count_dead + 1, count_alive + 1),
    dropped_p90  = qbeta(0.90, count_dead + 1, count_alive + 1)
    )

ggplot(plot_tbl) +
  geom_errorbar(aes(x = segment, ymin = dropped_p10, ymax = dropped_p90),
                width = 0, size = 1) +
  geom_errorbar(aes(x = segment, ymin = dropped_p25, ymax = dropped_p75),
                width = 0, size = 3) +
  geom_point(aes(x = segment, y = dropped_prop),
             size = 2, colour = "red") +
  expand_limits(y = 0) +
  scale_y_continuous(labels = label_percent()) +
  labs(
    x = "Customer Segment",
    y = "Churn Percentage",
    title = "Estimate of Customer Churn Percentage by Customer Segment"
    ) +
  theme(
    axis.text.x = element_text(size = 10, angle = 20, vjust = 0.5)
    )
```

For the purposes of checking both the frequency and monetary values, we filter
out those customers with no transactions after the cutoff date i.e. we only
look at 'alive' customers.

```{r plot_alive_customer_stats_boxplot, echo=TRUE}
plot_tbl <- segment_check_tbl %>%
  drop_na(most_recent_purchase)

ggplot(plot_tbl) +
  geom_boxplot(aes(x = segment, y = transaction_count, fill = segment_count)) +
  scale_y_log10(labels = label_comma()) +
  scale_fill_gradient(low = "blue", high = "red", labels = label_comma()) +
  expand_limits(y = 0) +
  labs(
    x = "Customer Segment",
    y = "Transaction Count",
    fill = "Segment Size",
    title = "Boxplot of Customer Transaction Count by Customer Segment"
    ) +
  theme(
    axis.text.x = element_text(size = 10, angle = 20, vjust = 0.5)
    )


ggplot(plot_tbl) +
  geom_boxplot(aes(x = segment, y = total_purchase)) +
  scale_y_log10(labels = label_comma()) +
  scale_fill_gradient(low = "blue", high = "red", labels = label_comma()) +
  expand_limits(y = 0) +
  labs(
    x = "Customer Segment",
    y = "Total Spend",
    fill = "Segment Size",
    title = "Boxplot of Customer Total Revenue by Customer Segment"
    ) +
  theme(
    axis.text.x = element_text(size = 10, angle = 20, vjust = 0.5)
    )

```





# R Environment

```{r show_session_info, echo=TRUE, message=TRUE}
sessioninfo::session_info()
```
