---
title: "Using RFM Models the Online Retail Dataset"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: yes
    css: styles.css

  html_document:
    fig_caption: yes
    theme: spacelab #sandstone #spacelab #flatly
    highlight: pygments
    number_sections: TRUE
    toc: TRUE
    toc_depth: 2
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---


```{r import_libraries, echo=FALSE, message=FALSE}
library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(directlabels)
library(magrittr)
library(rlang)
library(fs)
library(purrr)
library(furrr)
library(glue)
library(tidyquant)
library(rsample)
library(rfm)
library(xplorerr)
library(DT)


source("lib_utils.R")

resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2")
  )


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallel::detectCores()
  )

theme_set(theme_cowplot())

set.seed(42)

plan(multisession)
```


# Load Data

We first want to load our dataset and prepare them as before.


```{r load_transaction_data, echo=TRUE}
tnx_data_tbl <- read_rds("data/retail_data_cleaned_tbl.rds")

tnx_data_tbl %>% glimpse()
```

We filter out all returns data and various other transactions we excluded from
our dataset.

```{r prepare_data_arules, echo=TRUE}
tnx_purchase_tbl <- tnx_data_tbl %>%
  filter(
    quantity > 0,
    exclude == FALSE
    ) %>%
  select(
    invoice_date, invoice_id, stock_code, customer_id, description,
    quantity, price, stock_value
    )

tnx_purchase_tbl %>% glimpse()
```

Much as we did for the BTYD models, we want to look at individual transactions
so we aggregate across products and combine the data down to an individual
invoice.

```{r calculate_customer_daily_spend, echo=TRUE}
daily_spend_invoice_tbl <- tnx_purchase_tbl %>%
  drop_na(customer_id) %>%
  group_by(invoice_date, customer_id, invoice_id) %>%
  summarise(
    .groups = "drop",
    
    invoice_spend = sum(stock_value)
  )

daily_spend_invoice_tbl %>% glimpse()


daily_spend_tbl <- daily_spend_invoice_tbl %>%
  group_by(invoice_date, customer_id) %>%
  summarise(
    .groups = "drop",
    
    total_spend = sum(invoice_spend),
    tnx_count   = n()
    )

daily_spend_tbl %>% glimpse()
```


Finally, we want to set a date to make the end of the "fitting" data, which
will be used in the various estimation routines to fit parameters of our models.

```{r set_training_data_date, echo=TRUE}
training_data_date <- as.Date("2011-03-31")
```


# Basic RFM Analysis

We have some basic transaction data so we now add this to routines provided by
the `rfm` package.


```{r construct_basic_rfm_structures, echo=TRUE}
customer_rfmdata <- daily_spend_invoice_tbl %>%
  filter(invoice_date < training_data_date) %>%
  rfm_table_order(
    customer_id   = customer_id,
    order_date    = invoice_date,
    revenue       = invoice_spend,
    analysis_date = training_data_date
    )

customer_rfmdata %>% print()
```

## RFM Visualisations

We now want to look at a few provided visualisations of this data.

```{r plot_rfm_heatmap_data, echo=TRUE}
customer_rfmdata %>% rfm_heatmap()

customer_rfmdata %>% rfm_bar_chart()

customer_rfmdata %>% rfm_histograms()

customer_rfmdata %>% rfm_rm_plot()

customer_rfmdata %>% rfm_fm_plot()

customer_rfmdata %>% rfm_rf_plot()
```


# Customer Segments

We can use the RFM measures to segment our customer base, and to do so we need
to first define each of the segments.

```{r construct_customer_segments, echo=TRUE}
segment_names <- c(
  "Champions", "Loyal Customers", "Potential Loyalist", "New Customers",
  "Promising", "Need Attention", "About To Sleep", "At Risk",
  "Can't Lose Them", "Lost"
  )

recency_lower   <- c(4, 2, 3, 4, 3, 2, 2, 1, 1, 1)
recency_upper   <- c(5, 5, 5, 5, 4, 3, 3, 2, 1, 2)
frequency_lower <- c(4, 3, 1, 1, 1, 2, 1, 2, 4, 1)
frequency_upper <- c(5, 5, 3, 1, 1, 3, 2, 5, 5, 2)
monetary_lower  <- c(4, 3, 1, 1, 1, 2, 1, 2, 4, 1)
monetary_upper  <- c(5, 5, 3, 1, 1, 3, 2, 5, 5, 2)

segment_defs_tbl <- tibble(
  segment_names,
  recency_lower,
  recency_upper,
  frequency_lower,
  frequency_upper,
  monetary_lower,
  monetary_upper
  )

segment_defs_tbl %>% glimpse()
```

We first visually inspect these segment definitions and the bands.

```{r display_customer_segment_definitions, echo=TRUE}
segments_show_tbl <- segment_defs_tbl %>%
  mutate(
    recency   = glue("{recency_lower}-{recency_upper}")     %>% as.character(),
    frequency = glue("{frequency_lower}-{frequency_upper}") %>% as.character(),
    monetary  = glue("{monetary_lower}-{monetary_upper}")   %>% as.character()
    ) %>%
  select(
    segment_names, recency, frequency, monetary
    )

segments_show_tbl %>%
  datatable(
    colnames = c("Segment", "R", "F", "M"),
    options = list(
      columnDefs = list(list(className = 'dt-left', targets = 0:4))
      )
    )
```


We now want to check all the different possibilities for customers to check if
it is possible for some combinations to fit in multiple categories, as this
can help with segment definition.


```{r validate_segment_definitions, echo=TRUE}
select_customer_segment <- function(rc, fr, mn, data_tbl) {
  selection_tbl <- data_tbl %>%
    filter(
      recency_lower   <= rc, rc <= recency_upper,
      frequency_lower <= fr, fr <= frequency_upper,
      monetary_lower  <= mn, mn <= monetary_upper,
      )
  
  return(selection_tbl)
}

test_customer_tbl <- expand_grid(
    recency   = 1:5,
    frequency = 1:5,
    monetary  = 1:5
    ) %>%
  mutate(
    segment_id = 1:n(),
    
    .before = 1
    ) %>%
  mutate(
    select_data = pmap(
      list(rc = recency, fr = frequency, mn = monetary),
      select_customer_segment,
      data_tbl = segment_defs_tbl
      )
    )
  
test_customer_tbl %>% glimpse()
```

Now we check these definitions against our customers and in particular we 
need to focus on customers assigned to none of the segments, and customers
assigned to multiple segments.

```{r find_missing_customer_segments, echo=TRUE}
missing_customer_segments_tbl <- test_customer_tbl %>%
  filter(map_int(select_data, nrow) == 0) %>%
  select(-select_data)

missing_customer_segments_tbl %>% datatable()
```


## Analysing the Customer Segments

Now that we have checked our definitions, we categorise each of the customers
into one of these segments.

```{r segment_customer_base, echo=TRUE}
customer_segments_tbl <- customer_rfmdata %>%
  rfm_segment(
    segment_names   = segment_names,
    recency_lower   = recency_lower,
    recency_upper   = recency_upper,
    frequency_lower = frequency_lower,
    frequency_upper = frequency_upper,
    monetary_lower  = monetary_lower,
    monetary_upper  = monetary_upper
    )

customer_segments_tbl %>% glimpse()
```

Having assigned the customers to each segment we now can now use this data to
run some summary information.

```{r plot_segment_counts, echo=TRUE}
plot_tbl <- customer_segments_tbl %>% count(segment)

ggplot(plot_tbl) +
  geom_col(aes(x = segment, y = n, fill = segment), alpha = 0.25) +
  geom_text(aes(x = segment, y = n, label = n)) +
  labs(
    x = "Segment",
    y = "Count",
    title = "Plot of Size of Customer Segments"
    ) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 20, vjust = 0.5)
    )
```


Finally, we produce boxplots of the customer segments by the various metrics
we have calculated.

```{r construct_segment_boxplots, echo=TRUE}
ggplot(customer_segments_tbl) +
  geom_boxplot(aes(x = segment, y = recency_days)) +
  labs(
    x = "Customer Segment",
    y = "Days Since Last Transaction",
    title = "Boxplot of Transaction Recency by Customer Segment"
    ) +
  theme(axis.text.x = element_text(angle = 20, vjust = 0.5))


ggplot(customer_segments_tbl) +
  geom_boxplot(aes(x = segment, y = transaction_count)) +
  labs(
    x = "Customer Segment",
    y = "Count of Transactions",
    title = "Boxplot of Transaction Count by Customer Segment"
    ) +
  theme(axis.text.x = element_text(angle = 20, vjust = 0.5))


ggplot(customer_segments_tbl) +
  geom_boxplot(aes(x = segment, y = amount)) +
  labs(
    x = "Customer Segment",
    y = "Count of Transactions",
    title = "Boxplot of Customer Total Revenue by Customer Segment"
    ) +
  theme(axis.text.x = element_text(angle = 20, vjust = 0.5))

```




# R Environment

```{r show_session_info, echo=TRUE, message=TRUE}
sessioninfo::session_info()
```
