---
title: "Using Buy-Till-You-Die (BTYD) Models the Online Retail Dataset"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: yes

  html_document:
    fig_caption: yes
    theme: spacelab #sandstone #spacelab #flatly
    highlight: pygments
    number_sections: TRUE
    toc: TRUE
    toc_depth: 2
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---


```{r import_libraries, echo=FALSE, message=FALSE}
library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(BTYD)
library(BTYDplus)
library(CLVTools)


source("lib_utils.R")

resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2")
  )


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallel::detectCores()
  )

theme_set(theme_cowplot())

set.seed(42)
```


# Load Data

We first want to load our datasets and prepare them for some simple association
rules mining.

```{r load_transaction_data, echo=TRUE}
tnx_data_tbl <- read_rds("data/retail_data_cleaned_tbl.rds")

tnx_data_tbl %>% glimpse()
```

To use our rules mining we just need the invoice data and the stock code, so
we can ignore the rest. Also, we ignore the issue of returns and just look at
purchases.

```{r prepare_data_arules, echo=TRUE}
tnx_purchase_tbl <- tnx_data_tbl %>%
  filter(
    quantity > 0,
    exclude == FALSE
    ) %>%
  select(
    invoice_date, invoice_id, stock_code, customer_id, description,
    quantity, price, stock_value
    )

tnx_purchase_tbl %>% glimpse()
```

Use of BTYD models assumes a total spend over a period of day and those
differences 

```{r calculate_customer_daily_spend, echo=TRUE}
daily_spend_invoice_tbl <- tnx_purchase_tbl %>%
  drop_na(customer_id) %>%
  group_by(invoice_date, customer_id, invoice_id) %>%
  summarise(
    .groups = "drop",
    
    invoice_spend = sum(stock_value)
  )

daily_spend_invoice_tbl %>% glimpse()


daily_spend_tbl <- daily_spend_invoice_tbl %>%
  group_by(invoice_date, customer_id) %>%
  summarise(
    .groups = "drop",
    
    total_spend = sum(invoice_spend),
    tnx_count   = n()
  )

daily_spend_tbl %>% glimpse()
```


# Background Theory

Before we start working on fitting and using the various Buy-Till-You-Die
models, we first need to discuss the basic underlying theory and model.

In this model, we assume a customer becomes 'alive' to the business at the
first purchase and then makes purchases stochastically but at a steady-rate
for a period of time, and then 'dies' - i.e. becomes inactive to the business -
hence the use of "Buy-Till-You-Die".

Thus, at a high level these models decompose into modelling the transaction
events using distributions such as the Poisson or Negative Binomial, and then
modelling the 'dropout' process using some other method.

A number of BTYD models exist and for this workshop we will focus on the
BG/NBD model - the Beta-Geometric Negative Binomial Distribution model.

This model relies on a number of base assumptions:

  1. While active, the number of transactions made by a customer follows a
  Poisson process with transaction rate $\lambda$.
  1. Heterogeneity in $\lambda$ follows a gamma distribution
  $\Gamma(\lambda | \alpha, r)$ with parameters shape $\alpha$ and rate $r$. 
  1. After any transaction, a customer becomes inactive with probability $p$.
  1. Heterogeneity in $p$ follows a Beta distribution $B(p | a, b)$ with shape
  parameters $a$ and $b$.
  1. The transaction rate $\lambda$ and the dropout probability $p$ vary
  independently across customers.





# Create Initial BTYD Model


```{r create_initial_btyd_model, echo=TRUE}
daily_cbscbt <- daily_spend_tbl %>%
  select(cust = customer_id, date = invoice_date, sales = total_spend) %>%
  dc.ElogToCbsCbt(
    per             = "week",
    T.cal           = as.Date("2010-06-30"),
    merge.same.date = TRUE
    )

params_bgnbd <- daily_cbscbt$cal$cbs %>%
  bgnbd.EstimateParameters()


bgnbd.PlotFrequencyInCalibration(
  params  = params_bgnbd,
  cal.cbs = daily_cbscbt$cal$cbs,
  censor  = 15
  )

bgnbd.PlotTransactionRateHeterogeneity(params = params_bgnbd)

bgnbd.PlotDropoutRateHeterogeneity(params = params_bgnbd)


bgnbd.Expectation(params = params_bgnbd, t = 52)

bgnbd.PAlive(params = params_bgnbd, x = 5, t.x = 26, T.cal = 30)
```



# R Environment

```{r show_session_info, echo=TRUE, message=TRUE}
sessioninfo::session_info()
```
