---
title: "Data Science Soup to Nuts: Retrieving Retail Data"
author: "Mick Cooney <mickcooney@gmail.com>"
date: ""
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: TRUE
    code_folding: hide
    fig_caption: TRUE

  html_document:
    fig_caption: yes
    number_sections: yes
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---


```{r knit_opts, include = FALSE}
library(conflicted)
library(tidyverse)
library(magrittr)
library(readxl)
library(scales)
library(cowplot)
library(curl)
library(glue)
library(fs)


source("lib_utils.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2")
  )


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallel::detectCores()
  )

set.seed(42)

theme_set(theme_cowplot())
```



---


All code and data for this workshop is available at the following URL:

https://github.com/kaybenleroll/data_workshops

Code is available in the `ws_soupnuts_202101/` directory.


# Retrieve Dataset

We now want to retrieve the dataset used for this project, which is available
at the UCI Machine Learning Repository.

```{r retrieve_online_retail_data, echo=TRUE}
data_url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/00502/online_retail_II.xlsx"

xlsx_datafile     <- "data/online_retail_II.xlsx"

if(!file_exists(xlsx_datafile)) {
  curl_download(
    data_url,
    destfile = xlsx_datafile,
    quiet    = FALSE,
    mode     = "wb"
    )
} else {
  message(glue("Datafile {xlsx_datafile} found. Skipping download."))
}
```

Now that we have downloaded the XLSX file, we want to read in the data and
parse it.


## Parse the Data


```{r load_parse_data, echo=TRUE}
retrieve_datafile <- "data/retail_data_tbl.rds"

create_excel_datetime <- function(x)
  (x * (60 * 60 * 24)) %>% as.POSIXct(origin = "1899-12-30", tz = "GMT")

data_cols <- cols(
  .default      = col_character(),
  Quantity      = col_number(),
  InvoiceDate   = col_number(),
  Price         = col_number()
)

retail_data_tbl <- excel_sheets(xlsx_datafile) %>%
  enframe(name = NULL, value = "excel_sheet") %>%
  mutate(
    data = map(excel_sheet, read_excel,
               path      = xlsx_datafile,
               col_types = "text")
    ) %>%
  unnest(data) %>%
  format_csv() %>%
  read_csv(col_types = data_cols) %>%
  mutate(
    InvoiceDate = create_excel_datetime(InvoiceDate)
    )

retail_data_tbl %>% glimpse()
```


Finally, we output this data to the disk.

```{r write_data_to_disk, echo=TRUE}
retail_data_tbl %>% write_rds("data/retail_data_raw_tbl.rds")
```



# R Environment

```{r show_session_info, echo=TRUE, message=TRUE}
sessioninfo::session_info()
```
