---
title: "An Introduction to Bayesian Reasoning"
subtitle: "Society of Northern Ireland Actuaries"
author: "Mick Cooney <mcooney@describedata.com>"
date: "6 October 2020"
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: pygments
    center: true
---

```{r knit_opts, include=FALSE}
library(conflicted)
library(tidyverse)
library(scales)
library(magrittr)
library(cowplot)
library(rstan)
library(bayesplot)
library(tidybayes)

source("custom_functions.R")

resolve_conflicts(c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2"))



options(width = 80L,
        warn  =   1,
        mc.cores = parallel::detectCores()
        )

knitr::opts_chunk$set(tidy       = FALSE,
                      cache      = FALSE,
                      warning    = FALSE,
                      message    = FALSE,
                      fig.height =     7,
                      fig.width  =    11)


theme_set(theme_cowplot())

set.seed(42)
stan_seed <- 42
```




# Bayesian Inference Engine

---

Prior Knowledge

$+$

Data

\

$=$

\

Posterior Knowledge

---

## Discrete Variables

\

$$
P(A \, | B) = \frac{P(B \, | A) P(A)}{P(B)} 
$$


---

## Continuous Variables

\

Parameters, $\theta$

\

Data, $D$

---

Prior: $p(\theta)$

\

Likelihood: $p(D | \theta)$

\

Posterior: $p(\theta | D)$

---

$$
p(\theta \, | \, D) = \int p(\theta) \, p(D \, | \, \theta)
$$

\

Posterior calculation is high-dim integral

---

Use MCMC to sample posterior


# Quick Example

```{r load_monthly_returns, echo=FALSE}
spy_returns_tbl <- read_rds("data/spy_returns_tbl.rds")
```


## SPY Monthly Returns


```{r plot_asset_monthly_returns, echo=FALSE}
ggplot(spy_returns_tbl) +
  geom_histogram(aes(x = monthly_return), bins = 30) +
  scale_x_continuous(labels = label_percent()) +
  xlab("Asset Return") +
  ylab("Frequency Count") +
  ggtitle("Histogram of Monthly SPY Returns from Jan 2020")
```

---

```{r plot_cumulative_return, echo=FALSE}
plot_tbl <- spy_returns_tbl %>%
  mutate(cuml_return = monthly_return %>% add(1) %>% cumprod())

ggplot(plot_tbl) +
  geom_line(aes(x = date, y = cuml_return)) +
  scale_y_continuous(labels = label_percent()) +
  xlab("Date") +
  ylab("Cumulative Return") +
  ggtitle("Lineplot of the SPY Cumulative Returns")
```

---

```{r asset_returns_stanmodel, cache=FALSE, warning=TRUE, echo=FALSE, results='hide'}
stan_file <- "asset_returns.stan"

standata_lst <- list(
  prior_pd = 0,
  
  y = spy_returns_tbl %>% pull(monthly_return),
  N = spy_returns_tbl %>% nrow(),
  
  return_prior_mu = 0.0,
  return_prior_sd = 0.1,
  
  vol_prior_mu    = -1.6,
  vol_prior_sd    =  0.2
)


asset_returns_stanmodel <- stan_model(stan_file)

asset_returns_prior_stanfit <- sampling(
  object = asset_returns_stanmodel,
  data   = standata_lst %>% list_modify(prior_pd = 1),
  iter   = 1000,
  chains = 4,
  seed   = stan_seed
)

asset_returns_stanfit <- sampling(
  object = asset_returns_stanmodel,
  data   = standata_lst,
  iter   = 1000,
  chains = 4,
  seed   = stan_seed
)

asset_returns_draws_tbl <- asset_returns_stanfit %>%
  tidy_draws()

```

```{r plot_posterior_histograms, echo=FALSE}
plot_tbl <- asset_returns_draws_tbl %>%
  select(.draw, mu, sigma) %>%
  gather("param", "value", -.draw)

ggplot(plot_tbl) +
  geom_histogram(aes(x = value), bins = 30) +
  facet_wrap(vars(param), scales = "free_x") +
  xlab("Posterior Value") +
  ylab("Frequency Count") +
  ggtitle("Histogram of Parameter Posterior Distributions")
```

---

```{r plot_joint_posterior_distribution, echo=FALSE}
ggplot(asset_returns_draws_tbl) +
  geom_point(aes(x = mu, y = sigma)) +
  ggtitle("Plot of Joint Posterior Distribution of mu and sigma")
```


## Model Checks

```{r plot_data_posterior_distributions, echo=FALSE}
ggplot() +
  geom_line(aes(x = monthly_return), stat = "density",
            data = spy_returns_tbl) +
  geom_line(aes(x = y_sim), stat = "density",
            data = asset_returns_draws_tbl,
            colour = "red") +
  xlab("Monthly Return") +
  ylab("Probability Density") +
  ggtitle("Comparison PDFs for Empirical Data and Posterior Predictive Distribution")
```


# Thank You!

\

Mick Cooney

\

mcooney@describedata.com

\

https://github.com/kaybenleroll/data_workshops




