---
title: "Initial NLP Work on Film Scripts"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    theme:
      light: cerulean
      dark: cyborg
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---


```{r import_libraries}
#| include: false

library(conflicted)
library(tidyverse)
library(magrittr)
library(rlang)
library(fs)
library(purrr)
library(furrr)
library(glue)
library(tidytext)
library(quanteda)



source("lib_utils.R")


conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2")
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallelly::availableCores()
  )

set.seed(42)
```

In this workbook we perform the initial NLP pre-processing and simple
explorations and visualisations of the data.

# Load Data

```{r load_films_master_data}
#| echo: true

films_master_tbl <- read_rds("data/films_master_tbl.rds")

films_master_tbl %>% glimpse()
```

The parsed data is stored in the file listed in the column `parsed_file` and
contains to separate tibbles, one with the detailed parsing of the script and
the second which aggregates all the text for both scene directions and
dialogue.

```{r load_parsed_script_data}
#| echo: true

films_parsed_tbl <- films_master_tbl %>%
  mutate(
    parsed_data = map(parsed_file, read_rds)
    ) %>%
  unnest(parsed_data) %>%
  select(
    film_title, release_year, genre, title_cleaned, parsing_detailed,
    parsing_aggregated
    )

films_parsed_tbl %>% glimpse()
```





# R Environment

```{r show_session_info}
#| echo: false
#| message: false


options(width = 120L)
sessioninfo::session_info()
options(width = 80L)
```
