---
title: "Retrieve and Parse Selected Film Scripts"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    theme:
      light: cerulean
      dark: cyborg
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---


```{r import_libraries}
#| include: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(directlabels)
library(magrittr)
library(rlang)
library(vctrs)
library(fs)
library(purrr)
library(furrr)
library(glue)
library(polite)
library(rvest)
library(snakecase)


source("lib_utils.R")


conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2")
  )


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallel::detectCores()
  )

theme_set(theme_cowplot())

set.seed(42)

plan(multisession)
```

In this workbook we first retrieve the list of film scripts from a input table
provided and then we parse out the retrieved HTML files.

# Set Up Input Data

We have a CSV file that contains the list of film scripts along with some
meta-data about the film itself such as the year of release and the genre.

```{r load_film_scripts_table}
#! echo: true
#! message: false

film_cols <- cols(
  release_year = col_integer(),
  genre        = col_character()
  )


films_tbl <- read_csv(
  file      = "input_data/film_list.csv",
  col_types = film_cols,
  progress  = FALSE
  )

films_tbl %>% glimpse()
```

We now want to set up some base parameters for this worksheet.


```{r setup_scraping_parameters}
#| echo: true

base_url <- "https://imsdb.com/scripts"

scraped_raw_dir   <- "scraped_files/scraped_raw"
parsed_script_dir <- "scraped_files/scraped_script"
```


# Scrape Film Script HTML

```{r setup_film_script_scraping}
#| echo: true

films_scraped_tbl <- films_tbl %>%
  mutate(
    title_cleaned   = film_title %>% to_snake_case(),
    retrieve_url    = glue("{base_url}/{script_filename}.html"),
    cached_htmlfile = glue("{scraped_raw_dir}/scraped_raw_{title_cleaned}.rds")
    )

films_scraped_tbl
```



```{r scraping_new_film_scripts}
#| echo: true

scrape_session <- bow(base_url, force = TRUE)

scrape_film_scripts <- function(script_url, cached_htmlfile, use_session) {
  message(glue("Scraping URL {script_url}"))

  scrape_data <- nod(use_session, script_url, verbose = TRUE) %>%
    scrape(verbose = TRUE, content = "text/html;charset=iso-8859-1") %>%
    as.character()

  scrape_data %>% write_rds(cached_htmlfile)
  
  return(scrape_data)
}


prescraped_tbl <- dir_ls(scraped_raw_dir) %>%
  vec_cast(to = character()) %>%
  enframe(name = NULL, value = "cached_htmlfile")


new_scraped_tbl <- films_scraped_tbl %>%
  anti_join(prescraped_tbl, by = "cached_htmlfile") %>%
  mutate(
    scrape_data = map2(retrieve_url, cached_htmlfile, scrape_film_scripts, use_session = scrape_session)
    )


films_scraped_tbl %>% glimpse()
```


We now should have all the film scripts scraped and stored as a file on the
file system.

```{r load_film_scraped_files}
#| echo: true

parse_script_text <- function(film_title, scraped_data) {
  message(glue("Scraping script text for {film_title}"))
  
  scraped_text <- scraped_data %>%
    html_elements("pre") %>%
    html_text2() %>%
    tail(n = 1)

  return(scraped_text)
}

write_script_text <- function(script_text, script_file) {
  message(glue("Writing file {script_file}"))
  
  script_text %>% write_lines(file = script_file)
  
  return(TRUE)
}


films_scraped_tbl <- films_scraped_tbl %>%
  mutate(
    scraped_html   = map_chr(cached_htmlfile, read_rds),
    parsed_data    = map(scraped_html, read_html),
    script_text    = map2_chr(title_cleaned, parsed_data, parse_script_text),
    script_txtfile = glue("{parsed_script_dir}/script_{title_cleaned}.txt"),
    write_file     = map2_lgl(script_text, script_txtfile, write_script_text)
    )

films_scraped_tbl %>% glimpse()
```






# Write to Disk

We now write this data out to disk.

```{r write_data_disk}
#| echo: false

films_scraped_tbl %>%
  select(
    film_title, release_year, genre, title_cleaned, cached_htmlfile, script_txtfile
    ) %>%
  write_rds("data/films_scraped_tbl.rds")
```


# R Environment

```{r show_session_info}
#| echo: false
#| message: false


options(width = 120L)
sessioninfo::session_info()
options(width = 80L)
```
