---
title: "Retrieve and Extract Selected Film Scripts"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    theme:
      light: cerulean
      dark: cyborg
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---


```{r import_libraries}
#| include: false

library(conflicted)
library(tidyverse)
library(magrittr)
library(rlang)
library(vctrs)
library(jsonlite)
library(fs)
library(purrr)
library(furrr)
library(glue)
library(polite)
library(rvest)
library(snakecase)


source("lib_utils.R")
source("lib_manual_script_cleaners.R")
source("lib_script_parsers.R")


conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2")
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallelly::availableCores()
  )

set.seed(42)
```

In this workbook we first retrieve the list of film scripts from a input table
provided and then we parse out the retrieved HTML files.

# Set Up Input Data

We have a CSV file that contains the list of film scripts along with some
meta-data about the film itself such as the year of release and the genre.

```{r load_film_scripts_table}
#! echo: true
#! message: false

film_cols <- cols(
  release_year = col_integer(),
  genre        = col_character()
  )


films_tbl <- read_csv(
  file      = "input_data/film_list.csv",
  col_types = film_cols,
  progress  = FALSE
  )

films_tbl |> glimpse()
```

Each film script may require different parsing routines to convert it to a
common format so we can run our usual NLP preprocessing etc.


```{r load_film_script_parser_data, echo=TRUE}
#| echo: true
#| message: false

film_parser_loaded_tbl <- read_json(
    "input_data/script_parser_data.json"
    ) |>
  enframe(name = "title_cleaned", value = "parser_conf")

film_parser_loaded_tbl |> glimpse()
```



We now want to set up some base parameters for this worksheet.


```{r setup_scraping_parameters}
#| echo: true

base_url <- "https://imsdb.com/scripts"

raw_scraped_dir    <- "scraped_files/scraped_raw"
raw_script_dir     <- "scraped_files/scraped_raw"
cleaned_script_dir <- "scraped_files/cleaned_script"
parsed_script_dir  <- "data/parsed_scripts"
```


# Scrape Film Script HTML

```{r setup_film_script_scraping}
#| echo: true

films_master_tbl <- films_tbl |>
  mutate(
    title_cleaned   = film_title |> to_snake_case(),
    retrieve_url    = glue("{base_url}/{script_filename}.html"),
    cached_htmlfile = glue("{raw_scraped_dir}/scraped_raw_{title_cleaned}.rds"),
    script_txtfile  = glue("{raw_script_dir}/script_{title_cleaned}.txt"),
    cleaned_txtfile = glue("{cleaned_script_dir}/cleanedscript_{title_cleaned}.txt"),
    parsed_file     = glue("{parsed_script_dir}/parsedscript_{title_cleaned}.rds")
    ) |>
  arrange(title_cleaned)

films_master_tbl |> glimpse()
```



```{r scraping_new_film_scripts}
#| echo: true

scrape_session <- bow(base_url, force = TRUE)

scrape_film_scripts <- function(script_url, cached_htmlfile, use_session) {
  message(glue("Scraping URL {script_url}"))

  scrape_data <- nod(use_session, script_url, verbose = TRUE) |>
    scrape(verbose = TRUE, content = "text/html;charset=iso-8859-1") |>
    as.character()

  scrape_data |> write_rds(cached_htmlfile)
  
  return(scrape_data)
}


prescraped_tbl <- dir_ls(raw_scraped_dir) |>
  vec_cast(to = character()) |>
  enframe(name = NULL, value = "cached_htmlfile")


new_scraped_tbl <- films_master_tbl |>
  anti_join(prescraped_tbl, by = "cached_htmlfile") |>
  mutate(
    scrape_data = map2(
      retrieve_url, cached_htmlfile,
      scrape_film_scripts,
      use_session = scrape_session
      )
    )


new_scraped_tbl |> glimpse()
```


We now should have all the film scripts scraped and stored as a file on the
file system.

```{r load_film_scraped_files}
#| echo: true
#| message: true

parse_script_text <- function(film_title, scraped_data) {
  scraped_text <- scraped_data |>
    html_elements("pre") |>
    html_text2() |>
    tail(n = 1)

  return(scraped_text)
}

write_script_text <- function(script_text, script_file) {
  message(glue("Writing file {script_file}"))
  
  script_text |> write_lines(file = script_file)
  
  return(TRUE)
}


films_scraped_tbl <- films_master_tbl |>
  mutate(
    scraped_html    = map_chr(cached_htmlfile, read_rds),
    parsed_data     = map(scraped_html, read_html),
    script_text     = map2_chr(title_cleaned, parsed_data, parse_script_text),
    write_file      = map2_lgl(script_text, script_txtfile, write_script_text)
    )

films_scraped_tbl |> glimpse()
```


# Make Manual Edits to Scripts

In this section we make some manual edits to our scripts. We load each script
individually and manually perform any required edits.

```{r manually_edit_script_text}
#| echo: true
#| message: true

films_scraped_tbl <- films_scraped_tbl |>
  mutate(
    flag_cleaned = pmap_lgl(
      list(
        input_file  = script_txtfile,
        output_file = cleaned_txtfile,
        film_key    = title_cleaned 
        ),
      manually_clean_scripts
      )
    )

films_scraped_tbl |> glimpse()
```


We want to keep a cut down version of this file for ongoing processing.

```{r construct_films_master_table}
#| echo: true

films_master_tbl <- films_scraped_tbl |>
  select(
    film_title, release_year, genre, title_cleaned, cached_htmlfile,
    script_txtfile, cleaned_txtfile, flag_cleaned, parsed_file
    )

films_master_tbl |> glimpse()
```



c# Parse the Scripts


We now combine the tables to set up the film scripts parsing.

```{r combine_script_master_with_parser}
#| echo: true
#| message: false

default_parser_conf <- film_parser_loaded_tbl |>
  filter(title_cleaned == ".default_parser") |>
  pull(parser_conf) |>
  extract2(1)

film_parsers_tbl <- films_master_tbl |>
  left_join(film_parser_loaded_tbl, by = "title_cleaned") |>
  replace_na(
    list(parser_conf = list(default_parser_conf))
    ) |>
  mutate(
    parser_func = map(parser_conf, construct_film_script_parser)
    )

film_parsers_tbl |> glimpse()
```


As before, we expect to be running this code multiple times if we want to
add multiple film scripts so we only run the parser on scripts that do not
have a pre-parsed file in the 


```{r parse_film_scripts}
#! echo: true
#! message: false

read_parse_write_script <- function(parser_func, input_file, output_file) {
  
  message(glue("Parsing script {input_file}..."))
  
  parsed_lst <- parser_func(input_file)
  
  message(glue("Writing parsed data to {output_file}..."))
  
  parsed_lst |> write_rds(file = output_file)
  
  return(TRUE)
}


parsed_scripts_tbl <- dir_ls(parsed_script_dir, regexp = "\\.rds$") |>
  as.character() |>
  enframe(name = NULL, value = "parsed_file")

run_parsers_tbl <- film_parsers_tbl |>
  anti_join(parsed_scripts_tbl, by = "parsed_file") |>
  mutate(
    write_file = pmap_lgl(
      list(
        parser_func = parser_func,
        input_file  = cleaned_txtfile,
        output_file = parsed_file
        ),
      read_parse_write_script
      )
    )

run_parsers_tbl |> glimpse()
```

We now load up all our parsed data.

```{r loading_parsed_scripts}
#| echo: true
#| message: false

films_parsed_tbl <- films_master_tbl |>
  mutate(
    parsed_data = map(parsed_file, read_rds)    
    ) |>
  unnest(parsed_data)

films_parsed_tbl |> glimpse()
```



# Write to Disk

We now write this data out to disk.

```{r write_data_disk}
#| echo: true

films_master_tbl |> write_rds("data/films_master_tbl.rds")
```



# R Environment

```{r show_session_info}
#| echo: false
#| message: false


options(width = 120L)
sessioninfo::session_info()
options(width = 80L)
```
