---
title: "Workshop: Introduction to Information Theory — Part 1: Foundations"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---

```{r import_libraries}
#| echo: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(furrr)
library(glue)
library(fs)
library(entropy)
library(infotheo)
library(ggraph)
library(tidygraph)
library(broom)

source("lib_utils.R")
source("lib_information_theory.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "entropy", "infotheo")
)

options(
  width    = 80L,
  warn     = 1,
  mc.cores = parallelly::availableCores()
)

set.seed(42)

theme_set(theme_cowplot())

plan(multisession)
```


# The Big Idea: Quantifying Surprise

> *"Information is the resolution of uncertainty."*
> — Claude Shannon

In 1948, Claude Shannon published a paper called *A Mathematical Theory of
Communication* that changed the world. He set out to answer a deceptively
simple question: **how do you measure the amount of information in a message?**

The answer he found — *entropy* — turned out to be one of the most
far-reaching concepts in all of science. It connects probability theory,
thermodynamics, computer science, statistics, and even biology.

Before diving into formulas, let us build intuition.

## Surprise as the Core Unit

Consider two statements:

1. *"The sun rose this morning."*
2. *"A seven-headed dragon landed on the roof of Dublin Castle."*

Statement 1 contains almost no information — it is near-certain, so it
surprises you not at all. Statement 2 contains an enormous amount of
information precisely because it is so unexpected.

This is the key insight: **information content is inversely related to
probability**. Likely events carry little information; unlikely events carry a
lot. We formalise this with the *self-information* (or *surprisal*) of an
event $x$:

$$
I(x) = -\log_2 P(x)
$$

The choice of base 2 is conventional and gives us a unit called the **bit**
(binary digit). Base $e$ gives *nats*, base 10 gives *dits* (or *hartleys*).

```{r self_information_demo}
#| echo: true
#| fig-cap: "Self-information grows as probability falls. A certain event carries 0 bits; a 1-in-1024 event carries 10 bits."

p_seq <- seq(0.001, 1, by = 0.001)

surprisal_tbl <- tibble(
  probability      = p_seq,
  self_information = -log2(p_seq)
)

ggplot(surprisal_tbl, aes(x = probability, y = self_information)) +
  geom_line(colour = "steelblue", linewidth = 1.2) +
  geom_vline(
    xintercept = c(0.5, 0.25, 0.125),
    linetype   = "dashed",
    colour     = "firebrick",
    alpha      = 0.7
  ) +
  annotate(
    "text",
    x     = c(0.5, 0.25, 0.125),
    y     = 9,
    label = c("p=0.5\n(1 bit)", "p=0.25\n(2 bits)", "p=0.125\n(3 bits)"),
    hjust = -0.1,
    size  = 3.5
  ) +
  labs(
    title    = "Self-Information: How Surprising is an Event?",
    subtitle = "I(x) = -log₂ P(x)",
    x        = "Probability P(x)",
    y        = "Self-Information I(x) [bits]"
  )
```

### Why This Matters in Practice

Surprisal is the atomic unit behind many practical systems:

- In compression, rare symbols get longer codewords because they carry more
  surprisal.
- In anomaly detection, very low-probability events are flagged precisely
  because they contain high information content under the baseline model.
- In forecasting, a model that assigns high probability to what actually
  happens yields lower average surprisal (better predictive quality).

## Workshop Materials

All code and materials for this workshop are available on GitHub:
**https://github.com/kaybenleroll/data_workshops**

This workshop requires the following R packages:

- `entropy`: Entropy estimation from data
- `infotheo`: Mutual information and discretisation tools
- `ggraph` / `tidygraph`: Graph-based diagram visualisation
- `tidyverse`: Data manipulation and visualisation
- `cowplot`: Plot theming and arrangement
- `furrr`: Parallel purrr operations


# Shannon Entropy: Expected Surprise

The self-information $I(x)$ tells us how surprising *one specific event* is.
But what we usually want is: **on average, how surprising are outcomes of this
random variable?** That average surprise is **Shannon entropy**.

$$
H(X) = -\sum_{x \in \mathcal{X}} P(x) \log_2 P(x)
= \mathbb{E}\bigl[-\log_2 P(X)\bigr]
$$

The zero-probability convention $0 \cdot \log_2(0) = 0$ is standard and
consistent with the limit $\lim_{p \to 0^+} p \log p = 0$.

Shannon entropy has several deep properties that make it the canonical measure
of uncertainty:

- **Continuity**: tiny changes in probabilities produce tiny changes in entropy.
- **Maximum at uniformity**: for fixed support size, entropy is maximized by the
  uniform distribution.
- **Additivity for independent systems**: if $X \perp Y$, then
  $H(X, Y) = H(X) + H(Y)$.
- **Chain-rule compatibility**: uncertainty decomposes naturally via
  $H(X, Y) = H(X) + H(Y|X)$.

Together, these properties formalize why entropy is not just one possible
uncertainty score, but effectively the unique one (up to a scale constant)
under broad and intuitive axioms.

## The Coin Flip Thought Experiment

The most intuitive example is a biased coin with heads probability $p$.

```{r coin_entropy_plot}
#| echo: true
#| fig-cap: "Entropy of a biased coin. Maximum uncertainty at p=0.5 (exactly 1 bit). Entropy falls to 0 at both extremes — a certain outcome has no surprise."

p_vals  <- seq(0, 1, by = 0.001)
entropy_coin <- map_dbl(p_vals, function(p) {
  shannon_entropy(c(p, 1 - p))
})

tibble(p = p_vals, entropy = entropy_coin) |>
  ggplot(aes(x = p, y = entropy)) +
  geom_line(colour = "darkorange", linewidth = 1.3) +
  geom_point(
    data = tibble(p = 0.5, entropy = 1),
    colour = "firebrick", size = 4
  ) +
  annotate("text", x = 0.5, y = 1.05, label = "H = 1 bit\n(fair coin)", hjust = 0.5) +
  scale_x_continuous(labels = label_number(accuracy = 0.1)) +
  labs(
    title    = "Binary Entropy Function H(p)",
    subtitle = "H(p) = -p log₂(p) - (1-p) log₂(1-p)",
    x        = "P(Heads) = p",
    y        = "Entropy H(p) [bits]"
  )
```

Key observations from this plot:

- **Certainty → Zero entropy**: when $p = 0$ or $p = 1$, the outcome is
  perfectly predictable; there is nothing to learn.
- **Maximum uncertainty at $p = 0.5$**: a fair coin is maximally unpredictable
  and has $H = 1$ bit — exactly the amount of information that one binary
  question ("is it heads?") can resolve.
- **Symmetry**: the curve is symmetric around $p = 0.5$.

## Entropy Grows with the Number of Outcomes

For a die with $k$ equally likely faces, entropy is $H = \log_2 k$.

```{r entropy_vs_outcomes}
#| echo: true
#| fig-cap: "Entropy of a fair k-sided die grows as log₂(k)."

die_tbl <- tibble(
  k       = 2:64,
  entropy = map_dbl(2:64, function(k) shannon_entropy(rep(1/k, k)))
)

ggplot(die_tbl, aes(x = k, y = entropy)) +
  geom_line(colour = "steelblue", linewidth = 1.2) +
  geom_point(
    data   = filter(die_tbl, k %in% c(2, 4, 6, 8, 16, 32, 64)),
    colour = "firebrick",
    size   = 3
  ) +
  annotate(
    "text",
    x     = c(2, 6),
    y     = c(shannon_entropy(rep(1/2, 2)) - 0.25, shannon_entropy(rep(1/6, 6)) + 0.2),
    label = c("Coin (1 bit)", "Fair die (~2.6 bits)"),
    size  = 3.5
  ) +
  scale_x_continuous(breaks = c(2, 4, 8, 16, 32, 64)) +
  labs(
    title = "Entropy of a Fair k-Sided Die",
    x     = "Number of Outcomes k",
    y     = "H(X) [bits]"
  )
```

### Why This Matters in Practice

This curve is a direct complexity baseline:

- If your source has effectively $k$ equiprobable states, you should budget
  about $\log_2(k)$ bits/symbol for lossless encoding.
- Adding categories (states/classes) increases uncertainty sublinearly, not
  linearly.
- In modeling, this helps calibrate whether a prediction task is intrinsically
  low-entropy (easy) or high-entropy (hard).

## Uniform Distributions Maximise Entropy

A fundamental theorem of information theory states that, among all
distributions over a finite alphabet of size $k$, the **uniform distribution
maximises entropy** ($H = \log_2 k$). Any bias toward particular outcomes
reduces entropy.

```{r entropy_vs_bias}
#| echo: true
#| fig-cap: "Entropy drops as one outcome is made more probable. The uniform distribution (dashed line) is always the maximum."

compute_biased_entropy <- function(p_dominant, k) {
  remaining <- (1 - p_dominant) / (k - 1)
  shannon_entropy(c(p_dominant, rep(remaining, k - 1)))
}

k_vals    <- c(2, 4, 8, 16)
p_dom_seq <- seq(1/max(k_vals), 0.999, by = 0.01)

bias_tbl <- cross_df(list(k = k_vals, p_dominant = p_dom_seq)) |>
  filter(p_dominant >= 1/k) |>
  mutate(
    entropy_bits = map2_dbl(p_dominant, k, compute_biased_entropy),
    uniform_h    = log2(k),
    k_label      = glue("k = {k} outcomes")
  )

ggplot(bias_tbl, aes(x = p_dominant, y = entropy_bits, colour = factor(k))) +
  geom_line(linewidth = 1) +
  geom_hline(
    data    = distinct(bias_tbl, k, uniform_h, k_label),
    aes(yintercept = uniform_h, colour = factor(k)),
    linetype = "dashed",
    alpha    = 0.6
  ) +
  facet_wrap(~k_label, scales = "free_y") +
  scale_colour_brewer(palette = "Set1", guide = "none") +
  labs(
    title    = "Entropy Decreases as One Outcome Dominates",
    subtitle = "Dashed line = maximum entropy (uniform distribution)",
    x        = "Probability of Dominant Outcome",
    y        = "Entropy [bits]"
  )
```

### Why This Matters in Practice

Entropy reduction from dominance is exactly what compression exploits:

- If a few outcomes dominate, shorter expected code lengths are possible.
- If outcomes are near-uniform, gains from lossless compression are limited.
- In risk terms, dominance often means predictability concentration, which can
  be valuable for control but dangerous if the dominant state shifts.

## Simulating Entropy Estimation

In practice, we rarely know the true distribution. We estimate entropy from
data. How quickly does the estimate converge as we collect more samples?

```{r entropy_convergence}
#| echo: true
#| fig-cap: "Entropy estimates converge to the true value (red dashed line) but are biased downward at small samples — known as the Miller-Madow bias."

# True distribution: loaded 6-sided die
true_probs <- c(0.30, 0.20, 0.20, 0.15, 0.10, 0.05)
true_H     <- shannon_entropy(true_probs)

sample_sizes <- c(10, 25, 50, 100, 250, 500, 1000, 2500, 5000)

conv_tbl <- simulate_entropy_convergence(
  probs        = true_probs,
  sample_sizes = sample_sizes,
  n_reps       = 200L
)

conv_summary <- conv_tbl |>
  group_by(n) |>
  summarise(
    mean_h  = mean(entropy_estimate),
    lo      = quantile(entropy_estimate, 0.1),
    hi      = quantile(entropy_estimate, 0.9),
    .groups = "drop"
  )

ggplot(conv_summary, aes(x = n)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = mean_h), colour = "steelblue", linewidth = 1.2) +
  geom_hline(yintercept = true_H, colour = "firebrick", linetype = "dashed") +
  annotate("text", x = 5000, y = true_H + 0.02,
           label = glue("True H = {round(true_H, 3)} bits"), hjust = 1.1, colour = "firebrick") +
  scale_x_log10(labels = label_comma()) +
  labs(
    title    = "Convergence of Entropy Estimate with Sample Size",
    subtitle = "Shaded band: 10th–90th percentile over 200 replications",
    x        = "Sample Size n (log scale)",
    y        = "Estimated Entropy [bits]"
  )
```

The downward bias at small samples is well known — the empirical distribution
has fewer "occupied" bins, systematically underestimating entropy. Several
bias-corrected estimators exist (e.g., the Miller-Madow correction); the
`entropy` package implements many of them.

```{r entropy_package_demo}
#| echo: true

# Compare naive estimator vs. bias-corrected (MM = Miller-Madow, JS = James-Stein)
set.seed(123)
small_sample <- sample(1:6, size = 30, replace = TRUE, prob = true_probs)
counts_30    <- tabulate(small_sample, nbins = 6)

naive_h <- entropy_from_counts(counts_30)
mm_h    <- entropy::entropy(counts_30, method = "MM",  unit = "log2")
js_h    <- entropy::entropy(counts_30, method = "SG",  unit = "log2")

cat(glue(
  "True entropy        : {round(true_H, 4)} bits\n",
  "Naive (plug-in)     : {round(naive_h, 4)} bits\n",
  "Miller-Madow (MM)   : {round(mm_h,   4)} bits\n",
  "Shrinkage (SG)      : {round(js_h,   4)} bits\n"
))
```

### Why This Matters in Practice

Entropy estimation is often done under data scarcity. That means estimator
choice is not cosmetic:

- For small samples, naive estimates can understate uncertainty and lead to
  overconfident conclusions.
- Bias-corrected estimators are useful whenever you compare uncertainty across
  groups with different sample sizes.
- In feature screening, entropy bias can propagate directly into mutual
  information rankings.


# Rényi Entropy: A Family of Entropies

Shannon entropy is the most famous member of a broader family. The **Rényi
entropy** of order $\alpha$ is:

$$
H_\alpha(X) = \frac{1}{1 - \alpha} \log_2 \left( \sum_{x} P(x)^\alpha \right)
$$

Different values of $\alpha$ emphasise different aspects of a distribution:

| $\alpha$ | Name | Emphasis |
|----------|------|----------|
| 0 | Hartley entropy | Support size only — ignores probabilities |
| 1 (limit) | Shannon entropy | Average surprise |
| 2 | Collision entropy | Probability of drawing the same value twice |
| $\infty$ | Min-entropy | Worst-case / most likely outcome |

The family is ordered: $H_0 \geq H_1 \geq H_2 \geq \cdots \geq H_\infty$.
This ordering reflects that higher $\alpha$ places more weight on larger
probabilities (i.e., less uncertainty is measured).

```{r renyi_family_plot}
#| echo: true
#| fig-cap: "Rényi entropy as a function of alpha for two distributions. The family interpolates between Hartley entropy (alpha=0) and min-entropy (alpha→∞). Shannon entropy (alpha=1) is the most natural choice."

# Two illustrative distributions
p_uniform  <- rep(1/8, 8)
p_skewed   <- c(0.50, 0.25, 0.10, 0.05, 0.04, 0.03, 0.02, 0.01)

alpha_seq <- c(seq(0.1, 0.9, 0.05), seq(1.1, 10, 0.2))

renyi_tbl <- tibble(
  alpha   = alpha_seq,
  uniform = map_dbl(alpha_seq, ~renyi_entropy(p_uniform, .x)),
  skewed  = map_dbl(alpha_seq, ~renyi_entropy(p_skewed,  .x))
) |>
  pivot_longer(-alpha, names_to = "distribution", values_to = "entropy")

# Add Shannon points (alpha = 1)
shannon_points <- tibble(
  alpha        = 1,
  distribution = c("uniform", "skewed"),
  entropy      = c(shannon_entropy(p_uniform), shannon_entropy(p_skewed))
)

ggplot(renyi_tbl, aes(x = alpha, y = entropy, colour = distribution)) +
  geom_line(linewidth = 1.2) +
  geom_point(data = shannon_points, size = 4, shape = 18) +
  annotate("text", x = 1, y = 3.4, label = "Shannon\n(α=1)", size = 3.5, hjust = -0.1) +
  scale_colour_manual(
    values = c("uniform" = "steelblue", "skewed" = "darkorange"),
    labels = c("Uniform (8 outcomes)", "Skewed (8 outcomes)")
  ) +
  labs(
    title    = "Rényi Entropy Family",
    subtitle = "Higher α emphasises the most probable outcomes more heavily",
    x        = "Order α",
    y        = "Rényi Entropy [bits]",
    colour   = "Distribution"
  ) +
  theme(legend.position = "bottom")
```

### Why This Matters in Practice

Rényi entropy gives you a tunable lens on uncertainty:

- Lower $\alpha$ values emphasize coverage and support size.
- Higher $\alpha$ values emphasize concentration and worst-case dominance.
- Security and reliability applications often care more about high-$\alpha$
  regimes (collision/min-entropy) than Shannon entropy alone.

## Collision Entropy and Password Strength

Rényi-2 (collision entropy) has a natural security interpretation. If an
adversary tries to guess a password by drawing uniformly, the probability of
success in one guess is:

$$
P(\text{collision}) = \sum_x P(x)^2 = 2^{-H_2(X)}
$$

```{r renyi_security_demo}
#| echo: true
#| fig-cap: "Comparing password distributions via Rényi entropy. The collision probability determines how many guesses an attacker needs."

# Simulated password strength distributions
english_letter_freq <- c(8.2, 1.5, 2.8, 4.3, 12.7, 2.2, 2.0, 6.1, 7.0, 0.15,
                         0.77, 4.0, 2.4, 6.7, 7.5, 1.9, 0.1, 6.0, 6.3, 9.1,
                         2.8, 0.98, 2.4, 0.15, 2.0, 0.074)
english_letter_freq <- english_letter_freq / sum(english_letter_freq)

# For i.i.d. letters, entropy over 4-char strings is 4x letter entropy.
english_h1_char <- shannon_entropy(english_letter_freq)
english_h2_char <- renyi_entropy(english_letter_freq, alpha = 2)
english_hinf_char <- renyi_entropy(english_letter_freq, alpha = Inf)

password_tbl <- tribble(
  ~scheme,               ~support_size, ~hartley_bits,         ~shannon_bits,            ~collision_bits,             ~min_entropy_bits,            ~collision_prob,
  "PIN (0000-9999)",    10000,         log2(10000),           log2(10000),              log2(10000),                 log2(10000),                  1 / 10000,
  "4-letter English",   26^4,          log2(26^4),            4 * english_h1_char,      4 * english_h2_char,         4 * english_hinf_char,        2^(-4 * english_h2_char),
  "6-char alphanumeric",62^6,          log2(62^6),            log2(62^6),               log2(62^6),                  log2(62^6),                   1 / (62^6)
)

password_tbl |>
  select(-support_size, -collision_prob) |>
  pivot_longer(-scheme, names_to = "measure", values_to = "bits") |>
  mutate(
    measure = factor(measure, levels = c("hartley_bits", "shannon_bits", "collision_bits", "min_entropy_bits"),
                     labels = c("H₀ (Hartley)", "H₁ (Shannon)", "H₂ (Collision)", "H∞ (Min-entropy)"))
  ) |>
  ggplot(aes(x = scheme, y = bits, fill = measure)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title  = "Password Strength: Rényi Entropy Family",
    x      = "Password Scheme",
    y      = "Entropy [bits]",
    fill   = "Entropy Order"
  ) +
  theme(
    axis.text.x  = element_text(angle = 20, hjust = 1),
    legend.position = "bottom"
  )
```

### Why This Matters in Practice

Different entropy orders answer different security questions:

- Shannon entropy reflects average-case uncertainty.
- Collision entropy relates to repeated-guess collision risk.
- Min-entropy reflects worst-case predictability and is often the most
  conservative credential-strength metric.


# Joint and Conditional Entropy

So far we have measured uncertainty about a single variable. Information theory
extends naturally to **multiple variables** through joint and conditional
entropy.

## The Entropy Chain Rule

For two discrete random variables $X$ and $Y$ with joint distribution
$P(X, Y)$:

$$
H(X, Y) = -\sum_x \sum_y P(x, y) \log_2 P(x, y)
$$

$$
H(Y \mid X) = H(X, Y) - H(X)
$$

The chain rule $H(X, Y) = H(X) + H(Y \mid X)$ states: *total uncertainty
equals uncertainty about $X$ plus the remaining uncertainty about $Y$ once $X$
is known*.

## Simulating Joint Entropy

```{r joint_entropy_demo}
#| echo: true
#| fig-cap: "Joint distribution between weather and umbrella-carrying. The joint entropy H(X,Y) = H(X) + H(Y|X)."

# Weather: Sunny / Rainy — P(Sunny) = 0.7
# Umbrella given weather (conditional)
# P(Umbrella | Rainy) = 0.90, P(Umbrella | Sunny) = 0.05

weather_probs <- c(Sunny = 0.7, Rainy = 0.3)
umbrella_given_weather <- matrix(
  c(0.95, 0.05,   # Sunny: no umbrella, umbrella
    0.10, 0.90),  # Rainy: no umbrella, umbrella
  nrow = 2, byrow = TRUE,
  dimnames = list(c("Sunny", "Rainy"), c("No Umbrella", "Umbrella"))
)

# Joint distribution: P(W=w, U=u) = P(U|W) * P(W)
joint <- sweep(umbrella_given_weather, MARGIN = 1, weather_probs, `*`)

h_weather   <- shannon_entropy(weather_probs)
h_joint     <- joint_entropy(joint)
h_cond      <- conditional_entropy(joint)
h_umbrella  <- shannon_entropy(colSums(joint))
mi_wu       <- mutual_information(joint)

cat(glue(
  "H(Weather)         = {round(h_weather,  4)} bits\n",
  "H(Umbrella)        = {round(h_umbrella, 4)} bits\n",
  "H(Weather, Umbr.)  = {round(h_joint,    4)} bits\n",
  "H(Umbrella|Weather)= {round(h_cond,     4)} bits\n",
  "I(Weather;Umbrella)= {round(mi_wu,      4)} bits\n",
  "\nCheck: H(W) + H(U|W) = {round(h_weather + h_cond, 4)} ≈ H(W,U) = {round(h_joint, 4)}\n"
))
```

### Why This Matters in Practice

The chain rule is a blueprint for staged decision systems:

- In triage, you first resolve broad uncertainty (e.g., segment/risk band),
  then conditional uncertainty within that segment.
- In diagnostics, each measurement should ideally reduce conditional entropy.
- In process design, low $H(Y|X)$ means the upstream variable $X$ is genuinely
  informative, not just correlated by chance.

```{r joint_entropy_heatmap}
#| echo: true
#| fig-cap: "Joint distribution P(Weather, Umbrella). The off-diagonal cells (unexpected combinations) have low probability — the variables are highly dependent."

joint |>
  as_tibble(rownames = "weather") |>
  pivot_longer(-weather, names_to = "umbrella", values_to = "prob") |>
  mutate(
    label = glue("{round(prob * 100, 1)}%"),
    info  = round(-log2(prob), 2)
  ) |>
  ggplot(aes(x = umbrella, y = weather, fill = prob)) +
  geom_tile(colour = "white", linewidth = 1.5) +
  geom_text(aes(label = label), colour = "white", size = 7, fontface = "bold") +
  scale_fill_gradient(low = "steelblue", high = "firebrick", labels = label_percent()) +
  labs(
    title = "Joint Distribution P(Weather, Umbrella)",
    x     = "Umbrella",
    y     = "Weather",
    fill  = "Probability"
  )
```

## Independence vs. Dependence

When $X$ and $Y$ are independent, $H(X, Y) = H(X) + H(Y)$ — joint entropy is
purely additive. **Any dependence reduces joint entropy below this sum.**

```{r independence_vs_dependence}
#| echo: true
#| fig-cap: "Scanning over the conditional probability P(Y=1 | X=1), moving from independence to full dependence. Joint entropy decreases monotonically as dependence increases."

# X: fair coin. Y depends on X to varying degrees.
p_x1 <- 0.5

dep_seq <- seq(0.5, 1.0, by = 0.01)

dep_tbl <- tibble(
  p_y1_given_x1 = dep_seq
) |>
  mutate(
    # P(Y=1|X=0) mirrors so that P(Y=1) = 0.5
    p_y1_given_x0 = 1 - dep_seq,
    joint = map2(p_y1_given_x1, p_y1_given_x0, function(p11, p10) {
      matrix(
        c(p_x1 * (1 - p11), p_x1 * p11,
          (1 - p_x1) * (1 - p10), (1 - p_x1) * p10),
        nrow = 2, ncol = 2
      )
    }),
    h_joint  = map_dbl(joint, joint_entropy),
    h_x      = shannon_entropy(c(p_x1, 1 - p_x1)),
    h_y      = map_dbl(joint, ~shannon_entropy(colSums(.x))),
    h_indep  = h_x + h_y,
    mi       = map_dbl(joint, mutual_information)
  )

p1_dep <- ggplot(dep_tbl, aes(x = p_y1_given_x1)) +
  geom_line(aes(y = h_joint, colour = "H(X,Y)"), linewidth = 1.2) +
  geom_line(aes(y = h_indep, colour = "H(X)+H(Y) [independence]"),
            linetype = "dashed", linewidth = 1) +
  scale_colour_manual(values = c("steelblue", "darkorange")) +
  labs(x = "P(Y=1 | X=1)", y = "Entropy [bits]", colour = NULL) +
  theme(legend.position = "bottom")

p2_dep <- ggplot(dep_tbl, aes(x = p_y1_given_x1, y = mi)) +
  geom_line(colour = "firebrick", linewidth = 1.2) +
  labs(x = "P(Y=1 | X=1)", y = "Mutual Information [bits]")

plot_grid(p1_dep, p2_dep, ncol = 2, labels = c("A", "B"))
```

### Why This Matters in Practice

The dependence scan is useful for feature engineering and monitoring:

- As dependence increases, MI rises even when marginals are unchanged.
- Joint entropy drops relative to $H(X)+H(Y)$, indicating less total uncertainty
  than under an independence assumption.
- This is exactly the signal exploited in redundancy reduction and
  interaction-aware models.


# Mutual Information

Mutual information $I(X; Y)$ measures how much knowing $X$ tells us about
$Y$ (and vice versa — it is symmetric):

$$
I(X; Y) = H(X) + H(Y) - H(X, Y)
= H(X) - H(X \mid Y)
= H(Y) - H(Y \mid X)
$$

An equivalent KL-divergence form makes it clear that $I(X;Y) \geq 0$, with
equality if and only if $X$ and $Y$ are independent:

$$
I(X; Y) = \sum_{x,y} P(x, y) \log_2 \frac{P(x, y)}{P(x) P(y)}
= D_{\text{KL}}\bigl(P_{XY} \| P_X \otimes P_Y\bigr)
$$

Two complementary ways to think about MI are useful in practice:

- **Uncertainty reduction view**:
  $I(X;Y) = H(Y) - H(Y|X)$. Knowing $X$ reduces uncertainty about $Y$ by
  exactly $I(X;Y)$ bits on average.
- **Distribution gap view**:
  $I(X;Y)$ is the KL gap between the real joint distribution and the product of
  marginals. If the joint truly factorizes, there is no dependence to exploit,
  so MI is zero.

This is why MI is broadly useful in feature selection, representation learning,
and dependency diagnostics: it quantifies any predictable structure between two
variables, not just linear comovement.

Another key theoretical property is the **data processing inequality**:
if $X \to Z \to Y$ is a Markov chain (you only observe $X$ through a possibly
lossy transformation $Z$), then

$$
I(X;Y) \le I(Z;Y)
$$

In plain terms: post-processing cannot create information that was not already
present.

Why should we care operationally? Because MI is a direct answer to questions
like:

- "How informative is this feature about the target?"
- "How much signal survives after compression or transformation?"
- "Is this sensor/channel worth adding, or is it redundant?"

If you think in terms of prediction, MI is "how many bits of uncertainty about
the target disappear once this input is known". If MI is tiny, the variable may
still correlate weakly, but it carries little decision-relevant information.

## Toy MI Calculations

The fastest way to make MI intuitive is to compute it on three simple cases:

1. Independent variables (MI should be 0)
2. Perfectly coupled variables (MI should equal entropy of either variable)
3. Noisy coupling (MI should sit between 0 and the perfect-coupling value)

```{r toy_mi_examples}
#| echo: true

# 1) Independent binary variables
joint_indep <- matrix(c(0.25, 0.25,
                        0.25, 0.25), nrow = 2, byrow = TRUE)

# 2) Perfect coupling (Y = X)
joint_perfect <- matrix(c(0.50, 0.00,
                          0.00, 0.50), nrow = 2, byrow = TRUE)

# 3) Noisy coupling: Y equals X with prob 0.8
joint_noisy <- matrix(c(0.40, 0.10,
                        0.10, 0.40), nrow = 2, byrow = TRUE)

toy_mi_tbl <- tibble(
  case = c("Independent", "Perfect coupling", "Noisy coupling (80% accurate)"),
  h_x  = c(shannon_entropy(rowSums(joint_indep)),
           shannon_entropy(rowSums(joint_perfect)),
           shannon_entropy(rowSums(joint_noisy))),
  h_y  = c(shannon_entropy(colSums(joint_indep)),
           shannon_entropy(colSums(joint_perfect)),
           shannon_entropy(colSums(joint_noisy))),
  h_y_given_x = c(conditional_entropy(joint_indep),
                  conditional_entropy(joint_perfect),
                  conditional_entropy(joint_noisy)),
  mi_bits = c(mutual_information(joint_indep),
              mutual_information(joint_perfect),
              mutual_information(joint_noisy))
)

toy_mi_tbl |> knitr::kable(digits = 4, caption = "Toy Mutual Information Calculations")
```

You can read this table as a diagnostic playbook:

- If $H(Y|X)$ is almost as large as $H(Y)$, then $X$ tells you little about
  $Y$ (low MI).
- If $H(Y|X)$ is near zero, then $X$ nearly determines $Y$ (high MI).
- In practice, most real systems look like the noisy middle case.

## The Information Venn Diagram

```{r venn_diagram}
#| echo: true
#| fig-cap: "The information-theoretic Venn diagram. Overlapping area = I(X;Y). Non-overlapping areas = conditional entropies. Total area = H(X,Y)."

# Draw the classic Venn diagram manually
circle_data <- tibble(
  x      = c(-0.75, 0.75),
  y      = c(0, 0),
  r      = c(1.5, 1.5),
  label  = c("X", "Y"),
  lx     = c(-1.5, 1.5)
)

theta <- seq(0, 2 * pi, length.out = 200)

circle_tbl <- circle_data |>
  pmap_dfr(function(x, y, r, label, lx) {
    tibble(
      cx    = x + r * cos(theta),
      cy    = y + r * sin(theta),
      group = label
    )
  })

ggplot() +
  geom_polygon(
    data    = filter(circle_tbl, group == "X"),
    aes(x = cx, y = cy),
    fill    = "steelblue",
    alpha   = 0.4,
    colour  = "steelblue",
    linewidth = 1
  ) +
  geom_polygon(
    data    = filter(circle_tbl, group == "Y"),
    aes(x = cx, y = cy),
    fill    = "darkorange",
    alpha   = 0.4,
    colour  = "darkorange",
    linewidth = 1
  ) +
  annotate("text", x = -1.4, y =  0.0, label = "H(X|Y)", size = 5, fontface = "bold") +
  annotate("text", x =  0.0, y =  0.0, label = "I(X;Y)", size = 5, fontface = "bold") +
  annotate("text", x =  1.4, y =  0.0, label = "H(Y|X)", size = 5, fontface = "bold") +
  annotate("text", x = -0.75, y =  1.8, label = "X",  size = 8, fontface = "bold", colour = "steelblue") +
  annotate("text", x =  0.75, y =  1.8, label = "Y",  size = 8, fontface = "bold", colour = "darkorange") +
  coord_equal() +
  theme_void() +
  labs(title = "Information Venn Diagram", subtitle = "H(X,Y) = Total area of both circles")
```

### Why This Matters in Practice

The Venn diagram is a mnemonic, not a geometric proof. Its value is conceptual:

- Overlap reminds you that shared predictability is a finite resource.
- Non-overlap reminds you conditional uncertainty can remain high even with
  non-zero dependence.
- It is a quick diagnostic mental model when reasoning about feature sets and
  target leakage.

## Mutual Information as a Measure of Association

Unlike Pearson correlation, MI captures *all* statistical dependencies —
not just linear ones.

```{r mi_vs_correlation}
#| echo: true
#| fig-cap: "MI vs. Pearson correlation on three relationships. MI detects the quadratic and circular patterns that correlation misses."

n_pts <- 1000

linear_tbl    <- tibble(x = rnorm(n_pts), y = 0.8 * x + rnorm(n_pts, sd = 0.6), rel = "Linear")
quadratic_tbl <- tibble(x = runif(n_pts, -2, 2), y = x^2 + rnorm(n_pts, sd = 0.3), rel = "Quadratic")
circular_tbl  <- {
  theta_c <- runif(n_pts, 0, 2 * pi)
  tibble(x = cos(theta_c) + rnorm(n_pts, sd = 0.1),
         y = sin(theta_c) + rnorm(n_pts, sd = 0.1),
         rel = "Circular")
}

all_rel <- bind_rows(linear_tbl, quadratic_tbl, circular_tbl)

stats_tbl <- all_rel |>
  group_by(rel) |>
  summarise(
    pearson_r = cor(x, y),
    mi_bits   = estimate_mi_histogram(x, y, n_bins_x = 20L, n_bins_y = 20L),
    .groups   = "drop"
  ) |>
  mutate(label = glue("r = {round(pearson_r, 2)} | MI = {round(mi_bits, 2)} bits"))

ggplot(left_join(all_rel, stats_tbl, by = "rel"),
       aes(x = x, y = y)) +
  geom_point(alpha = 0.25, size = 0.8, colour = "steelblue") +
  geom_text(
    data = distinct(left_join(all_rel, stats_tbl, by = "rel"), rel, label),
    aes(x = Inf, y = Inf, label = label),
    hjust = 1.05, vjust = 1.5, size = 3.5, fontface = "bold"
  ) +
  facet_wrap(~rel, scales = "free") +
  labs(
    title    = "Mutual Information vs. Pearson Correlation",
    subtitle = "MI detects non-linear dependencies that Pearson misses",
    x        = "X",
    y        = "Y"
  )
```

### Why This Matters in Practice

A common modeling failure mode is dropping non-linear predictors because
Pearson correlation looks small. MI helps avoid that:

- Use correlation for quick linear diagnostics.
- Use MI when relationship form is unknown or potentially non-linear.
- If MI is high but correlation is low, flexible models or feature transforms
  are likely warranted.


# KL Divergence and Cross-Entropy

Up to this point, we measured uncertainty inside a single distribution
(entropy) and shared information between variables (mutual information). The
next natural question is different:

**What if the distribution we use is wrong?**

That question is central in real modeling work. We rarely know the true
distribution exactly; we approximate it. KL divergence and cross-entropy are
the tools that quantify the cost of that approximation.

This section answers three practical questions:

1. How far is my model distribution from reality?
2. Why does the direction of that comparison matter?
3. Why is cross-entropy the default loss in modern classification models?

You can think of this section as the bridge from "information as uncertainty"
to "information as model quality and decision cost".

## Kullback-Leibler Divergence

The **KL divergence** $D_{\text{KL}}(P \| Q)$ measures how much information
is lost when we use distribution $Q$ to approximate distribution $P$:

$$
D_{\text{KL}}(P \| Q) = \sum_x P(x) \log_2 \frac{P(x)}{Q(x)}
$$

It is always $\geq 0$ (Gibbs' inequality), with equality iff $P = Q$.
Critically, **it is not symmetric** — $D_{\text{KL}}(P \| Q) \neq D_{\text{KL}}(Q \| P)$.

An operational interpretation is **expected coding regret**. If outcomes are
actually generated from $P$ but we design code lengths for $Q$, then the
average excess bits per symbol we pay is exactly $D_{\text{KL}}(P\|Q)$.

The same quantity appears in statistics and ML as expected log-loss gap under
model misspecification. This is why KL is central in likelihood methods,
variational inference, and Bayesian updating.

Another way to remember KL: it is a **directional mismatch penalty**.

- $D_{KL}(P||Q)$ asks: "If nature is $P$, how costly is it to act as if it were
  $Q$?"
- The reverse asks a different question. Hence asymmetry is not a bug; it is a
  feature of the question being asked.

## Toy KL and Cross-Entropy Calculations

```{r toy_kl_crossentropy_examples}
#| echo: true

# True class distribution P and two candidate models Q_good / Q_bad
p_true <- c(0.70, 0.20, 0.10)
q_good <- c(0.65, 0.25, 0.10)
q_bad  <- c(0.40, 0.30, 0.30)

h_p <- shannon_entropy(p_true)

toy_kl_tbl <- tibble(
  model = c("Q_good", "Q_bad"),
  kl_p_q = c(kl_divergence(p_true, q_good), kl_divergence(p_true, q_bad)),
  kl_q_p = c(kl_divergence(q_good, p_true), kl_divergence(q_bad, p_true)),
  cross_entropy = c(cross_entropy(p_true, q_good), cross_entropy(p_true, q_bad)),
  decomposition_check = cross_entropy - (h_p + kl_p_q)
)

cat(glue("H(P) = {round(h_p, 4)} bits\n\n"))
toy_kl_tbl |> knitr::kable(digits = 4, caption = "Toy KL and Cross-Entropy Calculations")
```

Interpretation:

- `Q_good` has small KL, so cross-entropy is only slightly above $H(P)$.
- `Q_bad` has much larger KL, so coding/regret penalty is materially larger.
- `KL(P||Q)` and `KL(Q||P)` differ, reinforcing the directional nature.
- The decomposition check should be ~0 numerically, confirming
  $H(P,Q)=H(P)+D_{KL}(P||Q)$.

```{r kl_asymmetry_demo}
#| echo: true
#| fig-cap: "Asymmetry of KL divergence illustrated on Gaussians. KL(P||Q) penalises regions where P has mass but Q does not. KL(Q||P) is the reverse."

x_grid <- seq(-6, 10, by = 0.05)

# Two Gaussians: P is the truth, Q is an approximating model
p_dens <- dnorm(x_grid, mean = 1, sd = 1)
q_dens <- dnorm(x_grid, mean = 3, sd = 2)

kl_pq_integrand <- ifelse(p_dens < 1e-20, 0, p_dens * log2(p_dens / (q_dens + 1e-20)))
kl_qp_integrand <- ifelse(q_dens < 1e-20, 0, q_dens * log2(q_dens / (p_dens + 1e-20)))

dx         <- diff(x_grid)[1]
kl_pq_val  <- sum(kl_pq_integrand) * dx
kl_qp_val  <- sum(kl_qp_integrand) * dx

tibble(
  x     = x_grid,
  P     = p_dens,
  Q     = q_dens,
  kl_pq = kl_pq_integrand,
  kl_qp = kl_qp_integrand
) |>
  pivot_longer(-x, names_to = "series", values_to = "value") |>
  filter(series %in% c("P", "Q")) |>
  ggplot(aes(x = x, y = value, colour = series)) +
  geom_line(linewidth = 1.2) +
  annotate("text", x = 8, y = 0.35,
           label = glue("KL(P||Q) = {round(kl_pq_val, 3)} bits"), colour = "steelblue",  hjust = 1) +
  annotate("text", x = 8, y = 0.30,
           label = glue("KL(Q||P) = {round(kl_qp_val, 3)} bits"), colour = "darkorange", hjust = 1) +
  scale_colour_manual(
    values = c("P" = "steelblue", "Q" = "darkorange"),
    labels = c("P: Truth N(1,1)", "Q: Model N(3,2)")
  ) +
  labs(
    title    = "KL Divergence is Not Symmetric",
    subtitle = "P is the truth, Q is the approximating model",
    x        = "x",
    y        = "Density",
    colour   = NULL
  ) +
  theme(legend.position = "bottom")
```

## Forward vs. Reverse KL

The direction of the KL divergence matters enormously in practice:

- **Forward KL $D_{\text{KL}}(P \| Q)$ ("mean-seeking")**: The model $Q$
  must cover all regions where $P$ has mass. Zero-avoiding for $Q$. Used in
  maximum likelihood estimation.
- **Reverse KL $D_{\text{KL}}(Q \| P)$ ("mode-seeking")**: The model $Q$
  concentrates on the modes of $P$. Zero-forcing for $Q$. Used in variational
  inference.

```{r forward_vs_reverse_kl}
#| echo: true
#| fig-cap: "Forward KL finds an average fit; reverse KL locks onto one mode. Both directions minimised against a bimodal target."

x_grid <- seq(-8, 8, by = 0.05)
dx     <- diff(x_grid)[1]

# True distribution: bimodal
p_true <- 0.5 * dnorm(x_grid, mean = -3, sd = 1) +
          0.5 * dnorm(x_grid, mean =  3, sd = 1)
p_true <- p_true / (sum(p_true) * dx)

# Forward KL optimal: Gaussian with mean and variance matched to P
p_mean <- sum(x_grid * p_true) * dx
p_var  <- sum((x_grid - p_mean)^2 * p_true) * dx
q_forward <- dnorm(x_grid, mean = p_mean, sd = sqrt(p_var))

# Reverse KL optimal: concentrate on one mode
q_reverse <- dnorm(x_grid, mean = 3, sd = 1)

plot_tbl <- tibble(
  x         = x_grid,
  `P (Truth)` = p_true,
  `Q (Forward KL)` = q_forward,
  `Q (Reverse KL)` = q_reverse
) |>
  pivot_longer(-x, names_to = "series", values_to = "density")

ggplot(plot_tbl, aes(x = x, y = density, colour = series, linetype = series)) +
  geom_line(linewidth = 1.2) +
  scale_colour_manual(values = c("steelblue", "firebrick", "darkorange")) +
  scale_linetype_manual(values = c("solid", "dashed", "dotdash")) +
  labs(
    title    = "Forward vs. Reverse KL on a Bimodal Distribution",
    subtitle = "Forward KL is mean-seeking; Reverse KL is mode-seeking",
    x        = "x",
    y        = "Density",
    colour   = NULL,
    linetype = NULL
  ) +
  theme(legend.position = "bottom")
```

### Why This Matters in Practice

This choice affects model behavior materially:

- If missing any plausible region is costly, forward-KL-style objectives are
  safer.
- If concentrating on high-probability modes is acceptable, reverse-KL-style
  objectives can be more efficient.
- Many approximation algorithms differ less by architecture than by which
  divergence direction they optimize.

## Cross-Entropy and the ML Connection

Cross-entropy connects information theory directly to machine learning:

$$
H(P, Q) = -\sum_x P(x) \log_2 Q(x) = H(P) + D_{\text{KL}}(P \| Q)
$$

This decomposition is worth pausing on:

- $H(P)$ is the irreducible uncertainty in the true data-generating process.
- $D_{\text{KL}}(P\|Q)$ is the additional penalty from using an imperfect model.

So cross-entropy is "unavoidable uncertainty + model error". In training,
only the second term is controllable.

For binary classification with true label $y \in \{0,1\}$ and predicted
probability $q = P_\theta(y=1|x)$, the per-example loss is

$$
\ell_{CE}(y,q) = -\bigl[y\log q + (1-y)\log(1-q)\bigr]
$$

This explains why overconfident wrong predictions are punished severely: if
$y=1$ but $q \to 0$, the loss diverges.

When you minimise cross-entropy loss in a neural network, you are
simultaneously minimising the KL divergence between the true label
distribution $P$ and your model's predicted distribution $Q$. The entropy
$H(P)$ is a constant (given the data), so minimising $H(P, Q)$ is exactly
equivalent to minimising $D_{\text{KL}}(P \| Q)$.

```{r cross_entropy_loss}
#| echo: true
#| fig-cap: "Cross-entropy loss as a function of predicted probability q for a true class (p=1). The loss is unbounded as q→0 and zero only when q=1."

q_seq <- seq(0.01, 0.99, by = 0.01)

ce_tbl <- tibble(
  q                    = q_seq,
  cross_entropy_bits   = -log2(q_seq),
  cross_entropy_nats   = -log(q_seq)
)

p1_ce <- ggplot(ce_tbl, aes(x = q, y = cross_entropy_bits)) +
  geom_line(colour = "firebrick", linewidth = 1.3) +
  geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  labs(
    title = "Cross-Entropy Loss (Binary, True Class)",
    x     = "Predicted Probability q",
    y     = "Loss [bits]"
  )

p2_ce <- ggplot(ce_tbl, aes(x = q, y = cross_entropy_nats)) +
  geom_line(colour = "steelblue", linewidth = 1.3) +
  geom_vline(xintercept = 0.5, linetype = "dashed", alpha = 0.5) +
  labs(
    title = "Cross-Entropy Loss (Binary, True Class)",
    x     = "Predicted Probability q",
    y     = "Loss [nats]"
  )

plot_grid(p1_ce, p2_ce, ncol = 2)
```

### Why This Matters in Practice

Cross-entropy is not just a training loss; it is a decision-quality metric:

- It rewards calibrated confidence, not just correct class rank.
- It strongly penalizes catastrophic overconfidence.
- Improvements in cross-entropy usually translate into better probability
  forecasts and downstream decision thresholds.

## Jensen-Shannon Divergence: A Symmetric Alternative

The JSD symmetrises KL divergence by comparing each distribution to their
mixture $M = (P + Q)/2$:

$$
\text{JSD}(P \| Q) = \frac{1}{2} D_{\text{KL}}(P \| M) + \frac{1}{2} D_{\text{KL}}(Q \| M)
$$

$\text{JSD} \in [0, 1]$ (bits), and $\sqrt{\text{JSD}}$ is a proper metric
(satisfies the triangle inequality). It is widely used to compare language
model output distributions.

```{r jsd_demo}
#| echo: true
#| fig-cap: "JSD between two Gaussians as their means separate. JSD is bounded in [0,1]; it reaches its maximum as the distributions become disjoint."

sigma_fixed <- 1.0
mu_sep      <- seq(0, 8, by = 0.25)

jsd_tbl <- tibble(
  separation = mu_sep,
  jsd = map_dbl(mu_sep, function(d) {
    # Discretise to compute numerically
    x_g <- seq(-10, 10, by = 0.05)
    dx  <- diff(x_g)[1]
    p   <- dnorm(x_g, mean =  d/2, sd = sigma_fixed)
    q   <- dnorm(x_g, mean = -d/2, sd = sigma_fixed)
    p   <- p / (sum(p) * dx)
    q   <- q / (sum(q) * dx)
    m   <- (p + q) / 2
    # Numerical KL
    kl <- function(a, b) sum(ifelse(a < 1e-20, 0, a * log2(a / (b + 1e-20)))) * dx
    0.5 * kl(p, m) + 0.5 * kl(q, m)
  })
)

ggplot(jsd_tbl, aes(x = separation, y = jsd)) +
  geom_line(colour = "purple", linewidth = 1.3) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "firebrick") +
  annotate("text", x = 7, y = 0.97, label = "JSD = 1 (disjoint)", hjust = 1, colour = "firebrick") +
  labs(
    title    = "JSD Between Two Gaussians vs. Their Separation",
    subtitle = glue("Both N(μ, σ²) with σ = {sigma_fixed}; means separated by d"),
    x        = "Mean Separation d",
    y        = "JSD [bits]"
  )
```

### Why This Matters in Practice

JSD is often preferred for distribution comparison dashboards because it is
bounded, symmetric, and numerically stable. It is especially useful for:

- Drift detection between training and production distributions.
- Segment-to-segment comparison where direction should not matter.
- Monitoring language model output shifts over time.


# Differential Entropy: Continuous Variables

So far we have worked exclusively with discrete distributions. For continuous
random variables, we define **differential entropy**:

$$
h(X) = -\int_{-\infty}^{\infty} f(x) \log_2 f(x) \, dx
$$

**Important caveats**: differential entropy can be negative and is not
invariant to scaling. It cannot be directly interpreted as "the number of bits
needed" without specifying quantisation resolution. However, *differences*
and *mutual information* involving differential entropy are meaningful.

It helps to connect this back to discrete coding by introducing a quantisation
bin width $\Delta$. If $X_\Delta$ is a discretised version of $X$, then
approximately

$$
H(X_\Delta) \approx h(X) + \log_2\!\left(\frac{1}{\Delta}\right)
$$

as $\Delta \to 0$. The additive resolution term explains why absolute
differential entropy values can be counterintuitive, while relative quantities
remain stable and interpretable.

Motivation-wise, differential entropy still matters because many real signals
are naturally continuous (volatility, latency, temperature, biometrics,
sensor amplitudes). Even though absolute values can be tricky, changes in
entropy under transformations or constraints are highly meaningful.

Two useful mental anchors:

- **Spread effect**: wider continuous distributions generally have higher
  differential entropy.
- **Unit effect**: rescaling changes entropy by an additive constant. If
  $Y = aX$, then $h(Y) = h(X) + \log_2|a|$.

## Toy Differential Entropy Calculations

```{r toy_differential_entropy_examples}
#| echo: true

toy_diff_tbl <- tibble(
  case = c("Uniform(0,1)", "Uniform(0,4)", "Normal(0,1)", "Normal(0,0.2^2)"),
  h_bits = c(
    diff_entropy_uniform(0, 1),
    diff_entropy_uniform(0, 4),
    diff_entropy_normal(1),
    diff_entropy_normal(0.2)
  )
)

toy_diff_tbl |> knitr::kable(digits = 4, caption = "Toy Differential Entropy Calculations")

h_u01 <- diff_entropy_uniform(0, 1)
h_u04 <- diff_entropy_uniform(0, 4)
h_n1  <- diff_entropy_normal(1)
h_n2  <- diff_entropy_normal(2)

cat(glue(
  "Uniform scaling: h(U(0,4)) - h(U(0,1)) = {round(h_u04 - h_u01, 4)} bits (expected: log2(4)=2)\n",
  "Normal scaling : h(N(0,2^2)) - h(N(0,1)) = {round(h_n2 - h_n1, 4)} bits (expected: log2(2)=1)\n"
))
```

This is the key practical takeaway: differential entropy is less about raw
"bits required" and more about how uncertainty changes as distributions become
more spread out, more concentrated, or reparameterized.

## Analytical Results for Common Distributions

| Distribution | Parameters | Differential Entropy |
|---|---|---|
| Uniform $U(a, b)$ | $a, b$ | $\log_2(b - a)$ |
| Normal $N(\mu, \sigma^2)$ | $\sigma$ | $\frac{1}{2}\log_2(2\pi e \sigma^2)$ |
| Exponential $\text{Exp}(\lambda)$ | $\lambda$ | $1 - \log_2 \lambda$ |
| Laplace $(0, b)$ | $b$ | $\log_2(2eb)$ |

The **Normal distribution maximises differential entropy** among all
distributions with a fixed variance — a beautiful maximum-entropy result.

This table is most useful as a quick analytic toolkit:

- If you can model a process as approximately Normal/Exponential/Uniform,
  entropy can be computed instantly without simulation.
- It provides sanity checks for numerical estimators.
- It helps separate scale effects from shape effects when comparing datasets.

```{r diff_entropy_demo}
#| echo: true
#| fig-cap: "Differential entropy of the Normal distribution grows as log₂(σ). Doubling σ adds exactly 1 bit of uncertainty."

sigma_seq <- seq(0.1, 10, by = 0.1)

diff_entropy_tbl <- tibble(
  sigma             = sigma_seq,
  normal_h          = map_dbl(sigma_seq, diff_entropy_normal),
  exponential_h     = map_dbl(sigma_seq, ~diff_entropy_exp(1 / .x)),
  uniform_h         = map_dbl(sigma_seq, ~diff_entropy_uniform(-sqrt(3) * .x, sqrt(3) * .x))
)

pivot_longer(diff_entropy_tbl, -sigma, names_to = "distribution", values_to = "entropy") |>
  mutate(distribution = str_remove(distribution, "_h") |> str_to_title()) |>
  ggplot(aes(x = sigma, y = entropy, colour = distribution)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dotted", colour = "grey50") +
  scale_colour_brewer(palette = "Dark2") +
  labs(
    title    = "Differential Entropy vs. Scale Parameter",
    subtitle = "All distributions have the same standard deviation σ",
    x        = "Scale (σ)",
    y        = "Differential Entropy h(X) [bits]",
    colour   = "Distribution"
  ) +
  theme(legend.position = "bottom")
```

## KDE-Based Entropy Estimation

When the true distribution is unknown, we can estimate differential entropy
from samples using a kernel density estimator.

```{r kde_entropy_estimation}
#| echo: true
#| fig-cap: "KDE-based entropy estimates vs. analytical truth across varying sample sizes. The bias decreases as n grows, but variance remains."

# Beta(2,5) distribution — non-Gaussian, analytical h not trivial
set.seed(99)
n_sizes_de  <- c(50, 100, 500, 1000, 5000)
n_reps_de   <- 100L

# True h(Beta(2,5)) via numerical integration
x_b <- seq(0.001, 0.999, by = 0.001)
f_b <- dbeta(x_b, 2, 5)
true_h_beta <- -sum(f_b * log2(f_b)) * diff(x_b)[1]

kde_conv_tbl <- future_map_dfr(n_sizes_de, function(n) {
  map_dfr(seq_len(n_reps_de), function(r) {
    s <- rbeta(n, 2, 5)
    tibble(n = n, rep = r, h_est = estimate_diff_entropy_kde(s))
  })
}, .options = furrr_options(seed = TRUE))

kde_summary <- kde_conv_tbl |>
  group_by(n) |>
  summarise(mean_h = mean(h_est), lo = quantile(h_est, 0.1), hi = quantile(h_est, 0.9))

ggplot(kde_summary, aes(x = n)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = mean_h), colour = "steelblue", linewidth = 1.2) +
  geom_hline(yintercept = true_h_beta, colour = "firebrick", linetype = "dashed") +
  annotate("text", x = 5000, y = true_h_beta + 0.03,
           label = glue("True h ≈ {round(true_h_beta, 3)} bits"), hjust = 1.1, colour = "firebrick") +
  scale_x_log10(labels = label_comma()) +
  labs(
    title    = "KDE Entropy Estimate for Beta(2,5) Distribution",
    subtitle = "Shaded: 10th–90th percentile over 100 replications",
    x        = "Sample Size n (log scale)",
    y        = "Estimated h(X) [bits]"
  )
```

### Why This Matters in Practice

For continuous data, entropy estimation is sensitive to smoothing and sample
size. In production settings:

- Track trends and relative changes rather than over-interpreting a single
  absolute estimate.
- Keep estimator choice fixed across time to avoid methodological drift.
- Pair entropy with distribution plots so variance/bias artifacts are visible.

## Entropy and Gaussianisation

An important practical fact: of all distributions with a given variance, the
Gaussian has the *highest* differential entropy. This means that transforming
a non-Gaussian distribution to be more Gaussian-shaped increases its entropy.
The difference $h(X_{\text{Gaussian}}) - h(X)$ is called the
**negentropy** — a measure of non-Gaussianity heavily used in Independent
Component Analysis (ICA).

```{r negentropy_demo}
#| echo: true
#| fig-cap: "Differential entropy of several distributions with identical variance (σ²=1). The Gaussian always has the highest entropy — it is the maximum-entropy distribution for a fixed variance."

sigma_sq <- 1.0

dist_entropy_tbl <- tibble(
  distribution = c("Gaussian", "Uniform (same var)", "Exponential (same var)",
                   "Laplace (same var)", "Bimodal (same var)"),
  h_bits = c(
    diff_entropy_normal(sqrt(sigma_sq)),
    diff_entropy_uniform(-sqrt(3 * sigma_sq), sqrt(3 * sigma_sq)),
    diff_entropy_exp(1 / sqrt(sigma_sq)),   # E[X]=sqrt(s), Var[X]=s
    {b_lap <- sqrt(sigma_sq / 2); log2(2 * exp(1) * b_lap)},  # Laplace(0,b)
    {# Bimodal: two Gaussians at ±1, σ=0.3
      x_bm <- seq(-4, 4, by = 0.01)
      f_bm <- 0.5 * dnorm(x_bm, -1, 0.3) + 0.5 * dnorm(x_bm, 1, 0.3)
      -sum(f_bm * log2(f_bm + 1e-20)) * 0.01}
  )
)

ggplot(dist_entropy_tbl, aes(x = reorder(distribution, h_bits), y = h_bits, fill = h_bits)) +
  geom_col(show.legend = FALSE) +
  geom_hline(
    yintercept = diff_entropy_normal(sqrt(sigma_sq)),
    linetype   = "dashed",
    colour     = "firebrick"
  ) +
  annotate("text", x = 0.5, y = diff_entropy_normal(sqrt(sigma_sq)) + 0.05,
           label = "Gaussian bound", hjust = 0, colour = "firebrick") +
  scale_fill_gradient(low = "steelblue", high = "firebrick") +
  coord_flip() +
  labs(
    title    = "Maximum-Entropy Principle: Gaussian Wins",
    subtitle = "All distributions have variance σ² = 1",
    x        = NULL,
    y        = "Differential Entropy h(X) [bits]"
  )
```

### Why This Matters in Practice

Gaussianisation and negentropy show up in many pipelines:

- In ICA and blind source separation, maximizing non-Gaussianity helps recover
  latent components.
- In anomaly detection, strong deviations from Gaussian entropy structure can
  reveal regime changes.
- In feature preprocessing, this principle explains why variance alone is not a
  complete uncertainty descriptor.


# Summary of Part 1

We have covered the first pillar of information theory:

| Concept | Formula | Interpretation |
|---|---|---|
| Self-information | $I(x) = -\log_2 P(x)$ | Surprise of a single event |
| Shannon Entropy | $H(X) = \mathbb{E}[I(X)]$ | Average uncertainty |
| Rényi Entropy | $H_\alpha(X) = \frac{1}{1-\alpha}\log_2 \sum p^\alpha$ | Family indexed by order |
| Joint Entropy | $H(X,Y)$ | Uncertainty about two variables together |
| Conditional Entropy | $H(Y \mid X) = H(X,Y) - H(X)$ | Remaining uncertainty after observing $X$ |
| Mutual Information | $I(X;Y) = H(X) + H(Y) - H(X,Y)$ | Shared information |
| KL Divergence | $D_{\text{KL}}(P \| Q)$ | Cost of approximating $P$ with $Q$ |
| Cross-Entropy | $H(P,Q) = H(P) + D_{\text{KL}}(P\|Q)$ | Total coding cost under wrong model |
| JSD | $\frac{1}{2}[D_{\text{KL}}(P\|M) + D_{\text{KL}}(Q\|M)]$ | Symmetric bounded divergence |
| Differential Entropy | $h(X) = -\int f \log_2 f$ | Continuous-variable entropy |

**Part 2** covers the operational applications: channel capacity, source
coding, Huffman codes, and a set of applied case studies in data compression,
finance, and machine learning.

If Part 1 gives you the language, Part 2 gives you the engineering limits:
how few bits you can use, how much noise you can tolerate, and what trade-offs
are unavoidable.
