---
title: "Worksheet: Information Theory Exercises"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---

```{r import_libraries}
#| echo: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(glue)
library(entropy)
library(infotheo)

source("lib_utils.R")
source("lib_information_theory.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "entropy", "infotheo")
)

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallelly::availableCores()
)

theme_set(theme_cowplot())
set.seed(42)
```

# Instructions

This worksheet is designed to reinforce the concepts in:

- Part 1: Foundations (`01_it_basics.qmd`)
- Part 2: Applications (`02_it_applications.qmd`)

Each section contains:

- **Task**: what to do
- **Hint**: optional guidance
- **Starter code**: a scaffold you can run and complete
- **Check yourself**: expected properties of the result


# Exercise 1: Entropy Intuition

## Task

Create three 5-outcome distributions and compare their entropy:

1. Uniform
2. Moderately skewed
3. Highly concentrated (one dominant outcome)

Plot them as bar charts and report $H(X)$ for each.

## Starter Code

```{r ex1_starter}
# TODO: define three distributions that sum to 1
p_uniform <- rep(1/5, 5)
p_mod     <- c(0.35, 0.25, 0.20, 0.12, 0.08)
p_skew    <- c(0.80, 0.10, 0.05, 0.03, 0.02)

ex1_tbl <- tibble(
  distribution = c("Uniform", "Moderate", "Skewed"),
  entropy_bits = c(
    shannon_entropy(p_uniform),
    shannon_entropy(p_mod),
    shannon_entropy(p_skew)
  )
)

ex1_tbl
```

## Check Yourself

- Entropy should be highest for uniform, lowest for skewed.
- All values should be between $0$ and $\log_2(5)$.


# Exercise 2: Joint, Conditional, Mutual Information

## Task

Construct a $2 \times 2$ joint distribution where variables are dependent.
Compute:

- $H(X)$
- $H(Y)$
- $H(X,Y)$
- $H(Y|X)$
- $I(X;Y)$

Verify the identities:

- $H(X,Y) = H(X) + H(Y|X)$
- $I(X;Y) = H(Y) - H(Y|X)$

## Starter Code

```{r ex2_starter}
# Example dependent joint distribution
joint <- matrix(c(0.45, 0.05,
                  0.10, 0.40),
                nrow = 2,
                byrow = TRUE)

h_x  <- shannon_entropy(rowSums(joint))
h_y  <- shannon_entropy(colSums(joint))
h_xy <- joint_entropy(joint)
h_yx <- conditional_entropy(joint)
mi   <- mutual_information(joint)

cat(glue(
  "H(X)      = {round(h_x,  4)}\n",
  "H(Y)      = {round(h_y,  4)}\n",
  "H(X,Y)    = {round(h_xy, 4)}\n",
  "H(Y|X)    = {round(h_yx, 4)}\n",
  "I(X;Y)    = {round(mi,   4)}\n",
  "\nCheck 1: H(X)+H(Y|X) = {round(h_x + h_yx, 4)}\n",
  "Check 2: H(Y)-H(Y|X) = {round(h_y - h_yx, 4)}\n"
))
```

## Check Yourself

- $I(X;Y)$ must be non-negative.
- The two identities should hold up to numerical precision.


# Exercise 3: KL Divergence Asymmetry

## Task

Define two discrete distributions $P$ and $Q$ over the same support and show
numerically that:

$$
D_{KL}(P\|Q) \ne D_{KL}(Q\|P)
$$

Also compute JSD and explain why it is symmetric.

## Starter Code

```{r ex3_starter}
p <- c(0.60, 0.25, 0.10, 0.05)
q <- c(0.25, 0.25, 0.25, 0.25)

kl_pq <- kl_divergence(p, q)
kl_qp <- kl_divergence(q, p)
jsd   <- js_divergence(p, q)

cat(glue(
  "KL(P||Q) = {round(kl_pq, 4)} bits\n",
  "KL(Q||P) = {round(kl_qp, 4)} bits\n",
  "JSD(P,Q) = {round(jsd,   4)} bits\n"
))
```

## Check Yourself

- `KL(P||Q)` and `KL(Q||P)` should differ.
- `JSD(P,Q)` should stay between 0 and 1 bit.


# Exercise 4: Huffman Coding Efficiency

## Task

1. Define an 8-symbol source.
2. Build a Huffman code.
3. Compute entropy $H(X)$ and expected code length $L$.
4. Verify $H(X) \le L < H(X) + 1$.

## Starter Code

```{r ex4_starter}
probs <- c(0.32, 0.20, 0.14, 0.10, 0.08, 0.07, 0.05, 0.04)
syms  <- LETTERS[1:8]

code_tbl <- build_huffman_code(syms, probs)
H <- shannon_entropy(probs)
L <- huffman_expected_length(code_tbl)

code_tbl |>
  arrange(desc(probability)) |>
  knitr::kable(digits = 4)

cat(glue(
  "H(X) = {round(H, 4)}\n",
  "L    = {round(L, 4)}\n",
  "H <= L ? {H <= L}\n",
  "L < H+1 ? {L < H + 1}\n"
))
```

## Check Yourself

- Most probable symbol should have one of the shortest codewords.
- The Huffman inequality should hold.


# Exercise 5: Channel Capacity and Code Rate

## Task

For a BSC with several noise levels $p$, compare capacity $C$ against two
candidate rates:

- $R_1 = 1.0$ (uncoded)
- $R_2 = 1/3$ (repetition-3 style)

Determine for which $p$ values each rate is theoretically feasible.

## Starter Code

```{r ex5_starter}
p_grid <- seq(0.01, 0.49, by = 0.01)

cap_tbl <- tibble(
  p_error = p_grid,
  capacity = bsc_capacity(p_grid),
  feasible_uncoded = 1.0 < capacity,
  feasible_rep3 = (1/3) < capacity
)

cap_tbl |>
  summarise(
    p_max_uncoded = max(p_error[feasible_uncoded]),
    p_max_rep3    = max(p_error[feasible_rep3])
  )
```

## Check Yourself

- `p_max_rep3` should be larger than `p_max_uncoded`.
- At $p=0.5$, capacity should be 0.


# Exercise 6: MI for Feature Ranking

## Task

Simulate a binary target and five candidate features:

- 2 informative
- 3 noise

Estimate MI(feature, target) and rank features.

## Starter Code

```{r ex6_starter}
set.seed(101)
n <- 3000

x1 <- rbinom(n, 1, 0.5)  # strong signal
x2 <- rbinom(n, 1, 0.5)  # weak signal
x3 <- rbinom(n, 1, 0.5)  # noise
x4 <- rbinom(n, 1, 0.5)  # noise
x5 <- rbinom(n, 1, 0.5)  # noise

y <- ifelse(runif(n) < ifelse(x1 == 1, 0.80, 0.20), 1L, 0L)
y <- ifelse(runif(n) < ifelse(x2 == 1, 0.06, -0.06) + 0.5, y, 1L - y)

mi_bin <- function(x, y) {
  tab <- table(x, y)
  mutual_information(tab / sum(tab))
}

rank_tbl <- tibble(
  feature = c("x1", "x2", "x3", "x4", "x5"),
  mi_bits = c(mi_bin(x1, y), mi_bin(x2, y), mi_bin(x3, y), mi_bin(x4, y), mi_bin(x5, y))
) |>
  arrange(desc(mi_bits))

rank_tbl
```

## Check Yourself

- `x1` should rank first most of the time.
- Noise variables should have MI close to zero.


# Exercise 7: Insurance Dependence Under Stress

## Task

Simulate frequency and severity with a latent catastrophe regime and compare:

- Pearson correlation
- Mutual information

Then recompute both metrics after removing catastrophe months and compare.

## Starter Code

```{r ex7_starter}
set.seed(2027)
n <- 8000

cat_regime <- rbinom(n, 1, 0.10)
freq <- rpois(n, lambda = ifelse(cat_regime == 1, 10, 4))
sev  <- rgamma(n,
               shape = ifelse(cat_regime == 1, 2.0, 3.8),
               rate  = ifelse(cat_regime == 1, 0.0010, 0.0038))

pearson_all <- cor(freq, sev)
mi_all      <- estimate_mi_histogram(freq, sev, n_bins_x = 12, n_bins_y = 12)

idx_normal <- which(cat_regime == 0)
pearson_normal <- cor(freq[idx_normal], sev[idx_normal])
mi_normal      <- estimate_mi_histogram(freq[idx_normal], sev[idx_normal], n_bins_x = 12, n_bins_y = 12)

cat(glue(
  "All months: Pearson={round(pearson_all,3)}, MI={round(mi_all,3)} bits\n",
  "Normal only: Pearson={round(pearson_normal,3)}, MI={round(mi_normal,3)} bits\n"
))
```

## Check Yourself

- Dependence should typically be stronger with catastrophe months included.
- MI often reveals stronger structure than Pearson in non-linear mixtures.


# Stretch Problems

1. Implement arithmetic coding and compare average length with Huffman for the
   same source.
2. Estimate differential entropy using multiple methods (KDE, kNN, spacing)
   and compare bias/variance.
3. Simulate finite-blocklength channel coding trade-offs ($n$, $R$, error
   probability) around capacity.
4. Build a small tree-based classifier and compare split rankings by Gini vs.
   information gain.
5. Use `infotheo::discretize()` with different methods (`equalwidth`,
   `equalfreq`, `globalequalwidth`) and inspect MI sensitivity.
