---
title: "Monitoring Process Change with Bayesian Methods"
subtitle: "Insurely You're Joking"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "2017-08-21"
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: pygments
    center: true
    reveal_options:
      slideNumber: true
---

```{r knit_opts, include = FALSE}
rm(list = ls())

knitr::opts_chunk$set(tidy  = FALSE
                     ,cache = FALSE
                     ,fig.height =  8
                     ,fig.width  = 11)

library(tidyverse)
library(purrr)
library(scales)
library(xts)
library(cowplot)
library(directlabels)


options(width = 80L
       ,warn  = 1)

set.seed(42)
```

# Introduction

## Structure of Talk

\


* Discussion of Problem
* Bayesian analysis and the Beta distribution
* Adding layers of noise
* Distribution distances and f-divergences


## Problem Discussion

\

Not Change-point analysis

\

Measure change effect

\

Signal vs noise

---

Want generic technique


## Sales-call Conversions

\

Binary outcome (0 or 1)

\

Monthly summaries

\

Sales due to faster turnaround


---

```{r generate_theta_data, echo=FALSE}
generate_process_rates <- function(mu0 = 0.10, sd0 = 0.03, mu1 = 0.15, sd1 = 0.03,
                                   start_date  = as.Date("2010-01-01"),
                                   end_date    = as.Date("2017-06-01"),
                                   change_date = as.Date("2016-01-01")) {

    month_vector <- as.yearmon(seq(start_date, end_date, by = "month"))
    switch_month <- as.yearmon(change_date)

    switch_idx <- match(switch_month, month_vector)

    pre_rate  <- rnorm(switch_idx - 1, mu0, sd0)
    post_rate <- rnorm(length(month_vector) - switch_idx + 1, mu1, sd1)

    rate_tbl <- data_frame(rate_date       = as.Date(month_vector)
                          ,underlying_rate = c(pre_rate, post_rate))

    return(rate_tbl)
}
```

```{r plot_process_rate, echo=FALSE}
plot_rate_tbl <- generate_process_rates(mu0 = 0.10, sd0 = 0.02, mu1 = 0.15, sd1 = 0.03)

ggplot(plot_rate_tbl) +
    geom_line(aes(x = rate_date, y = underlying_rate)) +
    expand_limits(y = 0) +
    xlab('Date') +
    ylab('Conversion Rate') +
    ggtitle("Plot of Time-Series of Conversion Rates")
```



# Bayesian Analysis

## Bayes Rule

\

$$
P(A | B) = \frac{P(B|A) P(A)}{P(B)}
$$

## Continuous Form

\

$$
p(\theta | D) \propto \int p(D | \theta) \, p(\theta) \ d\theta
$$

\

where

\

\begin{eqnarray*}
p(\theta)     &=& \text{Prior distribution for $\theta$} \\
p(D | \theta) &=& \text{Probability of seeing data $D$ given value $\theta$} \\
p(\theta | D) &=& \text{Posterior distribution for $\theta$}
\end{eqnarray*}


## Binomial Likelihood

\

Single trial:

$$
p(y|\theta) = \theta^y (1 - \theta)^{1-y}
$$

---

$n$ trials, $k$ successes:

$$
p(k | \theta) = \binom{n}{k} \, \theta^k (1 - \theta)^{n-k}
$$

---

### Beta Distribution

\

$$
p(\theta) = Beta(\alpha, \beta)
$$

\

$$
p(\theta | D) = Beta(\alpha + k, \beta + n - k)
$$

---

```{r plot_beta_distributions, echo=FALSE}
theta_seq <- seq(0, 1, by = 0.001)

p1 <- ggplot() +
    geom_line(aes(x = theta_seq, y = dbeta(theta_seq, 1, 1))) +
    xlab(expression(theta)) +
    ylab("Probability Density") +
    ggtitle("Beta(1, 1) Density")

p2 <- ggplot() +
    geom_line(aes(x = theta_seq, y = dbeta(theta_seq, 5, 5))) +
    xlab(expression(theta)) +
    ylab("Probability Density") +
    ggtitle("Beta(5, 5) Density")

p3 <- ggplot() +
    geom_line(aes(x = theta_seq, y = dbeta(theta_seq, 10, 10))) +
    xlab(expression(theta)) +
    ylab("Probability Density") +
    ggtitle("Beta(10, 10) Density")

p4 <- ggplot() +
    geom_line(aes(x = theta_seq, y = dbeta(theta_seq, 5, 15))) +
    xlab(expression(theta)) +
    ylab("Probability Density") +
    ggtitle("Beta(5, 15) Density")

plot_grid(p1, p2, p3, p4, ncol = 2)
```


---

```{r generate_beta_plots, echo=FALSE}
generate_beta_plot_data <- function(a, b, label = NULL) {
    theta     <- seq(0, 1, by = 0.0001)
    prob_dens <- dbeta(theta, a, b)

    beta_data_tbl <- data_frame(theta = theta
                               ,prob_dens = prob_dens)
    
    if(!is.null(label)) { beta_data_tbl <- beta_data_tbl %>% mutate(label = label) }
    
    return(beta_data_tbl)
}


data1_tbl <- generate_beta_plot_data(2, 2)
data2_tbl <- generate_beta_plot_data(10, 10)
data3_tbl <- generate_beta_plot_data(100, 100)

plotdata_tbl = list(`Beta(2, 2)`   = data1_tbl
                   ,`Beta(10, 10)` = data2_tbl
                   ,`Beta(100, 100)` = data3_tbl) %>%
    bind_rows(.id = 'label')

ggplot(plotdata_tbl) +
    geom_line(aes(x = theta, y = prob_dens, colour = label)) +
    xlab(expression(theta)) +
    ylab("Probability Density") +
    ggtitle("Comparison Density Plot for Beta Distribution")
```


# Attacking the Problem

## First Attempt

\

Generate data

\

Calculate yearly posterior distributions

\

Graph it

---

```{r generate_count_data, echo=FALSE}
generate_counts <- function(rate_tbl, call_count) {
    rate_tbl <- rate_tbl %>%
        mutate(call_count      = call_count
              ,conversion_count = map2_int(call_count, underlying_rate, rbinom, n = 1)
              ,conversion_rate  = conversion_count / call_count
        )

    return(rate_tbl)
}

fixed_rate_data_tbl  <- generate_process_rates(mu0 = 0.10, mu1 = 0.15, sd0 = 0, sd1 = 0)
fixed_count_data_tbl <- generate_counts(fixed_rate_data_tbl, call_count = 500)
```

```{r plot_conversion_rate_data, echo=FALSE}
plot_tbl <- fixed_count_data_tbl %>%
    select(-call_count, -conversion_count) %>%
    gather('rate_type','rate', -rate_date)

fixed_rate_lineplot <- ggplot(plot_tbl) +
    geom_line(aes(x = rate_date, y = rate, colour = rate_type)) +
    expand_limits(y = 0) +
    xlab("Date") +
    ylab("Rate") +
    ggtitle("Plot of Underlying and Observed Rates (Fixed Call Count)")

fixed_rate_lineplot %>% plot
```

---

```{r plot_yearly_distributions, echo=FALSE, warning=FALSE, message=FALSE}
generate_yearly_data <- function(rate_tbl) {
    year_tbl <- rate_tbl %>%
        group_by(data_year = format(rate_date, '%Y')) %>%
        summarise(a = sum(conversion_count)
                 ,b = sum(call_count - conversion_count)
                ) %>%
        mutate(cuml_a  = cumsum(a) + 1
              ,cuml_b  = cumsum(b) + 1
              ,distrib = pmap(list(cuml_a, cuml_b, data_year), generate_beta_plot_data)
                )
    
    distrib_tbl <- year_tbl$distrib %>% bind_rows

    return(distrib_tbl)
}

fixed_yearly_data_tbl <- generate_yearly_data(fixed_count_data_tbl) %>%
    rename(year = label)

fixed_yearly_distribplot <- ggplot(fixed_yearly_data_tbl) +
    geom_line(aes(x = theta, y = prob_dens, colour = year)) +
    scale_x_continuous(limits = c(0.075, 0.125)) +
    xlab(expression(theta)) +
    ylab("Prob Density") +
    ggtitle("Comparison Density Plot for Data Years (Fixed Call Count)")

fixed_yearly_distribplot %>% plot
```


## Randomise Monthly Calls

\

Had 500 calls per month

\

Treat as Poisson process

\

$$
C \sim Pois(500)
$$

---

```{r generate_call_count_noise, echo=FALSE}
noise_month_rate_tbl  <- generate_process_rates(mu0 = 0.10, mu1 = 0.15, sd0 = 0, sd1 = 0)

n_month <- noise_month_rate_tbl %>% nrow

noise_month_count_tbl  <- generate_counts(noise_month_rate_tbl, call_count = rpois(n_month, 500))
noise_month_yearly_tbl <- generate_yearly_data(noise_month_count_tbl) %>%
    rename(year = label)
```

```{r plot_noise_month_conversion_rates, echo=FALSE}
plot_tbl <- noise_month_count_tbl %>%
    select(-call_count, -conversion_count) %>%
    gather('rate_type','rate', -rate_date)

noise_month_lineplot <- ggplot(plot_tbl) +
    geom_line(aes(x = rate_date, y = rate, colour = rate_type)) +
    expand_limits(y = 0) +
    xlab("Date") +
    ylab("Rate") +
    ggtitle("Plot of Underlying and Observed Rates (Noisy Call Count)")

noise_month_lineplot %>% plot
```

---

```{r compare_fixed_noise_lineplots, echo=FALSE}
plot_grid(fixed_rate_lineplot, noise_month_lineplot, nrow = 2)
```

---

```{r plot_noise_yearly_distributions, echo=FALSE, warning=FALSE}
noise_month_yearly_distribplot <- ggplot(noise_month_yearly_tbl) +
    geom_line(aes(x = theta, y = prob_dens, colour = year)) +
    scale_x_continuous(limits = c(0.075, 0.125)) +
    xlab(expression(theta)) +
    ylab("Prob Density") +
    ggtitle("Comparison Density Plot for Data Years (Noisy Call Count)")

noise_month_yearly_distribplot %>% plot
```

---

```{r compare_fixed_noise_distribplots, echo=FALSE, warning=FALSE}
plot_grid(fixed_yearly_distribplot, noise_month_yearly_distribplot, nrow = 2)
```

---

```{r compare_fixed_noise_2x2, echo=FALSE, warning=FALSE}
plot_grid(fixed_rate_lineplot
         ,noise_month_lineplot
         ,fixed_yearly_distribplot
         ,noise_month_yearly_distribplot
         ,nrow = 2)
```


## Stochastic Conversion Rate

\

Add noise to the underlying rate?

\


---

```{r generate_stochastic_rate_data, echo=FALSE}
stoc_rate_tbl  <- generate_process_rates(mu0 = 0.10, mu1 = 0.15, sd0 = 0.02, sd1 = 0.02)

n_month <- stoc_rate_tbl %>% nrow

stoc_count_tbl <- generate_counts(stoc_rate_tbl, call_count = rpois(n_month, 500))
stoc_yearly_tbl <- generate_yearly_data(stoc_count_tbl) %>%
    rename(year = label)
```

```{r plot_stoc_conversion_rates, echo=FALSE}
plot_tbl <- stoc_count_tbl %>%
    select(-call_count, -conversion_count) %>%
    gather('rate_type','rate', -rate_date)

stoc_lineplot <- ggplot(plot_tbl) +
    geom_line(aes(x = rate_date, y = rate, colour = rate_type)) +
    expand_limits(y = 0) +
    xlab("Date") +
    ylab("Rate") +
    ggtitle("Plot of Underlying and Observed Rates (Stochastic Rate)")

stoc_lineplot %>% plot
```

---

```{r plot_stoc_yearly_distributions, echo=FALSE, warning=FALSE}
stoc_yearly_distribplot <- ggplot(stoc_yearly_tbl) +
    geom_line(aes(x = theta, y = prob_dens, colour = year)) +
    scale_x_continuous(limits = c(0.075, 0.125)) +
    xlab(expression(theta)) +
    ylab("Prob Density") +
    ggtitle("Comparison Density Plot for Data Years (Stochastic Rate)")

stoc_yearly_distribplot %>% plot
```


## More Simple Approach

\

Aggregate yearly

\

Look at yearly conversions


---

```{r plot_yearly_conversions, echo=FALSE}
plot_tbl <- stoc_count_tbl %>%
    group_by(year = format(rate_date, '%Y')) %>%
    summarise(conv = sum(conversion_count)
             ,call = sum(call_count)
             ,rate = conv / call) %>%
    mutate(year = year %>% as.numeric)

ggplot(plot_tbl) +
    geom_line(aes(x = year, y = rate)) +
    expand_limits(y = 0) +
    xlab("Year") +
    ylab("Conversion Rate") +
    ggtitle("Plot of Yearly Conversion Rates")
```

---

Why the discrepancy in outputs?


# Prior Data

## Strength of Prior

\

Quantity of data accumulates

\

Prior very strong

---

Need to rethink priors

---

Prior represents knowledge

\

How confident are we?


## Constructing Priors

\

Balancing act

\

Estimate $\theta$, assign a strength

---

Reparameterise $Beta(\alpha, \beta)$

\

$$
Beta(\alpha, \beta) \rightarrow Beta(\mu K, (1 - \mu) K)
$$

\

\begin{eqnarray*}
\mu &=& \text{probability expectation} \\
K   &=& \text{strength of belief}
\end{eqnarray*}

---

```{r estimate_prior_mu, echo=TRUE}
stoc_count_tbl %>%
    filter(rate_date < as.Date('2016-01-01')) %>%
    summarise(conv_count = sum(conversion_count)
             ,call_count = sum(call_count)
             ,rate       = conv_count / call_count
              )

```

---

Assume 1 year of 'strength'

\

$K = 12 \times 500 = 6,000$

\

\begin{eqnarray*}
\mu &=& 0.0967607 \\
K   &=& 6,000
\end{eqnarray*}

---

```{r plot_data_new_prior, echo=FALSE, warning=FALSE, message=FALSE}
mu <- 0.0967607
K  <- 6000

sixmonths_tbl <- stoc_count_tbl %>%
    filter(rate_date >= as.Date('2016-01-01')
          ,rate_date <= as.Date('2016-06-01')
           ) %>%
    mutate(cuml_a = cumsum(conversion_count)
          ,cuml_b = cumsum(call_count) - cuml_a
          ,new_a  = mu * K       + cuml_a
          ,new_b  = (1 - mu) * K + cuml_b
          ,distrib = pmap(list(a = new_a
                              ,b = new_b
                              ,label = format(rate_date, '%Y%m')
                            )
                         ,generate_beta_plot_data
                        )
           )

prior_tbl <- generate_beta_plot_data(mu * K, (1 - mu) * K, label = 'prior')

sixmonths_distrib_tbl <- c(sixmonths_tbl$distrib, list(prior_tbl)) %>% bind_rows

ggplot(sixmonths_distrib_tbl %>% filter(theta >= 0.075, theta <= 0.125)) +
    geom_line(aes(x = theta, y = prob_dens, colour = label)) +
    geom_dl(aes(x = theta, y = prob_dens, colour = label, label = label), method = 'top.bumpup') +
    xlab(expression(theta)) +
    ylab("Prob Density") +
    ggtitle("Plot of Posterior Densities")
```


## Moving Signal/Noise Ratio

\

$$
\mathcal{N}(0.10, 0.02) \rightarrow \mathcal{N}(0.15, 0.02)
$$

\

Mean-shift well outside variance


---

### Increase Noise

\

$$
\mathcal{N}(0.40, 0.08) \rightarrow \mathcal{N}(0.45, 0.08)
$$

\

Can we see difference?

---

```{r highbase_rate_data, echo=FALSE}
highbase_rate_tbl  <- generate_process_rates(mu0 = 0.40, mu1 = 0.45, sd0 = 0.08, sd1 = 0.08)

n_data <- highbase_rate_tbl %>% nrow

highbase_count_tbl <- generate_counts(highbase_rate_tbl, call_count = rpois(n_data, 500))
```

```{r plot_highbase_conversion_rates, echo=FALSE}
plot_tbl <- highbase_count_tbl %>%
    select(-call_count, -conversion_count) %>%
    gather('rate_type','rate', -rate_date)

highbase_lineplot <- ggplot(plot_tbl) +
    geom_line(aes(x = rate_date, y = rate, colour = rate_type)) +
    expand_limits(y = 0) +
    xlab("Date") +
    ylab("Rate") +
    ggtitle("Plot of Underlying and Observed Rates (High Base Rate)")

highbase_lineplot %>% plot
```

---

Very hard to spot a change!

---

### Analysis for $\mu = 0.40$

```{r new_approach_highbase_data, echo=FALSE, fig.height=7.5}
mu <- highbase_count_tbl %>%
    filter(rate_date < as.Date('2016-01-01')) %>%
    summarise(rate = sum(conversion_count) / sum(call_count)) %>%
    pull(rate)

K  <- 6000

highbase_sixmn_tbl <- highbase_count_tbl %>%
    filter(rate_date >= as.Date('2016-01-01')
          ,rate_date <= as.Date('2016-06-01')
           ) %>%
    mutate(cuml_a = cumsum(conversion_count)
          ,cuml_b = cumsum(call_count) - cuml_a
          ,new_a  = floor(mu * K       + cuml_a)
          ,new_b  = floor((1 - mu) * K + cuml_b)
          ,distrib = pmap(list(a = new_a
                              ,b = new_b
                              ,label = format(rate_date, '%Y%m')
                            )
                         ,generate_beta_plot_data
                        )
           )

prior_tbl            <- generate_beta_plot_data(mu * K, (1 - mu) * K, label = 'prior')
highbase_distrib_tbl <- c(highbase_sixmn_tbl$distrib, list(prior_tbl)) %>% bind_rows

ggplot(highbase_distrib_tbl %>% filter(theta >= 0.375, theta <= 0.50)) +
    geom_line(aes(x = theta, y = prob_dens, colour = label)) +
    geom_dl(aes(x = theta, y = prob_dens, colour = label, label = label), method = 'top.bumpup') +
    xlab(expression(theta)) +
    ylab("Prob Density") +
    ggtitle("Plot of Posterior Densities for Higher Base Rate")
```

---

```{r plot_highbase_yearly_conversions, echo=FALSE}
plot_tbl <- highbase_count_tbl %>%
    group_by(year = format(rate_date, '%Y')) %>%
    summarise(conv = sum(conversion_count)
             ,call = sum(call_count)
             ,rate = conv / call) %>%
    mutate(year = year %>% as.numeric)

ggplot(plot_tbl) +
    geom_line(aes(x = year, y = rate)) +
    expand_limits(y = 0) +
    xlab("Year") +
    ylab("Conversion Rate") +
    ggtitle("Plot of Yearly Conversion Rates for High-Base Level")
```

---

How can we quantify differences?


# f-divergences

## Distributional Differences

\

A *metric* or *distance*:

$$
d : X \times X \rightarrow \mathbb{R}^{+}
$$
\

\begin{align*}
d(x, y) &\geq 0 \; \forall x, y \in X,                    && \text{ non-negativity}             \\
d(x, y) &=    0 \; \iff \; x = y \; \forall x, y \in X,   && \text{ identity of indiscernables} \\
d(x, y) &=    d(y, x) \; \forall x, y \in X,              && \text{ symmetry}                   \\
d(x, z) &\leq d(x, y) + d(y, z) \; \forall x, y, z \in X, && \text{ triangle inequality}        \\ 
\end{align*}


---

### Common-Area Metric

\

$$
D(P, Q) = \int^1_0 \text{min}(P(x), Q(x)) \, dx 
$$

---

```{r common_area_plot, echo=FALSE}
theta_seq <- seq(0, 1, by = 0.0001);

P <- dbeta(theta_seq, 50, 50);
Q <- dbeta(theta_seq, 40, 60);

ggplot() +
    geom_line(aes(x = theta_seq, y = P)) +
    geom_line(aes(x = theta_seq, y = Q), colour = 'red') +
    geom_area(aes(x = theta_seq, y = pmin(P, Q)), fill = 'grey', alpha = 0.5) +
    xlab(expression(theta)) +
    ylab("Prob Density") +
    ggtitle("The Common-Area Metric")
```    

---

### Kullback-Leibler Divergence

\

$$
D_{KL}(P||Q) = \int^1_0 p(x) \ln \frac{p(x)}{q(x)} \, dx
$$

\

Not symmetric

\

No triangle inequality

\

Intuitive information theory interpretation

---

### Hellinger Distance

\

$$
H^2(P, Q) = 1 - \int \sqrt{p(x) q(x)} \, dx
$$

$$
0 \leq H(P, Q) \leq 1
$$

$$
H^2(P, Q) \leq \delta(P, Q) \leq \sqrt{2} H(P, Q)
$$

---

### f-div Values for Beta Distribution

\

$$
\mu = 0.10 \;\; K_1 = 6,000 \;\; K_2 = 7,000 \;\; K_3 = 12,000
$$

```{r distance_beta_distribution, echo=FALSE, fig.height=6}
mu <- 0.1
K1 <- 6000
K2 <- 7000
K3 <- 12000

x_seq <- seq(0.05, 0.15, by = 0.0001)

Beta1 <- dbeta(x_seq, mu * K1, (1 - mu) * K1)
Beta2 <- dbeta(x_seq, mu * K2, (1 - mu) * K2)
Beta3 <- dbeta(x_seq, mu * K3, (1 - mu) * K3)

staticmu_1_plot <- ggplot() +
    geom_line(aes(x = x_seq, y = Beta2), colour = 'red') +
    geom_area(aes(x = x_seq, y = pmin(Beta1, Beta2)), fill = 'grey', alpha = 0.5) +
    xlab(expression(theta)) +
    ylab('Prob Density') +
    ggtitle("K = 6000 vs K = 7000")

staticmu_2_plot <- ggplot() +
    geom_line(aes(x = x_seq, y = Beta1)) +
    geom_line(aes(x = x_seq, y = Beta3), colour = 'red') +
    geom_area(aes(x = x_seq, y = pmin(Beta1, Beta3)), fill = 'grey', alpha = 0.5) +
    xlab(expression(theta)) +
    ylab('Prob Density') +
    ggtitle("K = 6000 vs K = 12000")

plot_grid(staticmu_1_plot, staticmu_2_plot, ncol = 2)
```

---

```{r calculate_fdiv_metrics_functions, echo=FALSE}
common_area_metric <- function(x, P, Q) {
    stopifnot(length(P) == length(Q))
    stopifnot(length(x) == length(P))

    N  <- length(x)
    dx <- diff(x)

    return(1 - sum(pmin(P, Q)[2:N] * dx))
}


hellinger_metric <- function(x, P, Q) {
    stopifnot(length(P) == length(Q))
    stopifnot(length(x) == length(P))

    N  <- length(x)
    dx <- diff(x)

    return(1 - sum(sqrt(P * Q)[2:N] * dx))
}


kl_metric <- function(x, P, Q) {
    stopifnot(length(P) == length(Q))
    stopifnot(length(x) == length(P))

    N  <- length(x)
    dx <- diff(x)
    dx <- c(dx[1], dx)

    ### Need to select out the non-zero indices for both P and Q
    idx <- intersect(which(P > 0),  which(Q > 0))

    return(sum((P * log(P/Q) * dx)[idx]))
}

beta_hellinger_metric <- function(alpha1, beta1, alpha2, beta2) {
    B1 <- beta(0.5 * (alpha1 + alpha2), 0.5 * (beta1 + beta2))
    B2 <- beta(alpha1, beta1)
    B3 <- beta(alpha2, beta2)

    return(sqrt(1 - (B1 / sqrt(B2 * B3))))
}


calculate_metrics <- function(x, P, Q) {
    stopifnot(length(P) == length(Q))
    stopifnot(length(x) == length(P))

    commonarea <- common_area_metric(x, P, Q)
    hellinger  <- hellinger_metric(x, P, Q)
    kl         <- kl_metric(x, P, Q)

    return(c(commonarea = commonarea, hellinger = hellinger, kl = kl))
}
```

```{r calculate_metrics, echo=TRUE}
calculate_metrics(x_seq, Beta1, Beta1) %>% print(digits = 2)
calculate_metrics(x_seq, Beta1, Beta2) %>% print(digits = 2)
calculate_metrics(x_seq, Beta1, Beta3) %>% print(digits = 2)
```

## Construct Ideal Data

\

Fix $\mu_1$, prior of 1 year data, $K_1 = 6,000$

\

Set new $\mu_2$: Check distribution:

\

Two months, $K_2 = 7,000$; one year, $K_3 = 12,000$

---

### Small Move: $\mu_1 = 0.10 \;\; \mu_2 = 0.11$

```{r create_toy_comparison_charts_smallmove, echo=FALSE, fig.height=7}
x_seq <- seq(0, 1, by = 0.0001)

mu1 <- 0.10
mu2 <- 0.11

K1 <- 6000
K2 <- 7000
K3 <- 12000

Beta1 <- dbeta(x_seq, (mu1 * K1),                     ((1 - mu1) * K1))
Beta2 <- dbeta(x_seq, (mu1 * K1) + (mu2 * (K2 - K1)), ((1 - mu1) * K1 + ((1 - mu2) * (K2 - K1))))
Beta3 <- dbeta(x_seq, (mu1 * K1) + (mu2 * (K3 - K1)), ((1 - mu1) * K1 + ((1 - mu2) * (K3 - K1))))

plot_tbl <- data_frame(
    x_seq = x_seq
   ,Beta1 = Beta1
   ,Beta2 = Beta2
   ,Beta3 = Beta3
)

ggplot(plot_tbl %>% filter(x_seq >= 0.075, x_seq <= 0.15)) +
    geom_line(aes(x = x_seq, y = Beta1)) +
    geom_line(aes(x = x_seq, y = Beta2), colour = 'red') +
    geom_line(aes(x = x_seq, y = Beta3), colour = 'blue') +
    xlab(expression(theta)) +
    ylab("Prob Density") +
    ggtitle("Comparison Plot of mu = 0.10 -> 0.11")
```

---

```{r calculate_metrics_smallmove, echo=TRUE}
calculate_metrics(x_seq, Beta1, Beta1) %>% print(digits = 4)
calculate_metrics(x_seq, Beta1, Beta2) %>% print(digits = 4)
calculate_metrics(x_seq, Beta1, Beta3) %>% print(digits = 4)
```

---

### Larger Move: $\mu_1 = 0.10 \;\; \mu_2 = 0.15$

```{r create_toy_comparison_charts_bigmove, echo=FALSE, fig.height=7}
x_seq <- seq(0, 1, by = 0.0001)

mu1 <- 0.10
mu2 <- 0.15

K1 <- 6000
K2 <- 7000
K3 <- 12000

Beta1 <- dbeta(x_seq, (mu1 * K1),                     ((1 - mu1) * K1))
Beta2 <- dbeta(x_seq, (mu1 * K1) + (mu2 * (K2 - K1)), ((1 - mu1) * K1 + ((1 - mu2) * (K2 - K1))))
Beta3 <- dbeta(x_seq, (mu1 * K1) + (mu2 * (K3 - K1)), ((1 - mu1) * K1 + ((1 - mu2) * (K3 - K1))))

plot_tbl <- data_frame(
    x_seq = x_seq
   ,Beta1 = Beta1
   ,Beta2 = Beta2
   ,Beta3 = Beta3
)

ggplot(plot_tbl %>% filter(x_seq >= 0.075, x_seq <= 0.15)) +
    geom_line(aes(x = x_seq, y = Beta1)) +
    geom_line(aes(x = x_seq, y = Beta2), colour = 'red') +
    geom_line(aes(x = x_seq, y = Beta3), colour = 'blue') +
    xlab(expression(theta)) +
    ylab("Prob Density") +
    ggtitle("Comparison Plot of mu = 0.10 -> 0.15")
```

---

```{r calculate_metrics_bigmove, echo=TRUE}
calculate_metrics(x_seq, Beta1, Beta1) %>% print(digits = 4)
calculate_metrics(x_seq, Beta1, Beta2) %>% print(digits = 4)
calculate_metrics(x_seq, Beta1, Beta3) %>% print(digits = 4)
```

---

What about our data?


## Analyse Conversion Data

\

We have monthly call data

\

Have posterior distributions

\

Calculate metrics as data updates

---

### Low-Noise Move: $\mu = 0.10 \rightarrow 0.15$

```{r plot_fdiv_metrics_lownoise_data, echo=FALSE, fig.height=7.5}
mu <- 0.0967607
K  <- 6000

prior_tbl <- generate_beta_plot_data(mu * K, (1 - mu) * K, label = 'prior') %>%
    select(theta, prior = prob_dens)


stoc_calc_tbl <- stoc_count_tbl %>%
    filter(rate_date >= as.Date('2016-01-01')) %>%
    mutate(cuml_a = cumsum(conversion_count)
          ,cuml_b = cumsum(call_count) - cuml_a
          ,new_a  = mu * K       + cuml_a
          ,new_b  = (1 - mu) * K + cuml_b
          ,distrib = pmap(list(a = new_a
                              ,b = new_b
                              ,label = format(rate_date, '%Y%m')
                            )
                         ,generate_beta_plot_data
                        )
           )

stoc_distrib_tbl <- stoc_calc_tbl$distrib %>%
    bind_rows %>%
    inner_join(prior_tbl, by = 'theta')

stoc_fdiv_tbl <- stoc_distrib_tbl %>%
    group_by(label) %>%
    summarise(fdiv = list(calculate_metrics(theta, prior, prob_dens))) %>%
    mutate(common_area = map_dbl(fdiv, 'commonarea')
          ,hellinger   = map_dbl(fdiv, 'hellinger')
          ,kl          = map_dbl(fdiv, 'kl')
    ) %>%
    select(-fdiv) %>%
    gather('fdiv','value', -label) %>%
    mutate(month_date = as.Date(paste0(label, '01'), format = '%Y%m%d'))

ggplot(stoc_fdiv_tbl) +
    geom_line(aes(x = month_date, y = value)) +
    expand_limits(y = 0) +
    facet_wrap(~ fdiv, nrow = 2, scales = 'free_y') +
    xlab("Month") +
    ylab("f-divergence Metric") +
    ggtitle("Plot of f-divergence Metrics by Incoming Month")
```

---

### High-Noise Move: $\mu = 0.40 \rightarrow 0.45$

```{r plot_fdiv_metrics_highnoise_data, echo=FALSE, fig.height=7.5}
mu <- highbase_count_tbl %>%
    filter(rate_date < as.Date('2016-01-01')) %>%
    summarise(rate = sum(conversion_count) / sum(call_count)) %>%
    pull(rate)

K  <- 6000

prior_tbl <- generate_beta_plot_data(mu * K, (1 - mu) * K, label = 'prior') %>%
    select(theta, prior = prob_dens)


highbase_calc_tbl <- highbase_count_tbl %>%
    filter(rate_date >= as.Date('2016-01-01')) %>%
    mutate(cuml_a = cumsum(conversion_count)
          ,cuml_b = cumsum(call_count) - cuml_a
          ,new_a  = mu * K       + cuml_a
          ,new_b  = (1 - mu) * K + cuml_b
          ,distrib = pmap(list(a = new_a
                              ,b = new_b
                              ,label = format(rate_date, '%Y%m')
                            )
                         ,generate_beta_plot_data
                        )
           )

highbase_distrib_tbl <- highbase_calc_tbl$distrib %>%
    bind_rows %>%
    inner_join(prior_tbl, by = 'theta')

highbase_fdiv_tbl <- highbase_distrib_tbl %>%
    group_by(label) %>%
    summarise(fdiv = list(calculate_metrics(theta, prior, prob_dens))) %>%
    mutate(common_area = map_dbl(fdiv, 'commonarea')
          ,hellinger   = map_dbl(fdiv, 'hellinger')
          ,kl          = map_dbl(fdiv, 'kl')
    ) %>%
    select(-fdiv) %>%
    gather('fdiv','value', -label) %>%
    mutate(month_date = as.Date(paste0(label, '01'), format = '%Y%m%d'))

ggplot(highbase_fdiv_tbl) +
    geom_line(aes(x = month_date, y = value)) +
    expand_limits(y = 0) +
    facet_wrap(~ fdiv, nrow = 2, scales = 'free_y') +
    xlab("Month") +
    ylab("f-divergence Metric") +
    ggtitle("Plot of f-divergence Metrics by Incoming Month")
```

# Conclusion

## Summary

\

Binomial process with known change point

\

Model with Beta distribution

\

Aggregate data appropriately

\

Distribution plots and f-divergence metrics

\

Decide on thresholds


## Future Extensions

\

Try with other processes / distributions

\

More comprehensive behaviour investigation

\

Look at statistical distance

\

Time-series methods


## Questions?

\

mickcooney@gmail.com

\

https://github.com/kaybenleroll/dublin_r_workshops

\

Blog post:

\

http://blog.applied.ai/a-bayesian-approach-to-monitoring-process-change/
