---
title: "Generate Sample Bond Portfolio"
author: "Mick Cooney"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
output:
  rmdformats::readthedown:
    toc_depth: 3
    use_bookdown: TRUE
    code_folding: hide
    fig_caption: TRUE

  html_document:
    fig_caption: yes
    theme: spacelab #sandstone #spacelab #flatly
    highlight: pygments
    number_sections: TRUE
    toc: TRUE
    toc_depth: 3
    toc_float:
      smooth_scroll: FALSE

  pdf_document: default
---


```{r import_libraries, echo=FALSE, message=FALSE}
library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(stringr)
library(glue)
library(purrr)
library(furrr)
library(DT)
library(lubridate)


source("lib_utils.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "DT")
  )


knitr::opts_chunk$set(
  tidy       = FALSE,
  cache      = FALSE,
  warning    = FALSE,
  message    = FALSE,
  fig.height =     8,
  fig.width  =    11
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallel::detectCores()
  )

theme_set(theme_cowplot())

set.seed(42)

plan(multisession)
```

The code and materials this workshop is available on GitHub at this URL:

https://github.com/kaybenleroll/data_workshops



# Load Data

We first want to load our datasets and prepare them for some simple association
rules mining.

## Load Company Data

We want to load up the company data that will represent the various bond
issues in our portfolio.


```{r load_company_data, echo=TRUE}
company_cols <- cols(
  gics_code = col_character()
)

company_information_tbl <- read_csv(
  file      = "data/company_information.csv",
  col_types = company_cols
)

company_information_tbl %>% glimpse()
```



# Construct Bond Portfolio

We now want to construct the first bond portfolio for use in the workshop.

```{r construct_bond_portfolio, echo=TRUE}
current_date  <- as.Date("2019-01-01")
coupon_values <- seq(5, 10, by = 0.125)

generate_maturity_date <- function(n_dates, min_maturity = 1, max_maturity = 5,
                                   base_date = current_date) {
  min_mat <- min_maturity * 365
  max_mat <- max_maturity * 365
  
  maturity_date <- sample(min_mat:max_mat, size = n_dates, replace = TRUE) %>%
    days() %>%
    add(base_date)
  
  return(maturity_date)
}

generate_issued_date <- function(maturity_date, base_date, max_years = 5) {
  expired_days <- sample(
    x       = 0:(max_years * 365),
    size    = length(maturity_date),
    replace = TRUE
    )
  
  term_years <- difftime(maturity_date, base_date) %>%
    add(expired_days) %>%
    divide_by(365) %>%
    ceiling()

  issued_date <- maturity_date - years(term_years) - 1
  
  return(issued_date)
}


generate_bond_portfolio <- function(n_size, company_data_tbl, current_date,
                                    position_rng,
                                    coupon_values = seq(5, 10, by = 0.125)
                                    ) {

  sample_portfolio_tbl <- company_data_tbl %>%
    slice_sample(n = nrow(.), replace = FALSE) %>%
    head(n_size) %>%
    mutate(
      position      = position_rng(n()),
      coupon        = sample(coupon_values, size = n(), replace = TRUE),
      price         = rlnorm(
        n(),
        meanlog = log(100) - (0.5 * 0.05 * 0.05),
        sdlog   = 0.05) %>% round(2),
      maturity_date = generate_maturity_date(
        n(),
        min_maturity = 1,
        max_maturity = 5,
        base_date    = current_date),
      issued_date   = generate_issued_date(
        maturity_date,
        base_date = current_date,
        max_years = 5),
      issuer = glue(
        "{ticker_symbol} {coupon_label} {mat_date}",
        coupon_label = round(coupon, 3) %>% str_replace("\\.0+$", ""),
        mat_date     = format(maturity_date, "%m/%d/%Y")
        ) %>% as.character(),
      currency = str_c(country_code, "D", sep = "")
      ) %>%
    select(
      issuer, currency, position, price, coupon, issued_date, maturity_date
      )

  return(sample_portfolio_tbl)    
}

bond_portfolio_tbl <- generate_bond_portfolio(
    n_size           = 250,
    company_data_tbl = company_information_tbl,
    current_date     = current_date,
    position_rng     = function(n_pos) rnbinom(n_pos, mu = 200, size = 25),
    coupon_values    = seq(5, 10, by = 0.125)
    )

bond_portfolio_tbl %>% write_csv("data/sample_bond_portfolio.csv")
```



# Construct Alternative Bond Portfolio

We now want to construct a completely different portfolio to illustrate the
ability of the code to work on multiple inputs.

```{r generate_alternative_portfolio, echo=TRUE}
alt_portfolio_tbl <- generate_bond_portfolio(
    n_size           = 500,
    company_data_tbl = company_information_tbl,
    current_date     = current_date,
    coupon_values    = seq(5, 10, by = 0.125),
    position_rng     = function(n) rnbinom(n, mu = 300, size = 25)
    )

alt_portfolio_tbl %>% write_csv("data/alt1_bond_portfolio.csv")
```



# R Environment
 
```{r show_session_info, echo=TRUE, message=TRUE}
sessioninfo::session_info()
```
