---
title: "Loss Curve Modelling"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "2018-07-17"
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: pygments
    center: true
    reveal_options:
      slideNumber: true
---

```{r knit_opts, include = FALSE}
rm(list = ls())

knitr::opts_chunk$set(tidy  = FALSE
                     ,cache = FALSE
                     ,message = FALSE
                     ,warning = FALSE
                     ,fig.height =  8
                     ,fig.width  = 11)

library(tidyverse)
library(scales)
library(rstan)
library(bayesplot)
library(cowplot)


options(width = 80L
       ,warn  = 1
        )


rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())


set.seed(42)

stan_seed <- 42

source("custom_functions.R")
```

# Introduction

## NAIC Schedule P Dataset

```{r load_data, echo=FALSE}
### File was downloaded from http://www.casact.org/research/reserve_data/ppauto_pos.csv
data_files <- dir("data/", pattern = "\\.csv", full.names = TRUE)

data_cols <- cols(GRCODE = col_character())

rawdata_tbl <- data_files %>%
    map(read_claim_datafile, col_type = data_cols) %>%
    bind_rows

glimpse(rawdata_tbl)
```

---

```{r reorganise_data, echo=FALSE}
claimdata_tbl <- rawdata_tbl %>%
    mutate(acc_year   = as.character(accidentyear)
          ,dev_year   = developmentyear
          ,dev_lag    = developmentlag
          ,premium    = earnedpremdir
          ,cum_loss   = cumpaidloss
          ,loss_ratio = cum_loss / premium) %>%
    select(grcode, grname, lob, acc_year, dev_year, dev_lag, premium, cum_loss, loss_ratio) %>%
    group_by(grcode, lob, acc_year) %>%
    mutate(dev_factor = cum_loss / cum_loss[10]) %>%
    ungroup()

claimdata_tbl %>% glimpse()
```

---

```{r plot_total_losses, echo=FALSE}
use_grcode <- '43'

carrier_full_tbl <- claimdata_tbl %>%
    filter(lob == 'ppauto')

snapshot_tbl <- claimdata_tbl %>%
    filter(lob      == 'ppauto'
          ,grcode   == use_grcode
          ,dev_year <  1998
           )

losscurves_plot <- ggplot(snapshot_tbl) +
    geom_line (aes(x = dev_lag, y = cum_loss, colour = acc_year)) +
    geom_point(aes(x = dev_lag, y = cum_loss, colour = acc_year)) +
    expand_limits(y = 0) +
    scale_y_continuous(labels = comma) +
    xlab("Development Year") +
    ylab("Cumulative Loss") +
    ggtitle("Plot of Cumulative Losses by Cohort Year")

losscurves_plot %>% plot
```

---

```{r plot_loss_ratio, echo=FALSE}
devfactors_plot <- ggplot(snapshot_tbl) +
    geom_line (aes(x = dev_lag, y = loss_ratio, colour = acc_year)) +
    geom_point(aes(x = dev_lag, y = loss_ratio, colour = acc_year)) +
    expand_limits(y = 0) +
    xlab("Development Year") +
    ylab("Development Factor") +
    ggtitle("Plot of Development Factors by Cohort Year")

devfactors_plot %>% plot
```


## Grid Approximation

```{r show_fully_developed_curve, echo=FALSE}
single_curve_tbl <- snapshot_tbl %>%
    filter(acc_year == 1988)

fully_developed_likelihood <- create_curve_likelihood_function(single_curve_tbl)


mle_est <- optim(c(1,1)
                ,function(x) fully_developed_likelihood(x[1],x[2])$loglik
                ,control = list(fnscale = -1)
                 )
```

```{r single_parameter_approximation, echo=FALSE, fig.height=7.5}
ulr_vals <- seq(0.65, 0.75, by = 0.001)

ulr_likelihood_func <-function(ulr)
    map_dbl(ulr, function(x) fully_developed_likelihood(mle_est$par[1], x)$loglik)

likelihood_vals <- ulr_likelihood_func(ulr_vals) %>% exp %>% (function (x) x * 0.001)

ulr_probs <- likelihood_vals / sum(likelihood_vals)

ulr_likelihood_plot <- ggplot() +
    geom_line(aes(x = ulr_vals, y = ulr_probs)) +
    xlab("Ultimate Loss Ratio") +
    ylab("Likelihood") +
    ggtitle("Marginal Probability of Observed Curve as Function of ULR")

ulr_likelihood_plot %>% plot
```

---

### 2D Grid


```{r calculate_grid_approximation, echo=FALSE, cache=FALSE}
# dlambda <- 0.01
# dulr    <- 0.01
dlambda <- 0.025
dulr    <- 0.025

l_seq <- seq(0.10, 1.50, by = dlambda)
u_seq <- seq(0.10, 1.50, by = dulr)

curve_sd <- 0.05

grid_tbl <- crossing(lambda = l_seq, ulr = u_seq) %>%
    filter(lambda >= 0.2, lambda <= 0.6
          ,ulr    >= 0.6, ulr    <= 0.8
           ) %>%
    mutate(output   = map2(lambda, ulr, fully_developed_likelihood, sd = curve_sd)
          ,loglik   = map_dbl(output, 'loglik')
          ,imp_prob = exp(loglik) * dlambda * dulr
          ,prob     = imp_prob / sum(imp_prob)
           )
```

```{r plot_grid_approximation, echo=FALSE, fig.height=7}
sample_tbl <- grid_tbl %>%
    filter(prob >= quantile(prob, 0.80)) %>%
#    sample_n(10, weight = prob)
    sample_n(5, weight = prob)

grid_approx_plot <- ggplot(grid_tbl) +
    geom_tile   (aes(x = lambda, y = ulr, fill = prob)) +
    geom_contour(aes(x = lambda, y = ulr, z = prob), binwidth = 1, colour = 'black', bins = 10) +
    geom_point  (aes(x = mle_est$par[1], y = mle_est$par[2]), colour = 'black') +
    scale_fill_gradient(low = 'blue', high = 'red') +
    xlab(expression(lambda)) +
    ylab("ULR") +
    ggtitle("Grid Approximation of Curve Fit")

grid_approx_plot %>% plot
```

---

```{r plot_grid_approximation_sample_points, echo=FALSE}
grid_approx_points_plot <- grid_approx_plot +
    geom_point(aes(x = lambda, y = ulr), data = sample_tbl, colour = 'grey')
    
grid_approx_points_plot %>% plot
```

---

```{r plot_possible_curves, echo=FALSE}
mlecurve_tbl <- fully_developed_likelihood(mle_est$par[1]
                                          ,mle_est$par[2]
                                          ,mle_est$par[3]
                                           )$curve_tbl

plot_tbl <- sample_tbl %>%
    pull(output) %>%
    map('curve_tbl') %>%
    bind_rows(.id = 'id_val')

mle_neighbour_plot <- ggplot(plot_tbl) +
    geom_line (aes(x = dev_lag, y = predicted, group = id_val),                           colour = 'black', alpha = 0.2) +
    geom_line (aes(x = dev_lag, y = loss_ratio), data = plot_tbl %>% filter(id_val == 1), colour = 'blue') +
    geom_point(aes(x = dev_lag, y = loss_ratio), data = plot_tbl %>% filter(id_val == 1), colour = 'blue') +
    geom_line (aes(x = dev_lag, y = predicted),  data = mlecurve_tbl,                     colour = 'red', size = 1) +
    expand_limits(y = c(0, 0.8)) +
    xlab("Time") +
    ylab("Development Factor") +
    ggtitle("Plot of Possible Curves vs Observed")

mle_neighbour_plot %>% plot
```

---

```{r calculate_five_year_development, echo=FALSE}
fiveyear_tbl <- snapshot_tbl %>% filter(acc_year == 1993)

fiveyear_lst <- create_dev_factor_diagnostics(fiveyear_tbl, l_seq, u_seq)

fiveyear_lst$gridapprox_plot %>% plot
```

---

```{r calculate_three_year_development, echo=FALSE}
threeyear_tbl <- snapshot_tbl %>% filter(acc_year == 1995)

threeyear_lst <- create_dev_factor_diagnostics(threeyear_tbl, l_seq, u_seq)

threeyear_lst$gridapprox_plot %>% plot
```

---

```{r calculate_ten_years, echo=FALSE, height=10, width=14}
allyears_lst <- snapshot_tbl %>%
    split(.$acc_year) %>%
    map(create_dev_factor_diagnostics, l_seq = l_seq, u_seq = u_seq, plot_title = '')

allyears_plot <- allyears_lst %>%
    map('gridapprox_plot') %>%
    plot_grid(plotlist = ., ncol = 4)

allyears_plot %>% plot
```

---

```{r plot_empirical_prior, echo=FALSE}
empirical_prior_tbl <- allyears_lst %>%
    map('grid_tbl') %>%
    bind_rows %>%
    group_by(lambda, ulr) %>%
    summarise(inter = sum(prob)) %>%
    ungroup() %>%
    mutate(new_prob = inter / sum(inter))


ggplot(empirical_prior_tbl) +
    geom_tile(aes(x = lambda, y = ulr, fill = new_prob)) +
    scale_fill_gradient(low = 'blue', high = 'red') +
    ggtitle('Empirical Prior') +
    theme(legend.position = 'none')
```


# Specifying the Model

## Core Concept

\


$$
\text{Loss}(t) = \text{Premium} \times \text{ULR} \times \text{GF}(t)
$$


## Full Specification

\


$$
\text{Loss}(Y, t) \sim \text{Normal}(\mu(Y, t), \, \sigma_Y)
$$

where

\begin{eqnarray*}
\mu(Y, t)       &=& \text{Premium}(Y) \times \text{LR}(Y) \times \text{GF}(t)  \\
\text{GF}(t)    &=& \text{growth function of } t \\
\sigma_Y        &=& \text{Premium}(Y) \times \sigma \\
\text{LR}_Y     &\sim& \text{Lognormal}(\mu_{\text{LR}}, \sigma_{\text{LR}}) \\
\mu_{\text{LR}} &\sim& \text{Normal}(0, 0.5)
\end{eqnarray*}


## Importance of the Functional Form

\


Does the choice of function matter?

---

```{r weibull_loglogistic_comparison_plot, echo=FALSE}
t_seq <- seq(0, 15, by = 0.01)

loglogistic_func <- function(t, om, th) 1 - exp(-(t/th)^om)
weibull_func     <- function(t, om, th) t^om / (t^om + th^om)


weibull_tbl <- tibble(
    label = 'Weibull'
   ,t = t_seq
   ,value = weibull_func(t_seq, 1.5, 2.2)
)

loglogistic_tbl <- tibble(
    label = 'Log-logistic'
   ,t = t_seq
   ,value = loglogistic_func(t_seq, 1.5, 2.2)
)

plot_tbl <- bind_rows(weibull_tbl, loglogistic_tbl)


ggplot(plot_tbl) +
    geom_line(aes(x = t, y = value, colour = label)) +
    xlab(expression(t)) +
    ylab(expression("Growth Factor for (" * omega * "=1.5, " * theta * "=2.2)")) +
    ggtitle("Sample Curves for Log-Logistic and Weibull Forms")
```

---

```{r fit_new_params_gf, echo=FALSE}
ll_vals <- loglogistic_tbl$value

new_param_func <- function(x) {
    omega <- x[1]
    theta <- x[2]

    new_vals <- weibull_func(t_seq, omega, theta)

    tot_ss <- sum((new_vals - ll_vals)^2)

    return(tot_ss)
}

optim_params <- optim(c(1, 1), new_param_func)


fittedweibull_tbl <- tibble(
    label = 'Weibull (fitted)'
   ,t = t_seq
   ,value = weibull_func(t_seq
                        ,optim_params$par[1]
                        ,optim_params$par[2])
)


plot_tbl <- bind_rows(weibull_tbl
                     ,loglogistic_tbl
                     ,fittedweibull_tbl)

ggplot(plot_tbl) +
    geom_line(aes(x = t, y = value, colour = label)) +
    xlab(expression(t)) +
    ylab(expression("Functional Forms for Growth/Development Factors")) +
    ggtitle("Comparison Plot for Weibull and Log-Logistic Curves")    
```

---

Not much difference...

---

(probably)


# Building the Stan Model

## Getting Started

```{r load_stan_file, echo=FALSE, comment=''}
stan_code <- read_lines('losscurves_sislob.stan')
```

```{r print_functions_block, echo=FALSE, comment=''}
stan_code[1:10] %>% cat(sep = '\n')
```

---

```{r print_data_block, echo=FALSE, comment=''}
stan_code[12:28] %>% cat(sep = '\n')
```

---

```{r print_params_block, echo=FALSE, comment=''}
stan_code[30:40] %>% cat(sep = '\n')
```

---

```{r print_transparams_block, echo=FALSE, comment=''}
stan_code[42:55] %>% cat(sep = '\n')
```

---

```{r print_model_block, echo=FALSE, comment=''}
stan_code[57:69] %>% cat(sep = '\n')
```


# Models Fits and Diagnostics

## Model Convergence

```{r stan_data, echo=FALSE}
modeldata_tbl <- claimdata_tbl %>%
    filter(lob    == 'ppauto'
          ,grcode == use_grcode)

usedata_tbl <- modeldata_tbl %>%
    filter(dev_year < 1998)

cohort_maxtime <- usedata_tbl %>%
    group_by(acc_year) %>%
    summarise(maxtime = max(dev_lag)) %>%
    arrange(acc_year) %>%
    pull(maxtime)

cohort_premium <- usedata_tbl %>%
    group_by(acc_year) %>%
    summarise(premium = unique(premium)) %>%
    pull(premium)

t_values <- usedata_tbl %>%
    select(dev_lag) %>%
    arrange(dev_lag) %>%
    unique %>%
    pull(dev_lag)

standata_lst <- list(
    growthmodel_id = 1   # Use weibull rather than loglogistic
   ,n_data         = usedata_tbl %>% nrow
   ,n_time         = usedata_tbl %>% select(dev_lag)  %>% unique %>% nrow
   ,n_cohort       = usedata_tbl %>% select(acc_year) %>% unique %>% nrow
   ,cohort_id      = get_character_index(usedata_tbl$acc_year)
   ,cohort_maxtime = cohort_maxtime
   ,t_value        = t_values
   ,t_idx          = get_character_index(usedata_tbl$dev_lag)
   ,premium        = cohort_premium
   ,loss           = usedata_tbl$cum_loss
)
```

```{r create_stan_model, echo=FALSE}
model_sislob_stanmodel <- stan_model('losscurves_sislob.stan')

model_sislob_stanfit <- sampling(
    object = model_sislob_stanmodel
   ,data   = standata_lst
   ,iter   = 500
   ,chains = 8
   ,seed   = stan_seed
)
```

```{r plot_parameter_traceplots, echo=FALSE}
traceplot(model_sislob_stanfit, pars = c('omega','theta','LR','mu_LR','sd_LR'))
```

---

```{r plot_parameter_traceplots_rest, echo=FALSE}
traceplot(model_sislob_stanfit, pars = c('loss_sd','gf'))
```

---

```{r plot_rhat_values, echo=FALSE}
model_sislob_stanfit %>%
    rhat(pars = c('omega','theta','LR','mu_LR','sd_LR','loss_sd')) %>%
    mcmc_rhat(.) +
        yaxis_text() +
        ggtitle("Parameter Plot of R-hat Statistic")
```

---

```{r plot_neff_values, echo=FALSE}
model_sislob_stanfit %>%
    neff_ratio(pars = c('omega','theta','LR','mu_LR','sd_LR','loss_sd')) %>%
    mcmc_neff(.) +
        yaxis_text() +
        ggtitle("Parameter Plot of Effective Sample Size")
```

## Parameter Values

```{r plot_parameter_intervals, echo=FALSE}
model_sislob_stanfit %>%
    extract(inc_warmup = FALSE, permuted = FALSE) %>%
    mcmc_intervals(regex_pars = c('omega','theta','LR\\[', 'mu_LR','sd_LR','loss_sd')) +
        expand_limits(x = c(-0.5, 2.5)) +
        ggtitle("Posterior Credibility Intervals for Sampler Parameters")
```

---

```{r plot_sanity_check, echo=FALSE}
fitted_curves_tbl <- extract(model_sislob_stanfit)$loss_sample[,1,] %>%
    as_data_frame() %>%
    mutate(iter = 1:n()) %>%
    gather("timelbl", "value", -iter) %>%
    mutate(time = gsub("V", "", timelbl) %>% as.numeric())

ggplot(snapshot_tbl %>% filter(acc_year == 1988)) +
    geom_line (aes(x = time, y = value, group = iter)
              ,data = fitted_curves_tbl, alpha = 0.005) +
    geom_line (aes(x = dev_lag, y = cum_loss), colour = 'red') +
    geom_point(aes(x = dev_lag, y = cum_loss), colour = 'blue') +
    expand_limits(y = 0) +
    scale_y_continuous(labels = dollar) +
    xlab("Time") +
    ylab("Loss") +
    ggtitle("Plot of 1988 Year Loss Development Against Posterior Distribution")
```

---

```{r plot_1993_predict_cone, echo=FALSE}
predict_cone_tbl <- extract(model_sislob_stanfit)$loss_prediction[,6,] %>%
    as_data_frame() %>%
    mutate(iter = 1:n()) %>%
    gather("timelbl", "value", -iter) %>%
    mutate(time = gsub("V", "", timelbl) %>% as.numeric())

plot_predict <- ggplot(carrier_full_tbl %>% filter(grcode == use_grcode, acc_year == '1993')) +
    geom_line (aes(x = time, y = value, group = iter)
              ,data = predict_cone_tbl, alpha = 0.005) +
    geom_line (aes(x = dev_lag, y = cum_loss), colour = 'red') +
    geom_point(aes(x = dev_lag, y = cum_loss), colour = 'blue') +
    expand_limits(y = 0) +
    scale_y_continuous(labels = dollar) +
    xlab("Time") +
    ylab("Loss") +
    ggtitle("Plot of 1993 Year Loss Prediction")

plot_predict %>% plot
```

---

```{r plot_1995_predict_code, echo=FALSE}
predict_cone_tbl <- extract(model_sislob_stanfit)$loss_prediction[,8,] %>%
    as_data_frame() %>%
    mutate(iter = 1:n()) %>%
    gather("timelbl", "value", -iter) %>%
    mutate(time = gsub("V", "", timelbl) %>% as.numeric())

plot_predict <- ggplot(carrier_full_tbl %>% filter(grcode == use_grcode, acc_year == '1995')) +
    geom_line (aes(x = time, y = value, group = iter)
              ,data = predict_cone_tbl, alpha = 0.005) +
    geom_line (aes(x = dev_lag, y = cum_loss), colour = 'red') +
    geom_point(aes(x = dev_lag, y = cum_loss), colour = 'blue') +
    expand_limits(y = 0) +
    scale_y_continuous(labels = dollar) +
    xlab("Time") +
    ylab("Loss") +
    ggtitle("Plot of 1995 Year Loss Prediction")

plot_predict %>% plot
```


# Constructing the Predictive Checks

## Range of Loss Ratios

\

Project out ULR values

\

Find smallest and largest value for each iteration

\

Compare to data

---

```{r plot_range_ulrs, echo=FALSE}
ppc_min_lr <- extract(model_sislob_stanfit)$ppc_minLR
ppc_max_lr <- extract(model_sislob_stanfit)$ppc_maxLR

lr_tbl <- carrier_full_tbl %>%
    filter(grcode == use_grcode
          ,dev_lag == 10) %>%
    summarise(min_lr = min(loss_ratio)
             ,max_lr = max(loss_ratio))

min_plot <- ggplot() +
    geom_line(aes(x = ppc_min_lr), stat = 'density') +
    geom_vline(aes(xintercept = lr_tbl$min_lr), colour = 'red') +
    xlab("Minimum Loss Ratio") +
    ylab("Probability Density") +
    ggtitle("Min Loss Ratio")

max_plot <- ggplot() +
    geom_line(aes(x = ppc_max_lr), stat = 'density') +
    geom_vline(aes(xintercept = lr_tbl$max_lr), colour = 'red') +
    xlab("Maximum Loss Ratio") +
    ylab("Probability Density") +
    ggtitle("Max Loss Ratio")

plot_grid(min_plot, max_plot, nrow = 2)
```


## Total Reserves

\

Project out ultimate loss ratios

\

Calculate difference to current loss ratio

\

Add up across accounting years

---

```{r plot_total_reserves_ppc, echo=FALSE}
tckc <- snapshot_tbl %>%
    filter(lob == 'ppauto'
          ,grcode == use_grcode) %>%
    group_by(acc_year) %>%
    filter(dev_lag == max(dev_lag)) %>%
    pull(cum_loss) %>%
    sum

afc <- carrier_full_tbl %>%
    filter(grcode == use_grcode) %>%
    group_by(acc_year) %>%
    filter(dev_lag == max(dev_lag)) %>%
    pull(cum_loss) %>%
    sum

future_claims <- afc - tckc


ggplot() +
    geom_line(aes(x = extract(model_sislob_stanfit)$ppc_EFC), stat = 'density') +
    geom_vline(aes(xintercept = future_claims), colour = 'red') +
    scale_x_continuous(labels = dollar) +
    xlab("Future Claims ('000s)") +
    ylab("Probability Density") +
    ggtitle("Forecasted Reserves for All Claims")
```




# Conclusion

## Further Iterations

\

Non-constant variance

\

Other functional forms (CDFs)

\

Multiple lines of business

\

Multiple coverholders

---

On that note...

\

https://magesblog.com/post/2018-07-15-hierarchical-loss-reserving-with-growth-cruves-using-brms/



## Questions

\

Thank You!!!

\

mickcooney@gmail.com

\

http://mc-stan.org/users/documentation/case-studies/losscurves_casestudy.html
