---
title: "An Introduction to Bayesian Regression and Hierarchical Models"
subtitle: "Dublin Data Science"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "2019-08-21"
output:
  revealjs::revealjs_presentation:
    theme: night
    highlight: pygments
    center: true
    reveal_options:
      slideNumber: true
---

```{r knit_opts, include=FALSE, warning=FALSE, message=FALSE}
rm(list = ls()); gc()

library(conflicted)
library(tidyverse)
library(magrittr)
library(scales)
library(cowplot)
library(knitr)
library(fs)
library(snakecase)
library(rstan)
library(rstanarm)
library(bayesplot)
library(tidybayes)


source("custom_functions.R")

conflict_prefer('filter',   'dplyr')
conflict_prefer('lag',      'dplyr')
conflict_prefer('select',   'dplyr')


knitr::opts_chunk$set(tidy  = FALSE
                     ,cache = FALSE
                     ,message = FALSE
                     ,warning = FALSE
                     ,fig.height =  8
                     ,fig.width  = 11)


options(width = 80L
       ,warn  = 1
       ,mc.cores = parallel::detectCores()
        )

theme_set(theme_cowplot())

rstan_options(auto_write = TRUE)

set.seed(42)
stan_seed <- 42
```



# Introduction


## Diamonds

\


```{r show_diamonds_data, echo=FALSE}
diamonds %>%
  head(n = 10) %>%
  kable()
```

---

```{r show_bivariate_plot, echo=FALSE}
price_carat_plot <- ggplot(diamonds) +
  geom_point(aes(x = carat, y = price), alpha = 0.1, size = 1) +
  scale_y_continuous(labels = comma) +
  xlab("Carat") +
  ylab("Price") +
  ggtitle("Carat vs Price Plot")

price_cut_plot <- ggplot(diamonds) +
  geom_boxplot(aes(x = cut, y = price)) +
  scale_y_continuous(labels = comma) +
  xlab("Cut") +
  ylab("Price") +
  ggtitle("Cut vs Price Plot")

price_color_plot <- ggplot(diamonds) +
  geom_boxplot(aes(x = color, y = price)) +
  scale_y_continuous(labels = comma) +
  xlab("Color") +
  ylab("Price") +
  ggtitle("Color vs Price Plot")

price_table_plot <- ggplot(diamonds) +
  geom_point(aes(x = table, y = price), alpha = 0.1, size = 1) +
  scale_y_continuous(labels = comma) +
  xlab("Table") +
  ylab("Price") +
  ggtitle("Table vs Price Plot")

plot_grid(price_carat_plot, price_cut_plot, price_color_plot, price_table_plot
         ,ncol = 2)
```


## US Radon Measurements

\


```{r load_radon_data, echo=FALSE}
radon_cols <- cols(
  .default = col_character(),
  floor    = col_integer(),
  room     = col_integer(),
  wave     = col_double(),
  activity = col_double(),
  pcterr   = col_double(),
  adjwt    = col_double()
)

radon_tbl <- read_csv("data/srrs2.dat",
                      col_types = radon_cols,
                      na        = c("", "NA", ".")
                      ) %>%
  mutate(log_radon = log(if_else(activity == 0, 0.1, activity)))

radon_mn_tbl <- radon_tbl %>%
  filter(state == 'MN')
```


```{r show_radon_data, echo=FALSE}
radon_tbl %>%
  print()
```

---

```{r show_county_radon_boxplots, echo=FALSE}
show_counties <- c('LAC QUI PARLE', 'AITKIN', 'KOOCHICHING', 'DOUGLAS', 'CLAY'
                  ,'STEARNS', 'RAMSEY', 'ST LOUIS')

radon_mn_eight_tbl <- radon_mn_tbl %>%
  filter(county %in% show_counties)


ggplot(radon_mn_eight_tbl) +
  geom_boxplot(aes(x = to_title_case(county), y = activity)) +
  xlab("County") +
  ylab("Radon Activity Reading") +
  ggtitle("Boxplot for Radon Measurements for Selected Counties")
```

---

```{r show_county_radon_points, echo=FALSE}
ggplot(radon_mn_eight_tbl) +
  geom_point(aes(x = to_title_case(county), y = activity)) +
  xlab("County") +
  ylab("Radon Activity Reading") +
  ggtitle("Boxplot for Radon Measurements for Selected Counties")
```



# Basic Regression Theory

\

Investigate relationships between quantities

\


\begin{eqnarray*}
y &=& \text{output variable}  \\
x_i &=& \text{input variables}
\end{eqnarray*}


## Formulating the Problem

\

Observations drawn from Normal distribution

\

$$
y \sim \mathcal{N}(\mu, \sigma)
$$

---

Predictors determine mean, $\mu$

\

Choose functional form for this relationship


---

### Linear Model

\

$$
f(x) \to \beta \mathbf{X} \to \mu
$$

---

### Consequence

\

Each point drawn from an individual distribution

\

$$
y_i \sim \mathcal{N}\left(\sum \beta_j x_{ji}, \, \sigma\right)
$$

---

Model fit $\rightarrow$ determine values for $\beta$



## Model Fitting

\

How do we determine $\beta$?

---

Calculate $\mu$ from data

\

Calculate log-likelihood of measurement, $\mathcal{L}(y \, | \, \mu, \sigma)$

\

Sum over all datapoints

---

```{r show_loglik_heatmap, echo=FALSE}
diam_loglik_func <- construct_loglik_function(diamonds)

loglik_heatmap_data_file <- 'data/loglik_heatmap_tbl.rds'

if(!file_exists(loglik_heatmap_data_file)) {
  intcpt_vals <- seq(-3000, -1000, by = 25)
  carat_vals  <- seq( 6500,  8500, by = 25)
  
  loglik_heatmap_tbl <- crossing(intcpt = intcpt_vals, carat = carat_vals) %>%
    mutate(loglik = map2_dbl(intcpt, carat
                            ,diam_loglik_func
                            ,fit_sd = 1548, show_prob = 0.001))
  
  loglik_heatmap_tbl %>% write_rds(loglik_heatmap_data_file)
} else {
  loglik_heatmap_tbl <- read_rds(loglik_heatmap_data_file)
}


ggplot(loglik_heatmap_tbl) +
  geom_tile(aes(x = intcpt, y = carat, fill = loglik)) +
  geom_contour(aes(x = intcpt, y = carat, z = loglik), colour = 'black') +
  geom_point(aes(x = -2256, y = 7756), size = 3) +
  scale_fill_gradient(low = 'blue', high = 'red') +
  xlab("Intercept Parameter") +
  ylab("Carat Parameter") +
  ggtitle("Heatmap of Log-likelihood Values from Intercept and Carat Parameters")
```

---

Parameter uncertainty


# Bayesian Regression


## Bayesian Inference Engine

\

Prior Knowledge

$+$

Data

\

$=$

\

Posterior Knowledge

---

Parameters, $\theta$

\

Data, $D$

---

Prior: $p(\theta)$

\

Likelihood: $p(D | \theta)$

\

Posterior: $p(\theta | D)$

---

$$
p(\theta \, | \, D) = \int p(\theta) \, p(D \, | \, \theta)
$$

\

Posterior calculation is high-dim integral

---

Use MCMC to sample posterior

---

## Stan, rstan, and rstanarm / brms {data-background="img/stan_logo_tm.png"}

\

Probabilistic Programming Language

\

CmdStan, PyStan, rstan

\

rstanarm and brms

---

```{r fit_bayesian_regression_model, echo=TRUE}
diamondprice_stanlm <- stan_lm(
  price ~ carat + cut + color + clarity
 ,data            = diamonds
 ,prior           = R2(0.7)
 ,chains          = 4
 ,seed            = stan_seed
)
```

---

```{r check_fit_diagnostics, echo=FALSE, message=TRUE}
diamondprice_stanlm %>%
  .$stanfit %>%
  check_hmc_diagnostics()
```

---

```{r show_fit_traceplots, echo=FALSE}
diamondprice_stanlm %>%
  plot('trace') +
  ggtitle("Parameter Traceplots") +
  theme(axis.text.x = element_text(size = 7))
```

---

```{r show_fit_histograms, echo=FALSE}
diamondprice_stanlm %>%
  plot('hist') +
  ggtitle("Parameter Histograms") +
  theme(axis.text.x = element_text(size = 7))
```


## Using Model Outputs

\


How do we use these posterior draws?


---

```{r plot_diamonds_comparison, echo=FALSE}
sample_12_tbl <- diamonds %>% sample_n(12)

sample_predict_tbl <- sample_12_tbl %>%
  add_predicted_draws(diamondprice_stanlm)
  
ggplot(sample_predict_tbl) +
  geom_histogram(aes(x = .prediction), bins = 50) +
  geom_vline(aes(xintercept = price), colour = 'red') +
  facet_wrap(vars(.row), ncol = 4, scales = 'free_x') +
  xlab("Diamond Price") +
  ylab("Count") +
  ggtitle("Posterior Predictive Comparion Plot") +
  theme(axis.text.x = element_text(size = 7))
```

---

Negative prices?

\

Use logs

---

```{r fit_diamlog_model, echo=TRUE}
diamonds_tbl <- diamonds %>% mutate(log_price = log(price))

diamlog_stanlm <- stan_lm(
  log_price ~ carat + cut + color + clarity
 ,data            = diamonds_tbl
 ,prior           = R2(0.7)
 ,chains          = 4
 ,seed            = stan_seed
)
```

---

```{r check_diamlog_diagnostics, echo=FALSE, message=TRUE}
diamlog_stanlm %>%
  .$stanfit %>%
  check_hmc_diagnostics()
```

---

```{r show_diamlog_traceplots, echo=FALSE}
diamlog_stanlm %>%
  plot('trace') +
  ggtitle("Parameter Traceplots") +
  theme(axis.text.x = element_text(size = 7))
```

---

```{r show_diamlog_histograms, echo=FALSE}
diamlog_stanlm %>%
  plot('hist') +
  ggtitle("Parameter Histograms") +
  theme(axis.text.x = element_text(size = 7))
```

---

```{r plot_diamlog_predictions, echo=FALSE}
sample_diamlog_predict_tbl <- sample_12_tbl %>%
  mutate(log_price = log(price)) %>%
  add_predicted_draws(diamlog_stanlm)
  
ggplot(sample_diamlog_predict_tbl) +
  geom_histogram(aes(x = .prediction), bins = 50) +
  geom_vline(aes(xintercept = log_price), colour = 'red') +
  facet_wrap(vars(.row), ncol = 4, scales = 'free_x') +
  xlab("Log Diamond Price") +
  ylab("Count") +
  ggtitle("Posterior Predictive Diamond Log Price Comparison Plot")
```

---

A word of caution...

---

Predicting `log(price)`

---

What about `price`?

---

```{r plot_diamlog_predictions_natural, echo=FALSE}
ggplot(sample_diamlog_predict_tbl %>% mutate(predict_price = exp(.prediction))) +
  geom_histogram(aes(x = predict_price), bins = 50) +
  geom_vline(aes(xintercept = price), colour = 'red') +
  facet_wrap(vars(.row), ncol = 4, scales = 'free_x') +
  xlab("Diamond Price") +
  ylab("Count") +
  ggtitle("Posterior Predictive Diamond Log Price Naturalised Comparison Plot") +
  theme(axis.text.x = element_text(size = 7))
```

---

Induce a uniform distribution?

---

```{r check_logprice_induced_uniform, echo=FALSE}
plot_tbl <- diamonds_tbl %>%
  sample_n(5000) %>%
  add_predicted_draws(diamlog_stanlm) %>%
  summarise(cumlprob = ecdf(.prediction)(log_price[1]))


ggplot(plot_tbl) +
  geom_histogram(aes(x = cumlprob), bins = 50) +
  xlab("Cumulative Probability") +
  ylab("Frequency") +
  ggtitle("Induced Cumulative Probability Distribution of Diamond Log-Price")
```

---

```{r check_exp_logprice_induced_uniform, echo=FALSE}
plot_tbl <- diamonds_tbl %>%
  sample_n(1000) %>%
  add_predicted_draws(diamlog_stanlm) %>%
  summarise(cumlprob = ecdf(exp(.prediction))(price[1]))


ggplot(plot_tbl) +
  geom_histogram(aes(x = cumlprob), bins = 50) +
  xlab("Cumulative Probability") +
  ylab("Frequency") +
  ggtitle("Induced Cumulative Probability Distribution of Naturalised Diamond Log-Price")
```





# Hierarchical Models

---

What about small data?

\

How do we use structure?


## Fully-Pooled Model

```{r fit_mn_radon_pooled_model, echo=TRUE}
radon_full_stanlm <- stan_lm(
  log_radon ~ floor
 ,data            = radon_mn_tbl
 ,prior           = R2(0.7, what = 'mean')
 ,chains          = 4
 ,seed            = stan_seed
)
```

---

```{r check_radon_pooled_diagnostics, echo=FALSE, message=TRUE}
radon_full_stanlm %>%
  .$stanfit %>%
  check_hmc_diagnostics()
```


---

```{r plot_pooled_predictions, echo=FALSE}
plot_tbl <- tibble(floor = c(0, 1)) %>%
  add_predicted_draws(radon_full_stanlm) %>%
  ungroup() %>%
  arrange(.draw) %>%
  select(.draw, floor, .prediction)
  
ggplot(plot_tbl) +
  geom_line(aes(x = floor, y = .prediction, group = .draw), alpha = 0.1) +
  geom_point(aes(x = floor, y = log_radon), data = radon_mn_tbl) +
  xlab("Floor") +
  ylab("Logarithm of Radon Reading") +
  ggtitle("Data and Predictions for Fully Pooled Model")
```


## Unpooled Model

```{r fit_mn_radon_unpooled_model, echo=TRUE}
radon_none_stanlm <- stan_lm(
  log_radon ~ floor + county
 ,data            = radon_mn_tbl
 ,prior           = R2(0.7)
 ,chains          = 4
 ,seed            = stan_seed
)
```


---


```{r check_radon_unpooled_diagnostics, echo=FALSE, message=TRUE}
radon_none_stanlm %>%
  .$stanfit %>%
  check_hmc_diagnostics()
```

---

```{r extract_unpooled_params, echo=FALSE}
radon_none_draws_tbl <- radon_none_stanlm %>%
  tidy_draws() %>%
  gather_variables()

params_county_summary_tbl <- radon_none_draws_tbl %>%
  filter(str_detect(.variable, 'county')) %>%
  ungroup() %>%
  mutate(county = str_replace(.variable, 'county', '')) %>%
  group_by(county) %>%
  summarise(mean_val   = mean(.value)
           ,median_val = median(.value)
           ,lower      = quantile(.value, 0.25)
           ,upper      = quantile(.value, 0.75)
           )

plot_tbl <- radon_mn_tbl %>%
  count(county, name = 'sample_count') %>%
  inner_join(params_county_summary_tbl, by = 'county') %>%
  mutate(county = county %>% to_title_case()
        ,size   = log(sample_count)
         )

ggplot(plot_tbl) +
  geom_point(aes(x = county, y = mean_val)) +
  geom_errorbar(aes(x = county, ymin = lower, ymax = upper, colour = size), width = 0) +
  scale_colour_gradient(low = 'blue', high = 'red') +
  xlab("County") +
  ylab("Parameter Value") +
  ggtitle("50% Credibility Intervals for County Parameter (Full Pooling)") +
  theme(axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5))
```

---

Larger estimates $\to$ smaller sample size

---

What about partial pooling?

---

\begin{eqnarray*}
\text{Larger samples}  &\to& \text{individual estimates} \\
\text{Smaller samples} &\to& \text{grouped estimates}
\end{eqnarray*}


## Partial Pooling

```{r fit_mn_radon_partial_model, echo=TRUE}
radon_partial_1_stanlm <- stan_lmer(
  log_radon ~ 1 + (1 | county)
 ,data            = radon_mn_tbl
 ,prior           = normal(0, 1)
 ,chains          = 4
 ,seed            = stan_seed
)
```

---

```{r check_radon_partial_diagnostics, echo=FALSE, message=TRUE}
radon_partial_1_stanlm %>%
  .$stanfit %>%
  check_hmc_diagnostics()
```

---

```{r extract_partial_1_params, echo=FALSE}
radon_partial_1_draws_tbl <- radon_partial_1_stanlm %>%
  tidy_draws() %>%
  gather_variables()

params_county_partial_1_summary_tbl <- radon_partial_1_stanlm %>%
  gather_draws(b[i, v]) %>%
  ungroup() %>%
  mutate(county = str_replace_all(v, '_', ' ') %>% str_replace('county:', '')) %>%
  group_by(county) %>%
  summarise(mean_val   = mean(.value)
           ,median_val = median(.value)
           ,lower      = quantile(.value, 0.25)
           ,upper      = quantile(.value, 0.75)
           )

plot_tbl <- radon_mn_tbl %>%
  count(county, name = 'sample_count') %>%
  inner_join(params_county_partial_1_summary_tbl, by = 'county') %>%
  mutate(county = county %>% to_title_case()
        ,size   = log(sample_count)
         )

ggplot(plot_tbl) +
  geom_point(aes(x = county, y = mean_val)) +
  geom_errorbar(aes(x = county, ymin = lower, ymax = upper, colour = size), width = 0) +
  scale_colour_gradient(low = 'blue', high = 'red') +
  xlab("County") +
  ylab("Parameter Value") +
  ggtitle("50% Credibility Intervals for County Parameter (Partial Pooling)") +
  theme(axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5))
```


## Partial Pooling with Predictors

```{r fit_mn_radon_partial_2_model, echo=TRUE}
radon_partial_2_stanlm <- stan_lmer(
  log_radon ~ floor + (1 | county)
 ,data            = radon_mn_tbl
 ,prior           = normal(0, 1)
 ,chains          = 4
 ,seed            = stan_seed
)
```

---

```{r check_radon_partial_2_diagnostics, echo=FALSE, message=TRUE}
radon_partial_2_stanlm %>%
  .$stanfit %>%
  check_hmc_diagnostics()
```

---

```{r extract_partial_2_params, echo=FALSE}
radon_partial_2_draws_tbl <- radon_partial_2_stanlm %>%
  tidy_draws() %>%
  gather_variables()

params_county_partial_2_summary_tbl <- radon_partial_2_stanlm %>%
  gather_draws(b[i, v]) %>%
  ungroup() %>%
  mutate(county = str_replace_all(v, '_', ' ') %>% str_replace('county:', '')) %>%
  group_by(county) %>%
  summarise(mean_val   = mean(.value)
           ,median_val = median(.value)
           ,lower      = quantile(.value, 0.25)
           ,upper      = quantile(.value, 0.75)
           )

plot_tbl <- radon_mn_tbl %>%
  count(county, name = 'sample_count') %>%
  inner_join(params_county_partial_2_summary_tbl, by = 'county') %>%
  mutate(county = county %>% to_title_case()
        ,size   = log(sample_count)
         )

ggplot(plot_tbl) +
  geom_point(aes(x = county, y = mean_val)) +
  geom_errorbar(aes(x = county, ymin = lower, ymax = upper, colour = size), width = 0) +
  scale_colour_gradient(low = 'blue', high = 'red') +
  xlab("County") +
  ylab("Parameter Value") +
  ggtitle("50% Credibility Intervals for County Parameter (Partial Pooling with floor)") +
  theme(axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5))
```


# Stan Programs

## Domain Specific Language

\

Compiles to C++

\

Documentation improving rapidly

---

```{r show_radon_nopool, echo=FALSE}
read_lines("radon_no_pool.stan") %>%
  cat(sep = '\n')
```

---

Flexible

\

Censored, truncated data

\

Generative modelling




# Conclusion


## Problems and Shortcomings

\

Not a magic bullet

\

Implementation more complex (by design)




## Future Improvements

\

Ecosystem developing rapidly

\

PyStan and arviz




## Further Resources

\

[Stan documentation](https://mc-stan.org/users/documentation/)

\

[Statistical Rethinking](https://xcelab.net/rm/statistical-rethinking/)

\

`rstanarm`, `tidybayes`, `bayesplot`, `shinystan`




## Questions?

\

Email:

mickcooney@gmail.com

\

GitHub:

https://github.com/kaybenleroll/data_workshops


