---
title: "Workshop: Advanced Queuing Theory"
author: "Mick Cooney <mickcooney@gmail.com>"
date: "Last updated: `r format(Sys.time(), '%B %d, %Y')`"
editor: source
execute:
  message: false
  warning: false
  error: false
format:
  html:
    light: superhero
    dark: darkly
    anchor-sections: true
    embed-resources: true
    number-sections: true
    smooth-scroll: true
    toc: true
    toc-depth: 3
    toc-location: left
    code-fold: true
    code-summary: "Show code"
    fig-width: 11
    fig-height: 8
---

```{r import_libraries}
#| echo: false

library(conflicted)
library(tidyverse)
library(scales)
library(cowplot)
library(magrittr)
library(rlang)
library(purrr)
library(glue)
library(simmer)
library(simmer.plot)
library(queueing)

source("lib_utils.R")
source("lib_queueing.R")

conflict_lst <- resolve_conflicts(
  c("magrittr", "rlang", "dplyr", "readr", "purrr", "ggplot2", "simmer", "queueing")
  )

options(
  width = 80L,
  warn  = 1,
  mc.cores = parallelly::availableCores()
  )

theme_set(theme_cowplot())

set.seed(42)
```

# Introduction

This second session of the queuing theory workshop focuses on more complex
scenarios where the basic M/M/1 or M/M/c assumptions are relaxed or extended.

**Topics covered:**
- **Finite Capacity (M/M/1/K)**: When the queue cannot grow indefinitely.
- **Impatience**: Balking (refusing to join) and Reneging (leaving after waiting).
- **Priority Queues**: Handling different classes of customers.
- **Networks of Queues**: Tandem queues and Jackson networks.
- **Optimal Staffing**: Balancing the cost of waiting vs. the cost of servers.


# Finite Capacity Queues (M/M/1/K)

In many real-world systems, the space for waiting is limited (e.g., a small
waiting room, a buffer in a network router, or a finite number of phone lines).

When a customer arrives and the system is at its capacity **K**, the customer is
**blocked** and leaves the system immediately.

## Key Metrics for M/M/1/K

For an M/M/1/K queue with arrival rate λ and service rate μ:

- **Blocking Probability ($P_K$)**: The probability that an arriving customer
  finds the system full.
- **Effective Arrival Rate ($\lambda_{eff}$)**: $\lambda_{eff} = \lambda(1 - P_K)$.
  Only these customers actually enter the system.
- **Utilization**: $ho_{eff} = \lambda_{eff} / \mu = 1 - P_0$.

```{r mm1k_example}
#| echo: true

lambda_router <- 10  # packets per ms
mu_router <- 12      # capacity to process 12 packets per ms
k_buffer <- 5        # buffer size 5 (including one being processed)

metrics_k5 <- calculate_mm1k_metrics(lambda_router, mu_router, k_buffer)
metrics_k20 <- calculate_mm1k_metrics(lambda_router, mu_router, 20)

write_lines(
  glue(
    "=== ROUTER BUFFER ANALYSIS (K=5) ===

",
    "Arrival rate: {lambda_router}
",
    "Service rate: {mu_router}
",
    "Blocking Prob (Pk): {percent(metrics_k5$Pk, accuracy = 0.01)}
",
    "Avg wait: {round(metrics_k5$W, 4)} ms

",
    "COMPARED TO K=20:
",
    "Blocking Prob (Pk): {percent(metrics_k20$Pk, accuracy = 0.01)}
",
    "Avg wait: {round(metrics_k20$W, 4)} ms"
    ),
  stdout()
  )
```

**Trade-off**: Increasing buffer size (K) reduces the probability of losing
packets but increases the average wait time for those that enter.


# Impatience: Balking and Reneging

In human-centric systems, customers are not infinitely patient.

## Balking
**Balking** occurs when a customer arrives, sees a long queue, and decides not
to join at all.

## Reneging
**Reneging** occurs when a customer joins the queue but leaves before being
served because the wait is too long.

```{r simulate_reneging}
#| echo: true

lambda <- 10
mu <- 8  # System is unstable (ρ > 1) without reneging!

# Patients will wait at most 0.5 hours on average
patience_mean <- 0.5

customer_traj <- trajectory("Patient") |>
  renege_in(function() rexp(1, 1/patience_mean), out = trajectory() |> log_("Reneging")) |>
  seize("doctor", 1) |>
  renege_abort() |>
  timeout(function() rexp(1, mu)) |>
  release("doctor", 1)

sim_env <- simmer() |>
  add_resource("doctor", 1) |>
  add_generator("patient", customer_traj, function() rexp(1, lambda)) |>
  run(until = 100)

arrivals <- get_mon_arrivals(sim_env)
n_reneged <- sum(!arrivals$finished)

write_lines(
  glue(
    "Simulation with λ={lambda}, μ={mu}
",
    "Total arrivals: {nrow(arrivals)}
",
    "Customers reneged: {n_reneged} ({percent(n_reneged/nrow(arrivals))})
",
    "Avg wait (finished): {round(mean(arrivals$wait_time[arrivals$finished]), 2)}"
    ),
  stdout()
  )
```


# Priority Queues

Often, some customers are more "important" than others (e.g., emergency room
triage, premium support tiers).

## Non-Preemptive Priority
A high-priority customer goes to the front of the line but does not interrupt a
service already in progress.

## Preemptive Priority
A high-priority customer immediately interrupts the current service. The
interrupted customer returns to the queue.

```{r priority_simulation}
#| echo: true

# Priority 2 is high, Priority 1 is low
traj_high <- trajectory() |>
  set_prioritization(c(2, 0, FALSE)) |>
  seize("server", 1) |>
  timeout(function() rexp(1, 10)) |>
  release("server", 1)

traj_low <- trajectory() |>
  set_prioritization(c(1, 0, FALSE)) |>
  seize("server", 1) |>
  timeout(function() rexp(1, 10)) |>
  release("server", 1)

sim_env_prio <- simmer() |>
  add_resource("server", 1) |>
  add_generator("high", traj_high, function() rexp(1, 2)) |>
  add_generator("low", traj_low, function() rexp(1, 6)) |>
  run(until = 500)

arrivals_prio <- get_mon_arrivals(sim_env_prio) |>
  mutate(priority = if_else(str_detect(name, "high"), "High", "Low"))

arrivals_prio |>
  group_by(priority) |>
  summarise(
    avg_wait = mean(wait_time),
    n = n()
  )
```


# Networks of Queues

In many systems, a customer must pass through multiple stages of service.

## Tandem Queues
Queue 1 → Queue 2 → ... → Exit.

**Burke's Theorem**: The output process of an M/M/c queue in steady state is a
Poisson process with the same rate λ as the input process.

This means if Stage 1 is M/M/1 and Stage 2 is M/M/1, we can analyze them
independently as long as Stage 1 is stable.

```{r tandem_simulation}
#| echo: true

tandem_traj <- trajectory() |>
  seize("stage1", 1) |>
  timeout(function() rexp(1, 15)) |>
  release("stage1", 1) |>
  seize("stage2", 1) |>
  timeout(function() rexp(1, 12)) |>
  release("stage2", 1)

sim_env_tandem <- simmer() |>
  add_resource("stage1", 1) |>
  add_resource("stage2", 1) |>
  add_generator("customer", tandem_traj, function() rexp(1, 10)) |>
  run(until = 1000)

get_mon_resources(sim_env_tandem) |>
  group_by(resource) |>
  summarise(utilization = mean(server))
```


# Optimal Staffing and Economics

We want to find the number of servers **c** that minimizes total cost:

$$
	ext{Total Cost} = 	ext{Cost of Service} + 	ext{Cost of Waiting}
$$
$$
TC(c) = c \cdot C_s + L \cdot C_w
$$

Where:
- $C_s$: Cost per server per unit time.
- $C_w$: Cost of customer waiting per unit time.

```{r optimization_example}
#| echo: true

lambda <- 20
mu <- 8
cost_server <- 50   # $50/hour per agent
cost_wait <- 100    # $100/hour per waiting customer (lost productivity/goodwill)

find_optimal_c <- function(lambda, mu, cs, cw) {
  results <- tibble(c = ceiling(lambda/mu + 1):15) |>
    mutate(
      metrics = map(c, ~calculate_mmc_metrics(lambda, mu, .x)),
      L = map_dbl(metrics, ~.x$L),
      service_cost = c * cs,
      wait_cost = L * cw,
      total_cost = service_cost + wait_cost
    )
  return(results)
}

opt_results <- find_optimal_c(lambda, mu, cost_server, cost_wait)

ggplot(opt_results, aes(x = c, y = total_cost)) +
  geom_line() +
  geom_point(aes(color = total_cost == min(total_cost)), size = 3) +
  scale_color_manual(values = c("black", "red"), guide = "none") +
  labs(title = "Total Cost vs. Number of Servers",
       subtitle = glue("Optimal c = {opt_results$c[which.min(opt_results$total_cost)]}"),
       x = "Number of Servers (c)", y = "Total Cost ($/hr)")
```


# Summary of Advanced Concepts

1. **Finite Capacity**: Leads to blocking; increases effective service rate but
   loses customers.
2. **Impatience**: Critical for modeling human behavior; can stabilize
   otherwise unstable systems (at a cost).
3. **Priority**: Crucial for SLA management; improves high-priority wait times
   at the expense of low-priority ones.
4. **Networks**: Complex dependencies; tandem systems can sometimes be
   decomposed.
5. **Optimization**: Queuing theory isn't just about math; it's about making
   business decisions.

# R Environment

```{r session_info}
#| echo: true

sessioninfo::session_info()
```
